### 2014-09-15 proposal/mprop.tex
Added proposal template.

\documentclass{mprop}
\usepackage{graphicx}

% alternative font if you prefer
%\usepackage{times}

% for alternative page numbering use the following package
% and see documentation for commands
%\usepackage{fancyheadings}

% other potentially useful packages
%\uspackage{amssymb,amsmath}
%\usepackage{url}
%\usepackage{fancyvrb}
%\usepackage[final]{pdfpages}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Title of project proposal placed here}
\author{Name of author placed here}
\date{Date of submission placed here}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\tableofcontents
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{intro}

briefly explain the context of the project problem

\subsection{A subsection}
Please note your proposal need not follow the included section headings - this is only a suggested structure. Also add subsections etc as required

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Statement of Problem}

clearly state the problem to be addressed in your forthcoming project. Explain why it would be worthwhile to solve this problem.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background Survey}

present an overview of relevant previous work including articles, books, and existing software products. Critically evaluate the strengths and weaknesses of the previous work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proposed Approach}

state how you propose to solve the software development problem. Show that your proposed approach is feasible, but identify any risks.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Work Plan}

show how you plan to organize your work, identifying intermediate deliverables and dates.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% it is fine to change the bibliography style if you want
\bibliographystyle{plain}
\bibliography{mprop}
\end{document}

### 2014-09-16 proposal/proposal.tex
Added proposal beginnings and structured directories.

\documentclass{mprop}

\usepackage{graphicx}
\usepackage[style=alphabetic,natbib=true]{biblatex}
\addbibresource{proposal}

% alternative font if you prefer
%\usepackage{times}

% for alternative page numbering use the following package
% and see documentation for commands
%\usepackage{fancyheadings}

% other potentially useful packages
%\uspackage{amssymb,amsmath}
%\usepackage{url}
%\usepackage{fancyvrb}
%\usepackage[final]{pdfpages}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Formalisation of Session-based Lambda Calculus Type Systems}
\author{Craig McLaughlin}
\date{\today}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\tableofcontents
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{intro}

briefly explain the context of the project problem

\subsection{A subsection}
Please note your proposal need not follow the included section headings - this
is only a suggested structure. Also add subsections etc as required

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Statement of Problem}

clearly state the problem to be addressed in your forthcoming project. Explain
why it would be worthwhile to solve this problem.

The aim of this project is to develop a framework for formalising lambda
calculus type systems involving session types using interactive theorem
proving (in particular, the system Coq). The motivation for the framework is
that it will enable a whole class of related languages and type systems to be
formalised with little extra proof effort. Inspired by previous work on a
mechanised framework of pi calculus type systems~\cite{Gay:2001:FFP}, we shift
to the lambda calculus to provide a basis on which session types for
mainstream programming languages can be studied whilst taking advantage of
previous proof effort and specialising only for those parts unique to the
particular system.

From the generic framework, a number of directions can be explored including
session polymorphism, session subtyping, and permitting different forms of
aliasing within the type system. The majority of previous session type systems
for mainstream programming languages has focussed exclusively on linear type
systems. One disadvantage of a linear type system is the need to re-bind
objects after each operation~\cite{Gay:2010:LTT}. Principally, we intend to
instantiate the framework with a type system that provides the benefits of
aliasing with the guarantees of linearity, termed ``adoption and focus''. The
``adoption and focus'' type system is non-trivial requiring special language
features, e.g. to manage capabilities which can be thought as a form of object
liveness mechanism, so the framework design must support such flexibility by,
e.g allowing the typing judgements to be extended such that the capabilities
can be included in soundness results. The instantiation will demonstrate that
the framework supports a wide range of languages and type systems providing
researchers the ability to harness the power of formal verification in the
design, implementation and documentation of session-based type systems.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background Survey}

present an overview of relevant previous work including articles, books, and
existing software products. Critically evaluate the strengths and weaknesses
of the previous work.

System F-pop, session polymorphism formalised in Coq, Simon's Isabelle/HOL
development.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proposed Approach}

state how you propose to solve the software development problem. Show that
your proposed approach is feasible, but identify any risks.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Work Plan}

show how you plan to organize your work, identifying intermediate deliverables
and dates.

I need to perform some formalisation of the system. I should identify the key
issues (linear typing, bindings (as claimed in the literature), etc).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\printbibliography

\end{document}

### 2014-09-16 proposal/proposal.tex
Added ramblings.

@@
 briefly explain the context of the project problem
+
+computing distributed communication rise of internet networked communication
+the protocols have an effect on the communication and if the protocol is wrong
+distributed communicatin now a propenent is most peoples' daily lives
+whether it be posting pictures on social media websites sharing videos
+or even so much as sending emails all require a well-defined protocol to
+specify the type and order of messages to be sent and received between the two
+or more communicating partners
+mainstream programming languages used to build such systems often fail to meet
+the strong guarantees that the correct type of messages will be sent or that
+the ordering between messages is preserved
+the rise of e-commerce and increasing reliance on software systems in the
+areas of aviation/transport mean that much confidence in these systems is placed thus it is of great importance that there exist strong guarantees of correctness
+
+it is thus clear that the need for
+construct an argument for the use of a proof assistant and interactive theorem
+proving technologies to formalise a framework for developing type systems
+involving session types that are based on lambda calculi so that we can use
+such a development to reason about a variety of type systems while minimising
+proof effort tie this into an introduction into the subfield of session types
+and what they are good for essentially motivating our project in this space
+additionally we can introduce the reader to the discipline of interactive
+theorem proving and formal methods more generally as a way to formalise
+programming languages highlighting the benefits of reusability increased
+correctness and something what else could we say i think it would appropriate
+to mention that the session types studied so far include both pi and lambda
+calculi and it may be beneficil if the reader was informed also the pi
+calculus paradigms do not fit well into the existing mainstream programming
+language paradigms this seems like it should be in the statement of the problem section down below also i should read more of jtod's slides because this writing just keep writing malarky might just be what i need to pound out the pages because it eliminates thoughts as they come into the head what else adopt and focus we have in the problem statement but perhaps those paragraphs should be re-jigged for up here multithreaded functional programming with linear typing for session types a number of these systems use different type systems but they are based on all the lambda calculus fundamentally so to have a generic framework to encode these variants would save on proof engineering effort as is championed by the POPLMARK challenge aimed at reducing the effort for researchers to code programming language metatheory it provides a library needed to perform various things define various parts of programming languages viewed to be difficult to formalise including handling binders and freshness they also mention handling of linear typing but renege on actually formulating this as part of their challenge they insist it is a future work here is the future
 

### 2014-10-01 Paper reviews/Summaries using RRCS Template/LightweightLinearTypes.tex
Started summarising a summary but it's perhaps more polished now.

\documentclass[11pt]{article}
\usepackage{gensymb}
% Set the paper size to A4
\setlength{\paperheight}{297mm}
\setlength{\paperwidth}{210mm}
% Define commands which allow the width and height of the text
% to be specified. Centre the text on the page.
\newcommand{\settextwidth}[1]{
\setlength{\textwidth}{#1}
\setlength{\oddsidemargin}{\paperwidth}
\addtolength{\oddsidemargin}{-\textwidth}
\setlength{\oddsidemargin}{0.5\oddsidemargin}
\addtolength{\oddsidemargin}{-1in}
}
\newcommand{\settextheight}[1]{
\setlength{\textheight}{#1}
\setlength{\headheight}{0mm}
\setlength{\headsep}{0mm}
\setlength{\topmargin}{\paperheight}
\addtolength{\topmargin}{-\textheight}
\setlength{\topmargin}{0.5\topmargin}
\addtolength{\topmargin}{-1in}
}
% Top and bottom margins should be 15mm, so text height is 267mm
\settextheight{267mm}
% Left and right margins should be 25mm, so text width is 160mm
\settextwidth{160mm}
% No page number
\pagestyle{empty}
\setlength{\parskip}{6pt}
\begin{document}
\begin{tabular*}{160mm}{|l@{\extracolsep{\fill}}c@{\extracolsep{0pt}}|l|}
\hline
Title:
Lightweight Linear Types in System F\degree

& &
Name:
Craig McLaughlin

\\ \hline
Author[s]:
Karl Mazurak, Jianzhou Zhao, and Steve Zdancewic

& &
Student number:

\\ \hline
\end{tabular*}

\begin{center}
\begin{minipage}{140mm}
\small
Abstract: %
There are various approaches to providing both linear and non-linear access to
resources within a programming language. However, they either hamper the
inclusion of desirable programming features, or do not adequately reflect the
common case. The present paper presents an extension to System F which avoids
these deficiencies by categorising types whilst maintaining semantics
familiar to the base language. Mechanised proofs for type soundness and
parametricity are presented, and a relationship between linear and non-linear
types is defined explicating the semantics of linearity. Extensions to the
system are suggested in order to, among other things, facilitate usage in a
general-purpose functional language. The system can encode a wide variety of
safe interfaces and simulate a range of reference management schemes,
illustrating the applicability of the technique.

\end{minipage}
\end{center}

Previous work on incoporating linearity into a type system has proposed two
approaches. One approach is to consider all types to be linear by default,
and provide a construct (let !, in the literature) for temporary non-linear
access to the resource. The problem with this approach is that non-linear
references and access patterns are more common than linear ones, resulting in
a rather cumbersome programming approach involving a lot of binders. A second
technique is to use type qualifiers (one for linear, one for non-linear) then
all types are a combination of a base (such as Int, or String) and a type
qualifier. However, to allow for desirable programming features such as
polymorphism, the technique requires that all polymorphic functions be
quantified over all type qualifiers, even for simple functions such as plus
(where it could be called on either a linear or non-linear Int). Additionally,
in some situations, there is a relationship between type qualifiers, burdening
the programmer with unnecessary detail and complexity.

WTD
Extend System F by categorising types as being of two different ``kinds''
(linear or non-linear)
Present operational semantics, and proofs of soundness, progress +
preservation, and respects parametricity. Includes a subsumption rule for
kinds allowing non-linear kinds to be treated as linear.
Linearity aware run-time semantics permit one substitution of a value into
a function by separating out uses of values into two categories: conscripted
and discharged. Upon a value's first use as a function argument it is
conscripted then any subsequent use of that value, in context where replacing
it with a free variable would result in a stuck configuration, it is
discharged. The reduction steps are augmented with the set of conscripted
and discharged values during the step.
Some extensions of the language are described to incorporate features similar
to those provided by practical functional programming languages. In addition,
support for programming using linear types is suggested; making use of
syntactic to simplify or eliminate the need for the programmer to thread the
linear references through the program or directly handle existential
quantification.
ROF
An example translation from the current method to a previously proposed method
(let !) is presented but it is noted that translation in the opposite
direction is difficult since, for example, any functions taking an
unrestricted type must now take a ! type and bind it so we can have
unrestricted access, and calls within the function will need to rebind the
variable under !. Note, the let ! typing rule:$ let !x = t_1 in t_2$ requires
$t_1$ to have type $!o$.
RFW
If the recommended extensions do not appear in the mechanisation then that
would be an appropriate future work.
Extensions present algebraic datatypes semantics with constructors with
value argument considered a value.
POC
However, the syntactic sugar description I found to be rather confusing and
the syntactic sugar specialised. What is mentioned is the need to ensure
the syntactic sugar correctly infers types. It appears to be straightforward
but is an additional overhead over the system.
It is not mentioned whether the suggested extensions appear in the
mechanisation or not.
Possible discrepancy in algebraic datatype extension, constructors can expect
``additional type parameters, which act as through [sic] they are
existentially bound'' but the semantics of case indicate they are universally
bound.

\end{document}

### 2014-10-08 Paper reviews/Summaries using RRCS Template/Lolliproc.tex
Added summary of summary for Lolliproc.

\documentclass[11pt]{article}
\usepackage{gensymb}
% Set the paper size to A4
\setlength{\paperheight}{297mm}
\setlength{\paperwidth}{210mm}
% Define commands which allow the width and height of the text
% to be specified. Centre the text on the page.
\newcommand{\settextwidth}[1]{
\setlength{\textwidth}{#1}
\setlength{\oddsidemargin}{\paperwidth}
\addtolength{\oddsidemargin}{-\textwidth}
\setlength{\oddsidemargin}{0.5\oddsidemargin}
\addtolength{\oddsidemargin}{-1in}
}
\newcommand{\settextheight}[1]{
\setlength{\textheight}{#1}
\setlength{\headheight}{0mm}
\setlength{\headsep}{0mm}
\setlength{\topmargin}{\paperheight}
\addtolength{\topmargin}{-\textheight}
\setlength{\topmargin}{0.5\topmargin}
\addtolength{\topmargin}{-1in}
}
% Top and bottom margins should be 15mm, so text height is 267mm
\settextheight{267mm}
% Left and right margins should be 25mm, so text width is 160mm
\settextwidth{160mm}
% No page number
\pagestyle{empty}
\setlength{\parskip}{6pt}
\begin{document}
\begin{tabular*}{160mm}{|l@{\extracolsep{\fill}}c@{\extracolsep{0pt}}|l|}
\hline
Title:
Lolliproc: to Concurrency from Classical Linear Logic via Curry-Howard and
Control

& &
Name:
Craig McLaughlin

\\ \hline
Author[s]:
Karl Mazurak and Steve Zdancewic

& &
Student number:

\\ \hline
\end{tabular*}

\begin{center}
\begin{minipage}{140mm}
\small
Abstract: %
There exist several approaches to utilising the power of linear logic in type
systems and programming languages, from full linear logic incorporating
classical components to the intuitionistic fragment. However, all approaches
suffer from either poor integration of concurrency support or a language which
is not in tune with mainstream functional programming. The current work
presents a language, Lolliproc, to fill the void which provides a surface
syntax familiar to the function programmer and simple primitives for
concurrency which hide the underlying process calculus from the programmer.
The language semantics ensure deadlock-freedom and absence of race conditions.
This work serves as an interesting base on which to build lambda calculus
based type systems which utilise linearity in order to represent interprocess
communcation via session types. Exploring extensions to the language to
increase its expressiveness could have far reaching implications for the way
in which communication protocols are designed and implemented.

\end{minipage}
\end{center}

Previous work on integrating the features of linear logic into type systems
and programming languages have focussed on three strands. First, construction
of functional programming languages that are similiar to the mainstream were
developed using the intuitionistic portion of linear logic (as opposed to
classical linear logic) which do not have a strong connection between the
linearity present in the calculi and the concurrency primitives provided.
The second approach harnessed the full power of classical linear logic within
the sequent calculus but it is difficult to map this system back to
conventional functional languages due to the symmetry inherent in the language
operations. The third approach invovles a natural deduction system for linear
logic, which has stronger connection to mainstream functional languages but
the semantics do not provide a simple view of the concurrency implications of
the logic and rely on techniques to simulate concurrent evaluation.

The current work presents a language based on natural deduction that bridges
the gap in the previous works. Specifically, a concurrent functional
programming language is presented as the source level syntax backed up with an
underlying process calculus to provide the concurrency mechanisms. The details
of processes are hidden from the programmer using novel adaptions on control
operators. The language encodes classical linear logic adding double negation
elimination as a language primitive. The operational semantics of the linear
lambda calculus are standard with the exception of performing reduction of
evaluation contexts at the process level.

The control operators facilitate spawning child processes and waiting in the
parent process for the child to complete execution. When a process is spawned
the parent and child are each given an endpoint of a channel (sending end is
given to the child). A closed channel token indicates that communication has
finished and that the child process can terminate (provided all linear
resources have been used). Intuition of control operators is that a
continuation passed to the spawning operator is a function executing a
protocol.

 intuition of go is that it accepts a function which is then spawned as a
child process which must execute as the protocol specified by the function
type; a yield then blocks the parent to receive information from the child
then both child and parent continue executing the protocol (this is
essentially an encoding of session types)
 The go operator defined to take a function which describes a protocol which
a child process will execute and the dual type of the protocol is returned to
the parent process on which to yield
 Annotate go operations with protocol type for process level semantics
 No new family of types are introduced to represent channels; regular linear
functions serve this purpose, consequence: bidirectional communication from
a seemingly unidirectional language
 Underneath source interface (linear lambda calculus) is process language (a
la pi calculus) but programmer isn't exposed to processes
 Seemingly only permits communication in one direction yet can mimic
bidirectional communication using session types -- elaboration needed
 Future example enodes sub-computations offloaded to child processes by
encapsulating work inside a function (the authors call it a ``thunk'')
 Linking endpoints without having to yield in the parent is possible with a
clever definition of a function which passes a completion token on to the
parent process; the first child spawn ends up with linked endpoints,
communication has thus been established
 An elaborate example shows how to reserve the direction of a channel rather
than the spawning processing having the source it gets the sink, additionally
the example shows that while values may cycle between processes there are
never any cycles in the communication; essential for soundness proof
 Processes only appear at runtime; no need to implement process typing rules
in typechecker (really? Is that what they are saying here pg.13)

ROF
 Type soundness (progress and preservation) implies deadlock-freedom and
correct session typing
 Strong normalisation: all reduction sequences reach a normal form meaning
no non-terminating programs
 Confluence imposing the same normal form on two reduction sequences starting
from the same redex prohibits race conditions
 Expression typing rules ``follow natural-deduction presentation of
intuitionistic linear calculi''
 Both term and channel contexts are linear (no weakening or contraction)
 Progress only considered for processes
 Coq proofs (need to deal with context permutation since they are linear) but
are not completed; up to preservation and progress on processes (graph based
reasoning the issue)
 Parallel processes form an acyclic communication graph, the restriction of
one channel between two halves of a parallel composition is essential
 Used a weight assignment to processes (and channels) to sketch proof of
strong normalisation
 Show confluence from strong normalisation and the diamond property (local
confluence)

RCL
RTC
RFW
* Unrestricted types, recursive types, nondeterminism
* Polymorphism (extension to System F-pop)
* Inclusion of protocol kinds into the kinds of System F-pop rather than the
syntactic constructs given in Lolliproc
  - has interesting notions e.g. sending (quantified) types between processes
* Claimed strong connection between channel endpoints definition and focused
proofs
POC
* Two processes can share only one channel with the send endpoint in one of
the processes and the receive in the other but is needed for type soundness
WIL
* Don't understand how go is both an identity function (therefore accepting
and returning the same function type) and a primitive which returns the dual
type of the protocol (I cannot get the dual from negating the function type)
* Can the processes exist independently? That is, can we construct two
separate process and link the endpoints? This is done with public access
capabilities in other works; what would the equivalent be here?
* How can linearly typed functions (powmod in alice+bob example) be used more
than once?

\end{document}

### 2014-11-05 proposal/proposal.tex
Added beginnings of typing rules for GV; need opening and substitution.
Specifically, need opening for the abstraction case (and the other
binding rules). These definitions need to respect linearity.
Added introductory paragraphs and some references.

@@
 \usepackage[style=alphabetic,natbib=true]{biblatex}
-\addbibresource{proposal}
+\addbibresource{proposal.bib}
 
@@
 
-computing distributed communication rise of internet networked communication
-the protocols have an effect on the communication and if the protocol is wrong
-distributed communicatin now a propenent is most peoples' daily lives
-whether it be posting pictures on social media websites sharing videos
-or even so much as sending emails all require a well-defined protocol to
-specify the type and order of messages to be sent and received between the two
-or more communicating partners
-mainstream programming languages used to build such systems often fail to meet
-the strong guarantees that the correct type of messages will be sent or that
-the ordering between messages is preserved
-the rise of e-commerce and increasing reliance on software systems in the
-areas of aviation/transport mean that much confidence in these systems is placed thus it is of great importance that there exist strong guarantees of correctness
+Distributed communication systems are becoming increasing prevalent in modern society.
+To facilitate communication between concurrently executing agents requires complex network protocols to define the type and sequencing of messages.
+Typically these protocols are implemented in systems programming languages, such as C, which provide no correctness guarantees of the implementation.
+Furthermore, programming in these languages is often extremely error-prone, as one has a myriad of unsafe features at one's disposal.
+As more businesses begin to conduct transactions over the Web, it is imperative that they can reason about the communication protocols upon which they rely.
+
+Encoding communication protocols as \textit{session types} is one approach to achieve guarantees about correctness.
+One particular variant of session types are \textit{binary} session types~\footnote{There are also \textit{multiparty} session types, but I do not discuss them here.}.
+A binary session type specifies communication between two agents, and it specifies the order and type of messages to be sent or received by an agent in the system.
+The two agents have dual session types such that, for example, when one is expecting to receive a message the other is expecting to send a message; both session types agree on the type of message to be exchanged.
+In this way, correctness of communication can be statically assured by providing communicating agents with \textit{channels} that have dual session types.
+A channel is defined as a bi-directional conduit for exchanging messages, not unlike a socket in network programming.
+In contrast to sockets, however, a channel is associated with an explicit type which provides a description of its protocol.
+Thus, in session-based systems a channel is usually provided as a primitive type.
 
@@
 formalised with little extra proof effort. Inspired by previous work on a
-mechanised framework of pi calculus type systems~\cite{Gay:2001:FFP}, we shift
-to the lambda calculus to provide a basis on which session types for
+mechanised framework for pi calculus type systems~\cite{Gay:2001:FFP}, we
+shift to the lambda calculus to provide a basis on which session types for
 mainstream programming languages can be studied whilst taking advantage of
@@
 systems. One disadvantage of a linear type system is the need to re-bind
-objects after each operation~\cite{Gay:2010:LTT}. Principally, we intend to
+objects after each operation~\cite{Gay:2010:LAST}. Principally, we intend to
 instantiate the framework with a type system that provides the benefits of
@@
 
-System F-pop, session polymorphism formalised in Coq, Simon's Isabelle/HOL
+System F-pop~\cite{Mazurak:2010:LLT}
+session polymorphism formalised in Coq, Simon's Isabelle/HOL
 development.
+Philip Wadler's Propositions as Sessions~\cite{Wadler:2012}
+Simon J. Gay and Vasco Vasconcelos (LAST)~\cite{Gay:2010:LAST}
+Lolliproc~\cite{Mazurak:2010:LCC}
+PhD thesis~\cite{Mazurak:2013:LPP}
 
@@
 your proposed approach is feasible, but identify any risks.
+
+HOAS Coq formalisation using Metatheory library from UPenn.
+Explain why this approach was taken versus intrinsic encoding of the terms
 

### 2014-11-05 proposal/proposal.tex
A structure developing.

@@
 
-briefly explain the context of the project problem
+Two aspects of programming language research are relevant for the problem.
+The first is concerned with static type systems for ensuring runtime safety properties for concurrent communication systems.
+The second is the increasing use of proof assistants and interactive theorem provers, for studying and mechanically verifying the semantics and properties of programming languages.
+
+\subsection{On Communication}
 
@@
 
-it is thus clear that the need for
-construct an argument for the use of a proof assistant and interactive theorem
-proving technologies to formalise a framework for developing type systems
-involving session types that are based on lambda calculi so that we can use
-such a development to reason about a variety of type systems while minimising
-proof effort tie this into an introduction into the subfield of session types
-and what they are good for essentially motivating our project in this space
-additionally we can introduce the reader to the discipline of interactive
-theorem proving and formal methods more generally as a way to formalise
-programming languages highlighting the benefits of reusability increased
+\subsection{On Verification}
+
+construct an argument for the use of a proof assistant and interactive theorem proving technologies to formalise a framework for developing type systems involving session types that are based on lambda calculi so that we can use such a development to reason about a variety of type systems while minimising proof effort tie this into an introduction into the subfield of session types and what they are good for essentially motivating our project in this space additionally we can introduce the reader to the discipline of interactivetheorem proving and formal methods more generally as a way to formalise programming languages highlighting the benefits of reusability increased
+
 correctness and something what else could we say i think it would appropriate
@@
 language paradigms this seems like it should be in the statement of the problem section down below also i should read more of jtod's slides because this writing just keep writing malarky might just be what i need to pound out the pages because it eliminates thoughts as they come into the head what else adopt and focus we have in the problem statement but perhaps those paragraphs should be re-jigged for up here multithreaded functional programming with linear typing for session types a number of these systems use different type systems but they are based on all the lambda calculus fundamentally so to have a generic framework to encode these variants would save on proof engineering effort as is championed by the POPLMARK challenge aimed at reducing the effort for researchers to code programming language metatheory it provides a library needed to perform various things define various parts of programming languages viewed to be difficult to formalise including handling binders and freshness they also mention handling of linear typing but renege on actually formulating this as part of their challenge they insist it is a future work here is the future
-
-\subsection{A subsection}
-Please note your proposal need not follow the included section headings - this
-is only a suggested structure. Also add subsections etc as required
 

### 2014-11-10 proposal/proposal.tex
Adding some more thoughts to the proposal.

@@
 
-Two aspects of programming language research are relevant for the problem.
+Two aspects of programming language research are relevant for this project.
 The first is concerned with static type systems for ensuring runtime safety properties for concurrent communication systems.
@@
 
-Distributed communication systems are becoming increasing prevalent in modern society.
+Distributed communication systems are becoming increasingly prevalent in modern society.
 To facilitate communication between concurrently executing agents requires complex network protocols to define the type and sequencing of messages.
@@
 
+The increasing use of proof assitants and interactive theorem proving in programming language research has spiked....
+POPLMARK
+save on proof engineering effort as is championed by the POPLMARK challenge aimed at reducing the effort for researchers to code programming language metatheory it provides a library needed to perform various things define various parts of programming languages viewed to be difficult to formalise including handling binders and freshness they also mention handling of linear typing but renege on actually formulating this as part of their challenge they insist it is a future work here is the future
+Metatheory library
+
 construct an argument for the use of a proof assistant and interactive theorem proving technologies to formalise a framework for developing type systems involving session types that are based on lambda calculi so that we can use such a development to reason about a variety of type systems while minimising proof effort tie this into an introduction into the subfield of session types and what they are good for essentially motivating our project in this space additionally we can introduce the reader to the discipline of interactivetheorem proving and formal methods more generally as a way to formalise programming languages highlighting the benefits of reusability increased
@@
 calculus paradigms do not fit well into the existing mainstream programming
-language paradigms this seems like it should be in the statement of the problem section down below also i should read more of jtod's slides because this writing just keep writing malarky might just be what i need to pound out the pages because it eliminates thoughts as they come into the head what else adopt and focus we have in the problem statement but perhaps those paragraphs should be re-jigged for up here multithreaded functional programming with linear typing for session types a number of these systems use different type systems but they are based on all the lambda calculus fundamentally so to have a generic framework to encode these variants would save on proof engineering effort as is championed by the POPLMARK challenge aimed at reducing the effort for researchers to code programming language metatheory it provides a library needed to perform various things define various parts of programming languages viewed to be difficult to formalise including handling binders and freshness they also mention handling of linear typing but renege on actually formulating this as part of their challenge they insist it is a future work here is the future
+language paradigms this seems like it should be in the statement of the problem section down below also i should read more of jtod's slides because this writing just keep writing malarky might just be what i need to pound out the pages because it eliminates thoughts as they come into the head what else adopt and focus we have in the problem statement but perhaps those paragraphs should be re-jigged for up here multithreaded functional programming with linear typing for session types a number of these systems use different type systems but they are based on all the lambda calculus fundamentally so to have a generic framework to encode these variants would
 
@@
 
+Roughly can be split into
+
+type systems (pi and lambda)
+
+session type mechanisations
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+First order approach taken - represent bound variables as de Bruijn indices
+In contrast to a higher-order approach
+though the first-order encoding has its disadvantages namely one has to reason about variable substitution since without a higher-order approach this is not handled automatically by the meta-language
+the locally nameless representation is chosen with a few simplifying assumptions regarding variable substitution and closed terms
+
+Engineering Formal Metatheory
+Abstract:
+
+Formalising programming languages within a proof assistant is becoming a
+more widely adopted convention within the research community in lieu of hand
+written proofs.
+
+Review:
+
+Discussing issues in engineering metatheory for programming languages.
+Focus was on the locally nameless representation for binders with names
+for free variables and de Brujin indices for bound variables. There was
+also novel use of cofinite quantification to rules with bindings to increase
+the strength of the induction principles and reduce the amount of renaming
+of bound variables (due to freshness issues) significantly. The theory was
+tested with formalisations of the simply-typed lambda calculus, System F_sub,
+the calculus of constructions. The formalisations followed the POPLMARK
+challenge (type soundness, preservation, progress).
+
 HOAS Coq formalisation using Metatheory library from UPenn.
 Explain why this approach was taken versus intrinsic encoding of the terms
+
+The intrinsic encoding as described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one approach for encoding programming languages based on the lambda calculus.
+The key idea here is that terms are indexed by their type so as to prevent ill-formed terms from being constructed.
+However, in the encoding for GV there are issues with using this encoding: $(1)$ it is not immediately clear how to handle linear contexts using the de Bruijn variable encoding, since in the instrinic setting the environment supports weakening; $(2)$ an instrinic encoding of terms would be complicated by the need to enforce session types occur in certain instances (by use of a predicate); and $(3)$ well-typed terms require extra assumptions about binder freshness which cannot reasonably be expressed as a function type.
+For these reasons, an instrinic approach does not offer much benefit since a well-typed term relation would still need to be defined.
+I chose the .....
+encoding
 

### 2014-11-11 proposal/proposal.tex
Working on a proper summary of Engineering Formal Metatheory.
This paper forms most of the backbone reasoning I'm using in the Coq
development.

@@
 
+Interactive theorem proving and proof assistants has seen an upsurge in recent years.
+Especially within the domain of programming language research, there is a strong emphasis now on providing mechanised proofs accompanying any work in the area.
+
+POPLMARK
+--------
+SPL (+CPL)
+* Most formalisations are informal, making it difficult to extend the development to a larger setting.
+* Mistakes can easily slip into a development nullifying work
+* Considerable manual effort must be expended to update all proofs to support an extension and it is not immediately obvious which definitions may need updated
+* A lot of proof assistants offer a platform with which to study languages rigorously
+WTD
+
 The increasing use of proof assitants and interactive theorem proving in programming language research has spiked....
+
+Most research on session type systems has consisted of an informal presentation of the languages being defined.
+That is, the semantics and proofs are performed manually rather than aided by a proof assistant.
+
 POPLMARK
@@
 
+Here I will briefly distinguish between pi and lambda calculi, since the majority of mechanisations have been in the pi calculus.
+I will then discuss the various systems which have been informally formalised.
+A possible outline of a plan:
+
 System F-pop~\cite{Mazurak:2010:LLT}
@@
 Lolliproc~\cite{Mazurak:2010:LCC}
-PhD thesis~\cite{Mazurak:2013:LPP}
+\citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}
 

### 2014-11-12 proposal/proposal.tex
Wrote up notes on Engineering Formal Metatheory paper.

@@
 --------
-SPL (+CPL)
-* Most formalisations are informal, making it difficult to extend the development to a larger setting.
-* Mistakes can easily slip into a development nullifying work
-* Considerable manual effort must be expended to update all proofs to support an extension and it is not immediately obvious which definitions may need updated
-* A lot of proof assistants offer a platform with which to study languages rigorously
-WTD
 

### 2014-11-12 proposal/proposal.tex
Provisionally finished the introduction draft.

@@
 
+\usepackage{hyperref}
 \usepackage{graphicx}
-\usepackage[style=alphabetic,natbib=true]{biblatex}
+\usepackage[style=alphabetic,natbib=true,maxbibnames=100]{biblatex}
 \addbibresource{proposal.bib}
@@
 
-Interactive theorem proving and proof assistants has seen an upsurge in recent years.
-Especially within the domain of programming language research, there is a strong emphasis now on providing mechanised proofs accompanying any work in the area.
+In programming language research, work on static type systems is usually presented on paper complete with informal proofs of correctness for the system properties.
+Most types systems in the session types community have been developed in this way.
+However, it is more difficult to extend work using this approach as may be unclear how changes affect proofs and one will typically have to re-consider all proofs to be convinced the properties still hold.
+More generally, unassisted formalisms are error-prone to develop and do not provide a firm basis upon which other researchers can experiment; often it will require ``rolling your own'' compiler for the language of interest.
 
-POPLMARK
---------
+Proof assistants such as Coq~\cite{Coq:manual} offer a platform on which to develop programming languages and type systems and associated proofs of properties with strong guarantees of correctness.
+Not only is proving with a proof assistant much less error-prone than the unaided approach, it also eases extension of the system since the proof assistant will highlight parts of proofs that have been affected by changes.
+However, one may have to prove more properties using this approach since this setting does not permit appealing to human intuition.
+For example, alpha-equivalence and freshness and binders are usually assumed in informal presentations, whereas in the mechanised setting one has to explicitly encode and reason about these properties~\cite{Aydemir:2005:MMM, Aydemir:2008:EFM}.
 
-The increasing use of proof assitants and interactive theorem proving in programming language research has spiked....
+Interactive theorem proving and proof assistants has seen an upsurge in use in recent years.
+Especially within the domain of programming language research, there is a strong emphasis on providing mechanised proofs accompanying work in the area.
+Indeed, \citeauthor{Aydemir:2005:MMM} have published a series of challenges on mechanising programming languages aimed at providing a starting point for comparing different representation techniques~\cite{Aydemir:2005:MMM}.
+Another aim is to provide reusable libraries for common reasoning needed across programming language developments, e.g.
+handling of typing environments.
+This effort has resulted in a number of different approaches for representing programming languages within proof assistants.
+For instance, \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} present a representation of lambda calculi where bound variables are represented as \textit{de Bruijn indices}~\footnote{A de Bruijn index is a number used to represent a bound variable which indicates the position of its binder (starting from zero for the innermost binder).
+For example, $\lambda~(\lambda~1)$ corresponds to $\lambda x.
+\lambda y.
+x$, i.e.
+the constant function.} and free variables are represented as named terms; the \textit{locally nameless} representation.
+Their work resulted in a library for handling commonly occurring aspects in programming lanuage metatheory which has been used since in other developments~\cite{Park:2014:MMW}.
+Resuable libaries reduce the proof engineering effort needed for later works, in contrast to the pen-and-paper approach, and some proof assistants (as is the case with Coq) allow one to extract a program from the development which can provide a typechecker or compiler for the language formalised~\footnote{Curry-Howard strikes again; a proof of decidability of typechecking corresponds to a program implementing a typechecker for the language.}
 
-Most research on session type systems has consisted of an informal presentation of the languages being defined.
+%% Most research on session type systems has consisted of an informal presentation of the languages being defined.
 That is, the semantics and proofs are performed manually rather than aided by a proof assistant.
 
-POPLMARK
-save on proof engineering effort as is championed by the POPLMARK challenge aimed at reducing the effort for researchers to code programming language metatheory it provides a library needed to perform various things define various parts of programming languages viewed to be difficult to formalise including handling binders and freshness they also mention handling of linear typing but renege on actually formulating this as part of their challenge they insist it is a future work here is the future
-Metatheory library
+%% POPLMARK
+%% save on proof engineering effort as is championed by the POPLMARK challenge aimed at reducing the effort for researchers to code programming language metatheory it provides a library needed to perform various things define various parts of programming languages viewed to be difficult to formalise including handling binders and freshness they also mention handling of linear typing but renege on actually formulating this as part of their challenge they insist it is a future work here is the future
+%% Metatheory library
 
-construct an argument for the use of a proof assistant and interactive theorem proving technologies to formalise a framework for developing type systems involving session types that are based on lambda calculi so that we can use such a development to reason about a variety of type systems while minimising proof effort tie this into an introduction into the subfield of session types and what they are good for essentially motivating our project in this space additionally we can introduce the reader to the discipline of interactivetheorem proving and formal methods more generally as a way to formalise programming languages highlighting the benefits of reusability increased
-
-correctness and something what else could we say i think it would appropriate
-to mention that the session types studied so far include both pi and lambda
-calculi and it may be beneficil if the reader was informed also the pi
-calculus paradigms do not fit well into the existing mainstream programming
-language paradigms this seems like it should be in the statement of the problem section down below also i should read more of jtod's slides because this writing just keep writing malarky might just be what i need to pound out the pages because it eliminates thoughts as they come into the head what else adopt and focus we have in the problem statement but perhaps those paragraphs should be re-jigged for up here multithreaded functional programming with linear typing for session types a number of these systems use different type systems but they are based on all the lambda calculus fundamentally so to have a generic framework to encode these variants would
+%% correctness and something what else could we say i think it would appropriate
+%% to mention that the session types studied so far include both pi and lambda
+%% calculi and it may be beneficil if the reader was informed also the pi
+%% calculus paradigms do not fit well into the existing mainstream programming
+%% language paradigms this seems like it should be in the statement of the problem section down below also i should read more of jtod's slides because this writing just keep writing malarky might just be what i need to pound out the pages because it eliminates thoughts as they come into the head what else adopt and focus we have in the problem statement but perhaps those paragraphs should be re-jigged for up here multithreaded functional programming with linear typing for session types a number of these systems use different type systems but they are based on all the lambda calculus fundamentally so to have a generic framework to encode these variants would
 
@@
 of bound variables (due to freshness issues) significantly. The theory was
-tested with formalisations of the simply-typed lambda calculus, System F_sub,
+tested with formalisations of the simply-typed lambda calculus,
+System $F_{<:}$,
 the calculus of constructions. The formalisations followed the POPLMARK

### 2014-11-12 proposal/proposal.tex
Tidy up some proposal sections.

@@
 
-clearly state the problem to be addressed in your forthcoming project. Explain
-why it would be worthwhile to solve this problem.
-
 The aim of this project is to develop a framework for formalising lambda
@@
 formalised with little extra proof effort. Inspired by previous work on a
-mechanised framework for pi calculus type systems~\cite{Gay:2001:FFP}, we
+mechanised framework for pi calculus type systems~\cite{Gay:2001:FFP} we
 shift to the lambda calculus to provide a basis on which session types for
@@
 particular system.
+The framework will be heavily based on calculi developed by \citeauthor{Walder:2012}~\cite{Wadler:2012} describing a relationship between classical linear logic (based on previous work by \citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI}) and a functional session-based language (based on previous work by \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST}).
+
+These ideas can be expanded by considering the recent work by \citeauthor{Lindley:2014:SAP} \cite{Lindley:2014:SAP} who extend GV by \cite{Wadler:2012} to HGV, supporting session forwarding, replication and polymorphism.
 
@@
 session polymorphism, session subtyping, and permitting different forms of
-aliasing within the type system. The majority of previous session type systems
-for mainstream programming languages has focussed exclusively on linear type
-systems. One disadvantage of a linear type system is the need to re-bind
-objects after each operation~\cite{Gay:2010:LAST}. Principally, we intend to
-instantiate the framework with a type system that provides the benefits of
-aliasing with the guarantees of linearity, termed ``adoption and focus''. The
-``adoption and focus'' type system is non-trivial requiring special language
-features, e.g. to manage capabilities which can be thought as a form of object
-liveness mechanism, so the framework design must support such flexibility by,
-e.g allowing the typing judgements to be extended such that the capabilities
-can be included in soundness results. The instantiation will demonstrate that
-the framework supports a wide range of languages and type systems providing
-researchers the ability to harness the power of formal verification in the
-design, implementation and documentation of session-based type systems.
+aliasing within the type system.
+Initially, it will be interesting to extend the work of \citeauthor{Walder:2012} to consider session polymorphism corresponding to the quantifiers in the CP process calculus
+
+The majority of previous session type systems for mainstream programming languages have focussed exclusively on linear type systems.
+One disadvantage of a linear type system is the need to re-bind objects after each operation~\cite{Gay:2010:LAST}.
+Principally, we intend to instantiate the framework with a type system that provides the benefits of aliasing with the guarantees of linearity, termed ``adoption and focus''.
+The ``adoption and focus'' type system is non-trivial requiring special language features, e.g.
+to manage capabilities which can be thought as a form of object liveness mechanism, so the framework design must support such flexibility by, e.g allowing the typing judgements to be extended such that the capabilities can be included in soundness results.
+The instantiation will demonstrate that the framework supports a wide range of languages and type systems providing researchers the ability to harness the power of formal verification in the design, implementation and documentation of session-based type systems.
 

### 2014-11-14 proposal/proposal.tex
Working on proposal background/approach/work plan.

@@
 particular system.
-The framework will be heavily based on calculi developed by \citeauthor{Walder:2012}~\cite{Wadler:2012} describing a relationship between classical linear logic (based on previous work by \citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI}) and a functional session-based language (based on previous work by \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST}).
+The framework will be heavily based on calculi developed by \citeauthor{Walder:2012}~\cite{Wadler:2012} describing a relationship between classical linear logic (CP) and a functional language with session types (GV).
+GV is a minimal functional language supporting linearity so I believe such a basis will be general enough to support reasoning about a large class of functional languages with similar base properties.
 
-These ideas can be expanded by considering the recent work by \citeauthor{Lindley:2014:SAP} \cite{Lindley:2014:SAP} who extend GV by \cite{Wadler:2012} to HGV, supporting session forwarding, replication and polymorphism.
-
-From the generic framework, a number of directions can be explored including
+From this framework, a number of directions can be explored including
 session polymorphism, session subtyping, and permitting different forms of
 aliasing within the type system.
-Initially, it will be interesting to extend the work of \citeauthor{Walder:2012} to consider session polymorphism corresponding to the quantifiers in the CP process calculus
+The recent work by \citeauthor{Lindley:2014:SAP} \cite{Lindley:2014:SAP}, who extend GV to HGV supporting session forwarding, replication and polymorphism, would be a interesting place to start when looking to add these extensions.
 
-The majority of previous session type systems for mainstream programming languages have focussed exclusively on linear type systems.
+Additionally, the majority of previous session type systems for mainstream programming languages have focussed exclusively on linear type systems.
 One disadvantage of a linear type system is the need to re-bind objects after each operation~\cite{Gay:2010:LAST}.
-Principally, we intend to instantiate the framework with a type system that provides the benefits of aliasing with the guarantees of linearity, termed ``adoption and focus''.
+We could look to extend the GV type system tho provide the benefits of aliasing with the guarantees of linearity, termed ``adoption and focus'' as introduced by~\citeauthor{Fahndrich:2002} \cite{Fahndrich:2002}.
 The ``adoption and focus'' type system is non-trivial requiring special language features, e.g.
-to manage capabilities which can be thought as a form of object liveness mechanism, so the framework design must support such flexibility by, e.g allowing the typing judgements to be extended such that the capabilities can be included in soundness results.
-The instantiation will demonstrate that the framework supports a wide range of languages and type systems providing researchers the ability to harness the power of formal verification in the design, implementation and documentation of session-based type systems.
+to manage capabilities which can be thought as a form of object liveness mechanism, so the GV typing judgements must be extended such that the capabilities can be included in soundness results.
+A translation to CP must also be defined for the ``adopt'' and ``focus'' constructs.
+
+These extensions will argue that the framework supports a wide range of languages and type systems, providing researchers the ability to harness the power of formal verification in the design, implementation and documentation of session-based type systems.
 
@@
 
-present an overview of relevant previous work including articles, books, and
-existing software products. Critically evaluate the strengths and weaknesses
-of the previous work.
+%% present an overview of relevant previous work including articles, books, and
+%% existing software products. Critically evaluate the strengths and weaknesses
+%% of the previous work.
 
-Here I will briefly distinguish between pi and lambda calculi, since the majority of mechanisations have been in the pi calculus.
-I will then discuss the various systems which have been informally formalised.
-A possible outline of a plan:
+The background for this project can be categorised roughly as work on session types (sub-divided into pi and lambda calculi variants, \S \ref{sec:pis} and \S \ref{sec:lam}, respectively) and work on formal verification using proof assistants (particularly focussed on programming language mechanisation).
+Finally, some recent work has merged these two disciplines by verifying session-based pi calculus systems (\S \ref{sec:asts})
 
+\subsection{Session-based Type Systems and Linear Logic}\label{sec:sts}
+
+\subsubsection{In the \pi Calculus}\label{sec:pis}
+
+So we have work on pi calculus and linear logic by \citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI}.
+
+\subsubsection{In the \lambda Calculus}\label{sec:lam}
+
+Simon Gay inter-process thingL
+Previous work by the same authors ... had nonstandard operational semantics due to handling of channels.
+
+The work on a linear functional language with session types by \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the current project from the body of work on session-based $\lambda$ calculus type systems.
+They provide a paper proofs of their systems type (and runtime?) safety.
+Given the systems non-trivial features such as recursive session types (?) and buffered channels it will more difficult to extend these proofs manually.
+A mechanised version would provide stronger guarantees of correctness and allow others to more easily alter proofs as a result of changes made to the language.
+
+\subsubsection{Fusion}
+
+A number of systems incorporate both a functional language surface syntax with and translation to a more primitive process calculus.
+\citeauthor{Wadler:2012}'s recent work~\cite{Wadler:2012} fuses together a process calculus, CP, based on earlier work, on a relationship between the $\pi$ calculus and linear logic, by \citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by \citeauthor{Gay:2010:LAST}.
+The work describes a continuation-passing style (CPS) translation from GV to CP treating session types as propositions in a classical linear logic.
+
+Other work similar in vain to ... is Lolliproc~\cite{Mazurak:2010:LCC} and classical extension to second-order logic by \citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}.
+
+\subsection{Mechanising Programming Language Metatheory}
+
+\subsubsection{Applications to Session-based Type Systems}\label{sec:asts}
+
+session polymorphism formalised in Coq
+
+\subsection{Formal Tidbits}
+
+Simon's pi calculus framework may be worth mentioning here
 System F-pop~\cite{Mazurak:2010:LLT}
-session polymorphism formalised in Coq, Simon's Isabelle/HOL
-development.
-Philip Wadler's Propositions as Sessions~\cite{Wadler:2012}
-Simon J. Gay and Vasco Vasconcelos (LAST)~\cite{Gay:2010:LAST}
-Lolliproc~\cite{Mazurak:2010:LCC}
-\citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}
-
-Roughly can be split into
-
-type systems (pi and lambda)
-
-session type mechanisations
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Proposed Approach}
+\section{Proposed Approach}\label{sec:approach}
 
@@
 the locally nameless representation is chosen with a few simplifying assumptions regarding variable substitution and closed terms
+
+One may need to avoid certain tricky formulations.
+For instance, instead of providing n-ary branch and choice in GV, we provide only binary versions of these operators to simplify the development.
+This change is not restrictive however, since CP itself provides only binary versions of $\par$ and $&$.
+Likewise, on a more technical note, we almagmate all types into one inductive definition and provide a predicate for restricting propositions to consider only valid session types.
+\citeauthor{Wadler:2012} defines the GV grammar as a mutually recursive; session types are types which may contain types as subcomponents (arguments to send, for example).
+Unfortunately, handling mutually inductive definitions in Coq can be quite involved; requiring one to rely on the Coq system being able to provide a stronger mutual induction principle or defining one manual.
+Such a definition is possible, but it complicates elimination via the induction principle.
+( since the standard induction tactic will not work?)
 
@@
 
-HOAS Coq formalisation using Metatheory library from UPenn.
-Explain why this approach was taken versus intrinsic encoding of the terms
-
 The intrinsic encoding as described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one approach for encoding programming languages based on the lambda calculus.
@@
 
-I need to perform some formalisation of the system. I should identify the key
-issues (linear typing, bindings (as claimed in the literature), etc).
+I aim to have completed the GV and CP definitions complete with reduction and commuting conversion rules for CP by the end of December.
+This will require definition of infrastructure for handling binders for both GV, CP channels and CP propositional variables; I hope to be able to generalise as much as possible with the help of the UPenn Metatheory library.
+
+It remains to define the translation between GV and CP which I envisage will be the bulk of the work, requiring several rules on typing derivations; I envisage being able to define these as inductive datatype constructors.
 

### 2014-11-14 proposal/proposal.tex
Finished review of piDILL paper.

@@
 
-So we have work on pi calculus and linear logic by \citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI}.
+\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} provide a $\pi$ calculus type system with session types.
+They use a labelled transition system to represent process reduction with observable actions on channels.
+The relationship with intuitionistic linear logic provides absence of deadlocks in session-based communication.
+There is structural and congruence rules for processes.
 
@@
 The work describes a continuation-passing style (CPS) translation from GV to CP treating session types as propositions in a classical linear logic.
+The process calculus is slightly different to that described in the $\pi$DILL apart from being classical in nature the constructs for server replication are different, CP provides a weakening and contraction semantics whereas this is encode in one rule for process reduction by caires
+Wadler also dispenses with the structural rules; subsumed by the Assoc and Swap typing derivations and the cut reduction (double check why he does this)
 

### 2014-11-14 proposal/proposal.tex
Adding discussion on lambda calculus based approaches.

@@
 \usepackage{graphicx}
+\usepackage{cmll}
 \usepackage[style=alphabetic,natbib=true,maxbibnames=100]{biblatex}
@@
 particular system.
-The framework will be heavily based on calculi developed by \citeauthor{Walder:2012}~\cite{Wadler:2012} describing a relationship between classical linear logic (CP) and a functional language with session types (GV).
+The framework will be heavily based on calculi developed by \citeauthor{Wadler:2012}~\cite{Wadler:2012} describing a relationship between classical linear logic (CP) and a functional language with session types (GV).
 GV is a minimal functional language supporting linearity so I believe such a basis will be general enough to support reasoning about a large class of functional languages with similar base properties.
@@
 
-The background for this project can be categorised roughly as work on session types (sub-divided into pi and lambda calculi variants, \S \ref{sec:pis} and \S \ref{sec:lam}, respectively) and work on formal verification using proof assistants (particularly focussed on programming language mechanisation).
-Finally, some recent work has merged these two disciplines by verifying session-based pi calculus systems (\S \ref{sec:asts})
+The background for this project can be categorised roughly as work on session types (sub-divided into $\pi$ and $\lambda$-calculi variants, \S \ref{sec:pis} and \S \ref{sec:lam}, respectively) and work on formal verification using proof assistants (particularly focussed on programming language mechanisation).
+Finally, some recent work has merged these two disciplines by verifying a session-based $\pi$-calculus system (\S \ref{sec:asts}).
 
@@
 
-\subsubsection{In the \pi Calculus}\label{sec:pis}
+\subsubsection{In the $\pi$-calculus}\label{sec:pis}
 
-\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} provide a $\pi$ calculus type system with session types.
+\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} present a connection between $\pi$-calculus type system with session types and intuitionistic linear logic.
+They equate linear propositions with session types of the process calculus via a bidirectional translation relation.
 They use a labelled transition system to represent process reduction with observable actions on channels.
-The relationship with intuitionistic linear logic provides absence of deadlocks in session-based communication.
-There is structural and congruence rules for processes.
+They represent dual intuitionistic linear logic as a sequent calculus providing logical connectives which have session-based interpretations.
+Process reduction steps are equated with proof reduction via cut elimination.
+The relationship with intuitionistic linear logic provides absence of deadlocks in session-based communication for the process calculus.
 
-\subsubsection{In the \lambda Calculus}\label{sec:lam}
+\subsubsection{In the $\lambda$-calculus}\label{sec:lam}
 
-Simon Gay inter-process thingL
-Previous work by the same authors ... had nonstandard operational semantics due to handling of channels.
+\citeauthor{Gay:2008:STI}~\cite{Gay:2008:STI} provide a $\lambda$-calculus based language with session types for channel primitives.
+The type system is nonstandard, featuring an additional environment for tracking the session types of channels; these types are modified during term reduction to reflect communication.
+The extra channel environment is provided to support possible aliasing of channels by treating channel variables as a pointers or references into the channel environment.
+So, for a channel $c$ of session type $S$, a variable $x$ referring to this channel would be declared to have type $Chan~c$.
+The typing judgements and term reduction relation must be augmented to handle these environments.
+Unfortunately, a side-effect of this is that function types must specify the pre and post-states of channel environments and therefore, the type of arguments cannot work up to alpha-equivalence of channel identifiers.
+That is, if two channels $c$ and $d$ have the same session type, one cannot define a function to operate on variables of either channel type since the function definition will explicitly mention the identifier for the channel (in this case either, $Chan~c$ or $Chan~d$).
+However, the system does provide an attractive non-linear channel type which has been neglected in favour of a linear treatment of channels in more recent work~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2012}.
+Certainly, it is the case that an aliased approach would fit better with mainstream programming languages and perhaps the ``adoption and focus'' system~\cite{Fahndrich:2002} could offer some insights.
 
-The work on a linear functional language with session types by \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the current project from the body of work on session-based $\lambda$ calculus type systems.
-They provide a paper proofs of their systems type (and runtime?) safety.
-Given the systems non-trivial features such as recursive session types (?) and buffered channels it will more difficult to extend these proofs manually.
+The work on a linear functional language with session types by \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the current project from the body of work on session-based $\lambda$-calculus type systems.
+They provide paper proofs of their systems type (and runtime?) safety.
+Given the systems non-trivial features such as recursive session types and buffered channels for supporting asynchronous communication it will more difficult to extend these proofs manually.
 A mechanised version would provide stronger guarantees of correctness and allow others to more easily alter proofs as a result of changes made to the language.
@@
 A number of systems incorporate both a functional language surface syntax with and translation to a more primitive process calculus.
-\citeauthor{Wadler:2012}'s recent work~\cite{Wadler:2012} fuses together a process calculus, CP, based on earlier work, on a relationship between the $\pi$ calculus and linear logic, by \citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by \citeauthor{Gay:2010:LAST}.
+\citeauthor{Wadler:2012}'s recent work~\cite{Wadler:2012} fuses together a process calculus, CP, based on earlier work, on a relationship between the $\pi$-calculus and linear logic, by \citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by \citeauthor{Gay:2010:LAST}.
 The work describes a continuation-passing style (CPS) translation from GV to CP treating session types as propositions in a classical linear logic.
@@
 For instance, instead of providing n-ary branch and choice in GV, we provide only binary versions of these operators to simplify the development.
-This change is not restrictive however, since CP itself provides only binary versions of $\par$ and $&$.
+This change is not restrictive however, since CP itself provides only binary versions of plus ($\oplus$) and with ($\with$) constructs.
 Likewise, on a more technical note, we almagmate all types into one inductive definition and provide a predicate for restricting propositions to consider only valid session types.

### 2014-11-14 proposal/proposal.tex
Tweaking the pi-calculus section.

@@
 The relationship with intuitionistic linear logic provides absence of deadlocks in session-based communication for the process calculus.
+The correspondence of session types to linear logic propositions is an important step in providing a solid logical foundation for a concurrent type theory; in the same way ``sequential'' type theory ($\lambda$-calculi) profitted from a logical perspective.
+However, for session types to become a more usable tool for programming distributed communication systems (the ultimate goal, of course, is to write more correct programs), we cannot be expected to program directly in the $\pi$-calculus; a more mainstream programming paradigm must be built upon these solid logical foundations.
+With that in mind, enter the $\lambda$-calculus!
 

### 2014-11-15 proposal/proposal.tex
Some more additions to the background and approach.

@@
 
-The background for this project can be categorised roughly as work on session types (sub-divided into $\pi$ and $\lambda$-calculi variants, \S \ref{sec:pis} and \S \ref{sec:lam}, respectively) and work on formal verification using proof assistants (particularly focussed on programming language mechanisation).
+The background for this project can be categorised roughly as previous work on session-based type systems (sub-divided into $\pi$ and $\lambda$-calculi variants, \S \ref{sec:pis} and \S \ref{sec:lam}, respectively) and previous work on formal verification using proof assistants (particularly focussed on programming language mechanisation).
 Finally, some recent work has merged these two disciplines by verifying a session-based $\pi$-calculus system (\S \ref{sec:asts}).
@@
 A number of systems incorporate both a functional language surface syntax with and translation to a more primitive process calculus.
-\citeauthor{Wadler:2012}'s recent work~\cite{Wadler:2012} fuses together a process calculus, CP, based on earlier work, on a relationship between the $\pi$-calculus and linear logic, by \citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by \citeauthor{Gay:2010:LAST}.
+We shall focus on two such systems presented recently which have a connection to linear logic.
+
+Lolliproc~\cite{Mazurak:2010:LCC} provides a functional language with concurrency primitives as a surface syntax on top of an underlying process calculus.
+The process calculus provides classial linear logical interpretation for the functional language, utilising double negation elimination to provide session type duality.
+Concurrency is provided by control operators which, respectively, spawn a child process and wait for a child process to completed.
+The spawning process creates a channel between parent and child process; a channel is represented as a continuation and the type denotes the communication protocol (i.e.
+session type) between parent and child.
+The surface syntax is an effective abstraction layer for the programmer since it does not expose the underlying process calculus.
+The system has proofs for type safety, strong normalisation and confluence.
+Confluence provides the guarantee that no race conditions can occur, strong normalisation prohibits nonterminating programs and type safety assures deadlock freedom.
+Deadlock-freedom is implied by the acyclic communication graphs (progress of process reduction) which occur between parallel processes; follows from permitting only one channel between two halves of parallel composition construct in $\pi$-calculus.
+Most of the development is supported by Coq proof scripts with remaining issues due to reasoning about communication graphs.
+The development does not provide primitive send and receives; instead, these are encoded as linear functions.
+%% and classical extension to second-order logic by \citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}.
+
+\citeauthor{Wadler:2012}'s recent work~\cite{Wadler:2012} brings together a process calculus, CP, based on the earlier work by \citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by \citeauthor{Gay:2010:LAST}.
 The work describes a continuation-passing style (CPS) translation from GV to CP treating session types as propositions in a classical linear logic.
-The process calculus is slightly different to that described in the $\pi$DILL apart from being classical in nature the constructs for server replication are different, CP provides a weakening and contraction semantics whereas this is encode in one rule for process reduction by caires
-Wadler also dispenses with the structural rules; subsumed by the Assoc and Swap typing derivations and the cut reduction (double check why he does this)
+The process calculus is slightly different to that described in the $\pi$DILL~\cite{Caires:2010:STI}, apart from being classical in nature, the constructs for server replication are modified; CP provides a weakening and contraction rules whereas these are encode in one rule for process reduction by \citeauthor{Caires:2010:STI}.
+Additionally, rather than a sequent calculus, \citeauthor{Walder:2012} uses a natural deduction style for defining the process calculus.
+As presented, GV does not support all features of CP and thus the translation is one way (GV to CP).
+\citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to provide support for polymorphism and replication, following a similar CPS translation scheme.
 
-Other work similar in vain to ... is Lolliproc~\cite{Mazurak:2010:LCC} and classical extension to second-order logic by \citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}.
+%% Wadler also dispenses with the structural rules; subsumed by the Assoc and Swap typing derivations and the cut reduction (double check why he does this)
 
@@
 
+metatheory/poplmark papers.
+
 \subsubsection{Applications to Session-based Type Systems}\label{sec:asts}
 
-session polymorphism formalised in Coq
+%\citeauthor{
+%An extensible approach to session polymorphism
+\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based type system providing session polymorphism.
+The system is more general than session polymorphism via subtyping~\cite{Gay:2005:SST}.
+The authors mechanise their type system and its properties in the Coq proof assistant.
+However, the system mechanised is restricted to the $\pi$-calculus and the main focus is proving soundness properties for the deduction rules allowing polymorphism on input, rather than a more general basis for studying programming with session types.
 
@@
 
-Simon's pi calculus framework may be worth mentioning here
-System F-pop~\cite{Mazurak:2010:LLT}
+While not related to session-based type systems
+\citeauthor{Gay:2001:FFP} provides a $\pi$-calculus framework in Isabelle/HOL
+had the aim of mechanising in mind for session type support with linear and non-linear type system design
+name introduction in proof development from custome user-defined tactics are an issue.
+in coq one can specify the names directly and allow custom tactics to receive arguments for names.
+generally setting the hypotheses up correctly will provide suitable names automatically and it is only during destruction of some hypotheses that Coq may generate unsuitable names
+
+% System F-pop~\cite{Mazurak:2010:LLT}
 
@@
 your proposed approach is feasible, but identify any risks.
+
+I intend to mechanise the calculi described by \citeauthor{Wadler:2012}~\cite{Wadler:2012} using the Coq proof assistant.
 

### 2014-11-15 proposal/proposal.tex
Completed draft of background: onto approach + work plan.

@@
 
+% Include Coq lstlisting language definition and things upon which it depends.
+\usepackage{macros}
+
 \usepackage{hyperref}
+\usepackage{listings}
+\usepackage{lstcoq}
 \usepackage{graphicx}
@@
 The process calculus provides classial linear logical interpretation for the functional language, utilising double negation elimination to provide session type duality.
+The surface syntax is an effective abstraction layer for the programmer since it does not expose the underlying process calculus.
+
 Concurrency is provided by control operators which, respectively, spawn a child process and wait for a child process to completed.
@@
 session type) between parent and child.
-The surface syntax is an effective abstraction layer for the programmer since it does not expose the underlying process calculus.
-The system has proofs for type safety, strong normalisation and confluence.
+
+\citeauthor{Mazurak:2010:LCC} present proofs for type safety, strong normalisation and confluence.
 Confluence provides the guarantee that no race conditions can occur, strong normalisation prohibits nonterminating programs and type safety assures deadlock freedom.
@@
 The process calculus is slightly different to that described in the $\pi$DILL~\cite{Caires:2010:STI}, apart from being classical in nature, the constructs for server replication are modified; CP provides a weakening and contraction rules whereas these are encode in one rule for process reduction by \citeauthor{Caires:2010:STI}.
-Additionally, rather than a sequent calculus, \citeauthor{Walder:2012} uses a natural deduction style for defining the process calculus.
+Additionally, rather than a sequent calculus, \citeauthor{Wadler:2012} uses a natural deduction style for defining the process calculus.
 As presented, GV does not support all features of CP and thus the translation is one way (GV to CP).
@@
 
-While not related to session-based type systems
-\citeauthor{Gay:2001:FFP} provides a $\pi$-calculus framework in Isabelle/HOL
-had the aim of mechanising in mind for session type support with linear and non-linear type system design
-name introduction in proof development from custome user-defined tactics are an issue.
-in coq one can specify the names directly and allow custom tactics to receive arguments for names.
-generally setting the hypotheses up correctly will provide suitable names automatically and it is only during destruction of some hypotheses that Coq may generate unsuitable names
+While not related to session-based type systems \citeauthor{Gay:2001:FFP} provides a $\pi$-calculus framework mechanised in Isabelle/HOL.
+The aim is to build a mechanisaton of session-based type systems on top of the $\pi$-calculus framework.
+The use of de Bruijn indices for both bound and free variables causes issues with variable substitution often requiring permutation of typing environments (see \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} for a discussion on binder representations) and lifting of free variables during substitution; we avoid this issue in our representation choice (\S \ref{sec:approach}).
+
+%% Channel endpoints are represented as a polarity on a name. The polarities are not constrained to have the same type as each other but must be unique in the typing enviornment.
+
+The introduction of names for hypotheses during execution of custom tactics are an issue in the proof development.
+These names are automatically generated by the Isabelle/HOL system which creates a dependency between the names chosen and the tactic.
+Thus, one must be careful to use the exact names expected by the tactic.
+In the Coq proof assistant, defining custom tactics can be achieved using the Ltac language~\cite{Delahaye:2000:TLS}.
+Among other things, Ltac supports for pattern matching on hypotheses and goal forms.
+Thus, if we wished to pattern match on an inductive type describing variables, we could do so without mentioning the actual variable name as follows:
+
+\begin{coq}
+Ltac my_tactic :=
+  match goal with
+  | [v: var |- _] => ... tactics here possibly mentioning `v' hypothesis ...
+  end.
+\end{coq}
 
@@
 your proposed approach is feasible, but identify any risks.
+
+ %% Additionally, we make some simplifying assumptions about substitution for bound variables to avoid the lifting constructs defined by~\citeauthor{Gay:2001:FFP} in this work.
 

### 2014-11-16 proposal/proposal.tex
Finished background; moved POPLMARK discussion to BG survey.

@@
 \usepackage{graphicx}
+\usepackage{xspace}
 \usepackage{cmll}
@@
 
+\newcommand{\fpop}{System F${}^\circ$\xspace}
+
 % other potentially useful packages
@@
 
-\title{Formalisation of Session-based Lambda Calculus Type Systems}
+\title{Mechanisation of Session-based Lambda Calculus Type Systems}
 \author{Craig McLaughlin}
@@
 However, one may have to prove more properties using this approach since this setting does not permit appealing to human intuition.
-For example, alpha-equivalence and freshness and binders are usually assumed in informal presentations, whereas in the mechanised setting one has to explicitly encode and reason about these properties~\cite{Aydemir:2005:MMM, Aydemir:2008:EFM}.
-
-Interactive theorem proving and proof assistants has seen an upsurge in use in recent years.
-Especially within the domain of programming language research, there is a strong emphasis on providing mechanised proofs accompanying work in the area.
-Indeed, \citeauthor{Aydemir:2005:MMM} have published a series of challenges on mechanising programming languages aimed at providing a starting point for comparing different representation techniques~\cite{Aydemir:2005:MMM}.
-Another aim is to provide reusable libraries for common reasoning needed across programming language developments, e.g.
-handling of typing environments.
-This effort has resulted in a number of different approaches for representing programming languages within proof assistants.
-For instance, \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} present a representation of lambda calculi where bound variables are represented as \textit{de Bruijn indices}~\footnote{A de Bruijn index is a number used to represent a bound variable which indicates the position of its binder (starting from zero for the innermost binder).
-For example, $\lambda~(\lambda~1)$ corresponds to $\lambda x.
-\lambda y.
-x$, i.e.
-the constant function.} and free variables are represented as named terms; the \textit{locally nameless} representation.
-Their work resulted in a library for handling commonly occurring aspects in programming lanuage metatheory which has been used since in other developments~\cite{Park:2014:MMW}.
-Resuable libaries reduce the proof engineering effort needed for later works, in contrast to the pen-and-paper approach, and some proof assistants (as is the case with Coq) allow one to extract a program from the development which can provide a typechecker or compiler for the language formalised~\footnote{Curry-Howard strikes again; a proof of decidability of typechecking corresponds to a program implementing a typechecker for the language.}
-
-%% Most research on session type systems has consisted of an informal presentation of the languages being defined.
-That is, the semantics and proofs are performed manually rather than aided by a proof assistant.
-
-%% POPLMARK
-%% save on proof engineering effort as is championed by the POPLMARK challenge aimed at reducing the effort for researchers to code programming language metatheory it provides a library needed to perform various things define various parts of programming languages viewed to be difficult to formalise including handling binders and freshness they also mention handling of linear typing but renege on actually formulating this as part of their challenge they insist it is a future work here is the future
-%% Metatheory library
-
-%% correctness and something what else could we say i think it would appropriate
-%% to mention that the session types studied so far include both pi and lambda
-%% calculi and it may be beneficil if the reader was informed also the pi
-%% calculus paradigms do not fit well into the existing mainstream programming
-%% language paradigms this seems like it should be in the statement of the problem section down below also i should read more of jtod's slides because this writing just keep writing malarky might just be what i need to pound out the pages because it eliminates thoughts as they come into the head what else adopt and focus we have in the problem statement but perhaps those paragraphs should be re-jigged for up here multithreaded functional programming with linear typing for session types a number of these systems use different type systems but they are based on all the lambda calculus fundamentally so to have a generic framework to encode these variants would
+For example, alpha-equivalence and freshness of binders are usually assumed in informal presentations, whereas in the mechanised setting one has to explicitly encode and reason about these properties~\cite{Aydemir:2005:MMM, Aydemir:2008:EFM}.
 
@@
 
-metatheory/poplmark papers.
+Interactive theorem proving and proof assistants have seen an upsurge in use in recent years.
+Especially within the domain of programming language research, there is a strong emphasis on providing mechanised proofs accompanying work in the area.
+Indeed, \citeauthor{Aydemir:2005:MMM} have published a series of challenges on mechanising programming languages aimed at providing a starting point for comparing different representation techniques~\cite{Aydemir:2005:MMM}.
+Another aim is to provide reusable libraries for common reasoning needed across programming language developments, e.g.
+handling of typing environments.
+This effort has resulted in a number of different approaches for representing programming languages within proof assistants.
+For instance, \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} present a representation of lambda calculi where bound variables are represented as \textit{de Bruijn indices}~\footnote{A de Bruijn index is a number used to represent a bound variable which indicates the position of its binder (starting from zero for the innermost binder).
+For example, $\lambda~(\lambda~1)$ corresponds to $\lambda x.
+\lambda y.
+x$, i.e.
+the constant function.} and free variables are represented as named terms; the \textit{locally nameless} representation.
+Their work resulted in a library for handling commonly occurring aspects in programming lanuage metatheory which has been used since in other developments~\cite{Park:2014:MMW}.
+Resuable libaries reduce the proof engineering effort needed for later works, in contrast to the pen-and-paper approach, and some proof assistants (as is the case with Coq) allow one to extract a program from the development which can provide a typechecker or compiler for the language formalised~\footnote{Curry-Howard strikes again; a proof of decidability of typechecking corresponds to a program implementing a typechecker for the language.}
 
-\subsubsection{Applications to Session-based Type Systems}\label{sec:asts}
+Some recent mechanisation effort focuses on providing a basis for studying linearity within type systems.
+This work is of interest to the current project since most session-based type systems assume a strictly linear type system which requires re-binding of channel identifiers, such as in the functional setting of \citeauthor{Gay:2010:LAST}.
+As discussed previously, it would be interesting to study the system described by~\citeauthor{Fahndrich:2002}~\cite{Fahndrich:2002}, yet the system is specific to their ``adopt'' and ``focus'' constructs and is not mechanised.
+\citeauthor{Mazurak:2010:LLT} present an extension to System F with a notion describing types as having linear or non-linear \textit{kind}~\cite{Mazurak:2010:LLT}.
+This extension is motivated by previous work on incorporating linearity into type systems.
+Overall, previous attempts either hamper the inclusion of desirable programming features (such as polymorphism), or do not adequately reflect non-linearity as the common case leading to awkward programming.
+\fpop avoids these deficiencies by categorising types into kinds whilst maintaining similar semantics to System F.
+Mechanised proofs for type soundness and parametricity are presented, and the semantics allow non-linear types to be treated as linear.
+Examples show the system can provide a wide range of permissions on interfaces, from full exclusive access to read-only shared references.
+It would be interesting to see if one could encode the ``adopt'' and ``focus'' constructs in this system.
+The ``adopt'' rule cannot be encoded similarly to the ShareRef example~\cite{Mazurak:2010:LLT}.
+It's not immediately clear how to encode the ``focus'' rule within the present semantics, as one needs to prevent non-linear access to its argument implying knowledge of aliasing within the typing judgements; absence of aliasing is not identified with linear kinded types in \fpop.
 
-%\citeauthor{
-%An extensible approach to session polymorphism
-\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based type system providing session polymorphism.
-The system is more general than session polymorphism via subtyping~\cite{Gay:2005:SST}.
-The authors mechanise their type system and its properties in the Coq proof assistant.
-However, the system mechanised is restricted to the $\pi$-calculus and the main focus is proving soundness properties for the deduction rules allowing polymorphism on input, rather than a more general basis for studying programming with session types.
-
-\subsection{Formal Tidbits}
-
-While not related to session-based type systems \citeauthor{Gay:2001:FFP} provides a $\pi$-calculus framework mechanised in Isabelle/HOL.
+While not related to programming language metatheory, \citeauthor{Gay:2001:FFP} provides a $\pi$-calculus framework mechanised in Isabelle/HOL.
 The aim is to build a mechanisaton of session-based type systems on top of the $\pi$-calculus framework.
 The use of de Bruijn indices for both bound and free variables causes issues with variable substitution often requiring permutation of typing environments (see \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} for a discussion on binder representations) and lifting of free variables during substitution; we avoid this issue in our representation choice (\S \ref{sec:approach}).
-
-%% Channel endpoints are represented as a polarity on a name. The polarities are not constrained to have the same type as each other but must be unique in the typing enviornment.
 
@@
 
-% System F-pop~\cite{Mazurak:2010:LLT}
+\subsubsection{Applications to Session-based Type Systems}\label{sec:asts}
+
+\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based type system providing session polymorphism.
+The system is more general than session polymorphism via subtyping~\cite{Gay:2005:SST}.
+The authors mechanise their type system and its properties in the Coq proof assistant.
+However, the system mechanised is restricted to the $\pi$-calculus and the main focus is proving soundness properties for the deduction rules allowing polymorphism on input, rather than a more general basis for studying programming with session types.
 
@@
 
-state how you propose to solve the software development problem. Show that
-your proposed approach is feasible, but identify any risks.
+%% state how you propose to solve the software development problem. Show that
+%% your proposed approach is feasible, but identify any risks.
 

### 2014-11-16 proposal/proposal.tex
Completed draft of approach.

@@
 
- %% Additionally, we make some simplifying assumptions about substitution for bound variables to avoid the lifting constructs defined by~\citeauthor{Gay:2001:FFP} in this work.
+I intend to mechanise the calculi described by \citeauthor{Wadler:2012}~\cite{Wadler:2012} using the Coq proof assistant.
+In the spirit of the POPLMARK challenge, I aim to utilise available libraries and infrastructure as much as possible.
+\citeauthor{Aydemir:2008:EFM} developed a metatheory library for handling, among other things, association lists (environments) and free variable representation.
+I intend to use the version distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which has been updated for Coq version 8.4pl4.
+\citeauthor{Park:2014:MMW} describe a technique for removing nonlinear contexts from typing judgements.
+While their work extended to \fpop it is not clear how to handle an environment containing both non-linear and linear types as in GV.
+I wish to maintain as close a relationship as possible to the paper system presented by \citeauthor{Wadler:2012}, so separating out the non-linear and linear components is not an option at this stage.
 
-I intend to mechanise the calculi described by \citeauthor{Wadler:2012}~\cite{Wadler:2012} using the Coq proof assistant.
+Following \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, I use the locally nameless first-order representation for binders and free variables, preventing variable capture issues present in other first-order representations.
+Likewise, I make some simplifying assumptions about variable substitution, namely that \textit{closed} terms (terms with no unbound indices) are substituted for variables, which avoids having to define lifting constructs on the substitute term.
+The typing rules will not contain a rule for an unbound index so this decision is sound.
+A consequence is that we must define a notion of opening language terms involving binders as described by~\citeauthor{Aydemir:2008:EFM}.
+For example, an abstraction consists of a term $t$ which is well-typed when its unbound index, representing the argument to the abstraction, is replaced by some fresh variable $x$, written $t^x$.
+Some terms in GV require multiple binders but I believe it will be straightforward to extend this notion.
+Unfortunately, I must define similar operations on CP terms and handle substitution and opening on them as well as propositions which can accept propositional variables.
+I hope to be able to generalise as much as possible with the help of the UPenn Metatheory library.
+Alternatively, I could have chosen a higher-order abstract syntax (HOAS) approach such as the parametric version (PHOAS) described by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}.
+PHOAS relies on the metalanguage (Coq, in this case) for handling binders and substitution so has less overhead in managing syntax than the locally nameless approach.
+However, variables are now parameters defined at the level of the metalanguage so can represent terms not valid in the language being mechanised.
+Further, a higher-order approach would prevent the use of the Metatheory library which provides a number of useful tactics for manipulating typing contexts.
 
-First order approach taken - represent bound variables as de Bruijn indices
-In contrast to a higher-order approach
-though the first-order encoding has its disadvantages namely one has to reason about variable substitution since without a higher-order approach this is not handled automatically by the meta-language
-the locally nameless representation is chosen with a few simplifying assumptions regarding variable substitution and closed terms
+For handling binder freshness, I chose to adopt the cofinite quantification approach described by~\cite{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, which excludes a finite set of variables from being considered as the binder.
+In contrast, the traditional ``exists-fresh'' approach, where the binder is only required to be fresh within the abstraction's body, does not produce strong enough induction principles in some cases.
 
-One may need to avoid certain tricky formulations.
+While the aim is to be faithful to \citeauthor{Wadler:2012}'s description, I may need to alter the approach slightly to avoid tricky formulations.
 For instance, instead of providing n-ary branch and choice in GV, we provide only binary versions of these operators to simplify the development.
 This change is not restrictive however, since CP itself provides only binary versions of plus ($\oplus$) and with ($\with$) constructs.
-Likewise, on a more technical note, we almagmate all types into one inductive definition and provide a predicate for restricting propositions to consider only valid session types.
-\citeauthor{Wadler:2012} defines the GV grammar as a mutually recursive; session types are types which may contain types as subcomponents (arguments to send, for example).
-Unfortunately, handling mutually inductive definitions in Coq can be quite involved; requiring one to rely on the Coq system being able to provide a stronger mutual induction principle or defining one manual.
+Likewise, on a more technical note, we almagmate all types into one inductive definition and provide a predicate for restricting typing rules to consider only valid session types.
+\citeauthor{Wadler:2012} defines GV types as a set of mutually recursive definitions; session types are types which may contain types as subcomponents (arguments to send, for example).
+Unfortunately, handling mutually inductive definitions in Coq can be quite involved; requiring one to either rely on the Coq system to provide a stronger mutual induction principle or defining one manually.
 Such a definition is possible, but it complicates elimination via the induction principle.
-( since the standard induction tactic will not work?)
+% ( since the standard induction tactic will not work?)
 
-Engineering Formal Metatheory
-Abstract:
+Another aspect of the encoding is how to define terms.
+The intrinsic encoding described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one approach, indexing terms by their type so as to prevent ill-formed terms from being constructed.
+However, in the encoding for GV there are issues with using this encoding: $(1)$ it is not immediately clear how to handle linear contexts using the de Bruijn variable encoding, since in the instrinic setting the environment supports weakening; $(2)$ an instrinic encoding of terms would be complicated by the need to enforce that session types occur in certain instances (by use of a predicate); and $(3)$ well-typed terms require extra assumptions about binder freshness which cannot reasonably be expressed as a function type.
+For these reasons, an instrinic approach does not offer much benefit since a well-typed term relation would still need to be defined.
+Therefore, I chose an extrinsic encoding for terms and define well-typed terms as a separate inductive type.
 
-Formalising programming languages within a proof assistant is becoming a
-more widely adopted convention within the research community in lieu of hand
-written proofs.
-
-Review:
-
-Discussing issues in engineering metatheory for programming languages.
-Focus was on the locally nameless representation for binders with names
-for free variables and de Brujin indices for bound variables. There was
-also novel use of cofinite quantification to rules with bindings to increase
-the strength of the induction principles and reduce the amount of renaming
-of bound variables (due to freshness issues) significantly. The theory was
-tested with formalisations of the simply-typed lambda calculus,
-System $F_{<:}$,
-the calculus of constructions. The formalisations followed the POPLMARK
-challenge (type soundness, preservation, progress).
-
-The intrinsic encoding as described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one approach for encoding programming languages based on the lambda calculus.
-The key idea here is that terms are indexed by their type so as to prevent ill-formed terms from being constructed.
-However, in the encoding for GV there are issues with using this encoding: $(1)$ it is not immediately clear how to handle linear contexts using the de Bruijn variable encoding, since in the instrinic setting the environment supports weakening; $(2)$ an instrinic encoding of terms would be complicated by the need to enforce session types occur in certain instances (by use of a predicate); and $(3)$ well-typed terms require extra assumptions about binder freshness which cannot reasonably be expressed as a function type.
-For these reasons, an instrinic approach does not offer much benefit since a well-typed term relation would still need to be defined.
-I chose the .....
-encoding
+\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of their language proof infrastructure is proportional to the number of binding constructs.
+In the case of a simply typed lambda calculus this is not onerous, but GV has four binding constructs, CP has six and propositions have two.
+It may be true that this approach results in too many auxiliary definitions and operations not relevant to the actual properties one wishes to prove.
+However, I believe that once this infrastructure has been defined, the system can be extended without concern for binders.
 
@@
 I aim to have completed the GV and CP definitions complete with reduction and commuting conversion rules for CP by the end of December.
-This will require definition of infrastructure for handling binders for both GV, CP channels and CP propositional variables; I hope to be able to generalise as much as possible with the help of the UPenn Metatheory library.
+This will require definition of infrastructure for handling binders for both GV, CP channels and CP propositional variables;
 

### 2014-11-16 proposal/proposal.tex
Finished first draft; just proofreading to be completed.

@@
 
-show how you plan to organize your work, identifying intermediate deliverables
-and dates.
+I aim to have completed the GV and CP definitions complete with reduction and commuting conversion rules for CP by the end of December.
+This will require definition of infrastructure for handling binders for both GV, CP channels and CP propositional variables and inductive definitions for well-typed terms, commuting conversions, and CP principal cut reductions.
+So far I have completed some of the definitions for well-typed terms and CP process rules; the remaining rules require the infrastructure for substitution and opening of binders.
 
-I aim to have completed the GV and CP definitions complete with reduction and commuting conversion rules for CP by the end of December.
-This will require definition of infrastructure for handling binders for both GV, CP channels and CP propositional variables;
-
-It remains to define the translation between GV and CP which I envisage will be the bulk of the work, requiring several rules on typing derivations; I envisage being able to define these as inductive datatype constructors.
+After December, it remains to define the translation between GV and CP which I envisage will be the bulk of the work, requiring several rules on typing derivations; I envisage being able to define these as inductive datatype constructors but I shall take care to reuse any suitable libraries that are available, e.g.
+if there is a library for handling CPS translations.
 

### 2014-11-17 proposal/proposal.tex
Proofread. Sending this version as the first draft.

@@
 The first is concerned with static type systems for ensuring runtime safety properties for concurrent communication systems.
-The second is the increasing use of proof assistants and interactive theorem provers, for studying and mechanically verifying the semantics and properties of programming languages.
+The second is the increasing use of proof assistants and interactive theorem provers for studying and mechanically verifying the semantics and properties of programming languages.
 
@@
 The two agents have dual session types such that, for example, when one is expecting to receive a message the other is expecting to send a message; both session types agree on the type of message to be exchanged.
-In this way, correctness of communication can be statically assured by providing communicating agents with \textit{channels} that have dual session types.
+In this way, correctness of communication can be statically assured by providing communicating agents with a \textit{channel} whose endpoints have dual session types.
 A channel is defined as a bi-directional conduit for exchanging messages, not unlike a socket in network programming.
@@
 Most types systems in the session types community have been developed in this way.
-However, it is more difficult to extend work using this approach as may be unclear how changes affect proofs and one will typically have to re-consider all proofs to be convinced the properties still hold.
+However, it is more difficult to extend work using this approach, as it may be unclear how changes affect proofs and one will typically have to re-consider all proofs to be convinced the properties still hold.
 More generally, unassisted formalisms are error-prone to develop and do not provide a firm basis upon which other researchers can experiment; often it will require ``rolling your own'' compiler for the language of interest.
@@
 formalised with little extra proof effort. Inspired by previous work on a
-mechanised framework for pi calculus type systems~\cite{Gay:2001:FFP} we
+mechanised framework for $\pi$-calculus type systems~\cite{Gay:2001:FFP}, we
 shift to the lambda calculus to provide a basis on which session types for
-mainstream programming languages can be studied whilst taking advantage of
+mainstream programming languages can be studied, whilst taking advantage of
 previous proof effort and specialising only for those parts unique to the
@@
 One disadvantage of a linear type system is the need to re-bind objects after each operation~\cite{Gay:2010:LAST}.
-We could look to extend the GV type system tho provide the benefits of aliasing with the guarantees of linearity, termed ``adoption and focus'' as introduced by~\citeauthor{Fahndrich:2002} \cite{Fahndrich:2002}.
+We could look to extend the GV type system to provide the benefits of aliasing with the guarantees of linearity, termed ``adoption and focus'' as introduced by~\citeauthor{Fahndrich:2002} \cite{Fahndrich:2002}.
 The ``adoption and focus'' type system is non-trivial requiring special language features, e.g.
@@
 The type system is nonstandard, featuring an additional environment for tracking the session types of channels; these types are modified during term reduction to reflect communication.
-The extra channel environment is provided to support possible aliasing of channels by treating channel variables as a pointers or references into the channel environment.
-So, for a channel $c$ of session type $S$, a variable $x$ referring to this channel would be declared to have type $Chan~c$.
+The extra channel environment is provided to support possible aliasing of channels by treating channel variables as pointers or references into the channel environment.
+So, for a channel $c$ of session type $S$, a variable $x$ referring to this channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the channel enviornment.
 The typing judgements and term reduction relation must be augmented to handle these environments.
@@
 The work on a linear functional language with session types by \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the current project from the body of work on session-based $\lambda$-calculus type systems.
-They provide paper proofs of their systems type (and runtime?) safety.
+They provide paper proofs of their systems type safety.
 Given the systems non-trivial features such as recursive session types and buffered channels for supporting asynchronous communication it will more difficult to extend these proofs manually.
@@
 
-A number of systems incorporate both a functional language surface syntax with and translation to a more primitive process calculus.
+A number of systems incorporate both a functional language surface syntax with a translation to a more primitive process calculus.
 We shall focus on two such systems presented recently which have a connection to linear logic.
@@
 Lolliproc~\cite{Mazurak:2010:LCC} provides a functional language with concurrency primitives as a surface syntax on top of an underlying process calculus.
-The process calculus provides classial linear logical interpretation for the functional language, utilising double negation elimination to provide session type duality.
+The process calculus provides classical linear logical interpretation for the functional language, utilising double negation elimination to provide session type duality.
 The surface syntax is an effective abstraction layer for the programmer since it does not expose the underlying process calculus.
 
-Concurrency is provided by control operators which, respectively, spawn a child process and wait for a child process to completed.
+Concurrency is provided by control operators which, respectively, spawn a child process and wait for a child process to complete.
 The spawning process creates a channel between parent and child process; a channel is represented as a continuation and the type denotes the communication protocol (i.e.
@@
 \citeauthor{Mazurak:2010:LCC} present proofs for type safety, strong normalisation and confluence.
-Confluence provides the guarantee that no race conditions can occur, strong normalisation prohibits nonterminating programs and type safety assures deadlock freedom.
+Confluence provides the guarantee that no race conditions can occur, strong normalisation prohibits nonterminating programs and type safety assures deadlock-freedom.
 Deadlock-freedom is implied by the acyclic communication graphs (progress of process reduction) which occur between parallel processes; follows from permitting only one channel between two halves of parallel composition construct in $\pi$-calculus.
@@
 The work describes a continuation-passing style (CPS) translation from GV to CP treating session types as propositions in a classical linear logic.
-The process calculus is slightly different to that described in the $\pi$DILL~\cite{Caires:2010:STI}, apart from being classical in nature, the constructs for server replication are modified; CP provides a weakening and contraction rules whereas these are encode in one rule for process reduction by \citeauthor{Caires:2010:STI}.
-Additionally, rather than a sequent calculus, \citeauthor{Wadler:2012} uses a natural deduction style for defining the process calculus.
+The process calculus is slightly different to that described in by $\pi$DILL~\cite{Caires:2010:STI}, apart from being classical in nature, the constructs for server replication are modified; CP provides a weakening and contraction rules whereas these are encoded in one rule for process reduction by \citeauthor{Caires:2010:STI}.
+Additionally, rather than a two-sided sequent calculus, \citeauthor{Wadler:2012} uses one-sided sequents for defining the process calculus.
 As presented, GV does not support all features of CP and thus the translation is one way (GV to CP).
@@
 the constant function.} and free variables are represented as named terms; the \textit{locally nameless} representation.
-Their work resulted in a library for handling commonly occurring aspects in programming lanuage metatheory which has been used since in other developments~\cite{Park:2014:MMW}.
-Resuable libaries reduce the proof engineering effort needed for later works, in contrast to the pen-and-paper approach, and some proof assistants (as is the case with Coq) allow one to extract a program from the development which can provide a typechecker or compiler for the language formalised~\footnote{Curry-Howard strikes again; a proof of decidability of typechecking corresponds to a program implementing a typechecker for the language.}
+Their work resulted in a library for handling commonly occurring aspects in programming language metatheory which has been used since in other developments~\cite{Park:2014:MMW}.
+Reusable libraries reduce the proof engineering effort needed for later works, in contrast to the pen-and-paper approach, and some proof assistants (as is the case with Coq) allow one to extract a program from the development which can provide a typechecker or compiler for the language formalised.~\footnote{Curry-Howard strikes again; a proof of decidability of typechecking corresponds to a program implementing a typechecker for the language.}
 
@@
 This work is of interest to the current project since most session-based type systems assume a strictly linear type system which requires re-binding of channel identifiers, such as in the functional setting of \citeauthor{Gay:2010:LAST}.
-As discussed previously, it would be interesting to study the system described by~\citeauthor{Fahndrich:2002}~\cite{Fahndrich:2002}, yet the system is specific to their ``adopt'' and ``focus'' constructs and is not mechanised.
-\citeauthor{Mazurak:2010:LLT} present an extension to System F with a notion describing types as having linear or non-linear \textit{kind}~\cite{Mazurak:2010:LLT}.
+As discussed previously, it would be interesting to study the system described by~\citeauthor{Fahndrich:2002}~\cite{Fahndrich:2002}, yet their system is specialised for their ``adopt'' and ``focus'' constructs and is not mechanised.
+\citeauthor{Mazurak:2010:LLT} present an extension to System F (termed \fpop) with a notion describing types as having linear or non-linear \textit{kind}~\cite{Mazurak:2010:LLT}.
 This extension is motivated by previous work on incorporating linearity into type systems.
@@
 It would be interesting to see if one could encode the ``adopt'' and ``focus'' constructs in this system.
-The ``adopt'' rule cannot be encoded similarly to the ShareRef example~\cite{Mazurak:2010:LLT}.
-It's not immediately clear how to encode the ``focus'' rule within the present semantics, as one needs to prevent non-linear access to its argument implying knowledge of aliasing within the typing judgements; absence of aliasing is not identified with linear kinded types in \fpop.
+The ``adopt'' rule can be encoded similarly to the ShareRef example~\cite{Mazurak:2010:LLT}.
+It is not immediately clear how to encode the ``focus'' rule within the present semantics, as one needs to prevent non-linear access to its argument implying the presence of knowledge about aliasing within the typing judgements; absence of aliasing is not identified with linear kinded types in \fpop.
 
-While not related to programming language metatheory, \citeauthor{Gay:2001:FFP} provides a $\pi$-calculus framework mechanised in Isabelle/HOL.
+While not related to programming language metatheory, \citeauthor{Gay:2001:FFP}~\cite{Gay:2001:FFP} provides a $\pi$-calculus framework mechanised in Isabelle/HOL.
 The aim is to build a mechanisaton of session-based type systems on top of the $\pi$-calculus framework.
@@
 In the Coq proof assistant, defining custom tactics can be achieved using the Ltac language~\cite{Delahaye:2000:TLS}.
-Among other things, Ltac supports for pattern matching on hypotheses and goal forms.
+Among other things, Ltac supports pattern matching on hypotheses and goal forms.
 Thus, if we wished to pattern match on an inductive type describing variables, we could do so without mentioning the actual variable name as follows:
@@
 In the spirit of the POPLMARK challenge, I aim to utilise available libraries and infrastructure as much as possible.
-\citeauthor{Aydemir:2008:EFM} developed a metatheory library for handling, among other things, association lists (environments) and free variable representation.
+\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a metatheory library for handling, among other things, association lists (environments) and free variable representation.
 I intend to use the version distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which has been updated for Coq version 8.4pl4.
@@
 While their work extended to \fpop it is not clear how to handle an environment containing both non-linear and linear types as in GV.
-I wish to maintain as close a relationship as possible to the paper system presented by \citeauthor{Wadler:2012}, so separating out the non-linear and linear components is not an option at this stage.
+I wish to maintain as close a relationship as possible to the paper system presented by \citeauthor{Wadler:2012}, so separating out the non-linear and linear components (as in \fpop) is not an option at this stage.
 
@@
 A consequence is that we must define a notion of opening language terms involving binders as described by~\citeauthor{Aydemir:2008:EFM}.
-For example, an abstraction consists of a term $t$ which is well-typed when its unbound index, representing the argument to the abstraction, is replaced by some fresh variable $x$, written $t^x$.
+For example, an abstraction with body term $t$ which is well-typed if $t$ is well-typed when its unbound index, representing the argument to the abstraction, is replaced by some fresh variable $x$, written $t^x$.
 Some terms in GV require multiple binders but I believe it will be straightforward to extend this notion.
 Unfortunately, I must define similar operations on CP terms and handle substitution and opening on them as well as propositions which can accept propositional variables.
-I hope to be able to generalise as much as possible with the help of the UPenn Metatheory library.
+I hope to be able to generalise as much as possible with the help of the Metatheory library.
 Alternatively, I could have chosen a higher-order abstract syntax (HOAS) approach such as the parametric version (PHOAS) described by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}.
@@
 This change is not restrictive however, since CP itself provides only binary versions of plus ($\oplus$) and with ($\with$) constructs.
-Likewise, on a more technical note, we almagmate all types into one inductive definition and provide a predicate for restricting typing rules to consider only valid session types.
+Likewise, on a more technical note, we amalgamate all types into one inductive definition and provide a predicate for restricting typing rules to consider only valid session types.
 \citeauthor{Wadler:2012} defines GV types as a set of mutually recursive definitions; session types are types which may contain types as subcomponents (arguments to send, for example).
@@
 The intrinsic encoding described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one approach, indexing terms by their type so as to prevent ill-formed terms from being constructed.
-However, in the encoding for GV there are issues with using this encoding: $(1)$ it is not immediately clear how to handle linear contexts using the de Bruijn variable encoding, since in the instrinic setting the environment supports weakening; $(2)$ an instrinic encoding of terms would be complicated by the need to enforce that session types occur in certain instances (by use of a predicate); and $(3)$ well-typed terms require extra assumptions about binder freshness which cannot reasonably be expressed as a function type.
-For these reasons, an instrinic approach does not offer much benefit since a well-typed term relation would still need to be defined.
+However, in the encoding for GV there are issues with using this encoding: $(1)$ it is not immediately clear how to handle linear contexts using the de Bruijn variable encoding, since in the intrinsic setting the environment supports weakening; $(2)$ an intrinsic encoding of terms would be complicated by the need to enforce that session types occur in certain instances (by use of a predicate); and $(3)$ well-typed terms require extra assumptions about binder freshness which cannot reasonably be expressed as a function type.
+For these reasons, an intrinsic approach does not offer much benefit since a well-typed term relation would still need to be defined.
 Therefore, I chose an extrinsic encoding for terms and define well-typed terms as a separate inductive type.
@@
 It may be true that this approach results in too many auxiliary definitions and operations not relevant to the actual properties one wishes to prove.
-However, I believe that once this infrastructure has been defined, the system can be extended without concern for binders.
+However, I believe that once this infrastructure has been defined, the system can be extended without concern for these binding constructs.
 
@@
 
-I aim to have completed the GV and CP definitions complete with reduction and commuting conversion rules for CP by the end of December.
-This will require definition of infrastructure for handling binders for both GV, CP channels and CP propositional variables and inductive definitions for well-typed terms, commuting conversions, and CP principal cut reductions.
+I aim to have completed the GV and CP definitions including reduction and commuting conversion rules for CP by the end of December.
+This will require definition of infrastructure for handling binders for both GV, CP channels and CP propositional variables, and inductive definitions for well-typed terms, commuting conversions and CP principal cut reductions.
 So far I have completed some of the definitions for well-typed terms and CP process rules; the remaining rules require the infrastructure for substitution and opening of binders.
@@
 After December, it remains to define the translation between GV and CP which I envisage will be the bulk of the work, requiring several rules on typing derivations; I envisage being able to define these as inductive datatype constructors but I shall take care to reuse any suitable libraries that are available, e.g.
-if there is a library for handling CPS translations.
+a library for handling CPS translations.
 

### 2014-11-25 proposal/proposal.tex
Working out final additions. Use xelatex so section refs fixed.
There seems to be a problem with section label references using the
-pdf command of latexmk requiring a second compile to fixed it.
-xelatex does not seem suffer from this deficiency.

I have plans for discussing a Caires and Pfenning-style example.
I have plans for presenting cofinite quantification versions of
some of Wadlers typing rules.
I have plans to have a couple more plans.

@@
 \usepackage{macros}
+\usepackage{hyperref}
 
-\usepackage{hyperref}
 \usepackage{listings}
@@
 \newcommand{\fpop}{System F${}^\circ$\xspace}
+\newcommand{\1}{\textbf{1}\xspace}
 
@@
 aliasing within the type system.
-The recent work by \citeauthor{Lindley:2014:SAP} \cite{Lindley:2014:SAP}, who extend GV to HGV supporting session forwarding, replication and polymorphism, would be a interesting place to start when looking to add these extensions.
+The recent work by \citeauthor{Lindley:2014:SAP} \cite{Lindley:2014:SAP}, who extend GV to HGV supporting session forwarding, replication and polymorphism, would be an interesting place to start when looking to add these extensions.
 
@@
 One disadvantage of a linear type system is the need to re-bind objects after each operation~\cite{Gay:2010:LAST}.
-We could look to extend the GV type system to provide the benefits of aliasing with the guarantees of linearity, termed ``adoption and focus'' as introduced by~\citeauthor{Fahndrich:2002} \cite{Fahndrich:2002}.
+One could extend the GV type system to provide the benefits of aliasing with the guarantees of linearity, termed ``adoption and focus'' as introduced by~\citeauthor{Fahndrich:2002} \cite{Fahndrich:2002}.
 The ``adoption and focus'' type system is non-trivial requiring special language features, e.g.
@@
 
-\subsubsection{In the $\pi$-calculus}\label{sec:pis}
+\subsubsection{In the \texorpdfstring{$\pi$}{pi}-calculus}\label{sec:pis}
 
-\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} present a connection between $\pi$-calculus type system with session types and intuitionistic linear logic.
+\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} present a connection between a $\pi$-calculus type system with session types ($\pi$DILL) and dual intuitionistic linear logic (DILL)~\cite{Barber:1996}.
 They equate linear propositions with session types of the process calculus via a bidirectional translation relation.
-They use a labelled transition system to represent process reduction with observable actions on channels.
-They represent dual intuitionistic linear logic as a sequent calculus providing logical connectives which have session-based interpretations.
+They use a binary reduction relation and a labelled transition system to represent process reduction and process action with its environment as is standard~\cite{Sangiorgi:2001}.
+
+They represent dual intuitionistic linear logic as a sequent calculus providing logical connectives which have session-based interpretations in $\pi$DILL.
+Consider their rules for branch and choice in $\pi$DILL
+
+... here are the rules ...
+
+Note that branch and choice both have some notion of branching (R rules for +/L rules for \&) and at least for \& this is counter-intuitive in session-based setting, since the intutition for \& is that of offering a choice between alternatives rather than making the choice.
+Here, however, the intuition should be extended to account for its location in the typing judgement.
+The channel is on the left of the sequent which caires and pfenning suggest as some service providing to the process.
+
+While in the case of branch and choice, which have different semantics this is mitigated somewhat but as can be seen, there are two rules for selecting a branch, one for branch and another for choice if we consider their server example
+
+As noted by Wadler...jfp there is a certain amount of redundancy in the two-sided nature of the rules forcing one to commit to a certain convention when typing a communication system.
+Consider, an equivalent system to the buy/quote example given by caires and pfen.
+In the following, the types Book, Quote and Receipt can be considered synoms for the type \1 as in caires and pfenning.
+
+... server example here ...
+
+The derivation for ClientBody is T1R, T implies R, T tensor R, T plus R2. Except for the final rule, these are exact opposites of the rules used for ...
+
+The final system is near identical to the one produced by the example given caires and pfenning.
+Here the client communicates along the channel $c$ on the right and the server utilises the channel $c$ on the left.
+This typing, while correct, violates the intuition given by caires and pfenning of the context providing the services used by the process.
+The example above suggests the server utilises some channel provided by the client.
+On the other hand, one could imagine such a channel being setup between two communicating agents; the ``owner'' of the channel is largely irrelevant.
+Note that clientproto is in fact the dual of serverproto in Wadler's classical process calculus.
+
+then it is clear the server is providing a choice to the client indicated by the branch (\&). Yet an equal derivation can be obtained
+
+Consider a shopper which operates according to a certain protocol.
+the idea being that the shopper can conduct transactions with any server which provides the desired service.
+Concretely, suppose the shopper operates according to the following protocol (a similar example to the one produced by caires and pfenning ...
+and inspired by gay and vasco)
+
+choice (buy : ! Book . ! Card . ? Receipt . end, quote : ! Book . ? Price . end)
+
+then in $\pi$DILL we can construct a shopper protocol (treating book, card, address
+
 Process reduction steps are equated with proof reduction via cut elimination.
 The relationship with intuitionistic linear logic provides absence of deadlocks in session-based communication for the process calculus.
+
 The correspondence of session types to linear logic propositions is an important step in providing a solid logical foundation for a concurrent type theory; in the same way ``sequential'' type theory ($\lambda$-calculi) profitted from a logical perspective.
@@
 
-\subsubsection{In the $\lambda$-calculus}\label{sec:lam}
+\subsubsection{In the \texorpdfstring{$\lambda$}{lambda}-calculus}
+\label{sec:lam}
 
@@
 In contrast, the traditional ``exists-fresh'' approach, where the binder is only required to be fresh within the abstraction's body, does not produce strong enough induction principles in some cases.
+This change will necessitate changes to the typing rules of GV and CP.
+Consider for instance the connect with rule in GV~\cite{Wadler:2012}:
+
+... connect with rule here ... ... connect rule using cofinite quantification
+
+The conclusion binds a variable
+
+Consider also the rule for output in CP~\cite{Wadler:2012} and its corresponding version used here:
+
+... output in CP ... ... version used here with cofinite quantification ...
+
+The rule here ... is this needed? What are we showing here with the different version?
 
@@
 
+ For this step of the work, I allocated one month. For the remainder, I shall look to extend the definitions of GV and CP using the definitions I have provided
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-11-26 proposal/proposal.tex
Added discussion of Alms/Adoption&Focus/moved things around.

@@
 \usepackage{cmll}
-\usepackage[style=alphabetic,natbib=true,maxbibnames=100]{biblatex}
+
+\usepackage[style=alphabetic,sorting=nymt,natbib=true,
+  maxbibnames=100]{biblatex}
+
+\DeclareSortingScheme{nymt}{
+  \sort{
+    \field{author}
+    \field{year}
+    \field{month}
+    \field{title}
+  }
+}
+
 \addbibresource{proposal.bib}
@@
 formalised with little extra proof effort. Inspired by previous work on a
-mechanised framework for $\pi$-calculus type systems~\cite{Gay:2001:FFP}, we
+mechanised framework for $\pi$-calculus type systems~\cite{Gay:2001:FFP}, I
 shift to the lambda calculus to provide a basis on which session types for
@@
 From this framework, a number of directions can be explored including
-session polymorphism, session subtyping, and permitting different forms of
+session polymorphism session subtyping, and permitting different forms of
 aliasing within the type system.
@@
 
+%remove the adoption and focus stuff and add something about typestate
+
 Additionally, the majority of previous session type systems for mainstream programming languages have focussed exclusively on linear type systems.
 One disadvantage of a linear type system is the need to re-bind objects after each operation~\cite{Gay:2010:LAST}.
-One could extend the GV type system to provide the benefits of aliasing with the guarantees of linearity, termed ``adoption and focus'' as introduced by~\citeauthor{Fahndrich:2002} \cite{Fahndrich:2002}.
-The ``adoption and focus'' type system is non-trivial requiring special language features, e.g.
-to manage capabilities which can be thought as a form of object liveness mechanism, so the GV typing judgements must be extended such that the capabilities can be included in soundness results.
-A translation to CP must also be defined for the ``adopt'' and ``focus'' constructs.
+One could extend the GV type system to provide the benefits of aliasing with the guarantees of linearity which has been studied in several systems using a variety of alias control forms~\cite{Fahndrich:2002,Mazurak:2010:LLT,Tov:2011}.
+Another approach which has been utilised in previous session-based $\lambda$-calculus type systems, is to support side-effecting functions by allowing functions to close over channels in the environment~\cite{Gay:2003:STI,Vasconcelos:2006:TCM}.
+An interesting extension to the base GV type system would be to incorporate the closure semantics into the type system which would permit functions to modify session-typed channels in the environment.
+The typing rules would be similar to that described by \citeauthor{Gay:2010:MST} for typing expressions in an object-oriented setting for session types~\cite{Gay:2010:MST}, e.g.
+the rule for lambda abstraction could be:
+
+... lambda abstraction ...
+
+The $\Gamma$, $\Gamma '$ environments represent pre and post-states of variables satisfied by term $t$.
+The typing for the function argument is inspired by typestate-oriented programming~\cite{Aldrich:2009} and says that the function expects a channel behaving according to protocol $A$ and alters the state of the channel internally producing a channel behaving according to protocol $B$.
+It remains to be seen whether such typing rules can be encoded as described in the GV setting.
 
@@
 
+most important: duality of sessions is obscured in $\pi$DILL because of the nature of the typing judgements
+
 then it is clear the server is providing a choice to the client indicated by the branch (\&). Yet an equal derivation can be obtained
@@
 Process reduction steps are equated with proof reduction via cut elimination.
-The relationship with intuitionistic linear logic provides absence of deadlocks in session-based communication for the process calculus.
+The relationship with proof normalisation in DILL provides absence of deadlocks in session-based communication for the process calculus.
 
@@
 
-\citeauthor{Gay:2008:STI}~\cite{Gay:2008:STI} provide a $\lambda$-calculus based language with session types for channel primitives.
+A series of works by \citeauthor{Gay:2003:STI}~\cite{Gay:2003:STI} and \citeauthor{Vasconcelos:2006:TCM}~\cite{Vasconcelos:2006:TCM} provide a $\lambda$-calculus based language with session types for channel primitives.
 The type system is nonstandard, featuring an additional environment for tracking the session types of channels; these types are modified during term reduction to reflect communication.
@@
 That is, if two channels $c$ and $d$ have the same session type, one cannot define a function to operate on variables of either channel type since the function definition will explicitly mention the identifier for the channel (in this case either, $Chan~c$ or $Chan~d$).
-However, the system does provide an attractive non-linear channel type which has been neglected in favour of a linear treatment of channels in more recent work~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2012}.
-Certainly, it is the case that an aliased approach would fit better with mainstream programming languages and perhaps the ``adoption and focus'' system~\cite{Fahndrich:2002} could offer some insights.
+However, the system does provide an attractive non-linear channel type which has been neglected in favour of a linear treatment of channels in more recent works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2012}.
+Certainly, it is the case that an aliased approach would fit better with mainstream programming languages and perhaps studying a similar system could offer some insights.
+
+The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of aliased access to linear resources.
+The type system requires special language features, e.g.
+to manage capabilities which can be thought as a form of object liveness mechanism, so GV typing judgements would need to be extended such that the capabilities can be included in soundness results.
+A extension to the translation to CP to incorporate the ``adopt'' and ``focus'' constructs would also need to be defined.
+It is not immediately clear how these features would assist in the management of session typed channels.
+The ``focus'' rule (and linearity generally) enforces an invariant on the focussed object (channel) such that the type is maintained during the focus operation:
+
+... copy of focus operation here ...
+
+Such a rule would appear to only benefit recursive session types.
+Indeed, focussing on a non-recursive channel would be pointless since one would be prevented from communicating across it since the focus rule requires maintenance of the channel type.
+Further, it would be undesirable to permit modification of a session type without guaranteeing exclusive access to the channel since this could lead to invalid communication e.g.
+trying to communicate on a closed channel.
+
+The Alms language~\cite{Aldrich:2009} provides a functional language with affine types based on linear logic.
+A notion of type \textit{kind} is used to distinguish between affine and unrestricted values.
+The typing judgements use two contexts; one for variables with unrestricted kind and another for variables of affine type and type variables.
+Functions are also labelled with a kind which may depend on their arguments e.g.
+a multiple argument function must have affine kinding if it accepts an affine typed resource as an argument other than the final argument.
+The reason for this is simply to prevent the function from using an affine resource more than once.
+Additionally, functions act as closures over their environments so their qualifier may be determined by a free variable in the environment, and a relation is provided to calculate the appropriate kind qualifier.
+Type variables are provided to support kind polymorphism of function qualifiers.
+The system is expressive enough to implement session types as a module using a similar approach to \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} although treating channels as affine rather than linearly typed.
+\citeauthor{Aldrich:2009} note that Alms' type system can support the ``adoption and focus'' system but do not specify details or whether any changes are necessary to their typing judgements.
+Moreover, \citeauthor{Fahndrich:2002} do not provide operational semantics or soundness results for the ``adoption and focus'' type system.
 
 The work on a linear functional language with session types by \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the current project from the body of work on session-based $\lambda$-calculus type systems.
-They provide paper proofs of their systems type safety.
-Given the systems non-trivial features such as recursive session types and buffered channels for supporting asynchronous communication it will more difficult to extend these proofs manually.
+They provide paper proofs of their system's type safety.
+Given the non-trivial features of the type system such as recursive session types and buffered channels for supporting asynchronous communication it will be challenging to extend these proofs manually.
 A mechanised version would provide stronger guarantees of correctness and allow others to more easily alter proofs as a result of changes made to the language.
@@
 \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to provide support for polymorphism and replication, following a similar CPS translation scheme.
+Lastly, \citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational semantics for GV and their system has greater similarities with LAST than prior GV works, for example, by treating send and receive as linear functions rather than language primitives.
 
@@
 The system is more general than session polymorphism via subtyping~\cite{Gay:2005:SST}.
+more general how?
+
  The authors mechanise their type system and its properties in the Coq proof assistant.
@@
 
+I am going to look into adapting adoption and focus for the GV setting to allow one to control aliasing on session typed channels independent of their session characteristics so the aim is not to manipulate a linear type but to provide a nonlinear view of some resource and when changes are to be made, enforce some non-aliased view such that there no references in the context
+perhaps the notion I'm looking for is to prevent sending on a guarded type or enforce the guarded type must be preserved in the adoption block
+the trick in the adoption and focus system is that the type will be deallocated when its adopter's lifetime ends; I suspect in our system this is not useful since we will only be permitted to use a channel in a non-aliased context
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-11-26 proposal/proposal.tex
Breaking the lines using fill-paragraph.

@@
 
-Two aspects of programming language research are relevant for this project.
-The first is concerned with static type systems for ensuring runtime safety properties for concurrent communication systems.
-The second is the increasing use of proof assistants and interactive theorem provers for studying and mechanically verifying the semantics and properties of programming languages.
+Two aspects of programming language research are relevant for this
+project. The first is concerned with static type systems for ensuring runtime
+safety properties for concurrent communication systems. The second is the
+increasing use of proof assistants and interactive theorem provers for
+studying and mechanically verifying the semantics and properties of
+programming languages.
 
@@
 
-Distributed communication systems are becoming increasingly prevalent in modern society.
-To facilitate communication between concurrently executing agents requires complex network protocols to define the type and sequencing of messages.
-Typically these protocols are implemented in systems programming languages, such as C, which provide no correctness guarantees of the implementation.
-Furthermore, programming in these languages is often extremely error-prone, as one has a myriad of unsafe features at one's disposal.
-As more businesses begin to conduct transactions over the Web, it is imperative that they can reason about the communication protocols upon which they rely.
+Distributed communication systems are becoming increasingly prevalent in
+modern society. To facilitate communication between concurrently executing
+agents requires complex network protocols to define the type and sequencing of
+messages. Typically these protocols are implemented in systems programming
+languages, such as C, which provide no correctness guarantees of the
+implementation. Furthermore, programming in these languages is often extremely
+error-prone, as one has a myriad of unsafe features at one's disposal. As more
+businesses begin to conduct transactions over the Web, it is imperative that
+they can reason about the communication protocols upon which they rely.
 
-Encoding communication protocols as \textit{session types} is one approach to achieve guarantees about correctness.
-One particular variant of session types are \textit{binary} session types~\footnote{There are also \textit{multiparty} session types, but I do not discuss them here.}.
-A binary session type specifies communication between two agents, and it specifies the order and type of messages to be sent or received by an agent in the system.
-The two agents have dual session types such that, for example, when one is expecting to receive a message the other is expecting to send a message; both session types agree on the type of message to be exchanged.
-In this way, correctness of communication can be statically assured by providing communicating agents with a \textit{channel} whose endpoints have dual session types.
-A channel is defined as a bi-directional conduit for exchanging messages, not unlike a socket in network programming.
-In contrast to sockets, however, a channel is associated with an explicit type which provides a description of its protocol.
-Thus, in session-based systems a channel is usually provided as a primitive type.
+Encoding communication protocols as \textit{session types} is one approach to
+achieve guarantees about correctness. One particular variant of session types
+are \textit{binary} session types~\footnote{There are also \textit{multiparty}
+ session types, but I do not discuss them here.}. A binary session type
+specifies communication between two agents, and it specifies the order and
+type of messages to be sent or received by an agent in the system. The two
+agents have dual session types such that, for example, when one is expecting
+to receive a message the other is expecting to send a message; both session
+types agree on the type of message to be exchanged. In this way, correctness
+of communication can be statically assured by providing communicating agents
+with a \textit{channel} whose endpoints have dual session types. A channel is
+defined as a bi-directional conduit for exchanging messages, not unlike a
+socket in network programming. In contrast to sockets, however, a channel is
+associated with an explicit type which provides a description of its
+protocol. Thus, in session-based systems a channel is usually provided as a
+primitive type.
 
@@
 
-In programming language research, work on static type systems is usually presented on paper complete with informal proofs of correctness for the system properties.
-Most types systems in the session types community have been developed in this way.
-However, it is more difficult to extend work using this approach, as it may be unclear how changes affect proofs and one will typically have to re-consider all proofs to be convinced the properties still hold.
-More generally, unassisted formalisms are error-prone to develop and do not provide a firm basis upon which other researchers can experiment; often it will require ``rolling your own'' compiler for the language of interest.
+In programming language research, work on static type systems is usually
+presented on paper complete with informal proofs of correctness for the system
+properties. Most types systems in the session types community have been
+developed in this way. However, it is more difficult to extend work using this
+approach, as it may be unclear how changes affect proofs and one will
+typically have to re-consider all proofs to be convinced the properties still
+hold. More generally, unassisted formalisms are error-prone to develop and do
+not provide a firm basis upon which other researchers can experiment; often it
+will require ``rolling your own'' compiler for the language of interest.
 
-Proof assistants such as Coq~\cite{Coq:manual} offer a platform on which to develop programming languages and type systems and associated proofs of properties with strong guarantees of correctness.
-Not only is proving with a proof assistant much less error-prone than the unaided approach, it also eases extension of the system since the proof assistant will highlight parts of proofs that have been affected by changes.
-However, one may have to prove more properties using this approach since this setting does not permit appealing to human intuition.
-For example, alpha-equivalence and freshness of binders are usually assumed in informal presentations, whereas in the mechanised setting one has to explicitly encode and reason about these properties~\cite{Aydemir:2005:MMM, Aydemir:2008:EFM}.
+Proof assistants such as Coq~\cite{Coq:manual} offer a platform on which to
+develop programming languages and type systems and associated proofs of
+properties with strong guarantees of correctness. Not only is proving with a
+proof assistant much less error-prone than the unaided approach, it also eases
+extension of the system since the proof assistant will highlight parts of
+proofs that have been affected by changes. However, one may have to prove more
+properties using this approach since this setting does not permit appealing to
+human intuition. For example, alpha-equivalence and freshness of binders are
+usually assumed in informal presentations, whereas in the mechanised setting
+one has to explicitly encode and reason about these
+properties~\cite{Aydemir:2005:MMM, Aydemir:2008:EFM}.
 
@@
 previous proof effort and specialising only for those parts unique to the
-particular system.
-The framework will be heavily based on calculi developed by \citeauthor{Wadler:2012}~\cite{Wadler:2012} describing a relationship between classical linear logic (CP) and a functional language with session types (GV).
-GV is a minimal functional language supporting linearity so I believe such a basis will be general enough to support reasoning about a large class of functional languages with similar base properties.
+particular system. The framework will be heavily based on calculi developed by
+\citeauthor{Wadler:2012}~\cite{Wadler:2012} describing a relationship between
+classical linear logic (CP) and a functional language with session types
+(GV). GV is a minimal functional language supporting linearity so I believe
+such a basis will be general enough to support reasoning about a large class
+of functional languages with similar base properties.
 
-From this framework, a number of directions can be explored including
-session polymorphism session subtyping, and permitting different forms of
-aliasing within the type system.
-The recent work by \citeauthor{Lindley:2014:SAP} \cite{Lindley:2014:SAP}, who extend GV to HGV supporting session forwarding, replication and polymorphism, would be an interesting place to start when looking to add these extensions.
-
-%remove the adoption and focus stuff and add something about typestate
+From this framework, a number of directions can be explored including session
+polymorphism session subtyping, and permitting different forms of aliasing
+within the type system. The recent work by \citeauthor{Lindley:2014:SAP}
+\cite{Lindley:2014:SAP}, who extend GV to HGV supporting session forwarding,
+replication and polymorphism, would be an interesting place to start when
+looking to add these extensions.
 

### 2014-11-26 proposal/proposal.tex
Added line breaks.

@@
 
-Additionally, the majority of previous session type systems for mainstream programming languages have focussed exclusively on linear type systems.
-One disadvantage of a linear type system is the need to re-bind objects after each operation~\cite{Gay:2010:LAST}.
-One could extend the GV type system to provide the benefits of aliasing with the guarantees of linearity which has been studied in several systems using a variety of alias control forms~\cite{Fahndrich:2002,Mazurak:2010:LLT,Tov:2011}.
-Another approach which has been utilised in previous session-based $\lambda$-calculus type systems, is to support side-effecting functions by allowing functions to close over channels in the environment~\cite{Gay:2003:STI,Vasconcelos:2006:TCM}.
-An interesting extension to the base GV type system would be to incorporate the closure semantics into the type system which would permit functions to modify session-typed channels in the environment.
-The typing rules would be similar to that described by \citeauthor{Gay:2010:MST} for typing expressions in an object-oriented setting for session types~\cite{Gay:2010:MST}, e.g.
-the rule for lambda abstraction could be:
+Additionally, the majority of previous session type systems for mainstream
+programming languages have focussed exclusively on linear type systems. One
+disadvantage of a linear type system is the need to re-bind objects after each
+operation~\cite{Gay:2010:LAST}. One could extend the GV type system to provide
+the benefits of aliasing with the guarantees of linearity which has been
+studied in several systems using a variety of alias control
+methods~\cite{Fahndrich:2002,Mazurak:2010:LLT,Tov:2011}. Another approach,
+which has been utilised in previous session-based $\lambda$-calculus type
+systems, is to support side-effecting functions by allowing functions to close
+over channels in the environment~\cite{Gay:2003:STI,Vasconcelos:2006:TCM}. An
+interesting extension to the base GV type system would be to incorporate the
+closure semantics into the type system which would permit functions to modify
+session-typed channels in the environment. The typing rules would be similar
+to that described by \citeauthor{Gay:2010:MST} for typing expressions in an
+object-oriented setting for session types~\cite{Gay:2010:MST}, e.g. the rule
+for lambda abstraction could be:
 
@@
 
-The $\Gamma$, $\Gamma '$ environments represent pre and post-states of variables satisfied by term $t$.
-The typing for the function argument is inspired by typestate-oriented programming~\cite{Aldrich:2009} and says that the function expects a channel behaving according to protocol $A$ and alters the state of the channel internally producing a channel behaving according to protocol $B$.
-It remains to be seen whether such typing rules can be encoded as described in the GV setting.
+The $\Gamma$, $\Gamma '$ environments represent pre and post-states of
+variables satisfied by term $t$. The typing for the function argument is
+inspired by typestate-oriented programming~\cite{Aldrich:2009} and says that
+the function expects a channel behaving according to protocol $A$ and alters
+the state of the channel internally producing a channel behaving according to
+protocol $B$. It remains to be seen whether such typing rules can be encoded
+as described in the GV setting.
 
-These extensions will argue that the framework supports a wide range of languages and type systems, providing researchers the ability to harness the power of formal verification in the design, implementation and documentation of session-based type systems.
+These extensions will argue that the framework supports a wide range of
+languages and type systems, providing researchers the ability to harness the
+power of formal verification in the design, implementation and documentation
+of session-based type systems.
 
@@
 
-%% present an overview of relevant previous work including articles, books, and
-%% existing software products. Critically evaluate the strengths and weaknesses
-%% of the previous work.
+%% present an overview of relevant previous work including articles, books,
+%% and existing software products. Critically evaluate the strengths and
+%% weaknesses of the previous work.
 
-The background for this project can be categorised roughly as previous work on session-based type systems (sub-divided into $\pi$ and $\lambda$-calculi variants, \S \ref{sec:pis} and \S \ref{sec:lam}, respectively) and previous work on formal verification using proof assistants (particularly focussed on programming language mechanisation).
-Finally, some recent work has merged these two disciplines by verifying a session-based $\pi$-calculus system (\S \ref{sec:asts}).
+The background for this project can be categorised roughly as previous work on
+session-based type systems (sub-divided into $\pi$ and $\lambda$-calculi
+variants, \S \ref{sec:pis} and \S \ref{sec:lam}, respectively) and previous
+work on formal verification using proof assistants (particularly focussed on
+programming language mechanisation). Finally, some recent work has merged
+these two disciplines by verifying a session-based $\pi$-calculus system (\S
+\ref{sec:asts}).
 
@@
 
-\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} present a connection between a $\pi$-calculus type system with session types ($\pi$DILL) and dual intuitionistic linear logic (DILL)~\cite{Barber:1996}.
-They equate linear propositions with session types of the process calculus via a bidirectional translation relation.
-They use a binary reduction relation and a labelled transition system to represent process reduction and process action with its environment as is standard~\cite{Sangiorgi:2001}.
+\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} present a connection
+between a $\pi$-calculus type system with session types ($\pi$DILL) and dual
+intuitionistic linear logic (DILL)~\cite{Barber:1996}. They equate linear
+propositions with session types of the process calculus via a bidirectional
+translation relation. They use a binary reduction relation and a labelled
+transition system to represent process reduction and process action with its
+environment as is standard~\cite{Sangiorgi:2001}.
 
-They represent dual intuitionistic linear logic as a sequent calculus providing logical connectives which have session-based interpretations in $\pi$DILL.
-Consider their rules for branch and choice in $\pi$DILL
+They represent dual intuitionistic linear logic as a sequent calculus
+providing logical connectives which have session-based interpretations in
+$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL:
 
@@
 
-Note that branch and choice both have some notion of branching (R rules for +/L rules for \&) and at least for \& this is counter-intuitive in session-based setting, since the intutition for \& is that of offering a choice between alternatives rather than making the choice.
-Here, however, the intuition should be extended to account for its location in the typing judgement.
-The channel is on the left of the sequent which caires and pfenning suggest as some service providing to the process.
+Note that branch and choice both have some notion of branching (R rules for
++/L rules for \&) and at least for \& this is counter-intuitive in
+session-based setting, since the intutition for \& is that of offering a
+choice between alternatives rather than making the choice. Here, however, the
+intuition should be extended to account for its location in the typing
+judgement. The channel is on the left of the sequent which caires and pfenning
+suggest as some service providing to the process.
 
-While in the case of branch and choice, which have different semantics this is mitigated somewhat but as can be seen, there are two rules for selecting a branch, one for branch and another for choice if we consider their server example
+While in the case of branch and choice, which have different semantics this is
+mitigated somewhat but as can be seen, there are two rules for selecting a
+branch, one for branch and another for choice if we consider their server
+example
 
-As noted by Wadler...jfp there is a certain amount of redundancy in the two-sided nature of the rules forcing one to commit to a certain convention when typing a communication system.
-Consider, an equivalent system to the buy/quote example given by caires and pfen.
-In the following, the types Book, Quote and Receipt can be considered synoms for the type \1 as in caires and pfenning.
+As noted by Wadler...jfp there is a certain amount of redundancy in the
+two-sided nature of the rules forcing one to commit to a certain convention
+when typing a communication system. Consider, an equivalent system to the
+buy/quote example given by caires and pfen. In the following, the types Book,
+Quote and Receipt can be considered synoms for the type \1 as in caires and
+pfenning.
 
@@
 
-The derivation for ClientBody is T1R, T implies R, T tensor R, T plus R2. Except for the final rule, these are exact opposites of the rules used for ...
+The derivation for ClientBody is T1R, T implies R, T tensor R, T plus
+R2. Except for the final rule, these are exact opposites of the rules used for
+...
 
-The final system is near identical to the one produced by the example given caires and pfenning.
-Here the client communicates along the channel $c$ on the right and the server utilises the channel $c$ on the left.
-This typing, while correct, violates the intuition given by caires and pfenning of the context providing the services used by the process.
-The example above suggests the server utilises some channel provided by the client.
-On the other hand, one could imagine such a channel being setup between two communicating agents; the ``owner'' of the channel is largely irrelevant.
-Note that clientproto is in fact the dual of serverproto in Wadler's classical process calculus.
+The final system is near identical to the one produced by the example given
+caires and pfenning. Here the client communicates along the channel $c$ on the
+right and the server utilises the channel $c$ on the left. This typing, while
+correct, violates the intuition given by caires and pfenning of the context
+providing the services used by the process. The example above suggests the
+server utilises some channel provided by the client. On the other hand, one
+could imagine such a channel being setup between two communicating agents; the
+``owner'' of the channel is largely irrelevant. Note that clientproto is in
+fact the dual of serverproto in Wadler's classical process calculus.
 
-most important: duality of sessions is obscured in $\pi$DILL because of the nature of the typing judgements
+most important: duality of sessions is obscured in $\pi$DILL because of the
+nature of the typing judgements
 
-then it is clear the server is providing a choice to the client indicated by the branch (\&). Yet an equal derivation can be obtained
-
-Consider a shopper which operates according to a certain protocol.
-the idea being that the shopper can conduct transactions with any server which provides the desired service.
-Concretely, suppose the shopper operates according to the following protocol (a similar example to the one produced by caires and pfenning ...
-and inspired by gay and vasco)
-
-choice (buy : ! Book . ! Card . ? Receipt . end, quote : ! Book . ? Price . end)
-
-then in $\pi$DILL we can construct a shopper protocol (treating book, card, address
-
-Process reduction steps are equated with proof reduction via cut elimination.
-The relationship with proof normalisation in DILL provides absence of deadlocks in session-based communication for the process calculus.
-
-The correspondence of session types to linear logic propositions is an important step in providing a solid logical foundation for a concurrent type theory; in the same way ``sequential'' type theory ($\lambda$-calculi) profitted from a logical perspective.
-However, for session types to become a more usable tool for programming distributed communication systems (the ultimate goal, of course, is to write more correct programs), we cannot be expected to program directly in the $\pi$-calculus; a more mainstream programming paradigm must be built upon these solid logical foundations.
-With that in mind, enter the $\lambda$-calculus!
+Process reduction steps are equated with proof reduction via cut
+elimination. The relationship with proof normalisation in DILL provides
+absence of deadlocks in session-based communication for the process
+calculus. The correspondence of session types to linear logic propositions is
+an important step in providing a solid logical foundation for a concurrent
+type theory; in the same way ``sequential'' type theory ($\lambda$-calculi)
+profitted from a logical perspective. However, for session types to become a
+more usable tool for programming distributed communication systems (the
+ultimate goal, of course, is to write more correct programs), we cannot be
+expected to program directly in the $\pi$-calculus; a more mainstream
+programming paradigm must be built upon these solid logical foundations. With
+that in mind, enter the $\lambda$-calculus!
 
@@
 
-A series of works by \citeauthor{Gay:2003:STI}~\cite{Gay:2003:STI} and \citeauthor{Vasconcelos:2006:TCM}~\cite{Vasconcelos:2006:TCM} provide a $\lambda$-calculus based language with session types for channel primitives.
-The type system is nonstandard, featuring an additional environment for tracking the session types of channels; these types are modified during term reduction to reflect communication.
-The extra channel environment is provided to support possible aliasing of channels by treating channel variables as pointers or references into the channel environment.
-So, for a channel $c$ of session type $S$, a variable $x$ referring to this channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the channel enviornment.
-The typing judgements and term reduction relation must be augmented to handle these environments.
-Unfortunately, a side-effect of this is that function types must specify the pre and post-states of channel environments and therefore, the type of arguments cannot work up to alpha-equivalence of channel identifiers.
-That is, if two channels $c$ and $d$ have the same session type, one cannot define a function to operate on variables of either channel type since the function definition will explicitly mention the identifier for the channel (in this case either, $Chan~c$ or $Chan~d$).
-However, the system does provide an attractive non-linear channel type which has been neglected in favour of a linear treatment of channels in more recent works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2012}.
-Certainly, it is the case that an aliased approach would fit better with mainstream programming languages and perhaps studying a similar system could offer some insights.
+A series of works by \citeauthor{Gay:2003:STI}~\cite{Gay:2003:STI} and
+\citeauthor{Vasconcelos:2006:TCM}~\cite{Vasconcelos:2006:TCM} provide a
+$\lambda$-calculus based language with session types for channel
+primitives. The type system is nonstandard, featuring an additional
+environment for tracking the session types of channels; these types are
+modified during term reduction to reflect communication. The extra channel
+environment is provided to support possible aliasing of channels by treating
+channel variables as pointers or references into the channel environment. So,
+for a channel $c$ of session type $S$, a variable $x$ referring to this
+channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the
+channel enviornment. The typing judgements and term reduction relation must be
+augmented to handle these environments. Unfortunately, a side-effect of this
+is that function types must specify the pre and post-states of channel
+environments and therefore, the type of arguments cannot work up to
+alpha-equivalence of channel identifiers. That is, if two channels $c$ and $d$
+have the same session type, one cannot define a function to operate on
+variables of either channel type since the function definition will explicitly
+mention the identifier for the channel (in this case either, $Chan~c$ or
+$Chan~d$). However, the system does provide an attractive non-linear channel
+type which has been neglected in favour of a linear treatment of channels in
+more recent
+works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2012}. Certainly, it is the
+case that an aliased approach would fit better with mainstream programming
+languages and perhaps studying a similar system could offer some insights.
 
-The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of aliased access to linear resources.
-The type system requires special language features, e.g.
-to manage capabilities which can be thought as a form of object liveness mechanism, so GV typing judgements would need to be extended such that the capabilities can be included in soundness results.
-A extension to the translation to CP to incorporate the ``adopt'' and ``focus'' constructs would also need to be defined.
-It is not immediately clear how these features would assist in the management of session typed channels.
-The ``focus'' rule (and linearity generally) enforces an invariant on the focussed object (channel) such that the type is maintained during the focus operation:
+The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of
+aliased access to linear resources. The type system requires special language
+features, e.g. to manage capabilities which can be thought as a form of object
+liveness mechanism, so GV typing judgements would need to be extended such
+that the capabilities can be included in soundness results. A extension to the
+translation to CP to incorporate the ``adopt'' and ``focus'' constructs would
+also need to be defined. It is not immediately clear how these features would
+assist in the management of session typed channels. The ``focus'' rule (and
+linearity generally) enforces an invariant on the focussed object (channel)
+such that the type is maintained during the focus operation:
 
@@
 
-Such a rule would appear to only benefit recursive session types.
-Indeed, focussing on a non-recursive channel would be pointless since one would be prevented from communicating across it since the focus rule requires maintenance of the channel type.
-Further, it would be undesirable to permit modification of a session type without guaranteeing exclusive access to the channel since this could lead to invalid communication e.g.
-trying to communicate on a closed channel.
+Such a rule would appear to only benefit recursive session types. Indeed,
+focussing on a non-recursive channel would be pointless since one would be
+prevented from communicating across it since the focus rule requires
+maintenance of the channel type. Further, it would be undesirable to permit
+modification of a session type without guaranteeing exclusive access to the
+channel since this could lead to invalid communication e.g. trying to
+communicate on a closed channel.
 
-The Alms language~\cite{Aldrich:2009} provides a functional language with affine types based on linear logic.
-A notion of type \textit{kind} is used to distinguish between affine and unrestricted values.
-The typing judgements use two contexts; one for variables with unrestricted kind and another for variables of affine type and type variables.
-Functions are also labelled with a kind which may depend on their arguments e.g.
-a multiple argument function must have affine kinding if it accepts an affine typed resource as an argument other than the final argument.
-The reason for this is simply to prevent the function from using an affine resource more than once.
-Additionally, functions act as closures over their environments so their qualifier may be determined by a free variable in the environment, and a relation is provided to calculate the appropriate kind qualifier.
-Type variables are provided to support kind polymorphism of function qualifiers.
-The system is expressive enough to implement session types as a module using a similar approach to \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} although treating channels as affine rather than linearly typed.
-\citeauthor{Aldrich:2009} note that Alms' type system can support the ``adoption and focus'' system but do not specify details or whether any changes are necessary to their typing judgements.
-Moreover, \citeauthor{Fahndrich:2002} do not provide operational semantics or soundness results for the ``adoption and focus'' type system.
+The Alms language~\cite{Aldrich:2009} provides a functional language with
+affine types based on linear logic. A notion of type \textit{kind} is used to
+distinguish between affine and unrestricted values. The typing judgements use
+two contexts; one for variables with unrestricted kind and another for
+variables of affine type and type variables. Functions are also labelled with
+a kind which may depend on their arguments e.g. a multiple argument function
+must have affine kinding if it accepts an affine typed resource as an argument
+other than the final argument. The reason for this is simply to prevent the
+function from using an affine resource more than once. Additionally, functions
+act as closures over their environments so their qualifier may be determined
+by a free variable in the environment, and a relation is provided to calculate
+the appropriate kind qualifier. Type variables are provided to support kind
+polymorphism of function qualifiers. The system is expressive enough to
+implement session types as a module using a similar approach to
+\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} although treating channels as
+affine rather than linearly typed. \citeauthor{Aldrich:2009} note that Alms'
+type system can support the ``adoption and focus'' system but do not specify
+details or whether any changes are necessary to their typing
+judgements. Moreover, \citeauthor{Fahndrich:2002} do not provide operational
+semantics or soundness results for the ``adoption and focus'' type system.
 
-The work on a linear functional language with session types by \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the current project from the body of work on session-based $\lambda$-calculus type systems.
-They provide paper proofs of their system's type safety.
-Given the non-trivial features of the type system such as recursive session types and buffered channels for supporting asynchronous communication it will be challenging to extend these proofs manually.
-A mechanised version would provide stronger guarantees of correctness and allow others to more easily alter proofs as a result of changes made to the language.
+The work on a linear functional language with session types by
+\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the
+current project from the body of work on session-based $\lambda$-calculus type
+systems. They provide paper proofs of their system's type safety. Given the
+non-trivial features of the type system such as recursive session types and
+buffered channels for supporting asynchronous communication it will be
+challenging to extend these proofs manually. A mechanised version would
+provide stronger guarantees of correctness and allow others to more easily
+alter proofs as a result of changes made to the language.
 
@@
 
-A number of systems incorporate both a functional language surface syntax with a translation to a more primitive process calculus.
-We shall focus on two such systems presented recently which have a connection to linear logic.
+A number of systems incorporate both a functional language surface syntax with
+a translation to a more primitive process calculus. We shall focus on two such
+systems presented recently which have a connection to linear logic.
 
-Lolliproc~\cite{Mazurak:2010:LCC} provides a functional language with concurrency primitives as a surface syntax on top of an underlying process calculus.
-The process calculus provides classical linear logical interpretation for the functional language, utilising double negation elimination to provide session type duality.
-The surface syntax is an effective abstraction layer for the programmer since it does not expose the underlying process calculus.
+Lolliproc~\cite{Mazurak:2010:LCC} provides a functional language with
+concurrency primitives as a surface syntax on top of an underlying process
+calculus. The process calculus provides a classical linear logic
+interpretation for the functional language, utilising double negation
+elimination to provide session type duality. The surface syntax is an
+effective abstraction layer for the programmer since it does not expose the
+underlying process calculus.
 
-Concurrency is provided by control operators which, respectively, spawn a child process and wait for a child process to complete.
-The spawning process creates a channel between parent and child process; a channel is represented as a continuation and the type denotes the communication protocol (i.e.
-session type) between parent and child.
+Concurrency is provided by control operators which, respectively, spawn a
+child process and wait for a child process to complete. The spawning process
+creates a channel between parent and child process; a channel is represented
+as a continuation and the type denotes the communication protocol
+(i.e. session type) between parent and child.
 
-\citeauthor{Mazurak:2010:LCC} present proofs for type safety, strong normalisation and confluence.
-Confluence provides the guarantee that no race conditions can occur, strong normalisation prohibits nonterminating programs and type safety assures deadlock-freedom.
-Deadlock-freedom is implied by the acyclic communication graphs (progress of process reduction) which occur between parallel processes; follows from permitting only one channel between two halves of parallel composition construct in $\pi$-calculus.
-Most of the development is supported by Coq proof scripts with remaining issues due to reasoning about communication graphs.
-The development does not provide primitive send and receives; instead, these are encoded as linear functions.
-%% and classical extension to second-order logic by \citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}.
+\citeauthor{Mazurak:2010:LCC} present proofs for type safety, strong
+normalisation and confluence. Confluence provides the guarantee that no race
+conditions can occur, strong normalisation prohibits nonterminating programs
+and type safety assures deadlock-freedom. Deadlock-freedom is implied by the
+acyclic communication graphs (progress of process reduction) which occur
+between parallel processes; follows from permitting only one channel between
+two halves of parallel composition construct in the process calculus. Most of
+the development is supported by Coq proof scripts with remaining issues due to
+reasoning about communication graphs. The development does not provide
+primitive send and receives; instead, these are encoded as linear functions.
+%% and classical extension to second-order logic by
+%% \citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}.
 
-\citeauthor{Wadler:2012}'s recent work~\cite{Wadler:2012} brings together a process calculus, CP, based on the earlier work by \citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by \citeauthor{Gay:2010:LAST}.
-The work describes a continuation-passing style (CPS) translation from GV to CP treating session types as propositions in a classical linear logic.
-The process calculus is slightly different to that described in by $\pi$DILL~\cite{Caires:2010:STI}, apart from being classical in nature, the constructs for server replication are modified; CP provides a weakening and contraction rules whereas these are encoded in one rule for process reduction by \citeauthor{Caires:2010:STI}.
-Additionally, rather than a two-sided sequent calculus, \citeauthor{Wadler:2012} uses one-sided sequents for defining the process calculus.
-As presented, GV does not support all features of CP and thus the translation is one way (GV to CP).
-\citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to provide support for polymorphism and replication, following a similar CPS translation scheme.
-Lastly, \citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational semantics for GV and their system has greater similarities with LAST than prior GV works, for example, by treating send and receive as linear functions rather than language primitives.
+\citeauthor{Wadler:2012}'s recent work~\cite{Wadler:2012} brings together a
+process calculus, CP, based on the earlier work by
+\citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
+\citeauthor{Gay:2010:LAST}. The work describes a continuation-passing style
+(CPS) translation from GV to CP treating session types as propositions in a
+classical linear logic. The process calculus is slightly different to that
+described by $\pi$DILL~\cite{Caires:2010:STI}, apart from being classical in
+nature, the constructs for server replication are modified; CP provides a
+weakening and contraction rules whereas these are encoded in one rule for
+process reduction by \citeauthor{Caires:2010:STI}. Additionally, rather than a
+two-sided sequent calculus, \citeauthor{Wadler:2012} uses one-sided sequents
+for defining the process calculus, which provides a more intuitive
+presentation of session type duality and clearer correspondence to a
+$\lambda$-calculi surface syntax. As presented, GV does not support all
+features of CP and thus the translation is one way (GV to
+CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to
+provide support for polymorphism and replication, following a similar CPS
+translation scheme. Lastly,
+\citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
+semantics for GV and their system has greater similarities with LAST than
+prior GV works, for example, by treating send and receive as linear functions
+rather than language primitives.
 
-%% Wadler also dispenses with the structural rules; subsumed by the Assoc and Swap typing derivations and the cut reduction (double check why he does this)
+%% Wadler also dispenses with the structural rules; subsumed by the Assoc and
+%% Swap typing derivations and the cut reduction (double check why he does
+%% this)
 
@@
 
-Interactive theorem proving and proof assistants have seen an upsurge in use in recent years.
-Especially within the domain of programming language research, there is a strong emphasis on providing mechanised proofs accompanying work in the area.
-Indeed, \citeauthor{Aydemir:2005:MMM} have published a series of challenges on mechanising programming languages aimed at providing a starting point for comparing different representation techniques~\cite{Aydemir:2005:MMM}.
-Another aim is to provide reusable libraries for common reasoning needed across programming language developments, e.g.
-handling of typing environments.
-This effort has resulted in a number of different approaches for representing programming languages within proof assistants.
-For instance, \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} present a representation of lambda calculi where bound variables are represented as \textit{de Bruijn indices}~\footnote{A de Bruijn index is a number used to represent a bound variable which indicates the position of its binder (starting from zero for the innermost binder).
-For example, $\lambda~(\lambda~1)$ corresponds to $\lambda x.
-\lambda y.
-x$, i.e.
-the constant function.} and free variables are represented as named terms; the \textit{locally nameless} representation.
-Their work resulted in a library for handling commonly occurring aspects in programming language metatheory which has been used since in other developments~\cite{Park:2014:MMW}.
-Reusable libraries reduce the proof engineering effort needed for later works, in contrast to the pen-and-paper approach, and some proof assistants (as is the case with Coq) allow one to extract a program from the development which can provide a typechecker or compiler for the language formalised.~\footnote{Curry-Howard strikes again; a proof of decidability of typechecking corresponds to a program implementing a typechecker for the language.}
+Interactive theorem proving and proof assistants have seen an upsurge in use
+in recent years. Especially within the domain of programming language
+research, there is a strong emphasis on providing mechanised proofs
+accompanying work in the area. Indeed, \citeauthor{Aydemir:2005:MMM} have
+published a series of challenges on mechanising programming languages aimed at
+providing a starting point for comparing different representation
+techniques~\cite{Aydemir:2005:MMM}. Another aim is to provide reusable
+libraries for common reasoning needed across programming language
+developments, e.g. handling of typing environments. This effort has resulted
+in a number of different approaches for representing programming languages
+within proof assistants. For instance,
+\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} present a representation
+of $\lambda$-calculi where bound variables are represented as \textit{de
+  Bruijn indices}~\footnote{A de Bruijn index is a number used to represent a
+  bound variable which indicates the position of its binder (starting from
+  zero for the innermost binder). For example, $\lambda~(\lambda~1)$
+  corresponds to $\lambda x. \lambda y. x$, i.e. the constant function.} and
+free variables are represented as named terms; the \textit{locally nameless}
+representation. Their work resulted in a library for handling commonly
+occurring aspects in programming language metatheory which has been used since
+in other developments~\cite{Park:2014:MMW}. Reusable libraries reduce the
+proof engineering effort needed for later works, in contrast to the
+pen-and-paper approach, and some proof assistants (as is the case with Coq)
+allow one to extract a program from the development which can provide a
+typechecker or compiler for the language formalised.~\footnote{Curry-Howard
+  strikes again; a proof of decidability of typechecking corresponds to a
+  program implementing a typechecker for the language.}
 
-Some recent mechanisation effort focuses on providing a basis for studying linearity within type systems.
-This work is of interest to the current project since most session-based type systems assume a strictly linear type system which requires re-binding of channel identifiers, such as in the functional setting of \citeauthor{Gay:2010:LAST}.
-As discussed previously, it would be interesting to study the system described by~\citeauthor{Fahndrich:2002}~\cite{Fahndrich:2002}, yet their system is specialised for their ``adopt'' and ``focus'' constructs and is not mechanised.
-\citeauthor{Mazurak:2010:LLT} present an extension to System F (termed \fpop) with a notion describing types as having linear or non-linear \textit{kind}~\cite{Mazurak:2010:LLT}.
-This extension is motivated by previous work on incorporating linearity into type systems.
-Overall, previous attempts either hamper the inclusion of desirable programming features (such as polymorphism), or do not adequately reflect non-linearity as the common case leading to awkward programming.
-\fpop avoids these deficiencies by categorising types into kinds whilst maintaining similar semantics to System F.
-Mechanised proofs for type soundness and parametricity are presented, and the semantics allow non-linear types to be treated as linear.
-Examples show the system can provide a wide range of permissions on interfaces, from full exclusive access to read-only shared references.
-It would be interesting to see if one could encode the ``adopt'' and ``focus'' constructs in this system.
-The ``adopt'' rule can be encoded similarly to the ShareRef example~\cite{Mazurak:2010:LLT}.
-It is not immediately clear how to encode the ``focus'' rule within the present semantics, as one needs to prevent non-linear access to its argument implying the presence of knowledge about aliasing within the typing judgements; absence of aliasing is not identified with linear kinded types in \fpop.
+Some recent mechanisation effort focuses on providing a basis for studying
+linearity within type systems. This work is of interest to the current project
+since most session-based type systems assume a strictly linear type system
+which requires re-binding of channel identifiers, such as in the functional
+setting of \citeauthor{Gay:2010:LAST}. As discussed previously, it would be
+interesting to study a system for supporting side-effecting functions (aka
+closures) for session channels. In order to achieve this, operational
+semantics for GV would need to be defined, and the recent draft by
+\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
+do this. \citeauthor{Mazurak:2010:LLT} present an extension to System F
+(termed \fpop) with a notion describing types as having linear or non-linear
+\textit{kind}~\cite{Mazurak:2010:LLT}. This extension is motivated by previous
+work on incorporating linearity into type systems. Overall, previous attempts
+either hamper the inclusion of desirable programming features (such as
+polymorphism), or do not adequately reflect non-linearity as the common case
+leading to awkward programming. \fpop avoids these deficiencies by
+categorising types into kinds whilst maintaining similar semantics to System
+F. Mechanised proofs for type soundness and parametricity are presented, and
+the semantics allow non-linear types to be treated as linear. Examples show
+the system can provide a wide range of permissions on interfaces, from full
+exclusive access to read-only shared references. The system is very close to
+the Alms language described by \citeauthor{Aldrich:2009} but has the advantage
+that the soundness results (type safety and parametricity) have been entirely
+mechanised in Coq. It would be interesting to see if one could encode the
+``adopt'' and ``focus'' constructs in this system to compare its
+expressiveness to Alms. The ``adopt'' rule can be encoded similarly to the
+ShareRef example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to
+encode the ``focus'' rule within the present semantics, as one needs to
+prevent non-linear access to its argument implying the presence of knowledge
+about aliasing within the typing judgements; absence of aliasing is not
+identified with linear kinded types in \fpop.
 
-While not related to programming language metatheory, \citeauthor{Gay:2001:FFP}~\cite{Gay:2001:FFP} provides a $\pi$-calculus framework mechanised in Isabelle/HOL.
-The aim is to build a mechanisaton of session-based type systems on top of the $\pi$-calculus framework.
-The use of de Bruijn indices for both bound and free variables causes issues with variable substitution often requiring permutation of typing environments (see \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} for a discussion on binder representations) and lifting of free variables during substitution; we avoid this issue in our representation choice (\S \ref{sec:approach}).
+While not related to programming language metatheory,
+\citeauthor{Gay:2001:FFP}~\cite{Gay:2001:FFP} provides a $\pi$-calculus
+framework mechanised in Isabelle/HOL. The aim is to build a mechanisaton of
+session-based type systems on top of the $\pi$-calculus framework. The use of
+de Bruijn indices for both bound and free variables causes issues with
+variable substitution often requiring permutation of typing environments (see
+\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} for a discussion on
+binder representations) and lifting of free variables during substitution; I
+avoid this issue in my representation choice (\S \ref{sec:approach}).
 
-The introduction of names for hypotheses during execution of custom tactics are an issue in the proof development.
-These names are automatically generated by the Isabelle/HOL system which creates a dependency between the names chosen and the tactic.
-Thus, one must be careful to use the exact names expected by the tactic.
-In the Coq proof assistant, defining custom tactics can be achieved using the Ltac language~\cite{Delahaye:2000:TLS}.
-Among other things, Ltac supports pattern matching on hypotheses and goal forms.
-Thus, if we wished to pattern match on an inductive type describing variables, we could do so without mentioning the actual variable name as follows:
+The introduction of names for hypotheses during execution of custom tactics
+are an issue in the proof development. These names are automatically generated
+by the Isabelle/HOL system which creates a dependency between the names chosen
+and the tactic. Thus, one must be careful to use the exact names expected by
+the tactic. In the Coq proof assistant, defining custom tactics can be
+achieved using the Ltac language~\cite{Delahaye:2000:TLS}. Among other things,
+Ltac supports pattern matching on hypotheses and goal forms. Thus, if we
+wished to pattern match on an inductive type describing variables, we could do
+so without mentioning the actual variable name as follows:
 
@@
 
-\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based type system providing session polymorphism.
-The system is more general than session polymorphism via subtyping~\cite{Gay:2005:SST}.
-more general how?
+\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based
+type system providing session polymorphism. The system is more general than
+session polymorphism via subtyping~\cite{Gay:2005:SST}. more general how?
 
- The authors mechanise their type system and its properties in the Coq proof assistant.
-However, the system mechanised is restricted to the $\pi$-calculus and the main focus is proving soundness properties for the deduction rules allowing polymorphism on input, rather than a more general basis for studying programming with session types.
+The authors mechanise their type system and its properties in the Coq proof
+assistant. However, the system mechanised is restricted to the $\pi$-calculus
+and the main focus is proving soundness properties for the deduction rules
+allowing polymorphism on input, rather than a more general basis for studying
+programming with session types.
 
@@
 
-I intend to mechanise the calculi described by \citeauthor{Wadler:2012}~\cite{Wadler:2012} using the Coq proof assistant.
-In the spirit of the POPLMARK challenge, I aim to utilise available libraries and infrastructure as much as possible.
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a metatheory library for handling, among other things, association lists (environments) and free variable representation.
-I intend to use the version distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which has been updated for Coq version 8.4pl4.
-\citeauthor{Park:2014:MMW} describe a technique for removing nonlinear contexts from typing judgements.
-While their work extended to \fpop it is not clear how to handle an environment containing both non-linear and linear types as in GV.
-I wish to maintain as close a relationship as possible to the paper system presented by \citeauthor{Wadler:2012}, so separating out the non-linear and linear components (as in \fpop) is not an option at this stage.
+I intend to mechanise the calculi described by
+\citeauthor{Wadler:2012}~\cite{Wadler:2012} using the Coq proof assistant. In
+the spirit of the POPLMARK challenge, I aim to utilise available libraries and
+infrastructure as much as
+possible. \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a
+metatheory library for handling, among other things, association lists
+(environments) and free variable representation. I intend to use the version
+distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which has been
+updated for Coq version 8.4pl4. \citeauthor{Park:2014:MMW} describe a
+technique for removing nonlinear contexts from typing judgements. While their
+work extended to \fpop it is not clear how to handle an environment containing
+both non-linear and linear types as in GV. I wish to maintain as close a
+relationship as possible to the paper system presented by
+\citeauthor{Wadler:2012}, so separating out the non-linear and linear
+components (as in \fpop) is not an option at this stage.
 
-Following \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, I use the locally nameless first-order representation for binders and free variables, preventing variable capture issues present in other first-order representations.
-Likewise, I make some simplifying assumptions about variable substitution, namely that \textit{closed} terms (terms with no unbound indices) are substituted for variables, which avoids having to define lifting constructs on the substitute term.
-The typing rules will not contain a rule for an unbound index so this decision is sound.
-A consequence is that we must define a notion of opening language terms involving binders as described by~\citeauthor{Aydemir:2008:EFM}.
-For example, an abstraction with body term $t$ which is well-typed if $t$ is well-typed when its unbound index, representing the argument to the abstraction, is replaced by some fresh variable $x$, written $t^x$.
-Some terms in GV require multiple binders but I believe it will be straightforward to extend this notion.
-Unfortunately, I must define similar operations on CP terms and handle substitution and opening on them as well as propositions which can accept propositional variables.
-I hope to be able to generalise as much as possible with the help of the Metatheory library.
-Alternatively, I could have chosen a higher-order abstract syntax (HOAS) approach such as the parametric version (PHOAS) described by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}.
-PHOAS relies on the metalanguage (Coq, in this case) for handling binders and substitution so has less overhead in managing syntax than the locally nameless approach.
-However, variables are now parameters defined at the level of the metalanguage so can represent terms not valid in the language being mechanised.
-Further, a higher-order approach would prevent the use of the Metatheory library which provides a number of useful tactics for manipulating typing contexts.
+Following \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, I use the
+locally nameless first-order representation for binders and free variables,
+preventing variable capture issues present in other first-order
+representations. Likewise, I make some simplifying assumptions about variable
+substitution, namely that \textit{closed} terms (terms with no unbound
+indices) are substituted for variables, which avoids having to define lifting
+constructs on the substitute term. The typing rules will not contain a rule
+for an unbound index so this decision is sound. A consequence is that we must
+define a notion of opening language terms involving binders as described
+by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
+$t$ which is well-typed if $t$ is well-typed when its unbound index,
+representing the argument to the abstraction, is replaced by some fresh
+variable $x$, written $t^x$. Some terms in GV require multiple binders but I
+believe it will be straightforward to extend this notion. Unfortunately, I
+must define similar operations on CP terms and handle substitution and opening
+on them as well as propositions which can accept propositional variables. I
+hope to be able to generalise as much as possible with the help of the
+Metatheory library. Alternatively, I could have chosen a higher-order abstract
+syntax (HOAS) approach such as the parametric version (PHOAS) described
+by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
+on the metalanguage (Coq, in this case) for handling binders and substitution
+so has less overhead in managing syntax than the locally nameless
+approach. However, variables are now parameters defined at the level of the
+metalanguage so can represent terms not valid in the language being
+mechanised. Further, a higher-order approach would prevent the use of the
+Metatheory library which provides a number of useful tactics for manipulating
+typing contexts.
 
-For handling binder freshness, I chose to adopt the cofinite quantification approach described by~\cite{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, which excludes a finite set of variables from being considered as the binder.
-In contrast, the traditional ``exists-fresh'' approach, where the binder is only required to be fresh within the abstraction's body, does not produce strong enough induction principles in some cases.
-This change will necessitate changes to the typing rules of GV and CP.
-Consider for instance the connect with rule in GV~\cite{Wadler:2012}:
+For handling binder freshness, I chose to adopt the cofinite quantification
+approach described by~\cite{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, which
+excludes a finite set of variables from being considered as the binder. In
+contrast, the traditional ``exists-fresh'' approach, where the binder is only
+required to be fresh within the abstraction's body, does not produce strong
+enough induction principles in some cases. This change will necessitate
+changes to the typing rules of GV and CP. Consider for instance the connect
+with rule in GV~\cite{Wadler:2012}:
 
@@
 
-Consider also the rule for output in CP~\cite{Wadler:2012} and its corresponding version used here:
+Consider also the rule for output in CP~\cite{Wadler:2012} and its
+corresponding version used here:
 
@@
 
-The rule here ... is this needed? What are we showing here with the different version?
+The rule here ... is this needed? What are we showing here with the different
+version?
 
-While the aim is to be faithful to \citeauthor{Wadler:2012}'s description, I may need to alter the approach slightly to avoid tricky formulations.
-For instance, instead of providing n-ary branch and choice in GV, we provide only binary versions of these operators to simplify the development.
-This change is not restrictive however, since CP itself provides only binary versions of plus ($\oplus$) and with ($\with$) constructs.
-Likewise, on a more technical note, we amalgamate all types into one inductive definition and provide a predicate for restricting typing rules to consider only valid session types.
-\citeauthor{Wadler:2012} defines GV types as a set of mutually recursive definitions; session types are types which may contain types as subcomponents (arguments to send, for example).
-Unfortunately, handling mutually inductive definitions in Coq can be quite involved; requiring one to either rely on the Coq system to provide a stronger mutual induction principle or defining one manually.
-Such a definition is possible, but it complicates elimination via the induction principle.
+While the aim is to be faithful to \citeauthor{Wadler:2012}'s description, I
+may need to alter the approach slightly to avoid tricky formulations. For
+instance, instead of providing n-ary branch and choice in GV, we provide only
+binary versions of these operators to simplify the development. This change is
+not restrictive however, since CP itself provides only binary versions of plus
+($\oplus$) and with ($\with$) constructs. Likewise, on a more technical note,
+we amalgamate all types into one inductive definition and provide a predicate
+for restricting typing rules to consider only valid session
+types. \citeauthor{Wadler:2012} defines GV types as a set of mutually
+recursive definitions; session types are types which may contain types as
+subcomponents (arguments to send, for example). Unfortunately, handling
+mutually inductive definitions in Coq can be quite involved; requiring one to
+either rely on the Coq system to provide a stronger mutual induction principle
+or defining one manually. Such a definition is possible, but it complicates
+elimination via the induction principle.
 % ( since the standard induction tactic will not work?)
 
-Another aspect of the encoding is how to define terms.
-The intrinsic encoding described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one approach, indexing terms by their type so as to prevent ill-formed terms from being constructed.
-However, in the encoding for GV there are issues with using this encoding: $(1)$ it is not immediately clear how to handle linear contexts using the de Bruijn variable encoding, since in the intrinsic setting the environment supports weakening; $(2)$ an intrinsic encoding of terms would be complicated by the need to enforce that session types occur in certain instances (by use of a predicate); and $(3)$ well-typed terms require extra assumptions about binder freshness which cannot reasonably be expressed as a function type.
-For these reasons, an intrinsic approach does not offer much benefit since a well-typed term relation would still need to be defined.
-Therefore, I chose an extrinsic encoding for terms and define well-typed terms as a separate inductive type.
+Another aspect of the encoding is how to define terms. The intrinsic encoding
+described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
+approach, indexing terms by their type so as to prevent ill-formed terms from
+being constructed. However, in the encoding for GV there are issues with using
+this encoding: $(1)$ it is not immediately clear how to handle linear contexts
+using the de Bruijn variable encoding, since in the intrinsic setting the
+environment supports weakening; $(2)$ an intrinsic encoding of terms would be
+complicated by the need to enforce that session types occur in certain
+instances (by use of a predicate); and $(3)$ well-typed terms require extra
+assumptions about binder freshness which cannot reasonably be expressed as a
+function type. For these reasons, an intrinsic approach does not offer much
+benefit since a well-typed term relation would still need to be
+defined. Therefore, I chose an extrinsic encoding for terms and define
+well-typed terms as a separate inductive type.
 
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of their language proof infrastructure is proportional to the number of binding constructs.
-In the case of a simply typed lambda calculus this is not onerous, but GV has four binding constructs, CP has six and propositions have two.
-It may be true that this approach results in too many auxiliary definitions and operations not relevant to the actual properties one wishes to prove.
-However, I believe that once this infrastructure has been defined, the system can be extended without concern for these binding constructs.
+\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
+their language proof infrastructure is proportional to the number of binding
+constructs. In the case of a simply typed lambda calculus this is not onerous,
+but GV has four binding constructs, CP has six and propositions have two. It
+may be true that this approach results in too many auxiliary definitions and
+operations not relevant to the actual properties one wishes to prove. However,
+I believe that once this infrastructure has been defined, the system can be
+extended without concern for these binding constructs.
 
@@
 
-I aim to have completed the GV and CP definitions including reduction and commuting conversion rules for CP by the end of December.
-This will require definition of infrastructure for handling binders for both GV, CP channels and CP propositional variables, and inductive definitions for well-typed terms, commuting conversions and CP principal cut reductions.
-So far I have completed some of the definitions for well-typed terms and CP process rules; the remaining rules require the infrastructure for substitution and opening of binders.
+I aim to have completed the GV and CP definitions including reduction and
+commuting conversion rules for CP by the end of December. This will require
+definition of infrastructure for handling binders for both GV, CP channels and
+CP propositional variables, and inductive definitions for well-typed terms,
+commuting conversions and CP principal cut reductions. So far I have completed
+some of the definitions for well-typed terms and CP process rules; the
+remaining rules require the infrastructure for substitution and opening of
+binders.
 
-After December, it remains to define the translation between GV and CP which I envisage will be the bulk of the work, requiring several rules on typing derivations; I envisage being able to define these as inductive datatype constructors but I shall take care to reuse any suitable libraries that are available, e.g.
-a library for handling CPS translations.
+After December, it remains to define the translation between GV and CP which I
+envisage will be the bulk of the work, requiring several rules on typing
+derivations; I envisage being able to define these as inductive datatype
+constructors but I shall take care to reuse any suitable libraries that are
+available, e.g. a library for handling CPS translations.
 
- For this step of the work, I allocated one month. For the remainder, I shall look to extend the definitions of GV and CP using the definitions I have provided
+For this step of the work, I allocated one month. For the remainder, I shall
+look to extend the definitions of GV and CP using the definitions I have
+provided
 
-I am going to look into adapting adoption and focus for the GV setting to allow one to control aliasing on session typed channels independent of their session characteristics so the aim is not to manipulate a linear type but to provide a nonlinear view of some resource and when changes are to be made, enforce some non-aliased view such that there no references in the context
-perhaps the notion I'm looking for is to prevent sending on a guarded type or enforce the guarded type must be preserved in the adoption block
-the trick in the adoption and focus system is that the type will be deallocated when its adopter's lifetime ends; I suspect in our system this is not useful since we will only be permitted to use a channel in a non-aliased context
+I am going to look into adapting adoption and focus for the GV setting to
+allow one to control aliasing on session typed channels independent of their
+session characteristics so the aim is not to manipulate a linear type but to
+provide a nonlinear view of some resource and when changes are to be made,
+enforce some non-aliased view such that there no references in the context
+perhaps the notion I'm looking for is to prevent sending on a guarded type or
+enforce the guarded type must be preserved in the adoption block the trick in
+the adoption and focus system is that the type will be deallocated when its
+adopter's lifetime ends; I suspect in our system this is not useful since we
+will only be permitted to use a channel in a non-aliased context
 

### 2014-11-26 proposal/proposal.tex
Expanded the section session polymorphism.

@@
 \addbibresource{proposal.bib}
+
+\input{macros}
 
@@
 which requires re-binding of channel identifiers, such as in the functional
-setting of \citeauthor{Gay:2010:LAST}. As discussed previously, it would be
-interesting to study a system for supporting side-effecting functions (aka
-closures) for session channels. In order to achieve this, operational
-semantics for GV would need to be defined, and the recent draft by
-\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
-do this. \citeauthor{Mazurak:2010:LLT} present an extension to System F
-(termed \fpop) with a notion describing types as having linear or non-linear
-\textit{kind}~\cite{Mazurak:2010:LLT}. This extension is motivated by previous
-work on incorporating linearity into type systems. Overall, previous attempts
-either hamper the inclusion of desirable programming features (such as
-polymorphism), or do not adequately reflect non-linearity as the common case
-leading to awkward programming. \fpop avoids these deficiencies by
-categorising types into kinds whilst maintaining similar semantics to System
-F. Mechanised proofs for type soundness and parametricity are presented, and
-the semantics allow non-linear types to be treated as linear. Examples show
-the system can provide a wide range of permissions on interfaces, from full
-exclusive access to read-only shared references. The system is very close to
-the Alms language described by \citeauthor{Aldrich:2009} but has the advantage
-that the soundness results (type safety and parametricity) have been entirely
-mechanised in Coq. It would be interesting to see if one could encode the
-``adopt'' and ``focus'' constructs in this system to compare its
-expressiveness to Alms. The ``adopt'' rule can be encoded similarly to the
-ShareRef example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to
-encode the ``focus'' rule within the present semantics, as one needs to
-prevent non-linear access to its argument implying the presence of knowledge
-about aliasing within the typing judgements; absence of aliasing is not
-identified with linear kinded types in \fpop.
+setting of \citeauthor{Gay:2010:LAST}. \citeauthor{Mazurak:2010:LLT} present
+an extension to System F (termed \fpop) with a notion describing types as
+having linear or non-linear \textit{kind}~\cite{Mazurak:2010:LLT}. This
+extension is motivated by previous work on incorporating linearity into type
+systems. Overall, previous attempts either hamper the inclusion of desirable
+programming features (such as polymorphism), or do not adequately reflect
+non-linearity as the common case leading to awkward programming. \fpop avoids
+these deficiencies by categorising types into kinds whilst maintaining similar
+semantics to System F. Mechanised proofs for type soundness and parametricity
+are presented, and the semantics allow non-linear types to be treated as
+linear. Examples show the system can provide a wide range of permissions on
+interfaces, from full exclusive access to read-only shared references. The
+system is very close to the Alms language described by
+\citeauthor{Aldrich:2009} but has the advantage that the soundness results
+(type safety and parametricity) have been entirely mechanised in Coq. It would
+be interesting to see if one could encode the ``adopt'' and ``focus''
+constructs in this system to compare its expressiveness to Alms. The ``adopt''
+rule can be encoded similarly to the ShareRef
+example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to encode the
+``focus'' rule within the present semantics, as one needs to prevent
+non-linear access to its argument implying the presence of knowledge about
+aliasing within the typing judgements; absence of aliasing is not identified
+with linear kinded types in \fpop.
 
@@
 type system providing session polymorphism. The system is more general than
-session polymorphism via subtyping~\cite{Gay:2005:SST}. more general how?
-
-The authors mechanise their type system and its properties in the Coq proof
-assistant. However, the system mechanised is restricted to the $\pi$-calculus
-and the main focus is proving soundness properties for the deduction rules
-allowing polymorphism on input, rather than a more general basis for studying
-programming with session types.
+session polymorphism via subtyping~\cite{Gay:2005:SST} in that it is capable
+of typing a generic forwarding process between mutually dual session
+types. Defining such a process is not possible using subtyping since one
+wishes to restrict the types to be dual, not a subtype of some other
+type. This more general form of polymorphism is achieved by permitting input
+on all session types (including \lstinline{end}, the terminal session type)
+and then employing transition hypotheses to ensure the type permits such
+input. The authors mechanise their type system and its properties in the Coq
+proof assistant. However, the mechanised system is restricted to the
+$\pi$-calculus and the main focus is proving soundness properties for the
+deduction rules allowing polymorphism, rather than a more general basis for
+studying programming with session types.
 
@@
 
+As discussed previously, it would be
+interesting to study a system for supporting side-effecting functions (aka
+closures) for session channels. In order to achieve this, operational
+semantics for GV would need to be defined, and the recent draft by
+\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
+do this.
+
 For this step of the work, I allocated one month. For the remainder, I shall

### 2014-11-26 proposal/proposal.tex
Finished read-through and update sections: just pictures left.

@@
 particular system. The framework will be heavily based on calculi developed by
-\citeauthor{Wadler:2012}~\cite{Wadler:2012} describing a relationship between
+\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
 classical linear logic (CP) and a functional language with session types
@@
 more recent
-works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2012}. Certainly, it is the
+works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2014}. Certainly, it is the
 case that an aliased approach would fit better with mainstream programming
@@
 channel since this could lead to invalid communication e.g. trying to
-communicate on a closed channel.
+communicate on a closed channel. Additionally, a linear object is consumed
+upon being adopted and will be deallocated when its adopter's lifetime ends,
+although an alternative semantics could be defined which returned the linear
+resource to the context. Moreover, \citeauthor{Fahndrich:2002} do not provide
+operational semantics or soundness results for the ``adoption and focus'' type
+system.
 
@@
 type system can support the ``adoption and focus'' system but do not specify
-details or whether any changes are necessary to their typing
-judgements. Moreover, \citeauthor{Fahndrich:2002} do not provide operational
-semantics or soundness results for the ``adoption and focus'' type system.
+details or whether any changes are necessary to their typing judgements.
 
@@
 
-\citeauthor{Wadler:2012}'s recent work~\cite{Wadler:2012} brings together a
+\citeauthor{Wadler:2014}'s recent work~\cite{Wadler:2014} brings together a
 process calculus, CP, based on the earlier work by
@@
 process reduction by \citeauthor{Caires:2010:STI}. Additionally, rather than a
-two-sided sequent calculus, \citeauthor{Wadler:2012} uses one-sided sequents
+two-sided sequent calculus, \citeauthor{Wadler:2014} uses one-sided sequents
 for defining the process calculus, which provides a more intuitive
@@
 I intend to mechanise the calculi described by
-\citeauthor{Wadler:2012}~\cite{Wadler:2012} using the Coq proof assistant. In
+\citeauthor{Wadler:2014}~\cite{Wadler:2014} using the Coq proof assistant. In
 the spirit of the POPLMARK challenge, I aim to utilise available libraries and
@@
 relationship as possible to the paper system presented by
-\citeauthor{Wadler:2012}, so separating out the non-linear and linear
+\citeauthor{Wadler:2014}, so separating out the non-linear and linear
 components (as in \fpop) is not an option at this stage.
@@
 hope to be able to generalise as much as possible with the help of the
-Metatheory library. Alternatively, I could have chosen a higher-order abstract
-syntax (HOAS) approach such as the parametric version (PHOAS) described
+Metatheory library.
+
+Alternatively, I could have chosen a higher-order abstract syntax (HOAS)
+approach such as the parametric version (PHOAS) described
 by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
@@
 enough induction principles in some cases. This change will necessitate
-changes to the typing rules of GV and CP. Consider for instance the connect
-with rule in GV~\cite{Wadler:2012}:
+changes to the typing rules of GV and CP. Consider for instance the Connect
+rule in GV~\cite{Wadler:2014} (Connect) and its corresponding cofinite
+quantification version (ConnectCO):
 
@@
 
-The conclusion binds a variable
-
-Consider also the rule for output in CP~\cite{Wadler:2012} and its
-corresponding version used here:
+The conclusion binds a variable in Connect but a type in ConnectCO. Thus, in
+the premise for ConnectCO one can choose a suitably fresh $x$.
+The same change is made to the rules for CP so that, for example, output which allocates a fresh variable $y$ no longer names $y$ in the conclusion:
 
@@
 
-The rule here ... is this needed? What are we showing here with the different
-version?
+Since the name $y$ is assumed to be fresh in the original presentation these
+changes do not alter the typing of processes nor the semantics of the
+translation from GV to CP. An equivalence between ``exists-fresh'' and
+cofinite quantification definitions of the simply-typed $\lambda$ calculus is
+given by \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. In general, I
+do not aim to formally prove adequacy of the mechanisation.
 
-While the aim is to be faithful to \citeauthor{Wadler:2012}'s description, I
+While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
 may need to alter the approach slightly to avoid tricky formulations. For
-instance, instead of providing n-ary branch and choice in GV, we provide only
-binary versions of these operators to simplify the development. This change is
-not restrictive however, since CP itself provides only binary versions of plus
-($\oplus$) and with ($\with$) constructs. Likewise, on a more technical note,
-we amalgamate all types into one inductive definition and provide a predicate
-for restricting typing rules to consider only valid session
-types. \citeauthor{Wadler:2012} defines GV types as a set of mutually
+instance, instead of providing $n$-ary branch and choice in GV, we provide
+only binary versions of these operators to simplify the development. This
+change is not restrictive however, since CP itself provides only binary
+versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
+technical note, I amalgamate all types into one inductive definition and
+provide a predicate for restricting typing rules to consider only valid
+session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
 recursive definitions; session types are types which may contain types as
@@
 envisage will be the bulk of the work, requiring several rules on typing
-derivations; I envisage being able to define these as inductive datatype
-constructors but I shall take care to reuse any suitable libraries that are
-available, e.g. a library for handling CPS translations.
+derivations; I plan to define these as inductive datatype constructors but I
+shall take care to reuse any suitable libraries that are available, e.g. a
+library for handling CPS translations. For this step of the work, I allocated
+one month but this is a cautious estimate accounting some slack time for any
+delays in preceding stages of the development; I suspect this stage will not
+require a full month.
 
-As discussed previously, it would be
+For the remainder, I shall look to extend the definitions of GV and CP using
+the definitions I have provided. As discussed previously, it would be
 interesting to study a system for supporting side-effecting functions (aka
@@
 \citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
-do this.
+do this~\cite{Lindley:2014:SPS}. I allocated the remaining time (approx. two
+months) to perform this extension because I believe it requires the most
+development: extending existing definitions to handle side-effecting
+functions, defining operational semantics, proving subject reduction and
+related properties. Additionally, it will also require a theoretical treatment
+prior to undertaking the development task, in contrast to the preceding
+development work which mostly utilises results from the literature.
 
-For this step of the work, I allocated one month. For the remainder, I shall
-look to extend the definitions of GV and CP using the definitions I have
-provided
-
-I am going to look into adapting adoption and focus for the GV setting to
-allow one to control aliasing on session typed channels independent of their
-session characteristics so the aim is not to manipulate a linear type but to
-provide a nonlinear view of some resource and when changes are to be made,
-enforce some non-aliased view such that there no references in the context
-perhaps the notion I'm looking for is to prevent sending on a guarded type or
-enforce the guarded type must be preserved in the adoption block the trick in
-the adoption and focus system is that the type will be deallocated when its
-adopter's lifetime ends; I suspect in our system this is not useful since we
-will only be permitted to use a channel in a non-aliased context
+I shall review and update the above plan throughtout the project as I proceed
+to the various milestones I have set out, and whenever I identify changing
+circumstances.
 

### 2014-11-27 proposal/proposal.tex
Added effecting lambda abs. and Caires & Pfenning &/+ rules.

@@
 \usepackage{cmll}
-
+\usepackage{prooftree}
+\usepackage{amsmath}
+\newcommand{\typestate}[2]{{#1}\gg{#2}}
 \usepackage[style=alphabetic,sorting=nymt,natbib=true,
@@
 
-\input{macros}
+\input{propmacros}
 
@@
 
-... lambda abstraction ...
+%% \begin{prooftree}
+%% \AxiomC{$\Gamma , x \colon A \triangleright t \colon T \triangleleft \Gamma ' , x \colon B$}
+%% \UnaryInfC{$\Gamma \triangleright t \colon (A \gg B) \rightarrow T \triangleleft \Gamma '$} 
+%% \end{prooftree}
+
+\begin{gather*}
+  \frac{
+    \typeds{\Gamma,\typedid{x}{A}}{t}{T}{\Gamma',\typedid{x}{B}}
+  }{
+    \typeds{\Gamma}{\abs{x}{t}}{\fn{\typestate{A}{B}}{T}}{\Gamma'}
+  }\quad
+\end{gather*}
 
@@
 
-... here are the rules ...
+\input{sill-branch-choice-rules}
 
 Note that branch and choice both have some notion of branching (R rules for
-+/L rules for \&) and at least for \& this is counter-intuitive in
-session-based setting, since the intutition for \& is that of offering a
-choice between alternatives rather than making the choice. Here, however, the
-intuition should be extended to account for its location in the typing
-judgement. The channel is on the left of the sequent which caires and pfenning
-suggest as some service providing to the process.
+$\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
+a session-based setting, since the intutition for \& is that of
+\emph{offering} a choice between alternatives rather than \emph{making} the
+choice. Here, however, the intuition should be extended to account for its
+location in the typing judgement. The channel is on the left of the sequent,
+and \citeauthor{Caires:2010:STI} suggest one consider the context as services
+being provided to the typed process. Thus, the type of channels in the context
+is really the type of the other endpoint of the channel, and the typed process
+behaves according to the dual protocol.
+For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to represent the dual of a protocol $A$, the typed process is behaving according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from alternatives (in this case, $\inrterm$) and then behaving according to the selected protocol (in this case, protocol $B^\bot$).
 
@@
 
-I shall review and update the above plan throughtout the project as I proceed
-to the various milestones I have set out, and whenever I identify changing
-circumstances.
+There is a risk of being unable to perform the side-effecting extension as
+sketched in \S \ref{sec:approach}. I shall review and update the above plan
+throughtout the project as I proceed to the various milestones I have set out,
+and if required will alter the above plans if necessary.
 

### 2014-11-28 proposal/proposal.tex
Added Buy/Quote Client-Server example to parallel C&P's.

@@
 
+% Thanks to Stefan Kottwitz on tex.stackexhange.com:
+% http://tex.stackexchange.com/questions/69662/how-to-globally-change-the-spacing-around-equations
+%% \expandafter\def\expandafter\normalsize\expandafter{%
+%%     \normalsize
+%%     \setlength\abovedisplayskip{0pt}
+%%     \setlength\belowdisplayskip{0pt}
+%%     \setlength\abovedisplayshortskip{0pt}
+%%     \setlength\belowdisplayshortskip{0pt}
+%% }
+
 \addbibresource{proposal.bib}
@@
 \input{propmacros}
+
+\allowdisplaybreaks
 
@@
 \newcommand{\fpop}{System F${}^\circ$\xspace}
-\newcommand{\1}{\textbf{1}\xspace}
 
@@
 for lambda abstraction could be:
-
-%% \begin{prooftree}
-%% \AxiomC{$\Gamma , x \colon A \triangleright t \colon T \triangleleft \Gamma ' , x \colon B$}
-%% \UnaryInfC{$\Gamma \triangleright t \colon (A \gg B) \rightarrow T \triangleleft \Gamma '$} 
-%% \end{prooftree}
-
 \begin{gather*}
@@
 $\pi$DILL. Consider their rules for branch and choice in $\pi$DILL:
-
 \input{sill-branch-choice-rules}
@@
 is really the type of the other endpoint of the channel, and the typed process
-behaves according to the dual protocol.
-For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to represent the dual of a protocol $A$, the typed process is behaving according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from alternatives (in this case, $\inrterm$) and then behaving according to the selected protocol (in this case, protocol $B^\bot$).
+behaves according to the dual protocol. For example, considering
+$\SILLbranchLTwo$, and taking $A^\bot$ to represent the dual of a protocol $A$
+as defined by \citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process
+is behaving according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting
+from alternatives (in this case, $\inrterm$) and then behaving according to
+the selected protocol (in this case, protocol $B^\bot$).
 
-While in the case of branch and choice, which have different semantics this is
-mitigated somewhat but as can be seen, there are two rules for selecting a
-branch, one for branch and another for choice if we consider their server
-example
+As noted by Wadler~\cite{Wadler:2014} there is a certain amount of redundancy
+in the two-sided nature of the rules forcing one to commit to a certain
+convention when typing a communication system. Consider, an equivalent system
+to the buy/quote example given by
+\citeauthor{Caires:2010:STI}~\cite[\S~3]{Caires:2010:STI}:
+\begin{gather*}
+\sdef{ClientProto}{\BChoice{(\tensor{Book}
+                            {(\tensor{Card}{(\lfn{Receipt}{\1})})})}
+                           {(\tensor{Book}{(\lfn{Quote}{\1})})}}
+\end{gather*}
 
-As noted by Wadler...jfp there is a certain amount of redundancy in the
-two-sided nature of the rules forcing one to commit to a certain convention
-when typing a communication system. Consider, an equivalent system to the
-buy/quote example given by caires and pfen. In the following, the types Book,
-Quote and Receipt can be considered synoms for the type \1 as in caires and
-pfenning.
+In the protocol defined above, a client selects between buying a book or
+receiving a price quote for a book. In the buy option case, the client sends
+the name of the book to the server and their card details, then the server
+sends the receipt for the purchase. In the quote option case, the client sends
+the name of the book and receives a price quote from the server. I have
+replaced the types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more
+meaningful type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate;
+all are synonyms for the type $\1$ which denotes the terminal session
+type. The client and server process bodies are as follows:
+\begin{gather*}
+\sdef{ClientBody_c}
+     {\CPinr{c}{\SLoutput{1984}{c}{\SLinput{c}{quote}{\0}}}}\\
+\sdef{ServerBody_c}
+     {\CPcase{c}
+             {\SLinput{c}{book}{\SLinput{c}{card}{\SLoutput{receipt}{c}{\0}}}}
+             {\SLinput{c}{book}{\SLoutput{quote}{c}{\0}}}}
+\end{gather*}
 
-... server example here ...
+\begin{samepage}
+The typings for the client and server processes are given by:
+\begin{gather*}
+\tsequent{\cdot}{\cdot}{ClientBody_c}{\typedid{c}{ClientProto}}
+\tag{\mkTrule{ClientBody}} \label{eq:cbtyp}
+\\ \tsequent{\cdot}{\typedid{c}{ClientProto}}{ServerBody_c}{\typedid{-}{\1}}
+\tag{\mkTrule{ServerBody}} \label{eq:sbtyp}
+\end{gather*}
+\end{samepage}
 
-The derivation for ClientBody is T1R, T implies R, T tensor R, T plus
-R2. Except for the final rule, these are exact opposites of the rules used for
-...
+The derivation for \eqref{eq:cbtyp} is $T\1 R$, $\lfn{T}{R}$, $\tensor{T}{R}$,
+$\BChoice{T}{R_2}$ and \eqref{eq:sbtyp} is derivable from
+$\BChoice{T}{L}$. Except for the final rule, these are exact dual rules of the
+rules used in example by \citeauthor{Caires:2010:STI}. Indeed, consider the
+composite process:
+\begin{gather*}
+\sdef{QSimple}{\Spar{c}{ClientBody_c}{ServerBody_c}}\\
+\tsequent{\cdot}{\cdot}{QSimple}{\typedid{-}{\1}}
+\end{gather*}
 
-The final system is near identical to the one produced by the example given
-caires and pfenning. Here the client communicates along the channel $c$ on the
-right and the server utilises the channel $c$ on the left. This typing, while
-correct, violates the intuition given by caires and pfenning of the context
-providing the services used by the process. The example above suggests the
-server utilises some channel provided by the client. On the other hand, one
-could imagine such a channel being setup between two communicating agents; the
-``owner'' of the channel is largely irrelevant. Note that clientproto is in
-fact the dual of serverproto in Wadler's classical process calculus.
-
-most important: duality of sessions is obscured in $\pi$DILL because of the
-nature of the typing judgements
+The final system is near identical to the one produced in the example given by
+\citeauthor{Caires:2010:STI} except the individual processes within the
+parallel composition have been swapped. Here the client communicates along the
+channel $c$ on the right-side of the sequent and the server utilises the
+channel $c$ on the left of the sequent. This typing, while correct, violates
+the intuition given by \citeauthor{Caires:2010:STI} of the context providing
+the services used by the process, since one would usually associate provision
+of services with the server not the client. Instead, the example above
+suggests the server utilises some channel provided by the client. On the other
+hand, one could imagine such a channel being setup between two communicating
+agents; the ``owner'' of the channel is largely irrelevant. Note that
+$ClientProto$ is in fact the dual of the $ServerProto$ definition by
+\citeauthor{Caires:2010:STI} using the definition of dual used in Wadler's
+classical process calculus~\cite{Wadler:2014}. In $pi$DILL, the duality
+between endpoints of the same channel are obscured by the nature of the typing
+judgements.
 

### 2014-11-28 proposal/proposal.tex
Missed pi symbol.
Added focus rule and associated description.

@@
 \citeauthor{Caires:2010:STI} using the definition of dual used in Wadler's
-classical process calculus~\cite{Wadler:2014}. In $pi$DILL, the duality
+classical process calculus~\cite{Wadler:2014}. In $\pi$DILL, the duality
 between endpoints of the same channel are obscured by the nature of the typing
@@
 liveness mechanism, so GV typing judgements would need to be extended such
-that the capabilities can be included in soundness results. A extension to the
+that the capabilities can be included in soundness results. An extension to the
 translation to CP to incorporate the ``adopt'' and ``focus'' constructs would
@@
 such that the type is maintained during the focus operation:
+\input{focus-rule}
 
-... copy of focus operation here ...
+The typing judgements include \emph{capabilities} that track aliasing to heap
+objects, e.g. $\AFcap{\rho}{h}$ denotes a mapping from a static name $\rho$ to
+a heap type $h$, and a variable $x$ may be typed as $\typedid{x}{\AFtr{\rho}}$
+indicating that $x$ refers to an object of type $h$. The purpose of the
+capabilities is to enforce restrictions on when an object can be
+referenced. In other words, if $\AFcap{\rho}{h}$ is in the capabilities at
+some program point then $x$ may be used, otherwise such use is prohibited. In
+the [focus] rule, evaluating $e_1$ retrieves the object to be focussed. Note
+the type of $e_1$ is a ``guarded'' type. The guard $G$ is the static name of
+some object which ``adopted''~\footnote{I omit the adopt typing rule for
+  brevity.} an object of heap type $h$. If a program point has
+$\AFcap{G}{h_1}$ in the capabilities for some $h_1$ then all objects guarded
+by $G$ may be used; $\capsequent{\Delta}{C_1}{G}$ denotes containment of a
+such a capability in $C_1$. The guarded type may be thought of as policing
+manipulation of linear components within $h$. Treating $\otimes$ as the
+disjoint union for capability sets, the typing for $e_2$ temporarily removes
+the guard $G$ from the set of capabilities, and adds a new variable $x$, and
+associated capability, with a type permitting access to the object of type $h$
+i.e. a non-guarded type. Removing $G$ will prohibit access to allow objects
+guarded by $G$ for the during of $e_2$. The final capabilities after executing
+$e_2$ are required to be $\tensor{C_3}{\AFcap{\rho}{h}}$, enforcing that the
+type must remain as $h$. The typing of the $\letterm$ expression ensures that
+the guard $G$ (contained in $C_1$) is restored.
 

### 2014-11-28 proposal/proposal.tex
Added cofinite quantification changes to Wadler's presentation.

@@
 changes to the typing rules of GV and CP. Consider for instance the Connect
-rule in GV~\cite{Wadler:2014} (Connect) and its corresponding cofinite
-quantification version (ConnectCO):
+rule in GV~\cite{Wadler:2014} (C) and its corresponding cofinite
+quantification version (CCO):
+\input{with-connect-rule}
 
-... connect with rule here ... ... connect rule using cofinite quantification
-
-The conclusion binds a variable in Connect but a type in ConnectCO. Thus, in
-the premise for ConnectCO one can choose a suitably fresh $x$.
-The same change is made to the rules for CP so that, for example, output which allocates a fresh variable $y$ no longer names $y$ in the conclusion:
-
-... output in CP ... ... version used here with cofinite quantification ...
+The conclusion binds a variable in (C) but a type in (CCO). Thus, in the
+premise for (CCO) one can choose a suitably fresh $x$ in each premise. Note
+that, $L$ can be chosen such that $\Psi \cup \Phi \cup \fv(M) \cup \fv(N)
+\subseteq L$, where $\fv(M)$ are the set of free variables in term $M$, and
+$x$ can be the same in both premises. Similar changes are made to the rules
+for CP so that, for example, output which allocates a fresh variable $y$ no
+longer names $y$ in the conclusion: \input{cp-output}
 
@@
 translation from GV to CP. An equivalence between ``exists-fresh'' and
-cofinite quantification definitions of the simply-typed $\lambda$ calculus is
+cofinite quantification definitions of the simply-typed $\lambda$-calculus is
 given by \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. In general, I

### 2014-11-28 proposal/proposal.tex
Second draft.

@@
 channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the
-channel enviornment. The typing judgements and term reduction relation must be
+channel environment. The typing judgements and term reduction relation must be
 augmented to handle these environments. Unfortunately, a side-effect of this
@@
 normalisation and confluence. Confluence provides the guarantee that no race
-conditions can occur, strong normalisation prohibits nonterminating programs
+conditions can occur, strong normalisation prohibits non-terminating programs
 and type safety assures deadlock-freedom. Deadlock-freedom is implied by the
@@
 sketched in \S \ref{sec:approach}. I shall review and update the above plan
-throughtout the project as I proceed to the various milestones I have set out,
+throughout the project as I proceed to the various milestones I have set out,
 and if required will alter the above plans if necessary.

### 2014-11-28 proposal/proposal.tex
Consistent pronoun usage: using "I" instead of "we" for now.

@@
 more usable tool for programming distributed communication systems (the
-ultimate goal, of course, is to write more correct programs), we cannot be
+ultimate goal, of course, is to write more correct programs), one cannot be
 expected to program directly in the $\pi$-calculus; a more mainstream
@@
 A number of systems incorporate both a functional language surface syntax with
-a translation to a more primitive process calculus. We shall focus on two such
+a translation to a more primitive process calculus. I shall focus on two such
 systems presented recently which have a connection to linear logic.
@@
 achieved using the Ltac language~\cite{Delahaye:2000:TLS}. Among other things,
-Ltac supports pattern matching on hypotheses and goal forms. Thus, if we
-wished to pattern match on an inductive type describing variables, we could do
-so without mentioning the actual variable name as follows:
+Ltac supports pattern matching on hypotheses and goal forms. Thus, if one
+wished to pattern match on an inductive type describing variables, one could
+do so without mentioning the actual variable name as follows:
 
@@
 constructs on the substitute term. The typing rules will not contain a rule
-for an unbound index so this decision is sound. A consequence is that we must
+for an unbound index so this decision is sound. A consequence is that I must
 define a notion of opening language terms involving binders as described
@@
 may need to alter the approach slightly to avoid tricky formulations. For
-instance, instead of providing $n$-ary branch and choice in GV, we provide
+instance, instead of providing $n$-ary branch and choice in GV, I provide
 only binary versions of these operators to simplify the development. This

### 2014-12-04 proposal/proposal.tex
Added (some) work plan, session types disc. and fixed "binds" anomaly.

@@
 
+\usepackage{tabularx}
+\usepackage{tabulary}
+
 \DeclareSortingScheme{nymt}{
@@
 }
+
+\newcommand{\Deliverable}[2]
+           {\multicolumn{2}{|c|}{\textbf{D{#1}: {#2}}} \\ \hline}
+\newcommand{\DSubD}[3]{D{#1}.{#2} & #3 \\ \hline}
+\newcommand{\DTask}[6]{D#1T#2 & #3 & #4 & #5 & #6 \\ \hline}
+\newcommand{\IDTsk}[2]{Task D{#1}T{#2}}
 
@@
 
+\subsubsection{Preamble}
+
+The definition of session types will be important for the remainder:
+\input{session-types.tex}
+
+Here, I use a modified version of the definition given by
+\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} where I have omitted recursive
+session types. The type \lstinline{end} denotes the terminal session type and
+a channel typed as such may not perform any communication. A communication
+channel with session type $\inpt{T}{S}$ expects to receive a term of type $T$
+and then continues with session type $S$. Conversely, a channel with type
+$\outpt{T}{S}$ can send a term of type $T$ and then has type $S$. Thus, input
+and output are considered \emph{dual} session types, in the sense that a
+channel with one endpoint having type $\inpt{T}{S}$ and the other endpoint
+having type $\outpt{T}{\Sbar{S}}$, where $\Sbar{S}$ is the dual of $S$,
+permits communication between two agents, allowing one to send a term of type
+$T$ and the other to receive it with continuing endpoint types $\Sbar{S}$ and
+$S$, respectively. Similarly, the remaining types $\Branch{l_i}{S_i}{i\in I}$
+and $\Choice{l_i}{S_i}{i\in I}$ are duals of each other. A channel of type
+$\Branch{l_i}{S_i}{i\in I}$ can receive a label $l_i$ which will select the
+corresponding session type $S_i$ as the continuing type of the channel. While
+a channel with type $\Choice{l_i}{S_i}{i\in I}$ can send a label $l_i$ and
+continues as $S_i$.
+
 \subsubsection{In the \texorpdfstring{$\pi$}{pi}-calculus}\label{sec:pis}
@@
 
-In the protocol defined above, a client selects between buying a book or
-receiving a price quote for a book. In the buy option case, the client sends
-the name of the book to the server and their card details, then the server
-sends the receipt for the purchase. In the quote option case, the client sends
-the name of the book and receives a price quote from the server. I have
-replaced the types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more
-meaningful type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate;
-all are synonyms for the type $\1$ which denotes the terminal session
-type. The client and server process bodies are as follows:
+Note, I am using $\tensor{}{}$ for session output and $\lfn{}{}$ for session
+input following the presentation by \citeauthor{Caires:2010:STI}. In the
+protocol defined above, a client selects between buying a book or receiving a
+price quote for a book. In the buy option case, the client sends the name of
+the book to the server and their card details, then the server sends the
+receipt for the purchase. In the quote option case, the client sends the name
+of the book and receives a price quote from the server. I have replaced the
+types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more meaningful
+type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate; all are
+synonyms for the type $\1$ which denotes the terminal session type. The client
+and server process bodies are as follows:
 \begin{gather*}
@@
 liveness mechanism, so GV typing judgements would need to be extended such
-that the capabilities can be included in soundness results. An extension to the
-translation to CP to incorporate the ``adopt'' and ``focus'' constructs would
-also need to be defined. It is not immediately clear how these features would
-assist in the management of session typed channels. The ``focus'' rule (and
-linearity generally) enforces an invariant on the focussed object (channel)
-such that the type is maintained during the focus operation:
+that the capabilities can be included in soundness results. An extension to
+the translation to CP to incorporate the ``adopt'' and ``focus'' constructs
+would also need to be defined. It is not immediately clear how these features
+would assist in the management of session typed channels. The ``focus'' rule
+(and linearity generally) enforces an invariant on the focussed object
+(channel) such that the type is maintained during the focus operation:
 \input{focus-rule}
@@
 
-The conclusion binds a variable in (C) but a type in (CCO). Thus, in the
+The conclusion specifies a variable in (C) but a type in (CCO). Thus, in the
 premise for (CCO) one can choose a suitably fresh $x$ in each premise. Note
@@
 
-I aim to have completed the GV and CP definitions including reduction and
-commuting conversion rules for CP by the end of December. This will require
-definition of infrastructure for handling binders for both GV, CP channels and
-CP propositional variables, and inductive definitions for well-typed terms,
-commuting conversions and CP principal cut reductions. So far I have completed
-some of the definitions for well-typed terms and CP process rules; the
-remaining rules require the infrastructure for substitution and opening of
-binders.
+The deliverables can be split into three main areas as indicated in the
+following table:
 
-After December, it remains to define the translation between GV and CP which I
-envisage will be the bulk of the work, requiring several rules on typing
-derivations; I plan to define these as inductive datatype constructors but I
-shall take care to reuse any suitable libraries that are available, e.g. a
-library for handling CPS translations. For this step of the work, I allocated
-one month but this is a cautious estimate accounting some slack time for any
-delays in preceding stages of the development; I suspect this stage will not
-require a full month.
+\begin{tabularx}{\linewidth}{|c|X|}\hline
+Deliverable \# &  Description \\ \hline
+\Deliverable{1}{Mechanisation of Wadler's system}
+\DSubD{1}{*}{Mechanisation of Wadler's system}
+\Deliverable{2}{Closure side-effecting extension}
+\DSubD{2}{1}{Implementation of the GV operational semantics}
+\DSubD{2}{2}{Implementation of the closure side-effecting extension}
+\Deliverable{3}{Report write-up}
+\DSubD{3}{1}{Description of approach, Wadler mechanisation and related work}
+\end{tabularx}
+
+The current proposal will form part of D3.1 (related work, etc.). The
+remaining parts of D3 will be written up as the project progresses, conducted
+in tandem with D1 and D2. The estimated work plan is given in the following
+table and sub-divides the above deliverable stages into tasks. The tasks
+marking the completion of a deliverable are noted. The tasks appear in
+chronological order.
+
+\begin{center}
+\begin{tabular}{|*{5}{c|}}\hline
+\multicolumn{5}{|c|}{\textbf{Estimated Work Plan}}\\ \hline
+Task & Start Date & Duration (days) & End Date & Deliverables \\ \hline
+\DTask{1}{1}{8th December}{15}{23rd December}{-}
+\DTask{3}{1}{20th December}{3}{23rd December}{-}
+\DTask{1}{2}{24th December}{17}{10th January}{D1}
+\DTask{3}{2}{7th January}{11}{18th January}{D3.1}
+\DTask{2}{1}{11th January}{7}{18th January}{-}
+\DTask{2}{2}{19th January}{18}{6th Feburary}{D2.1}
+%\DTask{3}{3}{
+\end{tabular}
+\end{center}
+
+\IDTsk{1}{1} involves completing the GV and CP definitions including reduction
+and commuting conversion rules for CP which I aim to complete during
+December. This will require definition of infrastructure for handling binders
+for both GV, CP channels and CP propositional variables, and inductive
+definitions for well-typed terms, commuting conversions and CP principal cut
+reductions. So far I have completed some of the definitions for well-typed
+terms and CP process rules; the remaining rules require the infrastructure for
+substitution and opening of binders. \IDTsk{3}{1} will consist of an initial
+structuring and write up of the mechanisation effort of Wadler's system. It
+will include a detailed description of the approach taken (which can, in part,
+be formed from the current proposal) and the related work (again, partly
+formed from the current document). Additionally, it will describe any issues
+which have arisen, including changes made to any of the definitions as
+presented by Wadler, including the changes already detailed here.
+
+For D1, it remains to define the translation between GV and CP which I
+envisage will require several rules on typing derivations; I plan to define
+these as inductive datatype constructors but I shall take care to reuse any
+suitable libraries that are available, e.g. a library for handling CPS
+translations. For this step of the work (\IDTsk{1}{2}), I made a cautious
+estimate on duration, accounting some slack time for any delays in preceding
+stages of the development. \IDTsk{3}{2} will document the translation phase
+and will also be used to document the initial approach for the extension (see
+table for overlapping task allocation). \IDTsk{3}{2} represents delivery of
+the first draft (D3.1) of the report describing the mechanisation of Wadler's
+system but omitting the theoretical treatment for the proposed extension.
 
@@
 closures) for session channels. In order to achieve this, operational
-semantics for GV would need to be defined, and the recent draft by
+semantics for GV would need to be defined and the recent draft by
 \citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
-do this~\cite{Lindley:2014:SPS}. I allocated the remaining time (approx. two
-months) to perform this extension because I believe it requires the most
-development: extending existing definitions to handle side-effecting
-functions, defining operational semantics, proving subject reduction and
-related properties. Additionally, it will also require a theoretical treatment
-prior to undertaking the development task, in contrast to the preceding
-development work which mostly utilises results from the literature.
+do this~\cite{Lindley:2014:SPS}. The task allocation time indicated in the
+work plan table. \IDTsk{2}{1} is a theoretical investigation into the
+operational semantics, deciding whether the system by
+\citeauthor{Lindley:2014:SPS} offers a suitable basis for the extension, and
+investigating the details of the side-effecting closure extension (assuming an
+appropriate base operational semantics for GV). \IDTsk{2}{2} is then the
+implementation of the suitable operational semantics (without side-effecting
+closure), culminating in the first part of the second deliverable
+(D2.1). Lastly, \IDTsk{2}{3} is an allocation of time to the task of
+implementing the extension permitting closures with side-effects. In my work
+plan, I have assumed \IDTsk{2}{2} implemented an operational semantics that
+eases this particular extension which justifies the task allocation time. Both
+\IDTsk{2}{2} and \IDTsk{2}{3} may also include revisions to the theoretical
+treatment started in \IDTsk{2}{1} as a consequence of initial design
+oversights.
 

### 2014-12-05 proposal/proposal.tex
Finished 3rd (and possibly final) draft.

@@
 
-\subsubsection{Preamble}
+\subsubsection{Preamble}\label{sec:pre}
 
@@
 providing logical connectives which have session-based interpretations in
-$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL:
+$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL, which have
+only two alternatives (cf. multiple alternatives in the standard definition in
+\S~\ref{sec:pre}):
 \input{sill-branch-choice-rules}
@@
 a session-based setting, since the intutition for \& is that of
-\emph{offering} a choice between alternatives rather than \emph{making} the
-choice. Here, however, the intuition should be extended to account for its
-location in the typing judgement. The channel is on the left of the sequent,
-and \citeauthor{Caires:2010:STI} suggest one consider the context as services
-being provided to the typed process. Thus, the type of channels in the context
-is really the type of the other endpoint of the channel, and the typed process
-behaves according to the dual protocol. For example, considering
-$\SILLbranchLTwo$, and taking $A^\bot$ to represent the dual of a protocol $A$
-as defined by \citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process
-is behaving according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting
-from alternatives (in this case, $\inrterm$) and then behaving according to
-the selected protocol (in this case, protocol $B^\bot$).
+\emph{offering} (receiving) a choice between alternatives rather than
+\emph{making} (sending) the choice. Here, however, the intuition should be
+extended to account for its location in the typing judgement. The channel is
+on the left of the sequent, and \citeauthor{Caires:2010:STI} suggest one
+consider the context as services being provided to the typed process. Thus,
+the type of channels in the context is really the type of the other endpoint
+of the channel, and the typed process behaves according to the dual
+protocol. For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to
+represent the dual of a protocol $A$ as defined by
+\citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process is behaving
+according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from
+alternatives (in this case, $\inrterm$) and then behaving according to the
+selected protocol (in this case, protocol $B^\bot$).
 
@@
 type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate; all are
-synonyms for the type $\1$ which denotes the terminal session type. The client
-and server process bodies are as follows:
+synonyms for the type $\1$ which denotes the terminal session type in this
+system. The client and server process bodies are as follows:
 \begin{gather*}
@@
 \citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
-\citeauthor{Gay:2010:LAST}. The work describes a continuation-passing style
-(CPS) translation from GV to CP treating session types as propositions in a
-classical linear logic. The process calculus is slightly different to that
-described by $\pi$DILL~\cite{Caires:2010:STI}, apart from being classical in
-nature, the constructs for server replication are modified; CP provides a
-weakening and contraction rules whereas these are encoded in one rule for
-process reduction by \citeauthor{Caires:2010:STI}. Additionally, rather than a
-two-sided sequent calculus, \citeauthor{Wadler:2014} uses one-sided sequents
-for defining the process calculus, which provides a more intuitive
-presentation of session type duality and clearer correspondence to a
-$\lambda$-calculi surface syntax. As presented, GV does not support all
-features of CP and thus the translation is one way (GV to
-CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to
-provide support for polymorphism and replication, following a similar CPS
-translation scheme. Lastly,
+\citeauthor{Gay:2010:LAST}. GV presents session types using the standard
+presentation as described in \S~\ref{sec:pre}. CP presents linear logical
+connectives with session type interpretations inspired by
+\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI}. \citeauthor{Wadler:2014}
+describes a continuation-passing style (CPS) translation from GV session types
+to CP classical linear logic propositions. The process calculus CP is slightly
+different to $\pi$DILL by \citeauthor{Caires:2010:STI}, apart from being
+classical in nature, the constructs for server replication are modified; CP
+provides a weakening and contraction rules whereas these are encoded in one
+rule for process reduction by \citeauthor{Caires:2010:STI}. Additionally,
+rather than a two-sided sequent calculus, \citeauthor{Wadler:2014} uses
+one-sided sequents for defining the process calculus, which provides a more
+intuitive presentation of session type duality and clearer correspondence to a
+$\lambda$-calculi surface syntax, simplifying the translation presentation. As
+presented, GV does not support all features of CP and thus the translation is
+one way (GV to CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP}
+extend GV to provide support for polymorphism and replication, following a
+similar CPS translation scheme. Lastly,
 \citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
@@
 rather than language primitives.
-
-%% Wadler also dispenses with the structural rules; subsumed by the Assoc and
-%% Swap typing derivations and the cut reduction (double check why he does
-%% this)
 
@@
 elimination via the induction principle.
-% ( since the standard induction tactic will not work?)
 
@@
 \Deliverable{1}{Mechanisation of Wadler's system}
-\DSubD{1}{*}{Mechanisation of Wadler's system}
+D1 & Mechanisation of Wadler's system \\ \hline
 \Deliverable{2}{Closure side-effecting extension}
 \DSubD{2}{1}{Implementation of the GV operational semantics}
-\DSubD{2}{2}{Implementation of the closure side-effecting extension}
+\DSubD{2}{2}{Implementation of the side-effecting closure extension}
 \Deliverable{3}{Report write-up}
 \DSubD{3}{1}{Description of approach, Wadler mechanisation and related work}
+\DSubD{3}{2}
+      {Description of GV operational semantics (w/o side-effecting closure)}
+\DSubD{3}{3}{Description of side-effecting closure and final version}
 \end{tabularx}
@@
 \DTask{3}{2}{7th January}{11}{18th January}{D3.1}
-\DTask{2}{1}{11th January}{7}{18th January}{-}
-\DTask{2}{2}{19th January}{18}{6th Feburary}{D2.1}
-%\DTask{3}{3}{
+\DTask{2}{1}{11th January}{14}{25th January}{-}
+\DTask{2}{2}{23th January}{21}{16th Feburary}{D2.1}
+\DTask{3}{3}{13th Feburary}{10}{23rd Feburary}{D3.2}
+\DTask{2}{3}{17th Feburary}{14}{3rd March}{D2.2}
+\DTask{3}{4}{1st March}{14}{15th March}{D3.3}
+\DTask{*}{*}{16th March}{39}{24th April}{-}
 \end{tabular}
@@
 \citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
-do this~\cite{Lindley:2014:SPS}. The task allocation time indicated in the
-work plan table. \IDTsk{2}{1} is a theoretical investigation into the
-operational semantics, deciding whether the system by
+do this~\cite{Lindley:2014:SPS}. \IDTsk{2}{1} is a theoretical investigation
+into the operational semantics, deciding whether the system by
 \citeauthor{Lindley:2014:SPS} offers a suitable basis for the extension, and
@@
 closure), culminating in the first part of the second deliverable
-(D2.1). Lastly, \IDTsk{2}{3} is an allocation of time to the task of
-implementing the extension permitting closures with side-effects. In my work
-plan, I have assumed \IDTsk{2}{2} implemented an operational semantics that
-eases this particular extension which justifies the task allocation time. Both
-\IDTsk{2}{2} and \IDTsk{2}{3} may also include revisions to the theoretical
-treatment started in \IDTsk{2}{1} as a consequence of initial design
-oversights.
+(D2.1). Following this implementation work, I plan to write up the operational
+semantics in the report (\IDTsk{3}{3}) which constitutes the second draft of
+the report deliverable (D3.2). Lastly, \IDTsk{2}{3} is an allocation of time
+to the task of implementing the extension permitting closures with
+side-effects which includes necessary changes to the typing judgements and
+translation between GV and CP. In my work plan, I have assumed \IDTsk{2}{2}
+implemented an operational semantics that eases this particular extension
+which justifies the task allocation time. Both \IDTsk{2}{2} and \IDTsk{2}{3}
+may also include revisions to the theoretical treatment started in
+\IDTsk{2}{1} as a consequence of initial design oversights. \IDTsk{3}{4}
+constitutes the final version of the report, including discussion of the
+side-effecting closure extension.
 
 There is a risk of being unable to perform the side-effecting extension as
-sketched in \S \ref{sec:approach}. I shall review and update the above plan
-throughout the project as I proceed to the various milestones I have set out,
-and if required will alter the above plans if necessary.
+sketched in \S \ref{sec:approach}. Additionally, there is considerable slack
+time identified in the work plan, indicated by \IDTsk{*}{*}. I believe these
+task allocation time estimates to be an accurate reflection of the work
+involved. However, the feasibility study is necessarily limited in scope, so I
+shall review and update the above plan throughout the project as I proceed to
+the various milestones I have set out, and if required will alter the above
+plans if necessary.
 

### 2014-12-11 paper/paper.tex
Back to regular append; comma too special. Added paper beginnings.

--- 2014-12-05 proposal/proposal.tex
@@
-\documentclass{mprop}
+\documentclass{mpaper}
 
-% Include Coq lstlisting language definition and things upon which it depends.
 \usepackage{macros}
@@
   maxbibnames=100]{biblatex}
-
-\usepackage{tabularx}
-\usepackage{tabulary}
+\addbibresource{paper.bib}
 
@@
 
-\newcommand{\Deliverable}[2]
-           {\multicolumn{2}{|c|}{\textbf{D{#1}: {#2}}} \\ \hline}
-\newcommand{\DSubD}[3]{D{#1}.{#2} & #3 \\ \hline}
-\newcommand{\DTask}[6]{D#1T#2 & #3 & #4 & #5 & #6 \\ \hline}
-\newcommand{\IDTsk}[2]{Task D{#1}T{#2}}
-
-% Thanks to Stefan Kottwitz on tex.stackexhange.com:
-% http://tex.stackexchange.com/questions/69662/how-to-globally-change-the-spacing-around-equations
-%% \expandafter\def\expandafter\normalsize\expandafter{%
-%%     \normalsize
-%%     \setlength\abovedisplayskip{0pt}
-%%     \setlength\belowdisplayskip{0pt}
-%%     \setlength\abovedisplayshortskip{0pt}
-%%     \setlength\belowdisplayshortskip{0pt}
-%% }
-
-\addbibresource{proposal.bib}
-
 \input{propmacros}
-
-\allowdisplaybreaks
-
-% alternative font if you prefer
-%\usepackage{times}
-
-% for alternative page numbering use the following package
-% and see documentation for commands
-%\usepackage{fancyheadings}
 
@@
 
-% other potentially useful packages
-%\uspackage{amssymb,amsmath}
-%\usepackage{url}
-%\usepackage{fancyvrb}
-%\usepackage[final]{pdfpages}
-
 \begin{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \author{Craig McLaughlin}
-\date{\today}
+\matricnum{1002524}
+
 \maketitle
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{abstract}
+According to Simon Peyton Jones, an abstract should address
+four key questions. First, what is the problem that this
+paper tackles? Second, why is this an interesting problem?
+Third, what is the solution this paper proposes?
+Finally, why is the proposed solution a good one?
+\end{abstract}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\tableofcontents
-\newpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}\label{intro}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Statement of Problem}
-
-The aim of this project is to develop a framework for formalising lambda
-calculus type systems involving session types using interactive theorem
-proving (in particular, the system Coq). The motivation for the framework is
-that it will enable a whole class of related languages and type systems to be
-formalised with little extra proof effort. Inspired by previous work on a
-mechanised framework for $\pi$-calculus type systems~\cite{Gay:2001:FFP}, I
-shift to the lambda calculus to provide a basis on which session types for
-mainstream programming languages can be studied, whilst taking advantage of
-previous proof effort and specialising only for those parts unique to the
-particular system. The framework will be heavily based on calculi developed by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
-classical linear logic (CP) and a functional language with session types
-(GV). GV is a minimal functional language supporting linearity so I believe
-such a basis will be general enough to support reasoning about a large class
-of functional languages with similar base properties.
-
-From this framework, a number of directions can be explored including session
-polymorphism session subtyping, and permitting different forms of aliasing
-within the type system. The recent work by \citeauthor{Lindley:2014:SAP}
-\cite{Lindley:2014:SAP}, who extend GV to HGV supporting session forwarding,
-replication and polymorphism, would be an interesting place to start when
-looking to add these extensions.
-
-Additionally, the majority of previous session type systems for mainstream
-programming languages have focussed exclusively on linear type systems. One
-disadvantage of a linear type system is the need to re-bind objects after each
-operation~\cite{Gay:2010:LAST}. One could extend the GV type system to provide
-the benefits of aliasing with the guarantees of linearity which has been
-studied in several systems using a variety of alias control
-methods~\cite{Fahndrich:2002,Mazurak:2010:LLT,Tov:2011}. Another approach,
-which has been utilised in previous session-based $\lambda$-calculus type
-systems, is to support side-effecting functions by allowing functions to close
-over channels in the environment~\cite{Gay:2003:STI,Vasconcelos:2006:TCM}. An
-interesting extension to the base GV type system would be to incorporate the
-closure semantics into the type system which would permit functions to modify
-session-typed channels in the environment. The typing rules would be similar
-to that described by \citeauthor{Gay:2010:MST} for typing expressions in an
-object-oriented setting for session types~\cite{Gay:2010:MST}, e.g. the rule
-for lambda abstraction could be:
-\begin{gather*}
-  \frac{
-    \typeds{\Gamma,\typedid{x}{A}}{t}{T}{\Gamma',\typedid{x}{B}}
-  }{
-    \typeds{\Gamma}{\abs{x}{t}}{\fn{\typestate{A}{B}}{T}}{\Gamma'}
-  }\quad
-\end{gather*}
-
-The $\Gamma$, $\Gamma '$ environments represent pre and post-states of
-variables satisfied by term $t$. The typing for the function argument is
-inspired by typestate-oriented programming~\cite{Aldrich:2009} and says that
-the function expects a channel behaving according to protocol $A$ and alters
-the state of the channel internally producing a channel behaving according to
-protocol $B$. It remains to be seen whether such typing rules can be encoded
-as described in the GV setting.
-
-These extensions will argue that the framework supports a wide range of
-languages and type systems, providing researchers the ability to harness the
-power of formal verification in the design, implementation and documentation
-of session-based type systems.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Background Survey}
-
-%% present an overview of relevant previous work including articles, books,
-%% and existing software products. Critically evaluate the strengths and
-%% weaknesses of the previous work.
+\section{Background}
 
@@
 The definition of session types will be important for the remainder:
-\input{session-types.tex}
+\input{../proposal/session-types.tex}
 
@@
 \S~\ref{sec:pre}):
-\input{sill-branch-choice-rules}
+\input{../proposal/sill-branch-choice-rules}
 
@@
 (channel) such that the type is maintained during the focus operation:
-\input{focus-rule}
+\input{../proposal/focus-rule}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Proposed Approach}\label{sec:approach}
+\section{Evaluation}
 
-%% state how you propose to solve the software development problem. Show that
-%% your proposed approach is feasible, but identify any risks.
+Graphs are always good. I recommend getting to grips with Matlab, R or
+gnuplot rather than exporting horribly Excel bitmapped graphs.
 
-I intend to mechanise the calculi described by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014} using the Coq proof assistant. In
-the spirit of the POPLMARK challenge, I aim to utilise available libraries and
-infrastructure as much as
-possible. \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a
-metatheory library for handling, among other things, association lists
-(environments) and free variable representation. I intend to use the version
-distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which has been
-updated for Coq version 8.4pl4. \citeauthor{Park:2014:MMW} describe a
-technique for removing nonlinear contexts from typing judgements. While their
-work extended to \fpop it is not clear how to handle an environment containing
-both non-linear and linear types as in GV. I wish to maintain as close a
-relationship as possible to the paper system presented by
-\citeauthor{Wadler:2014}, so separating out the non-linear and linear
-components (as in \fpop) is not an option at this stage.
+\begin{quotation}
+ The Assyrian came down like the wolf on the fold,
+ And his cohorts were gleaming in purple and gold;
+ And the sheen of their spears was like stars on the sea,
+ When the blue wave rolls nightly on deep Galilee.
 
-Following \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, I use the
-locally nameless first-order representation for binders and free variables,
-preventing variable capture issues present in other first-order
-representations. Likewise, I make some simplifying assumptions about variable
-substitution, namely that \textit{closed} terms (terms with no unbound
-indices) are substituted for variables, which avoids having to define lifting
-constructs on the substitute term. The typing rules will not contain a rule
-for an unbound index so this decision is sound. A consequence is that I must
-define a notion of opening language terms involving binders as described
-by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
-$t$ which is well-typed if $t$ is well-typed when its unbound index,
-representing the argument to the abstraction, is replaced by some fresh
-variable $x$, written $t^x$. Some terms in GV require multiple binders but I
-believe it will be straightforward to extend this notion. Unfortunately, I
-must define similar operations on CP terms and handle substitution and opening
-on them as well as propositions which can accept propositional variables. I
-hope to be able to generalise as much as possible with the help of the
-Metatheory library.
+ Like the leaves of the forest when Summer is green,
+ That host with their banners at sunset were seen:
+ Like the leaves of the forest when Autumn hath blown,
+ That host on the morrow lay withered and strown.
+\end{quotation}
 
-Alternatively, I could have chosen a higher-order abstract syntax (HOAS)
-approach such as the parametric version (PHOAS) described
-by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
-on the metalanguage (Coq, in this case) for handling binders and substitution
-so has less overhead in managing syntax than the locally nameless
-approach. However, variables are now parameters defined at the level of the
-metalanguage so can represent terms not valid in the language being
-mechanised. Further, a higher-order approach would prevent the use of the
-Metatheory library which provides a number of useful tactics for manipulating
-typing contexts.
+\section{Conclusions}
 
-For handling binder freshness, I chose to adopt the cofinite quantification
-approach described by~\cite{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, which
-excludes a finite set of variables from being considered as the binder. In
-contrast, the traditional ``exists-fresh'' approach, where the binder is only
-required to be fresh within the abstraction's body, does not produce strong
-enough induction principles in some cases. This change will necessitate
-changes to the typing rules of GV and CP. Consider for instance the Connect
-rule in GV~\cite{Wadler:2014} (C) and its corresponding cofinite
-quantification version (CCO):
-\input{with-connect-rule}
-
-The conclusion specifies a variable in (C) but a type in (CCO). Thus, in the
-premise for (CCO) one can choose a suitably fresh $x$ in each premise. Note
-that, $L$ can be chosen such that $\Psi \cup \Phi \cup \fv(M) \cup \fv(N)
-\subseteq L$, where $\fv(M)$ are the set of free variables in term $M$, and
-$x$ can be the same in both premises. Similar changes are made to the rules
-for CP so that, for example, output which allocates a fresh variable $y$ no
-longer names $y$ in the conclusion: \input{cp-output}
-
-Since the name $y$ is assumed to be fresh in the original presentation these
-changes do not alter the typing of processes nor the semantics of the
-translation from GV to CP. An equivalence between ``exists-fresh'' and
-cofinite quantification definitions of the simply-typed $\lambda$-calculus is
-given by \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. In general, I
-do not aim to formally prove adequacy of the mechanisation.
-
-While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
-may need to alter the approach slightly to avoid tricky formulations. For
-instance, instead of providing $n$-ary branch and choice in GV, I provide
-only binary versions of these operators to simplify the development. This
-change is not restrictive however, since CP itself provides only binary
-versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
-technical note, I amalgamate all types into one inductive definition and
-provide a predicate for restricting typing rules to consider only valid
-session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
-recursive definitions; session types are types which may contain types as
-subcomponents (arguments to send, for example). Unfortunately, handling
-mutually inductive definitions in Coq can be quite involved; requiring one to
-either rely on the Coq system to provide a stronger mutual induction principle
-or defining one manually. Such a definition is possible, but it complicates
-elimination via the induction principle.
-
-Another aspect of the encoding is how to define terms. The intrinsic encoding
-described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
-approach, indexing terms by their type so as to prevent ill-formed terms from
-being constructed. However, in the encoding for GV there are issues with using
-this encoding: $(1)$ it is not immediately clear how to handle linear contexts
-using the de Bruijn variable encoding, since in the intrinsic setting the
-environment supports weakening; $(2)$ an intrinsic encoding of terms would be
-complicated by the need to enforce that session types occur in certain
-instances (by use of a predicate); and $(3)$ well-typed terms require extra
-assumptions about binder freshness which cannot reasonably be expressed as a
-function type. For these reasons, an intrinsic approach does not offer much
-benefit since a well-typed term relation would still need to be
-defined. Therefore, I chose an extrinsic encoding for terms and define
-well-typed terms as a separate inductive type.
-
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
-their language proof infrastructure is proportional to the number of binding
-constructs. In the case of a simply typed lambda calculus this is not onerous,
-but GV has four binding constructs, CP has six and propositions have two. It
-may be true that this approach results in too many auxiliary definitions and
-operations not relevant to the actual properties one wishes to prove. However,
-I believe that once this infrastructure has been defined, the system can be
-extended without concern for these binding constructs.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Work Plan}
-
-The deliverables can be split into three main areas as indicated in the
-following table:
-
-\begin{tabularx}{\linewidth}{|c|X|}\hline
-Deliverable \# &  Description \\ \hline
-\Deliverable{1}{Mechanisation of Wadler's system}
-D1 & Mechanisation of Wadler's system \\ \hline
-\Deliverable{2}{Closure side-effecting extension}
-\DSubD{2}{1}{Implementation of the GV operational semantics}
-\DSubD{2}{2}{Implementation of the side-effecting closure extension}
-\Deliverable{3}{Report write-up}
-\DSubD{3}{1}{Description of approach, Wadler mechanisation and related work}
-\DSubD{3}{2}
-      {Description of GV operational semantics (w/o side-effecting closure)}
-\DSubD{3}{3}{Description of side-effecting closure and final version}
-\end{tabularx}
-
-The current proposal will form part of D3.1 (related work, etc.). The
-remaining parts of D3 will be written up as the project progresses, conducted
-in tandem with D1 and D2. The estimated work plan is given in the following
-table and sub-divides the above deliverable stages into tasks. The tasks
-marking the completion of a deliverable are noted. The tasks appear in
-chronological order.
-
-\begin{center}
-\begin{tabular}{|*{5}{c|}}\hline
-\multicolumn{5}{|c|}{\textbf{Estimated Work Plan}}\\ \hline
-Task & Start Date & Duration (days) & End Date & Deliverables \\ \hline
-\DTask{1}{1}{8th December}{15}{23rd December}{-}
-\DTask{3}{1}{20th December}{3}{23rd December}{-}
-\DTask{1}{2}{24th December}{17}{10th January}{D1}
-\DTask{3}{2}{7th January}{11}{18th January}{D3.1}
-\DTask{2}{1}{11th January}{14}{25th January}{-}
-\DTask{2}{2}{23th January}{21}{16th Feburary}{D2.1}
-\DTask{3}{3}{13th Feburary}{10}{23rd Feburary}{D3.2}
-\DTask{2}{3}{17th Feburary}{14}{3rd March}{D2.2}
-\DTask{3}{4}{1st March}{14}{15th March}{D3.3}
-\DTask{*}{*}{16th March}{39}{24th April}{-}
-\end{tabular}
-\end{center}
-
-\IDTsk{1}{1} involves completing the GV and CP definitions including reduction
-and commuting conversion rules for CP which I aim to complete during
-December. This will require definition of infrastructure for handling binders
-for both GV, CP channels and CP propositional variables, and inductive
-definitions for well-typed terms, commuting conversions and CP principal cut
-reductions. So far I have completed some of the definitions for well-typed
-terms and CP process rules; the remaining rules require the infrastructure for
-substitution and opening of binders. \IDTsk{3}{1} will consist of an initial
-structuring and write up of the mechanisation effort of Wadler's system. It
-will include a detailed description of the approach taken (which can, in part,
-be formed from the current proposal) and the related work (again, partly
-formed from the current document). Additionally, it will describe any issues
-which have arisen, including changes made to any of the definitions as
-presented by Wadler, including the changes already detailed here.
-
-For D1, it remains to define the translation between GV and CP which I
-envisage will require several rules on typing derivations; I plan to define
-these as inductive datatype constructors but I shall take care to reuse any
-suitable libraries that are available, e.g. a library for handling CPS
-translations. For this step of the work (\IDTsk{1}{2}), I made a cautious
-estimate on duration, accounting some slack time for any delays in preceding
-stages of the development. \IDTsk{3}{2} will document the translation phase
-and will also be used to document the initial approach for the extension (see
-table for overlapping task allocation). \IDTsk{3}{2} represents delivery of
-the first draft (D3.1) of the report describing the mechanisation of Wadler's
-system but omitting the theoretical treatment for the proposed extension.
-
-For the remainder, I shall look to extend the definitions of GV and CP using
-the definitions I have provided. As discussed previously, it would be
-interesting to study a system for supporting side-effecting functions (aka
-closures) for session channels. In order to achieve this, operational
-semantics for GV would need to be defined and the recent draft by
-\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
-do this~\cite{Lindley:2014:SPS}. \IDTsk{2}{1} is a theoretical investigation
-into the operational semantics, deciding whether the system by
-\citeauthor{Lindley:2014:SPS} offers a suitable basis for the extension, and
-investigating the details of the side-effecting closure extension (assuming an
-appropriate base operational semantics for GV). \IDTsk{2}{2} is then the
-implementation of the suitable operational semantics (without side-effecting
-closure), culminating in the first part of the second deliverable
-(D2.1). Following this implementation work, I plan to write up the operational
-semantics in the report (\IDTsk{3}{3}) which constitutes the second draft of
-the report deliverable (D3.2). Lastly, \IDTsk{2}{3} is an allocation of time
-to the task of implementing the extension permitting closures with
-side-effects which includes necessary changes to the typing judgements and
-translation between GV and CP. In my work plan, I have assumed \IDTsk{2}{2}
-implemented an operational semantics that eases this particular extension
-which justifies the task allocation time. Both \IDTsk{2}{2} and \IDTsk{2}{3}
-may also include revisions to the theoretical treatment started in
-\IDTsk{2}{1} as a consequence of initial design oversights. \IDTsk{3}{4}
-constitutes the final version of the report, including discussion of the
-side-effecting closure extension.
-
-There is a risk of being unable to perform the side-effecting extension as
-sketched in \S \ref{sec:approach}. Additionally, there is considerable slack
-time identified in the work plan, indicated by \IDTsk{*}{*}. I believe these
-task allocation time estimates to be an accurate reflection of the work
-involved. However, the feasibility study is necessarily limited in scope, so I
-shall review and update the above plan throughout the project as I proceed to
-the various milestones I have set out, and if required will alter the above
-plans if necessary.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+{\bf Acknowledgments.}
+This is optional; it is a location for you to thank people
 

### 2015-02-23 paper/paper.tex
Removed execution permission from *.bib/*.tex files in paper folder.

@@
 \usepackage{hyperref}
-
+\usepackage{comment}
 \usepackage{listings}
@@
 
-\title{Mechanisation of Session-based Lambda Calculus Type Systems}
+\title{Propositions as Sessions, Mechanically}
 \author{Craig McLaughlin}
@@
 \begin{abstract}
-According to Simon Peyton Jones, an abstract should address
-four key questions. First, what is the problem that this
-paper tackles? Second, why is this an interesting problem?
-Third, what is the solution this paper proposes?
-Finally, why is the proposed solution a good one?
+\begin{comment}
+SPL (+CPL)
+
+WTD
+
+ROF
+
+IMP
+\end{comment}
 \end{abstract}
 
-\section{Introduction}\label{intro}
+\begin{comment}
+\section{Issues}
 
-Two aspects of programming language research are relevant for this
-project. The first is concerned with static type systems for ensuring runtime
-safety properties for concurrent communication systems. The second is the
-increasing use of proof assistants and interactive theorem provers for
-studying and mechanically verifying the semantics and properties of
-programming languages.
+\begin{itemize}
+\item What is required of a Metatheory library for linear type systems?
+  Permutation reasoning? Weakening/Contraction for non-linear components?
+\item Discuss the different approaches that were taken to maintain close
+  correspondence with the paper presentation
+\item Make sure to note that the paper leaves details (a lot of the ``cruft'')
+  out of the proofs e.g. where weakening is applied in the typing derivations
+\item Mention the use of Metatheory and its limitations when mechanising
+  linear type systems like GV/CP
+\end{itemize}
+\end{comment}
 
-\subsection{On Communication}
+\section{Introduction}\label{sec:intro}
 
@@
 
-\subsection{On Verification}
-
 In programming language research, work on static type systems is usually
@@
 hold. More generally, unassisted formalisms are error-prone to develop and do
-not provide a firm basis upon which other researchers can experiment; often it
-will require ``rolling your own'' compiler for the language of interest.
+not provide a firm basis upon which other researchers can experiment.
 
@@
 extension of the system since the proof assistant will highlight parts of
-proofs that have been affected by changes. However, one may have to prove more
-properties using this approach since this setting does not permit appealing to
-human intuition. For example, alpha-equivalence and freshness of binders are
-usually assumed in informal presentations, whereas in the mechanised setting
-one has to explicitly encode and reason about these
+proofs that have been affected by changes. On the other hand, one may have to
+prove more properties using this approach since this setting does not permit
+appealing to human intuition. For example, alpha-equivalence and freshness of
+binders are usually assumed in informal presentations, whereas in the
+mechanised setting one has to explicitly encode and reason about these
 properties~\cite{Aydemir:2005:MMM, Aydemir:2008:EFM}.
 
-\section{Background}
+This project describes work towards the development of a framework for
+formalising lambda calculus type systems with session types using interactive
+theorem proving (in particular, the system Coq). The framework should enable a
+whole class of related languages and type systems to be formalised with little
+extra proof effort. Inspired by previous work on a mechanised framework for
+$\pi$-calculus type systems~\cite{Gay:2001:FFP}, the current work focuses on
+the lambda calculus to provide a basis on which session types for mainstream
+programming languages can be studied, whilst taking advantage of previous
+proof effort and specialising only for those parts unique to the particular
+system. The framework is heavily based on calculi developed by
+\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
+classical linear logic (CP) and a functional language with session types (GV).
+
+\section{Related Work}\label{sec:rw}
 
@@
 
-\section{Evaluation}
+\section{Formalisation}
 
-Graphs are always good. I recommend getting to grips with Matlab, R or
-gnuplot rather than exporting horribly Excel bitmapped graphs.
+I describe the formalisation as it currently stands in this section.
 
-\begin{quotation}
- The Assyrian came down like the wolf on the fold,
- And his cohorts were gleaming in purple and gold;
- And the sheen of their spears was like stars on the sea,
- When the blue wave rolls nightly on deep Galilee.
+\subsection{Future Work}
 
- Like the leaves of the forest when Summer is green,
- That host with their banners at sunset were seen:
- Like the leaves of the forest when Autumn hath blown,
- That host on the morrow lay withered and strown.
-\end{quotation}
+There remains much to be done.
 

### 2015-02-24 paper/paper.tex
Structuring the discussion of the development.

@@
 \end{abstract}
-
-\begin{comment}
-\section{Issues}
-
-\begin{itemize}
-\item What is required of a Metatheory library for linear type systems?
-  Permutation reasoning? Weakening/Contraction for non-linear components?
-\item Discuss the different approaches that were taken to maintain close
-  correspondence with the paper presentation
-\item Make sure to note that the paper leaves details (a lot of the ``cruft'')
-  out of the proofs e.g. where weakening is applied in the typing derivations
-\item Mention the use of Metatheory and its limitations when mechanising
-  linear type systems like GV/CP
-\end{itemize}
-\end{comment}
 
@@
 proof effort and specialising only for those parts unique to the particular
-system. The framework is heavily based on calculi developed by
+system. The starting point for the framework is the calculi developed by
 \citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
@@
 The definition of session types will be important for the remainder:
-\input{../proposal/session-types.tex}
+\input{../paper/session-types.tex}
 
@@
 
+\input{../paper/sill-branch-choice-rules}
+
 They represent dual intuitionistic linear logic as a sequent calculus
 providing logical connectives which have session-based interpretations in
-$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL, which have
-only two alternatives (cf. multiple alternatives in the standard definition in
-\S~\ref{sec:pre}):
-\input{../proposal/sill-branch-choice-rules}
+$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL in
+Figure~\ref{fig:sill-branching}, which have only two alternatives
+(cf. multiple alternatives in the standard definition in \S~\ref{sec:pre}).
 
@@
 $\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
-a session-based setting, since the intutition for \& is that of
+a session-based setting, since the intuition for \& is that of
 \emph{offering} (receiving) a choice between alternatives rather than
@@
 
-I describe the formalisation as it currently stands in this section.
+The formalisation was performed using the Coq proof
+assisstant~\cite{Coq:manual}. In the spirit of the POPLMARK
+challenge~\cite{Aydemir:2005:MMM}, the development utilises available
+libraries and infrastructure as much as possible. In particular,
+\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a metatheory
+library for handling, among other things, association lists (environments) and
+free variable representation. The current development uses the version of the
+library distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which
+has been updated for Coq version 8.4pl4. One goal of the formalisation effort
+was to minimise the extent to which the development differs from the
+pen-and-paper presentation and avoid questions of adequacy of the
+representation. Some changes were made to ease the mechanisation and are
+highlighted where they occur.
+
+The locally nameless first-order representation for binders and free variables
+was adopted for both GV and CP, following
+\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. This first-order
+variant prevents variable-capture issues when compared with other first-order
+approaches. Alternatively, I could have chosen a higher-order abstract syntax
+(HOAS) approach such as the parametric version (PHOAS) described
+by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
+on the metalanguage (Coq, in this case) for handling binders and substitution
+so has less overhead in managing syntax than the locally nameless
+approach. However, it was less clear how to handle linearity using this
+approach and has the disadvantage of being able to represent terms not valid
+in the language being mechanised (so called ``exotic'' terms). Further, a
+higher-order approach would prevent the use of the Metatheory library which
+provides a number of useful tactics for manipulating typing contexts. Previous
+efforts to mechanise \fpop~\cite{Park:2014:MMW} illustrate how to handle a
+language containing both linear and non-linear contexts. However, to stay true
+to CP/GV, I maintained a single context. Interestingly, it became clear that
+the mechanised version of CP required explicit weakening and contraction to
+facilitate the reduction rules.
+
+For handling binder freshness, I chose to adopt the cofinite quantification
+approach described by~\cite{Aydemir:2008:EFM}, which excludes a finite set of
+variables from being considered as the binder. In contrast, the traditional
+``exists-fresh'' approach, where the binder is only required to be fresh
+within the abstraction's body, does not produce strong enough induction
+principles in some cases. This change will necessitate changes to the typing
+rules of GV and CP. Consider for instance the Connect rule in
+GV~\cite{Wadler:2014} (C) and its corresponding cofinite quantification
+version (CCO): \input{with-connect-rule}
+
+The conclusion specifies a variable in (C) but a type in (CCO) (de Bruijn
+index rather than a name). Thus, in the premise for (CCO) one can choose a
+suitably fresh $x$ in each premise. Note that, $L$ can be chosen such that
+$\Psi \cup \Phi \cup \fv(M) \cup \fv(N) \subseteq L$, where $\fv(M)$ are the
+set of free variables in term $M$, and $x$ can be the same in both
+premises. Similar changes are made to the rules for CP (see
+\S~\ref{sec:cp}). These changes do not alter the typing of processes nor the
+semantics of the translation from GV to CP. An equivalence between
+``exists-fresh'' and cofinite quantification definitions of the simply-typed
+$\lambda$-calculus is given by
+\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
+result could be derived straightforwardly (but tediously) for the CP/GV
+system.
+
+\begin{comment}
+\section{Issues}
+
+\begin{itemize}
+\item What is required of a Metatheory library for linear type systems?
+  Permutation reasoning? Weakening/Contraction for non-linear components?
+\item Discuss the different approaches that were taken to maintain close
+  correspondence with the paper presentation
+\item Make sure to note that the paper leaves details (a lot of the ``cruft'')
+  out of the proofs e.g. where weakening is applied in the typing derivations
+\item Mention the use of Metatheory and its limitations when mechanising
+  linear type systems like GV/CP
+\end{itemize}
+\end{comment}
+
+\subsection{The process calculus CP}\label{sec:cp}
+
+\begin{comment}
+FIGURES
+
+propositions
+
+processes
+
+typing judgements
+
+equivalences and axiomatic cut
+
+principal cut rules
+
+commuting conversions
+\end{comment}
+
+Formalising the process calculus CP provided many challenges for the
+metatheory library since to my knowledge, the library has never been applied
+to a process calculus before. So for instance, in the lambda calculus one
+typically does not work up to permutation of binders, whereas in a process
+calculus these kinds of structural equivalences are standard.
+
+Note that the rules for polymorphism, $\exists$ and $\forall$, are defined but
+cut elimination on them is not. Since GV (\S~\ref{sec:gv}) does not support
+polymorphism the reduction rules are unnecessary for the translation. Future
+work is to support reduction under these terms.
+
+\subsection{The functional language GV}\label{sec:gv}
+
+\begin{comment}
+FIGURES
+
+types
+
+terms
+
+typing judgements
+\end{comment}
+
+\subsection{CPS translation GV to CP}
+
+Currently the formalisation does not handle this. Future work.
+
+for an unbound index so this decision is sound. A consequence is that I must
+define a notion of opening language terms involving binders as described
+by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
+$t$ which is well-typed if $t$ is well-typed when its unbound index,
+representing the argument to the abstraction, is replaced by some fresh
+variable $x$, written $t^x$. Some terms in GV require multiple binders but I
+believe it will be straightforward to extend this notion. Unfortunately, I
+must define similar operations on CP terms and handle substitution and opening
+on them as well as propositions which can accept propositional variables. I
+hope to be able to generalise as much as possible with the help of the
+Metatheory library. say something about gmeta here and its automatic
+generation of these infrastructure lemmas
+
+While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
+may need to alter the approach slightly to avoid tricky formulations. For
+instance, instead of providing $n$-ary branch and choice in GV, I provide
+only binary versions of these operators to simplify the development. This
+change is not restrictive however, since CP itself provides only binary
+versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
+technical note, I amalgamate all types into one inductive definition and
+provide a predicate for restricting typing rules to consider only valid
+session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
+recursive definitions; session types are types which may contain types as
+subcomponents (arguments to send, for example). Unfortunately, handling
+mutually inductive definitions in Coq can be quite involved; requiring one to
+either rely on the Coq system to provide a stronger mutual induction principle
+or defining one manually. Such a definition is possible, but it complicates
+elimination via the induction principle.
+
+Another aspect of the encoding is how to define terms. The intrinsic encoding
+described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
+approach, indexing terms by their type so as to prevent ill-formed terms from
+being constructed. However, in the encoding for GV there are issues with using
+this encoding: $(1)$ it is not immediately clear how to handle linear contexts
+using the de Bruijn variable encoding, since in the intrinsic setting the
+environment supports weakening; $(2)$ an intrinsic encoding of terms would be
+complicated by the need to enforce that session types occur in certain
+instances (by use of a predicate); and $(3)$ well-typed terms require extra
+assumptions about binder freshness which cannot reasonably be expressed as a
+function type. For these reasons, an intrinsic approach does not offer much
+benefit since a well-typed term relation would still need to be
+defined. Therefore, I chose an extrinsic encoding for terms and define
+well-typed terms as a separate inductive type.
+
+\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
+their language proof infrastructure is proportional to the number of binding
+constructs. In the case of a simply typed lambda calculus this is not onerous,
+but GV has four binding constructs, CP has six and propositions have two. It
+may be true that this approach results in too many auxiliary definitions and
+operations not relevant to the actual properties one wishes to prove. However,
+I believe that once this infrastructure has been defined, the system can be
+extended without concern for these binding constructs.
+
+\citeauthor{Park:2014:MMW} describe a technique for
+removing nonlinear contexts from typing judgements. While their work extended
+to \fpop it is not clear how to handle an environment containing both
+non-linear and linear types as in GV. I wish to maintain as close a
+relationship as possible to the paper system presented by
+\citeauthor{Wadler:2014}, so separating out the non-linear and linear
+components (as in \fpop) is not an option at this stage.
+
+I describe the formalisation as it currently stands in this section. The use
+of the metatheory library from UPenn provided useful tactics and properties
+regarding freshness of free variables and environments.
 
@@
 
-There remains much to be done.
+In addition to that already mentioned: there remains much to be done.
 

### 2015-02-24 paper/paper.tex
Tweaking content further. Hyperref and bibliography acting up.
Reference overlaps page boundary causing hyperref to explode. I
need to investigate this further.

@@
 \usepackage{macros}
-\usepackage{hyperref}
 \usepackage{comment}
@@
   match goal with
-  | [v: var |- _] => ... tactics here possibly mentioning `v' hypothesis ...
+  | [v: var |- _] =>
+    ... tactics here possibly mentioning `v' hypothesis ...
   end.
@@
 
+\subsection{Preliminaries}\label{sec:approach}
+
 The locally nameless first-order representation for binders and free variables
@@
 higher-order approach would prevent the use of the Metatheory library which
-provides a number of useful tactics for manipulating typing contexts. Previous
-efforts to mechanise \fpop~\cite{Park:2014:MMW} illustrate how to handle a
-language containing both linear and non-linear contexts. However, to stay true
-to CP/GV, I maintained a single context. Interestingly, it became clear that
-the mechanised version of CP required explicit weakening and contraction to
-facilitate the reduction rules.
+provides a number of useful tactics for manipulating typing contexts.
+
+Previous efforts to mechanise \fpop~\cite{Park:2014:MMW} illustrate how to
+handle a language containing both linear and non-linear contexts. However, to
+stay true to CP/GV, I maintained a single context. Interestingly, it became
+clear that the mechanised version of CP required explicit weakening and
+contraction to facilitate the reduction rules (\S~\ref{sec:cp}).
+
+\input{with-connect-rule}
 
 For handling binder freshness, I chose to adopt the cofinite quantification
-approach described by~\cite{Aydemir:2008:EFM}, which excludes a finite set of
-variables from being considered as the binder. In contrast, the traditional
-``exists-fresh'' approach, where the binder is only required to be fresh
-within the abstraction's body, does not produce strong enough induction
+approach described by~\citeauthor{Aydemir:2008:EFM}, which excludes a finite
+set of variables from being considered as the binder. In contrast, the
+traditional ``exists-fresh'' approach, where the binder is only required to be
+fresh within the abstraction's body, does not produce strong enough induction
 principles in some cases. This change will necessitate changes to the typing
-rules of GV and CP. Consider for instance the Connect rule in
-GV~\cite{Wadler:2014} (C) and its corresponding cofinite quantification
-version (CCO): \input{with-connect-rule}
+rules of GV and CP. Consider for instance the Connect rule in GV (C) and its
+corresponding cofinite quantification version (CCO) in
+Figure~\ref{fig:connect}. The conclusion specifies a variable in (C) but a
+type in (CCO) (de Bruijn index rather than a name). Thus, in the premise for
+(CCO) one can choose a suitably fresh $x$ in each premise. Note that, $L$ can
+be chosen such that $\Psi \cup \Phi \cup \fv(M) \cup \fv(N) \subseteq L$,
+where $\fv(M)$ are the set of free variables in term $M$, and $x$ can be the
+same in both premises. Similar changes are made to the rules for CP. These
+changes do not alter the typing of processes nor the semantics of the
+translation from GV to CP. An equivalence between ``exists-fresh'' and
+cofinite quantification definitions of the simply-typed $\lambda$-calculus is
+given by \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect
+a similar result could be derived straightforwardly (but tediously) for the
+CP/GV system.
 
-The conclusion specifies a variable in (C) but a type in (CCO) (de Bruijn
-index rather than a name). Thus, in the premise for (CCO) one can choose a
-suitably fresh $x$ in each premise. Note that, $L$ can be chosen such that
-$\Psi \cup \Phi \cup \fv(M) \cup \fv(N) \subseteq L$, where $\fv(M)$ are the
-set of free variables in term $M$, and $x$ can be the same in both
-premises. Similar changes are made to the rules for CP (see
-\S~\ref{sec:cp}). These changes do not alter the typing of processes nor the
-semantics of the translation from GV to CP. An equivalence between
-``exists-fresh'' and cofinite quantification definitions of the simply-typed
-$\lambda$-calculus is given by
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
-result could be derived straightforwardly (but tediously) for the CP/GV
-system.
+Another aspect of the encoding is how to define terms. The intrinsic encoding
+described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
+approach, indexing terms by their type so as to prevent ill-formed terms from
+being constructed. However, in the encoding for GV there are issues with using
+this encoding: $(1)$ it is not immediately clear how to handle linear contexts
+using the de Bruijn variable encoding, since in the intrinsic setting the
+environment supports weakening; $(2)$ an intrinsic encoding of terms would be
+complicated by the need to enforce that session types occur in certain
+instances (by use of a predicate); and $(3)$ well-typed terms require extra
+assumptions about binder freshness which cannot reasonably be expressed as a
+function type. For these reasons, an intrinsic approach does not offer much
+benefit since a well-typed term relation would still need to be
+defined. Therefore, I chose an extrinsic encoding for terms and define
+well-typed terms as a separate inductive type.
+
+\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
+their language proof infrastructure is proportional to the number of binding
+constructs. In the case of a simply typed lambda calculus this is not onerous,
+but GV has four binding constructs, CP has six and propositions have two. It
+may be true that this approach results in too many auxiliary definitions and
+operations not relevant to the actual properties one wishes to prove. However,
+I believe that once this infrastructure has been defined, the system can be
+extended without concern for these binding constructs.
+
+\citeauthor{Park:2014:MMW} describe a technique for
+removing nonlinear contexts from typing judgements. While their work extended
+to \fpop it is not clear how to handle an environment containing both
+non-linear and linear types as in GV. I wish to maintain as close a
+relationship as possible to the paper system presented by
+\citeauthor{Wadler:2014}, so separating out the non-linear and linear
+components (as in \fpop) is not an option at this stage.
 
@@
 typically does not work up to permutation of binders, whereas in a process
-calculus these kinds of structural equivalences are standard.
+calculus these kinds of structural equivalences are standard. Additionally,
+substitution and opening are only defined on processes for names. In other
+words, one cannot substitute an arbitrary process term for a name, unlike in
+the lambda calculus. Thus frameworks which assume a lambda calculus style
+substitution operation are not immediately amenable to the CP
+setting~\cite{Lee:2012}.
 
@@
 work is to support reduction under these terms.
+
+Explicit weakening and contraction were required to differentiate the
+reduction rules for servers and clients. In contrast, \citeauthor{Wadler:2014}
+defines these operations explicitly.
 
@@
 \end{comment}
+
+While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
+may need to alter the approach slightly to avoid tricky formulations. For
+instance, instead of providing $n$-ary branch and choice in GV, I provide
+only binary versions of these operators to simplify the development. This
+change is not restrictive however, since CP itself provides only binary
+versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
+technical note, I amalgamate all types into one inductive definition and
+provide a predicate for restricting typing rules to consider only valid
+session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
+recursive definitions; session types are types which may contain types as
+subcomponents (arguments to send, for example). Unfortunately, handling
+mutually inductive definitions in Coq can be quite involved; requiring one to
+either rely on the Coq system to provide a stronger mutual induction principle
+or defining one manually. Such a definition is possible, but it complicates
+elimination via the induction principle.
 
@@
 
-While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
-may need to alter the approach slightly to avoid tricky formulations. For
-instance, instead of providing $n$-ary branch and choice in GV, I provide
-only binary versions of these operators to simplify the development. This
-change is not restrictive however, since CP itself provides only binary
-versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
-technical note, I amalgamate all types into one inductive definition and
-provide a predicate for restricting typing rules to consider only valid
-session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
-recursive definitions; session types are types which may contain types as
-subcomponents (arguments to send, for example). Unfortunately, handling
-mutually inductive definitions in Coq can be quite involved; requiring one to
-either rely on the Coq system to provide a stronger mutual induction principle
-or defining one manually. Such a definition is possible, but it complicates
-elimination via the induction principle.
-
-Another aspect of the encoding is how to define terms. The intrinsic encoding
-described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
-approach, indexing terms by their type so as to prevent ill-formed terms from
-being constructed. However, in the encoding for GV there are issues with using
-this encoding: $(1)$ it is not immediately clear how to handle linear contexts
-using the de Bruijn variable encoding, since in the intrinsic setting the
-environment supports weakening; $(2)$ an intrinsic encoding of terms would be
-complicated by the need to enforce that session types occur in certain
-instances (by use of a predicate); and $(3)$ well-typed terms require extra
-assumptions about binder freshness which cannot reasonably be expressed as a
-function type. For these reasons, an intrinsic approach does not offer much
-benefit since a well-typed term relation would still need to be
-defined. Therefore, I chose an extrinsic encoding for terms and define
-well-typed terms as a separate inductive type.
-
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
-their language proof infrastructure is proportional to the number of binding
-constructs. In the case of a simply typed lambda calculus this is not onerous,
-but GV has four binding constructs, CP has six and propositions have two. It
-may be true that this approach results in too many auxiliary definitions and
-operations not relevant to the actual properties one wishes to prove. However,
-I believe that once this infrastructure has been defined, the system can be
-extended without concern for these binding constructs.
-
-\citeauthor{Park:2014:MMW} describe a technique for
-removing nonlinear contexts from typing judgements. While their work extended
-to \fpop it is not clear how to handle an environment containing both
-non-linear and linear types as in GV. I wish to maintain as close a
-relationship as possible to the paper system presented by
-\citeauthor{Wadler:2014}, so separating out the non-linear and linear
-components (as in \fpop) is not an option at this stage.
-
-I describe the formalisation as it currently stands in this section. The use
-of the metatheory library from UPenn provided useful tactics and properties
-regarding freshness of free variables and environments.
-
-\subsection{Future Work}
+\section{Future Work}
 
@@
 
-\printbibliography
+%\printbibliography
 

### 2015-02-25 paper/paper.tex
Fixed issue with bib/hyperref (I think). More thoughts on CP.

@@
 approach, indexing terms by their type so as to prevent ill-formed terms from
-being constructed. However, in the encoding for GV there are issues with using
-this encoding: $(1)$ it is not immediately clear how to handle linear contexts
-using the de Bruijn variable encoding, since in the intrinsic setting the
-environment supports weakening; $(2)$ an intrinsic encoding of terms would be
-complicated by the need to enforce that session types occur in certain
-instances (by use of a predicate); and $(3)$ well-typed terms require extra
-assumptions about binder freshness which cannot reasonably be expressed as a
-function type. For these reasons, an intrinsic approach does not offer much
-benefit since a well-typed term relation would still need to be
+being constructed. However, in the encoding for CP/GV there are issues with
+using this encoding: $(1)$ it is not immediately clear how to handle linear
+contexts using the de Bruijn variable encoding, since in the intrinsic setting
+the environment automatically supports weakening; $(2)$ an intrinsic encoding
+of GV terms would be complicated by the need to enforce that session types
+occur in certain instances (by use of a predicate); and $(3)$ well-typed terms
+require extra assumptions about binder freshness which cannot reasonably be
+expressed as a function type. For these reasons, an intrinsic approach does
+not offer much benefit since a well-typed term relation would still need to be
 defined. Therefore, I chose an extrinsic encoding for terms and define
@@
 constructs. In the case of a simply typed lambda calculus this is not onerous,
-but GV has four binding constructs, CP has six and propositions have two. It
-may be true that this approach results in too many auxiliary definitions and
-operations not relevant to the actual properties one wishes to prove. However,
-I believe that once this infrastructure has been defined, the system can be
-extended without concern for these binding constructs.
+but GV has four binding constructs, CP has six and propositions have
+two. Indeed, a lot of effort was expended setting up this
+infrastructure. However, I believe one of the contributions of this work is to
+articulate what one needs from state of the art metatheory libraries in order
+to handle session-based languages, and more generally, linear type systems.
 
@@
 words, one cannot substitute an arbitrary process term for a name, unlike in
-the lambda calculus. Thus frameworks which assume a lambda calculus style
+the lambda calculus. Thus frameworks which assume a lambda calculi style
 substitution operation are not immediately amenable to the CP
@@
 
-Explicit weakening and contraction were required to differentiate the
-reduction rules for servers and clients. In contrast, \citeauthor{Wadler:2014}
-defines these operations explicitly.
+Initially, an implicit notion of weakening and contraction were chosen so as
+to be in line with \citeauthor{Wadler:2014}'s definitions. However, explicit
+weakening and contraction were required to differentiate the reduction rules
+for the various interactions between servers and clients.
+
+Ordering in environments is ignored and the development required lemmas
+allowing the permutation of environments. Support for permutating environments
+is not currently supported in the Metatheory library but some general lemmas
+were added as a result of this work. Additionally, custom tactics were
+developed to automatatically solve some simple permutation goals. However,
+these tactics are not a ``catch all'' and some generalising still needs to be
+done.
+
+The paper presentation omits details: the principal cut rules do not specify
+weakening could occur: since weakening is implicit is tacitly assumed
+weakening can always be pushed further up the derivation tree e.g. AxCut rule
+
+another interesting aspect which the formalism reveal is the rule for
+associativity is constructed in a way to permit associativity in a certain way
+pre-determined
+
+metatheory library and the library of tactics from software foundations are
+used extensively: metatheory does not handle multiple binders (check this?)
+maybe add a generic tactic to handle multiple binders... or perhaps add a
+lemmas for handling binders
 
@@
 
-for an unbound index so this decision is sound. A consequence is that I must
-define a notion of opening language terms involving binders as described
-by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
-$t$ which is well-typed if $t$ is well-typed when its unbound index,
-representing the argument to the abstraction, is replaced by some fresh
-variable $x$, written $t^x$. Some terms in GV require multiple binders but I
-believe it will be straightforward to extend this notion. Unfortunately, I
-must define similar operations on CP terms and handle substitution and opening
-on them as well as propositions which can accept propositional variables. I
-hope to be able to generalise as much as possible with the help of the
-Metatheory library. say something about gmeta here and its automatic
-generation of these infrastructure lemmas
-
 \section{Future Work}
@@
 
-%\printbibliography
+\sloppy
+\printbibliography
 

### 2015-04-02 paper/abstract.tex
Updated paper. Separated out sections and added CP figure.

--- 2015-02-24 paper/paper.tex
@@
-\documentclass{mpaper}
-
-\usepackage{macros}
-\usepackage{hyperref}
-\usepackage{comment}
-\usepackage{listings}
-\usepackage{lstcoq}
-\usepackage{graphicx}
-\usepackage{xspace}
-\usepackage{cmll}
-\usepackage{prooftree}
-\usepackage{amsmath}
-\newcommand{\typestate}[2]{{#1}\gg{#2}}
-\usepackage[style=alphabetic,sorting=nymt,natbib=true,
-  maxbibnames=100]{biblatex}
-\addbibresource{paper.bib}
-
-\DeclareSortingScheme{nymt}{
-  \sort{
-    \field{author}
-    \field{year}
-    \field{month}
-    \field{title}
-  }
-}
-
-\input{propmacros}
-
-\newcommand{\fpop}{System F${}^\circ$\xspace}
-
-\begin{document}
-
-\title{Propositions as Sessions, Mechanically}
-\author{Craig McLaughlin}
-\matricnum{1002524}
-
-\maketitle
-
 \begin{abstract}
@@
 \end{abstract}
-
-\section{Introduction}\label{sec:intro}
-
-Distributed communication systems are becoming increasingly prevalent in
-modern society. To facilitate communication between concurrently executing
-agents requires complex network protocols to define the type and sequencing of
-messages. Typically these protocols are implemented in systems programming
-languages, such as C, which provide no correctness guarantees of the
-implementation. Furthermore, programming in these languages is often extremely
-error-prone, as one has a myriad of unsafe features at one's disposal. As more
-businesses begin to conduct transactions over the Web, it is imperative that
-they can reason about the communication protocols upon which they rely.
-
-Encoding communication protocols as \textit{session types} is one approach to
-achieve guarantees about correctness. One particular variant of session types
-are \textit{binary} session types~\footnote{There are also \textit{multiparty}
- session types, but I do not discuss them here.}. A binary session type
-specifies communication between two agents, and it specifies the order and
-type of messages to be sent or received by an agent in the system. The two
-agents have dual session types such that, for example, when one is expecting
-to receive a message the other is expecting to send a message; both session
-types agree on the type of message to be exchanged. In this way, correctness
-of communication can be statically assured by providing communicating agents
-with a \textit{channel} whose endpoints have dual session types. A channel is
-defined as a bi-directional conduit for exchanging messages, not unlike a
-socket in network programming. In contrast to sockets, however, a channel is
-associated with an explicit type which provides a description of its
-protocol. Thus, in session-based systems a channel is usually provided as a
-primitive type.
-
-In programming language research, work on static type systems is usually
-presented on paper complete with informal proofs of correctness for the system
-properties. Most types systems in the session types community have been
-developed in this way. However, it is more difficult to extend work using this
-approach, as it may be unclear how changes affect proofs and one will
-typically have to re-consider all proofs to be convinced the properties still
-hold. More generally, unassisted formalisms are error-prone to develop and do
-not provide a firm basis upon which other researchers can experiment.
-
-Proof assistants such as Coq~\cite{Coq:manual} offer a platform on which to
-develop programming languages and type systems and associated proofs of
-properties with strong guarantees of correctness. Not only is proving with a
-proof assistant much less error-prone than the unaided approach, it also eases
-extension of the system since the proof assistant will highlight parts of
-proofs that have been affected by changes. On the other hand, one may have to
-prove more properties using this approach since this setting does not permit
-appealing to human intuition. For example, alpha-equivalence and freshness of
-binders are usually assumed in informal presentations, whereas in the
-mechanised setting one has to explicitly encode and reason about these
-properties~\cite{Aydemir:2005:MMM, Aydemir:2008:EFM}.
-
-This project describes work towards the development of a framework for
-formalising lambda calculus type systems with session types using interactive
-theorem proving (in particular, the system Coq). The framework should enable a
-whole class of related languages and type systems to be formalised with little
-extra proof effort. Inspired by previous work on a mechanised framework for
-$\pi$-calculus type systems~\cite{Gay:2001:FFP}, the current work focuses on
-the lambda calculus to provide a basis on which session types for mainstream
-programming languages can be studied, whilst taking advantage of previous
-proof effort and specialising only for those parts unique to the particular
-system. The starting point for the framework is the calculi developed by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
-classical linear logic (CP) and a functional language with session types (GV).
-
-\section{Related Work}\label{sec:rw}
-
-The background for this project can be categorised roughly as previous work on
-session-based type systems (sub-divided into $\pi$ and $\lambda$-calculi
-variants, \S \ref{sec:pis} and \S \ref{sec:lam}, respectively) and previous
-work on formal verification using proof assistants (particularly focussed on
-programming language mechanisation). Finally, some recent work has merged
-these two disciplines by verifying a session-based $\pi$-calculus system (\S
-\ref{sec:asts}).
-
-\subsection{Session-based Type Systems and Linear Logic}\label{sec:sts}
-
-\subsubsection{Preamble}\label{sec:pre}
-
-The definition of session types will be important for the remainder:
-\input{../paper/session-types.tex}
-
-Here, I use a modified version of the definition given by
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} where I have omitted recursive
-session types. The type \lstinline{end} denotes the terminal session type and
-a channel typed as such may not perform any communication. A communication
-channel with session type $\inpt{T}{S}$ expects to receive a term of type $T$
-and then continues with session type $S$. Conversely, a channel with type
-$\outpt{T}{S}$ can send a term of type $T$ and then has type $S$. Thus, input
-and output are considered \emph{dual} session types, in the sense that a
-channel with one endpoint having type $\inpt{T}{S}$ and the other endpoint
-having type $\outpt{T}{\Sbar{S}}$, where $\Sbar{S}$ is the dual of $S$,
-permits communication between two agents, allowing one to send a term of type
-$T$ and the other to receive it with continuing endpoint types $\Sbar{S}$ and
-$S$, respectively. Similarly, the remaining types $\Branch{l_i}{S_i}{i\in I}$
-and $\Choice{l_i}{S_i}{i\in I}$ are duals of each other. A channel of type
-$\Branch{l_i}{S_i}{i\in I}$ can receive a label $l_i$ which will select the
-corresponding session type $S_i$ as the continuing type of the channel. While
-a channel with type $\Choice{l_i}{S_i}{i\in I}$ can send a label $l_i$ and
-continues as $S_i$.
-
-\subsubsection{In the \texorpdfstring{$\pi$}{pi}-calculus}\label{sec:pis}
-
-\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} present a connection
-between a $\pi$-calculus type system with session types ($\pi$DILL) and dual
-intuitionistic linear logic (DILL)~\cite{Barber:1996}. They equate linear
-propositions with session types of the process calculus via a bidirectional
-translation relation. They use a binary reduction relation and a labelled
-transition system to represent process reduction and process action with its
-environment as is standard~\cite{Sangiorgi:2001}.
-
-\input{../paper/sill-branch-choice-rules}
-
-They represent dual intuitionistic linear logic as a sequent calculus
-providing logical connectives which have session-based interpretations in
-$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL in
-Figure~\ref{fig:sill-branching}, which have only two alternatives
-(cf. multiple alternatives in the standard definition in \S~\ref{sec:pre}).
-
-Note that branch and choice both have some notion of branching (R rules for
-$\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
-a session-based setting, since the intuition for \& is that of
-\emph{offering} (receiving) a choice between alternatives rather than
-\emph{making} (sending) the choice. Here, however, the intuition should be
-extended to account for its location in the typing judgement. The channel is
-on the left of the sequent, and \citeauthor{Caires:2010:STI} suggest one
-consider the context as services being provided to the typed process. Thus,
-the type of channels in the context is really the type of the other endpoint
-of the channel, and the typed process behaves according to the dual
-protocol. For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to
-represent the dual of a protocol $A$ as defined by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process is behaving
-according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from
-alternatives (in this case, $\inrterm$) and then behaving according to the
-selected protocol (in this case, protocol $B^\bot$).
-
-As noted by Wadler~\cite{Wadler:2014} there is a certain amount of redundancy
-in the two-sided nature of the rules forcing one to commit to a certain
-convention when typing a communication system. Consider, an equivalent system
-to the buy/quote example given by
-\citeauthor{Caires:2010:STI}~\cite[\S~3]{Caires:2010:STI}:
-\begin{gather*}
-\sdef{ClientProto}{\BChoice{(\tensor{Book}
-                            {(\tensor{Card}{(\lfn{Receipt}{\1})})})}
-                           {(\tensor{Book}{(\lfn{Quote}{\1})})}}
-\end{gather*}
-
-Note, I am using $\tensor{}{}$ for session output and $\lfn{}{}$ for session
-input following the presentation by \citeauthor{Caires:2010:STI}. In the
-protocol defined above, a client selects between buying a book or receiving a
-price quote for a book. In the buy option case, the client sends the name of
-the book to the server and their card details, then the server sends the
-receipt for the purchase. In the quote option case, the client sends the name
-of the book and receives a price quote from the server. I have replaced the
-types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more meaningful
-type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate; all are
-synonyms for the type $\1$ which denotes the terminal session type in this
-system. The client and server process bodies are as follows:
-\begin{gather*}
-\sdef{ClientBody_c}
-     {\CPinr{c}{\SLoutput{1984}{c}{\SLinput{c}{quote}{\0}}}}\\
-\sdef{ServerBody_c}
-     {\CPcase{c}
-             {\SLinput{c}{book}{\SLinput{c}{card}{\SLoutput{receipt}{c}{\0}}}}
-             {\SLinput{c}{book}{\SLoutput{quote}{c}{\0}}}}
-\end{gather*}
-
-\begin{samepage}
-The typings for the client and server processes are given by:
-\begin{gather*}
-\tsequent{\cdot}{\cdot}{ClientBody_c}{\typedid{c}{ClientProto}}
-\tag{\mkTrule{ClientBody}} \label{eq:cbtyp}
-\\ \tsequent{\cdot}{\typedid{c}{ClientProto}}{ServerBody_c}{\typedid{-}{\1}}
-\tag{\mkTrule{ServerBody}} \label{eq:sbtyp}
-\end{gather*}
-\end{samepage}
-
-The derivation for \eqref{eq:cbtyp} is $T\1 R$, $\lfn{T}{R}$, $\tensor{T}{R}$,
-$\BChoice{T}{R_2}$ and \eqref{eq:sbtyp} is derivable from
-$\BChoice{T}{L}$. Except for the final rule, these are exact dual rules of the
-rules used in example by \citeauthor{Caires:2010:STI}. Indeed, consider the
-composite process:
-\begin{gather*}
-\sdef{QSimple}{\Spar{c}{ClientBody_c}{ServerBody_c}}\\
-\tsequent{\cdot}{\cdot}{QSimple}{\typedid{-}{\1}}
-\end{gather*}
-
-The final system is near identical to the one produced in the example given by
-\citeauthor{Caires:2010:STI} except the individual processes within the
-parallel composition have been swapped. Here the client communicates along the
-channel $c$ on the right-side of the sequent and the server utilises the
-channel $c$ on the left of the sequent. This typing, while correct, violates
-the intuition given by \citeauthor{Caires:2010:STI} of the context providing
-the services used by the process, since one would usually associate provision
-of services with the server not the client. Instead, the example above
-suggests the server utilises some channel provided by the client. On the other
-hand, one could imagine such a channel being setup between two communicating
-agents; the ``owner'' of the channel is largely irrelevant. Note that
-$ClientProto$ is in fact the dual of the $ServerProto$ definition by
-\citeauthor{Caires:2010:STI} using the definition of dual used in Wadler's
-classical process calculus~\cite{Wadler:2014}. In $\pi$DILL, the duality
-between endpoints of the same channel are obscured by the nature of the typing
-judgements.
-
-Process reduction steps are equated with proof reduction via cut
-elimination. The relationship with proof normalisation in DILL provides
-absence of deadlocks in session-based communication for the process
-calculus. The correspondence of session types to linear logic propositions is
-an important step in providing a solid logical foundation for a concurrent
-type theory; in the same way ``sequential'' type theory ($\lambda$-calculi)
-profitted from a logical perspective. However, for session types to become a
-more usable tool for programming distributed communication systems (the
-ultimate goal, of course, is to write more correct programs), one cannot be
-expected to program directly in the $\pi$-calculus; a more mainstream
-programming paradigm must be built upon these solid logical foundations. With
-that in mind, enter the $\lambda$-calculus!
-
-\subsubsection{In the \texorpdfstring{$\lambda$}{lambda}-calculus}
-\label{sec:lam}
-
-A series of works by \citeauthor{Gay:2003:STI}~\cite{Gay:2003:STI} and
-\citeauthor{Vasconcelos:2006:TCM}~\cite{Vasconcelos:2006:TCM} provide a
-$\lambda$-calculus based language with session types for channel
-primitives. The type system is nonstandard, featuring an additional
-environment for tracking the session types of channels; these types are
-modified during term reduction to reflect communication. The extra channel
-environment is provided to support possible aliasing of channels by treating
-channel variables as pointers or references into the channel environment. So,
-for a channel $c$ of session type $S$, a variable $x$ referring to this
-channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the
-channel environment. The typing judgements and term reduction relation must be
-augmented to handle these environments. Unfortunately, a side-effect of this
-is that function types must specify the pre and post-states of channel
-environments and therefore, the type of arguments cannot work up to
-alpha-equivalence of channel identifiers. That is, if two channels $c$ and $d$
-have the same session type, one cannot define a function to operate on
-variables of either channel type since the function definition will explicitly
-mention the identifier for the channel (in this case either, $Chan~c$ or
-$Chan~d$). However, the system does provide an attractive non-linear channel
-type which has been neglected in favour of a linear treatment of channels in
-more recent
-works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2014}. Certainly, it is the
-case that an aliased approach would fit better with mainstream programming
-languages and perhaps studying a similar system could offer some insights.
-
-The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of
-aliased access to linear resources. The type system requires special language
-features, e.g. to manage capabilities which can be thought as a form of object
-liveness mechanism, so GV typing judgements would need to be extended such
-that the capabilities can be included in soundness results. An extension to
-the translation to CP to incorporate the ``adopt'' and ``focus'' constructs
-would also need to be defined. It is not immediately clear how these features
-would assist in the management of session typed channels. The ``focus'' rule
-(and linearity generally) enforces an invariant on the focussed object
-(channel) such that the type is maintained during the focus operation:
-\input{../proposal/focus-rule}
-
-The typing judgements include \emph{capabilities} that track aliasing to heap
-objects, e.g. $\AFcap{\rho}{h}$ denotes a mapping from a static name $\rho$ to
-a heap type $h$, and a variable $x$ may be typed as $\typedid{x}{\AFtr{\rho}}$
-indicating that $x$ refers to an object of type $h$. The purpose of the
-capabilities is to enforce restrictions on when an object can be
-referenced. In other words, if $\AFcap{\rho}{h}$ is in the capabilities at
-some program point then $x$ may be used, otherwise such use is prohibited. In
-the [focus] rule, evaluating $e_1$ retrieves the object to be focussed. Note
-the type of $e_1$ is a ``guarded'' type. The guard $G$ is the static name of
-some object which ``adopted''~\footnote{I omit the adopt typing rule for
-  brevity.} an object of heap type $h$. If a program point has
-$\AFcap{G}{h_1}$ in the capabilities for some $h_1$ then all objects guarded
-by $G$ may be used; $\capsequent{\Delta}{C_1}{G}$ denotes containment of a
-such a capability in $C_1$. The guarded type may be thought of as policing
-manipulation of linear components within $h$. Treating $\otimes$ as the
-disjoint union for capability sets, the typing for $e_2$ temporarily removes
-the guard $G$ from the set of capabilities, and adds a new variable $x$, and
-associated capability, with a type permitting access to the object of type $h$
-i.e. a non-guarded type. Removing $G$ will prohibit access to allow objects
-guarded by $G$ for the during of $e_2$. The final capabilities after executing
-$e_2$ are required to be $\tensor{C_3}{\AFcap{\rho}{h}}$, enforcing that the
-type must remain as $h$. The typing of the $\letterm$ expression ensures that
-the guard $G$ (contained in $C_1$) is restored.
-
-Such a rule would appear to only benefit recursive session types. Indeed,
-focussing on a non-recursive channel would be pointless since one would be
-prevented from communicating across it since the focus rule requires
-maintenance of the channel type. Further, it would be undesirable to permit
-modification of a session type without guaranteeing exclusive access to the
-channel since this could lead to invalid communication e.g. trying to
-communicate on a closed channel. Additionally, a linear object is consumed
-upon being adopted and will be deallocated when its adopter's lifetime ends,
-although an alternative semantics could be defined which returned the linear
-resource to the context. Moreover, \citeauthor{Fahndrich:2002} do not provide
-operational semantics or soundness results for the ``adoption and focus'' type
-system.
-
-The Alms language~\cite{Aldrich:2009} provides a functional language with
-affine types based on linear logic. A notion of type \textit{kind} is used to
-distinguish between affine and unrestricted values. The typing judgements use
-two contexts; one for variables with unrestricted kind and another for
-variables of affine type and type variables. Functions are also labelled with
-a kind which may depend on their arguments e.g. a multiple argument function
-must have affine kinding if it accepts an affine typed resource as an argument
-other than the final argument. The reason for this is simply to prevent the
-function from using an affine resource more than once. Additionally, functions
-act as closures over their environments so their qualifier may be determined
-by a free variable in the environment, and a relation is provided to calculate
-the appropriate kind qualifier. Type variables are provided to support kind
-polymorphism of function qualifiers. The system is expressive enough to
-implement session types as a module using a similar approach to
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} although treating channels as
-affine rather than linearly typed. \citeauthor{Aldrich:2009} note that Alms'
-type system can support the ``adoption and focus'' system but do not specify
-details or whether any changes are necessary to their typing judgements.
-
-The work on a linear functional language with session types by
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the
-current project from the body of work on session-based $\lambda$-calculus type
-systems. They provide paper proofs of their system's type safety. Given the
-non-trivial features of the type system such as recursive session types and
-buffered channels for supporting asynchronous communication it will be
-challenging to extend these proofs manually. A mechanised version would
-provide stronger guarantees of correctness and allow others to more easily
-alter proofs as a result of changes made to the language.
-
-\subsubsection{Fusion}
-
-A number of systems incorporate both a functional language surface syntax with
-a translation to a more primitive process calculus. I shall focus on two such
-systems presented recently which have a connection to linear logic.
-
-Lolliproc~\cite{Mazurak:2010:LCC} provides a functional language with
-concurrency primitives as a surface syntax on top of an underlying process
-calculus. The process calculus provides a classical linear logic
-interpretation for the functional language, utilising double negation
-elimination to provide session type duality. The surface syntax is an
-effective abstraction layer for the programmer since it does not expose the
-underlying process calculus.
-
-Concurrency is provided by control operators which, respectively, spawn a
-child process and wait for a child process to complete. The spawning process
-creates a channel between parent and child process; a channel is represented
-as a continuation and the type denotes the communication protocol
-(i.e. session type) between parent and child.
-
-\citeauthor{Mazurak:2010:LCC} present proofs for type safety, strong
-normalisation and confluence. Confluence provides the guarantee that no race
-conditions can occur, strong normalisation prohibits non-terminating programs
-and type safety assures deadlock-freedom. Deadlock-freedom is implied by the
-acyclic communication graphs (progress of process reduction) which occur
-between parallel processes; follows from permitting only one channel between
-two halves of parallel composition construct in the process calculus. Most of
-the development is supported by Coq proof scripts with remaining issues due to
-reasoning about communication graphs. The development does not provide
-primitive send and receives; instead, these are encoded as linear functions.
-%% and classical extension to second-order logic by
-%% \citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}.
-
-\citeauthor{Wadler:2014}'s recent work~\cite{Wadler:2014} brings together a
-process calculus, CP, based on the earlier work by
-\citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
-\citeauthor{Gay:2010:LAST}. GV presents session types using the standard
-presentation as described in \S~\ref{sec:pre}. CP presents linear logical
-connectives with session type interpretations inspired by
-\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI}. \citeauthor{Wadler:2014}
-describes a continuation-passing style (CPS) translation from GV session types
-to CP classical linear logic propositions. The process calculus CP is slightly
-different to $\pi$DILL by \citeauthor{Caires:2010:STI}, apart from being
-classical in nature, the constructs for server replication are modified; CP
-provides a weakening and contraction rules whereas these are encoded in one
-rule for process reduction by \citeauthor{Caires:2010:STI}. Additionally,
-rather than a two-sided sequent calculus, \citeauthor{Wadler:2014} uses
-one-sided sequents for defining the process calculus, which provides a more
-intuitive presentation of session type duality and clearer correspondence to a
-$\lambda$-calculi surface syntax, simplifying the translation presentation. As
-presented, GV does not support all features of CP and thus the translation is
-one way (GV to CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP}
-extend GV to provide support for polymorphism and replication, following a
-similar CPS translation scheme. Lastly,
-\citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
-semantics for GV and their system has greater similarities with LAST than
-prior GV works, for example, by treating send and receive as linear functions
-rather than language primitives.
-
-\subsection{Mechanising Programming Language Metatheory}
-
-Interactive theorem proving and proof assistants have seen an upsurge in use
-in recent years. Especially within the domain of programming language
-research, there is a strong emphasis on providing mechanised proofs
-accompanying work in the area. Indeed, \citeauthor{Aydemir:2005:MMM} have
-published a series of challenges on mechanising programming languages aimed at
-providing a starting point for comparing different representation
-techniques~\cite{Aydemir:2005:MMM}. Another aim is to provide reusable
-libraries for common reasoning needed across programming language
-developments, e.g. handling of typing environments. This effort has resulted
-in a number of different approaches for representing programming languages
-within proof assistants. For instance,
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} present a representation
-of $\lambda$-calculi where bound variables are represented as \textit{de
-  Bruijn indices}~\footnote{A de Bruijn index is a number used to represent a
-  bound variable which indicates the position of its binder (starting from
-  zero for the innermost binder). For example, $\lambda~(\lambda~1)$
-  corresponds to $\lambda x. \lambda y. x$, i.e. the constant function.} and
-free variables are represented as named terms; the \textit{locally nameless}
-representation. Their work resulted in a library for handling commonly
-occurring aspects in programming language metatheory which has been used since
-in other developments~\cite{Park:2014:MMW}. Reusable libraries reduce the
-proof engineering effort needed for later works, in contrast to the
-pen-and-paper approach, and some proof assistants (as is the case with Coq)
-allow one to extract a program from the development which can provide a
-typechecker or compiler for the language formalised.~\footnote{Curry-Howard
-  strikes again; a proof of decidability of typechecking corresponds to a
-  program implementing a typechecker for the language.}
-
-Some recent mechanisation effort focuses on providing a basis for studying
-linearity within type systems. This work is of interest to the current project
-since most session-based type systems assume a strictly linear type system
-which requires re-binding of channel identifiers, such as in the functional
-setting of \citeauthor{Gay:2010:LAST}. \citeauthor{Mazurak:2010:LLT} present
-an extension to System F (termed \fpop) with a notion describing types as
-having linear or non-linear \textit{kind}~\cite{Mazurak:2010:LLT}. This
-extension is motivated by previous work on incorporating linearity into type
-systems. Overall, previous attempts either hamper the inclusion of desirable
-programming features (such as polymorphism), or do not adequately reflect
-non-linearity as the common case leading to awkward programming. \fpop avoids
-these deficiencies by categorising types into kinds whilst maintaining similar
-semantics to System F. Mechanised proofs for type soundness and parametricity
-are presented, and the semantics allow non-linear types to be treated as
-linear. Examples show the system can provide a wide range of permissions on
-interfaces, from full exclusive access to read-only shared references. The
-system is very close to the Alms language described by
-\citeauthor{Aldrich:2009} but has the advantage that the soundness results
-(type safety and parametricity) have been entirely mechanised in Coq. It would
-be interesting to see if one could encode the ``adopt'' and ``focus''
-constructs in this system to compare its expressiveness to Alms. The ``adopt''
-rule can be encoded similarly to the ShareRef
-example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to encode the
-``focus'' rule within the present semantics, as one needs to prevent
-non-linear access to its argument implying the presence of knowledge about
-aliasing within the typing judgements; absence of aliasing is not identified
-with linear kinded types in \fpop.
-
-While not related to programming language metatheory,
-\citeauthor{Gay:2001:FFP}~\cite{Gay:2001:FFP} provides a $\pi$-calculus
-framework mechanised in Isabelle/HOL. The aim is to build a mechanisaton of
-session-based type systems on top of the $\pi$-calculus framework. The use of
-de Bruijn indices for both bound and free variables causes issues with
-variable substitution often requiring permutation of typing environments (see
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} for a discussion on
-binder representations) and lifting of free variables during substitution; I
-avoid this issue in my representation choice (\S \ref{sec:approach}).
-
-The introduction of names for hypotheses during execution of custom tactics
-are an issue in the proof development. These names are automatically generated
-by the Isabelle/HOL system which creates a dependency between the names chosen
-and the tactic. Thus, one must be careful to use the exact names expected by
-the tactic. In the Coq proof assistant, defining custom tactics can be
-achieved using the Ltac language~\cite{Delahaye:2000:TLS}. Among other things,
-Ltac supports pattern matching on hypotheses and goal forms. Thus, if one
-wished to pattern match on an inductive type describing variables, one could
-do so without mentioning the actual variable name as follows:
-
-\begin{coq}
-Ltac my_tactic :=
-  match goal with
-  | [v: var |- _] => ... tactics here possibly mentioning `v' hypothesis ...
-  end.
-\end{coq}
-
-\subsubsection{Applications to Session-based Type Systems}\label{sec:asts}
-
-\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based
-type system providing session polymorphism. The system is more general than
-session polymorphism via subtyping~\cite{Gay:2005:SST} in that it is capable
-of typing a generic forwarding process between mutually dual session
-types. Defining such a process is not possible using subtyping since one
-wishes to restrict the types to be dual, not a subtype of some other
-type. This more general form of polymorphism is achieved by permitting input
-on all session types (including \lstinline{end}, the terminal session type)
-and then employing transition hypotheses to ensure the type permits such
-input. The authors mechanise their type system and its properties in the Coq
-proof assistant. However, the mechanised system is restricted to the
-$\pi$-calculus and the main focus is proving soundness properties for the
-deduction rules allowing polymorphism, rather than a more general basis for
-studying programming with session types.
-
-\section{Formalisation}
-
-The formalisation was performed using the Coq proof
-assisstant~\cite{Coq:manual}. In the spirit of the POPLMARK
-challenge~\cite{Aydemir:2005:MMM}, the development utilises available
-libraries and infrastructure as much as possible. In particular,
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a metatheory
-library for handling, among other things, association lists (environments) and
-free variable representation. The current development uses the version of the
-library distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which
-has been updated for Coq version 8.4pl4. One goal of the formalisation effort
-was to minimise the extent to which the development differs from the
-pen-and-paper presentation and avoid questions of adequacy of the
-representation. Some changes were made to ease the mechanisation and are
-highlighted where they occur.
-
-The locally nameless first-order representation for binders and free variables
-was adopted for both GV and CP, following
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. This first-order
-variant prevents variable-capture issues when compared with other first-order
-approaches. Alternatively, I could have chosen a higher-order abstract syntax
-(HOAS) approach such as the parametric version (PHOAS) described
-by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
-on the metalanguage (Coq, in this case) for handling binders and substitution
-so has less overhead in managing syntax than the locally nameless
-approach. However, it was less clear how to handle linearity using this
-approach and has the disadvantage of being able to represent terms not valid
-in the language being mechanised (so called ``exotic'' terms). Further, a
-higher-order approach would prevent the use of the Metatheory library which
-provides a number of useful tactics for manipulating typing contexts. Previous
-efforts to mechanise \fpop~\cite{Park:2014:MMW} illustrate how to handle a
-language containing both linear and non-linear contexts. However, to stay true
-to CP/GV, I maintained a single context. Interestingly, it became clear that
-the mechanised version of CP required explicit weakening and contraction to
-facilitate the reduction rules.
-
-For handling binder freshness, I chose to adopt the cofinite quantification
-approach described by~\cite{Aydemir:2008:EFM}, which excludes a finite set of
-variables from being considered as the binder. In contrast, the traditional
-``exists-fresh'' approach, where the binder is only required to be fresh
-within the abstraction's body, does not produce strong enough induction
-principles in some cases. This change will necessitate changes to the typing
-rules of GV and CP. Consider for instance the Connect rule in
-GV~\cite{Wadler:2014} (C) and its corresponding cofinite quantification
-version (CCO): \input{with-connect-rule}
-
-The conclusion specifies a variable in (C) but a type in (CCO) (de Bruijn
-index rather than a name). Thus, in the premise for (CCO) one can choose a
-suitably fresh $x$ in each premise. Note that, $L$ can be chosen such that
-$\Psi \cup \Phi \cup \fv(M) \cup \fv(N) \subseteq L$, where $\fv(M)$ are the
-set of free variables in term $M$, and $x$ can be the same in both
-premises. Similar changes are made to the rules for CP (see
-\S~\ref{sec:cp}). These changes do not alter the typing of processes nor the
-semantics of the translation from GV to CP. An equivalence between
-``exists-fresh'' and cofinite quantification definitions of the simply-typed
-$\lambda$-calculus is given by
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
-result could be derived straightforwardly (but tediously) for the CP/GV
-system.
-
-\begin{comment}
-\section{Issues}
-
-\begin{itemize}
-\item What is required of a Metatheory library for linear type systems?
-  Permutation reasoning? Weakening/Contraction for non-linear components?
-\item Discuss the different approaches that were taken to maintain close
-  correspondence with the paper presentation
-\item Make sure to note that the paper leaves details (a lot of the ``cruft'')
-  out of the proofs e.g. where weakening is applied in the typing derivations
-\item Mention the use of Metatheory and its limitations when mechanising
-  linear type systems like GV/CP
-\end{itemize}
-\end{comment}
-
-\subsection{The process calculus CP}\label{sec:cp}
-
-\begin{comment}
-FIGURES
-
-propositions
-
-processes
-
-typing judgements
-
-equivalences and axiomatic cut
-
-principal cut rules
-
-commuting conversions
-\end{comment}
-
-Formalising the process calculus CP provided many challenges for the
-metatheory library since to my knowledge, the library has never been applied
-to a process calculus before. So for instance, in the lambda calculus one
-typically does not work up to permutation of binders, whereas in a process
-calculus these kinds of structural equivalences are standard.
-
-Note that the rules for polymorphism, $\exists$ and $\forall$, are defined but
-cut elimination on them is not. Since GV (\S~\ref{sec:gv}) does not support
-polymorphism the reduction rules are unnecessary for the translation. Future
-work is to support reduction under these terms.
-
-\subsection{The functional language GV}\label{sec:gv}
-
-\begin{comment}
-FIGURES
-
-types
-
-terms
-
-typing judgements
-\end{comment}
-
-\subsection{CPS translation GV to CP}
-
-Currently the formalisation does not handle this. Future work.
-
-for an unbound index so this decision is sound. A consequence is that I must
-define a notion of opening language terms involving binders as described
-by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
-$t$ which is well-typed if $t$ is well-typed when its unbound index,
-representing the argument to the abstraction, is replaced by some fresh
-variable $x$, written $t^x$. Some terms in GV require multiple binders but I
-believe it will be straightforward to extend this notion. Unfortunately, I
-must define similar operations on CP terms and handle substitution and opening
-on them as well as propositions which can accept propositional variables. I
-hope to be able to generalise as much as possible with the help of the
-Metatheory library. say something about gmeta here and its automatic
-generation of these infrastructure lemmas
-
-While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
-may need to alter the approach slightly to avoid tricky formulations. For
-instance, instead of providing $n$-ary branch and choice in GV, I provide
-only binary versions of these operators to simplify the development. This
-change is not restrictive however, since CP itself provides only binary
-versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
-technical note, I amalgamate all types into one inductive definition and
-provide a predicate for restricting typing rules to consider only valid
-session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
-recursive definitions; session types are types which may contain types as
-subcomponents (arguments to send, for example). Unfortunately, handling
-mutually inductive definitions in Coq can be quite involved; requiring one to
-either rely on the Coq system to provide a stronger mutual induction principle
-or defining one manually. Such a definition is possible, but it complicates
-elimination via the induction principle.
-
-Another aspect of the encoding is how to define terms. The intrinsic encoding
-described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
-approach, indexing terms by their type so as to prevent ill-formed terms from
-being constructed. However, in the encoding for GV there are issues with using
-this encoding: $(1)$ it is not immediately clear how to handle linear contexts
-using the de Bruijn variable encoding, since in the intrinsic setting the
-environment supports weakening; $(2)$ an intrinsic encoding of terms would be
-complicated by the need to enforce that session types occur in certain
-instances (by use of a predicate); and $(3)$ well-typed terms require extra
-assumptions about binder freshness which cannot reasonably be expressed as a
-function type. For these reasons, an intrinsic approach does not offer much
-benefit since a well-typed term relation would still need to be
-defined. Therefore, I chose an extrinsic encoding for terms and define
-well-typed terms as a separate inductive type.
-
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
-their language proof infrastructure is proportional to the number of binding
-constructs. In the case of a simply typed lambda calculus this is not onerous,
-but GV has four binding constructs, CP has six and propositions have two. It
-may be true that this approach results in too many auxiliary definitions and
-operations not relevant to the actual properties one wishes to prove. However,
-I believe that once this infrastructure has been defined, the system can be
-extended without concern for these binding constructs.
-
-\citeauthor{Park:2014:MMW} describe a technique for
-removing nonlinear contexts from typing judgements. While their work extended
-to \fpop it is not clear how to handle an environment containing both
-non-linear and linear types as in GV. I wish to maintain as close a
-relationship as possible to the paper system presented by
-\citeauthor{Wadler:2014}, so separating out the non-linear and linear
-components (as in \fpop) is not an option at this stage.
-
-I describe the formalisation as it currently stands in this section. The use
-of the metatheory library from UPenn provided useful tactics and properties
-regarding freshness of free variables and environments.
-
-\subsection{Future Work}
-
-In addition to that already mentioned: there remains much to be done.
-
-\section{Conclusions}
-
-{\bf Acknowledgments.}
-This is optional; it is a location for you to thank people
-
-\printbibliography
-
-\end{document}

### 2015-04-02 paper/background.tex
Updated paper. Separated out sections and added CP figure.

--- 2015-02-24 paper/paper.tex
@@
-\documentclass{mpaper}
-
-\usepackage{macros}
-\usepackage{hyperref}
-\usepackage{comment}
-\usepackage{listings}
-\usepackage{lstcoq}
-\usepackage{graphicx}
-\usepackage{xspace}
-\usepackage{cmll}
-\usepackage{prooftree}
-\usepackage{amsmath}
-\newcommand{\typestate}[2]{{#1}\gg{#2}}
-\usepackage[style=alphabetic,sorting=nymt,natbib=true,
-  maxbibnames=100]{biblatex}
-\addbibresource{paper.bib}
-
-\DeclareSortingScheme{nymt}{
-  \sort{
-    \field{author}
-    \field{year}
-    \field{month}
-    \field{title}
-  }
-}
-
-\input{propmacros}
-
-\newcommand{\fpop}{System F${}^\circ$\xspace}
-
-\begin{document}
-
-\title{Propositions as Sessions, Mechanically}
-\author{Craig McLaughlin}
-\matricnum{1002524}
-
-\maketitle
-
-\begin{abstract}
-\begin{comment}
-SPL (+CPL)
-
-WTD
-
-ROF
-
-IMP
-\end{comment}
-\end{abstract}
-
-\section{Introduction}\label{sec:intro}
-
-Distributed communication systems are becoming increasingly prevalent in
-modern society. To facilitate communication between concurrently executing
-agents requires complex network protocols to define the type and sequencing of
-messages. Typically these protocols are implemented in systems programming
-languages, such as C, which provide no correctness guarantees of the
-implementation. Furthermore, programming in these languages is often extremely
-error-prone, as one has a myriad of unsafe features at one's disposal. As more
-businesses begin to conduct transactions over the Web, it is imperative that
-they can reason about the communication protocols upon which they rely.
-
-Encoding communication protocols as \textit{session types} is one approach to
-achieve guarantees about correctness. One particular variant of session types
-are \textit{binary} session types~\footnote{There are also \textit{multiparty}
- session types, but I do not discuss them here.}. A binary session type
-specifies communication between two agents, and it specifies the order and
-type of messages to be sent or received by an agent in the system. The two
-agents have dual session types such that, for example, when one is expecting
-to receive a message the other is expecting to send a message; both session
-types agree on the type of message to be exchanged. In this way, correctness
-of communication can be statically assured by providing communicating agents
-with a \textit{channel} whose endpoints have dual session types. A channel is
-defined as a bi-directional conduit for exchanging messages, not unlike a
-socket in network programming. In contrast to sockets, however, a channel is
-associated with an explicit type which provides a description of its
-protocol. Thus, in session-based systems a channel is usually provided as a
-primitive type.
-
-In programming language research, work on static type systems is usually
-presented on paper complete with informal proofs of correctness for the system
-properties. Most types systems in the session types community have been
-developed in this way. However, it is more difficult to extend work using this
-approach, as it may be unclear how changes affect proofs and one will
-typically have to re-consider all proofs to be convinced the properties still
-hold. More generally, unassisted formalisms are error-prone to develop and do
-not provide a firm basis upon which other researchers can experiment.
-
-Proof assistants such as Coq~\cite{Coq:manual} offer a platform on which to
-develop programming languages and type systems and associated proofs of
-properties with strong guarantees of correctness. Not only is proving with a
-proof assistant much less error-prone than the unaided approach, it also eases
-extension of the system since the proof assistant will highlight parts of
-proofs that have been affected by changes. On the other hand, one may have to
-prove more properties using this approach since this setting does not permit
-appealing to human intuition. For example, alpha-equivalence and freshness of
-binders are usually assumed in informal presentations, whereas in the
-mechanised setting one has to explicitly encode and reason about these
-properties~\cite{Aydemir:2005:MMM, Aydemir:2008:EFM}.
-
-This project describes work towards the development of a framework for
-formalising lambda calculus type systems with session types using interactive
-theorem proving (in particular, the system Coq). The framework should enable a
-whole class of related languages and type systems to be formalised with little
-extra proof effort. Inspired by previous work on a mechanised framework for
-$\pi$-calculus type systems~\cite{Gay:2001:FFP}, the current work focuses on
-the lambda calculus to provide a basis on which session types for mainstream
-programming languages can be studied, whilst taking advantage of previous
-proof effort and specialising only for those parts unique to the particular
-system. The starting point for the framework is the calculi developed by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
-classical linear logic (CP) and a functional language with session types (GV).
-
 \section{Related Work}\label{sec:rw}
@@
 
-\input{../paper/sill-branch-choice-rules}
+%\input{../paper/sill-branch-choice-rules}
 
@@
 studying programming with session types.
-
-\section{Formalisation}
-
-The formalisation was performed using the Coq proof
-assisstant~\cite{Coq:manual}. In the spirit of the POPLMARK
-challenge~\cite{Aydemir:2005:MMM}, the development utilises available
-libraries and infrastructure as much as possible. In particular,
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a metatheory
-library for handling, among other things, association lists (environments) and
-free variable representation. The current development uses the version of the
-library distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which
-has been updated for Coq version 8.4pl4. One goal of the formalisation effort
-was to minimise the extent to which the development differs from the
-pen-and-paper presentation and avoid questions of adequacy of the
-representation. Some changes were made to ease the mechanisation and are
-highlighted where they occur.
-
-The locally nameless first-order representation for binders and free variables
-was adopted for both GV and CP, following
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. This first-order
-variant prevents variable-capture issues when compared with other first-order
-approaches. Alternatively, I could have chosen a higher-order abstract syntax
-(HOAS) approach such as the parametric version (PHOAS) described
-by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
-on the metalanguage (Coq, in this case) for handling binders and substitution
-so has less overhead in managing syntax than the locally nameless
-approach. However, it was less clear how to handle linearity using this
-approach and has the disadvantage of being able to represent terms not valid
-in the language being mechanised (so called ``exotic'' terms). Further, a
-higher-order approach would prevent the use of the Metatheory library which
-provides a number of useful tactics for manipulating typing contexts. Previous
-efforts to mechanise \fpop~\cite{Park:2014:MMW} illustrate how to handle a
-language containing both linear and non-linear contexts. However, to stay true
-to CP/GV, I maintained a single context. Interestingly, it became clear that
-the mechanised version of CP required explicit weakening and contraction to
-facilitate the reduction rules.
-
-For handling binder freshness, I chose to adopt the cofinite quantification
-approach described by~\cite{Aydemir:2008:EFM}, which excludes a finite set of
-variables from being considered as the binder. In contrast, the traditional
-``exists-fresh'' approach, where the binder is only required to be fresh
-within the abstraction's body, does not produce strong enough induction
-principles in some cases. This change will necessitate changes to the typing
-rules of GV and CP. Consider for instance the Connect rule in
-GV~\cite{Wadler:2014} (C) and its corresponding cofinite quantification
-version (CCO): \input{with-connect-rule}
-
-The conclusion specifies a variable in (C) but a type in (CCO) (de Bruijn
-index rather than a name). Thus, in the premise for (CCO) one can choose a
-suitably fresh $x$ in each premise. Note that, $L$ can be chosen such that
-$\Psi \cup \Phi \cup \fv(M) \cup \fv(N) \subseteq L$, where $\fv(M)$ are the
-set of free variables in term $M$, and $x$ can be the same in both
-premises. Similar changes are made to the rules for CP (see
-\S~\ref{sec:cp}). These changes do not alter the typing of processes nor the
-semantics of the translation from GV to CP. An equivalence between
-``exists-fresh'' and cofinite quantification definitions of the simply-typed
-$\lambda$-calculus is given by
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
-result could be derived straightforwardly (but tediously) for the CP/GV
-system.
-
-\begin{comment}
-\section{Issues}
-
-\begin{itemize}
-\item What is required of a Metatheory library for linear type systems?
-  Permutation reasoning? Weakening/Contraction for non-linear components?
-\item Discuss the different approaches that were taken to maintain close
-  correspondence with the paper presentation
-\item Make sure to note that the paper leaves details (a lot of the ``cruft'')
-  out of the proofs e.g. where weakening is applied in the typing derivations
-\item Mention the use of Metatheory and its limitations when mechanising
-  linear type systems like GV/CP
-\end{itemize}
-\end{comment}
-
-\subsection{The process calculus CP}\label{sec:cp}
-
-\begin{comment}
-FIGURES
-
-propositions
-
-processes
-
-typing judgements
-
-equivalences and axiomatic cut
-
-principal cut rules
-
-commuting conversions
-\end{comment}
-
-Formalising the process calculus CP provided many challenges for the
-metatheory library since to my knowledge, the library has never been applied
-to a process calculus before. So for instance, in the lambda calculus one
-typically does not work up to permutation of binders, whereas in a process
-calculus these kinds of structural equivalences are standard.
-
-Note that the rules for polymorphism, $\exists$ and $\forall$, are defined but
-cut elimination on them is not. Since GV (\S~\ref{sec:gv}) does not support
-polymorphism the reduction rules are unnecessary for the translation. Future
-work is to support reduction under these terms.
-
-\subsection{The functional language GV}\label{sec:gv}
-
-\begin{comment}
-FIGURES
-
-types
-
-terms
-
-typing judgements
-\end{comment}
-
-\subsection{CPS translation GV to CP}
-
-Currently the formalisation does not handle this. Future work.
-
-for an unbound index so this decision is sound. A consequence is that I must
-define a notion of opening language terms involving binders as described
-by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
-$t$ which is well-typed if $t$ is well-typed when its unbound index,
-representing the argument to the abstraction, is replaced by some fresh
-variable $x$, written $t^x$. Some terms in GV require multiple binders but I
-believe it will be straightforward to extend this notion. Unfortunately, I
-must define similar operations on CP terms and handle substitution and opening
-on them as well as propositions which can accept propositional variables. I
-hope to be able to generalise as much as possible with the help of the
-Metatheory library. say something about gmeta here and its automatic
-generation of these infrastructure lemmas
-
-While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
-may need to alter the approach slightly to avoid tricky formulations. For
-instance, instead of providing $n$-ary branch and choice in GV, I provide
-only binary versions of these operators to simplify the development. This
-change is not restrictive however, since CP itself provides only binary
-versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
-technical note, I amalgamate all types into one inductive definition and
-provide a predicate for restricting typing rules to consider only valid
-session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
-recursive definitions; session types are types which may contain types as
-subcomponents (arguments to send, for example). Unfortunately, handling
-mutually inductive definitions in Coq can be quite involved; requiring one to
-either rely on the Coq system to provide a stronger mutual induction principle
-or defining one manually. Such a definition is possible, but it complicates
-elimination via the induction principle.
-
-Another aspect of the encoding is how to define terms. The intrinsic encoding
-described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
-approach, indexing terms by their type so as to prevent ill-formed terms from
-being constructed. However, in the encoding for GV there are issues with using
-this encoding: $(1)$ it is not immediately clear how to handle linear contexts
-using the de Bruijn variable encoding, since in the intrinsic setting the
-environment supports weakening; $(2)$ an intrinsic encoding of terms would be
-complicated by the need to enforce that session types occur in certain
-instances (by use of a predicate); and $(3)$ well-typed terms require extra
-assumptions about binder freshness which cannot reasonably be expressed as a
-function type. For these reasons, an intrinsic approach does not offer much
-benefit since a well-typed term relation would still need to be
-defined. Therefore, I chose an extrinsic encoding for terms and define
-well-typed terms as a separate inductive type.
-
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
-their language proof infrastructure is proportional to the number of binding
-constructs. In the case of a simply typed lambda calculus this is not onerous,
-but GV has four binding constructs, CP has six and propositions have two. It
-may be true that this approach results in too many auxiliary definitions and
-operations not relevant to the actual properties one wishes to prove. However,
-I believe that once this infrastructure has been defined, the system can be
-extended without concern for these binding constructs.
-
-\citeauthor{Park:2014:MMW} describe a technique for
-removing nonlinear contexts from typing judgements. While their work extended
-to \fpop it is not clear how to handle an environment containing both
-non-linear and linear types as in GV. I wish to maintain as close a
-relationship as possible to the paper system presented by
-\citeauthor{Wadler:2014}, so separating out the non-linear and linear
-components (as in \fpop) is not an option at this stage.
-
-I describe the formalisation as it currently stands in this section. The use
-of the metatheory library from UPenn provided useful tactics and properties
-regarding freshness of free variables and environments.
-
-\subsection{Future Work}
-
-In addition to that already mentioned: there remains much to be done.
-
-\section{Conclusions}
-
-{\bf Acknowledgments.}
-This is optional; it is a location for you to thank people
-
-\printbibliography
-
-\end{document}

### 2015-04-02 paper/formalisation.tex
Updated paper. Separated out sections and added CP figure.

--- 2015-02-25 paper/paper.tex
@@
-\documentclass{mpaper}
-
-\usepackage{macros}
-\usepackage{comment}
-\usepackage{listings}
-\usepackage{lstcoq}
-\usepackage{graphicx}
-\usepackage{xspace}
-\usepackage{cmll}
-\usepackage{prooftree}
-\usepackage{amsmath}
-\newcommand{\typestate}[2]{{#1}\gg{#2}}
-\usepackage[style=alphabetic,sorting=nymt,natbib=true,
-  maxbibnames=100]{biblatex}
-\addbibresource{paper.bib}
-
-\DeclareSortingScheme{nymt}{
-  \sort{
-    \field{author}
-    \field{year}
-    \field{month}
-    \field{title}
-  }
-}
-
-\input{propmacros}
-
-\newcommand{\fpop}{System F${}^\circ$\xspace}
-
-\begin{document}
-
-\title{Propositions as Sessions, Mechanically}
-\author{Craig McLaughlin}
-\matricnum{1002524}
-
-\maketitle
-
-\begin{abstract}
-\begin{comment}
-SPL (+CPL)
-
-WTD
-
-ROF
-
-IMP
-\end{comment}
-\end{abstract}
-
-\section{Introduction}\label{sec:intro}
-
-Distributed communication systems are becoming increasingly prevalent in
-modern society. To facilitate communication between concurrently executing
-agents requires complex network protocols to define the type and sequencing of
-messages. Typically these protocols are implemented in systems programming
-languages, such as C, which provide no correctness guarantees of the
-implementation. Furthermore, programming in these languages is often extremely
-error-prone, as one has a myriad of unsafe features at one's disposal. As more
-businesses begin to conduct transactions over the Web, it is imperative that
-they can reason about the communication protocols upon which they rely.
-
-Encoding communication protocols as \textit{session types} is one approach to
-achieve guarantees about correctness. One particular variant of session types
-are \textit{binary} session types~\footnote{There are also \textit{multiparty}
- session types, but I do not discuss them here.}. A binary session type
-specifies communication between two agents, and it specifies the order and
-type of messages to be sent or received by an agent in the system. The two
-agents have dual session types such that, for example, when one is expecting
-to receive a message the other is expecting to send a message; both session
-types agree on the type of message to be exchanged. In this way, correctness
-of communication can be statically assured by providing communicating agents
-with a \textit{channel} whose endpoints have dual session types. A channel is
-defined as a bi-directional conduit for exchanging messages, not unlike a
-socket in network programming. In contrast to sockets, however, a channel is
-associated with an explicit type which provides a description of its
-protocol. Thus, in session-based systems a channel is usually provided as a
-primitive type.
-
-In programming language research, work on static type systems is usually
-presented on paper complete with informal proofs of correctness for the system
-properties. Most types systems in the session types community have been
-developed in this way. However, it is more difficult to extend work using this
-approach, as it may be unclear how changes affect proofs and one will
-typically have to re-consider all proofs to be convinced the properties still
-hold. More generally, unassisted formalisms are error-prone to develop and do
-not provide a firm basis upon which other researchers can experiment.
-
-Proof assistants such as Coq~\cite{Coq:manual} offer a platform on which to
-develop programming languages and type systems and associated proofs of
-properties with strong guarantees of correctness. Not only is proving with a
-proof assistant much less error-prone than the unaided approach, it also eases
-extension of the system since the proof assistant will highlight parts of
-proofs that have been affected by changes. On the other hand, one may have to
-prove more properties using this approach since this setting does not permit
-appealing to human intuition. For example, alpha-equivalence and freshness of
-binders are usually assumed in informal presentations, whereas in the
-mechanised setting one has to explicitly encode and reason about these
-properties~\cite{Aydemir:2005:MMM, Aydemir:2008:EFM}.
-
-This project describes work towards the development of a framework for
-formalising lambda calculus type systems with session types using interactive
-theorem proving (in particular, the system Coq). The framework should enable a
-whole class of related languages and type systems to be formalised with little
-extra proof effort. Inspired by previous work on a mechanised framework for
-$\pi$-calculus type systems~\cite{Gay:2001:FFP}, the current work focuses on
-the lambda calculus to provide a basis on which session types for mainstream
-programming languages can be studied, whilst taking advantage of previous
-proof effort and specialising only for those parts unique to the particular
-system. The starting point for the framework is the calculi developed by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
-classical linear logic (CP) and a functional language with session types (GV).
-
-\section{Related Work}\label{sec:rw}
-
-The background for this project can be categorised roughly as previous work on
-session-based type systems (sub-divided into $\pi$ and $\lambda$-calculi
-variants, \S \ref{sec:pis} and \S \ref{sec:lam}, respectively) and previous
-work on formal verification using proof assistants (particularly focussed on
-programming language mechanisation). Finally, some recent work has merged
-these two disciplines by verifying a session-based $\pi$-calculus system (\S
-\ref{sec:asts}).
-
-\subsection{Session-based Type Systems and Linear Logic}\label{sec:sts}
-
-\subsubsection{Preamble}\label{sec:pre}
-
-The definition of session types will be important for the remainder:
-\input{../paper/session-types.tex}
-
-Here, I use a modified version of the definition given by
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} where I have omitted recursive
-session types. The type \lstinline{end} denotes the terminal session type and
-a channel typed as such may not perform any communication. A communication
-channel with session type $\inpt{T}{S}$ expects to receive a term of type $T$
-and then continues with session type $S$. Conversely, a channel with type
-$\outpt{T}{S}$ can send a term of type $T$ and then has type $S$. Thus, input
-and output are considered \emph{dual} session types, in the sense that a
-channel with one endpoint having type $\inpt{T}{S}$ and the other endpoint
-having type $\outpt{T}{\Sbar{S}}$, where $\Sbar{S}$ is the dual of $S$,
-permits communication between two agents, allowing one to send a term of type
-$T$ and the other to receive it with continuing endpoint types $\Sbar{S}$ and
-$S$, respectively. Similarly, the remaining types $\Branch{l_i}{S_i}{i\in I}$
-and $\Choice{l_i}{S_i}{i\in I}$ are duals of each other. A channel of type
-$\Branch{l_i}{S_i}{i\in I}$ can receive a label $l_i$ which will select the
-corresponding session type $S_i$ as the continuing type of the channel. While
-a channel with type $\Choice{l_i}{S_i}{i\in I}$ can send a label $l_i$ and
-continues as $S_i$.
-
-\subsubsection{In the \texorpdfstring{$\pi$}{pi}-calculus}\label{sec:pis}
-
-\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} present a connection
-between a $\pi$-calculus type system with session types ($\pi$DILL) and dual
-intuitionistic linear logic (DILL)~\cite{Barber:1996}. They equate linear
-propositions with session types of the process calculus via a bidirectional
-translation relation. They use a binary reduction relation and a labelled
-transition system to represent process reduction and process action with its
-environment as is standard~\cite{Sangiorgi:2001}.
-
-\input{../paper/sill-branch-choice-rules}
-
-They represent dual intuitionistic linear logic as a sequent calculus
-providing logical connectives which have session-based interpretations in
-$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL in
-Figure~\ref{fig:sill-branching}, which have only two alternatives
-(cf. multiple alternatives in the standard definition in \S~\ref{sec:pre}).
-
-Note that branch and choice both have some notion of branching (R rules for
-$\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
-a session-based setting, since the intuition for \& is that of
-\emph{offering} (receiving) a choice between alternatives rather than
-\emph{making} (sending) the choice. Here, however, the intuition should be
-extended to account for its location in the typing judgement. The channel is
-on the left of the sequent, and \citeauthor{Caires:2010:STI} suggest one
-consider the context as services being provided to the typed process. Thus,
-the type of channels in the context is really the type of the other endpoint
-of the channel, and the typed process behaves according to the dual
-protocol. For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to
-represent the dual of a protocol $A$ as defined by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process is behaving
-according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from
-alternatives (in this case, $\inrterm$) and then behaving according to the
-selected protocol (in this case, protocol $B^\bot$).
-
-As noted by Wadler~\cite{Wadler:2014} there is a certain amount of redundancy
-in the two-sided nature of the rules forcing one to commit to a certain
-convention when typing a communication system. Consider, an equivalent system
-to the buy/quote example given by
-\citeauthor{Caires:2010:STI}~\cite[\S~3]{Caires:2010:STI}:
-\begin{gather*}
-\sdef{ClientProto}{\BChoice{(\tensor{Book}
-                            {(\tensor{Card}{(\lfn{Receipt}{\1})})})}
-                           {(\tensor{Book}{(\lfn{Quote}{\1})})}}
-\end{gather*}
-
-Note, I am using $\tensor{}{}$ for session output and $\lfn{}{}$ for session
-input following the presentation by \citeauthor{Caires:2010:STI}. In the
-protocol defined above, a client selects between buying a book or receiving a
-price quote for a book. In the buy option case, the client sends the name of
-the book to the server and their card details, then the server sends the
-receipt for the purchase. In the quote option case, the client sends the name
-of the book and receives a price quote from the server. I have replaced the
-types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more meaningful
-type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate; all are
-synonyms for the type $\1$ which denotes the terminal session type in this
-system. The client and server process bodies are as follows:
-\begin{gather*}
-\sdef{ClientBody_c}
-     {\CPinr{c}{\SLoutput{1984}{c}{\SLinput{c}{quote}{\0}}}}\\
-\sdef{ServerBody_c}
-     {\CPcase{c}
-             {\SLinput{c}{book}{\SLinput{c}{card}{\SLoutput{receipt}{c}{\0}}}}
-             {\SLinput{c}{book}{\SLoutput{quote}{c}{\0}}}}
-\end{gather*}
-
-\begin{samepage}
-The typings for the client and server processes are given by:
-\begin{gather*}
-\tsequent{\cdot}{\cdot}{ClientBody_c}{\typedid{c}{ClientProto}}
-\tag{\mkTrule{ClientBody}} \label{eq:cbtyp}
-\\ \tsequent{\cdot}{\typedid{c}{ClientProto}}{ServerBody_c}{\typedid{-}{\1}}
-\tag{\mkTrule{ServerBody}} \label{eq:sbtyp}
-\end{gather*}
-\end{samepage}
-
-The derivation for \eqref{eq:cbtyp} is $T\1 R$, $\lfn{T}{R}$, $\tensor{T}{R}$,
-$\BChoice{T}{R_2}$ and \eqref{eq:sbtyp} is derivable from
-$\BChoice{T}{L}$. Except for the final rule, these are exact dual rules of the
-rules used in example by \citeauthor{Caires:2010:STI}. Indeed, consider the
-composite process:
-\begin{gather*}
-\sdef{QSimple}{\Spar{c}{ClientBody_c}{ServerBody_c}}\\
-\tsequent{\cdot}{\cdot}{QSimple}{\typedid{-}{\1}}
-\end{gather*}
-
-The final system is near identical to the one produced in the example given by
-\citeauthor{Caires:2010:STI} except the individual processes within the
-parallel composition have been swapped. Here the client communicates along the
-channel $c$ on the right-side of the sequent and the server utilises the
-channel $c$ on the left of the sequent. This typing, while correct, violates
-the intuition given by \citeauthor{Caires:2010:STI} of the context providing
-the services used by the process, since one would usually associate provision
-of services with the server not the client. Instead, the example above
-suggests the server utilises some channel provided by the client. On the other
-hand, one could imagine such a channel being setup between two communicating
-agents; the ``owner'' of the channel is largely irrelevant. Note that
-$ClientProto$ is in fact the dual of the $ServerProto$ definition by
-\citeauthor{Caires:2010:STI} using the definition of dual used in Wadler's
-classical process calculus~\cite{Wadler:2014}. In $\pi$DILL, the duality
-between endpoints of the same channel are obscured by the nature of the typing
-judgements.
-
-Process reduction steps are equated with proof reduction via cut
-elimination. The relationship with proof normalisation in DILL provides
-absence of deadlocks in session-based communication for the process
-calculus. The correspondence of session types to linear logic propositions is
-an important step in providing a solid logical foundation for a concurrent
-type theory; in the same way ``sequential'' type theory ($\lambda$-calculi)
-profitted from a logical perspective. However, for session types to become a
-more usable tool for programming distributed communication systems (the
-ultimate goal, of course, is to write more correct programs), one cannot be
-expected to program directly in the $\pi$-calculus; a more mainstream
-programming paradigm must be built upon these solid logical foundations. With
-that in mind, enter the $\lambda$-calculus!
-
-\subsubsection{In the \texorpdfstring{$\lambda$}{lambda}-calculus}
-\label{sec:lam}
-
-A series of works by \citeauthor{Gay:2003:STI}~\cite{Gay:2003:STI} and
-\citeauthor{Vasconcelos:2006:TCM}~\cite{Vasconcelos:2006:TCM} provide a
-$\lambda$-calculus based language with session types for channel
-primitives. The type system is nonstandard, featuring an additional
-environment for tracking the session types of channels; these types are
-modified during term reduction to reflect communication. The extra channel
-environment is provided to support possible aliasing of channels by treating
-channel variables as pointers or references into the channel environment. So,
-for a channel $c$ of session type $S$, a variable $x$ referring to this
-channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the
-channel environment. The typing judgements and term reduction relation must be
-augmented to handle these environments. Unfortunately, a side-effect of this
-is that function types must specify the pre and post-states of channel
-environments and therefore, the type of arguments cannot work up to
-alpha-equivalence of channel identifiers. That is, if two channels $c$ and $d$
-have the same session type, one cannot define a function to operate on
-variables of either channel type since the function definition will explicitly
-mention the identifier for the channel (in this case either, $Chan~c$ or
-$Chan~d$). However, the system does provide an attractive non-linear channel
-type which has been neglected in favour of a linear treatment of channels in
-more recent
-works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2014}. Certainly, it is the
-case that an aliased approach would fit better with mainstream programming
-languages and perhaps studying a similar system could offer some insights.
-
-The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of
-aliased access to linear resources. The type system requires special language
-features, e.g. to manage capabilities which can be thought as a form of object
-liveness mechanism, so GV typing judgements would need to be extended such
-that the capabilities can be included in soundness results. An extension to
-the translation to CP to incorporate the ``adopt'' and ``focus'' constructs
-would also need to be defined. It is not immediately clear how these features
-would assist in the management of session typed channels. The ``focus'' rule
-(and linearity generally) enforces an invariant on the focussed object
-(channel) such that the type is maintained during the focus operation:
-\input{../proposal/focus-rule}
-
-The typing judgements include \emph{capabilities} that track aliasing to heap
-objects, e.g. $\AFcap{\rho}{h}$ denotes a mapping from a static name $\rho$ to
-a heap type $h$, and a variable $x$ may be typed as $\typedid{x}{\AFtr{\rho}}$
-indicating that $x$ refers to an object of type $h$. The purpose of the
-capabilities is to enforce restrictions on when an object can be
-referenced. In other words, if $\AFcap{\rho}{h}$ is in the capabilities at
-some program point then $x$ may be used, otherwise such use is prohibited. In
-the [focus] rule, evaluating $e_1$ retrieves the object to be focussed. Note
-the type of $e_1$ is a ``guarded'' type. The guard $G$ is the static name of
-some object which ``adopted''~\footnote{I omit the adopt typing rule for
-  brevity.} an object of heap type $h$. If a program point has
-$\AFcap{G}{h_1}$ in the capabilities for some $h_1$ then all objects guarded
-by $G$ may be used; $\capsequent{\Delta}{C_1}{G}$ denotes containment of a
-such a capability in $C_1$. The guarded type may be thought of as policing
-manipulation of linear components within $h$. Treating $\otimes$ as the
-disjoint union for capability sets, the typing for $e_2$ temporarily removes
-the guard $G$ from the set of capabilities, and adds a new variable $x$, and
-associated capability, with a type permitting access to the object of type $h$
-i.e. a non-guarded type. Removing $G$ will prohibit access to allow objects
-guarded by $G$ for the during of $e_2$. The final capabilities after executing
-$e_2$ are required to be $\tensor{C_3}{\AFcap{\rho}{h}}$, enforcing that the
-type must remain as $h$. The typing of the $\letterm$ expression ensures that
-the guard $G$ (contained in $C_1$) is restored.
-
-Such a rule would appear to only benefit recursive session types. Indeed,
-focussing on a non-recursive channel would be pointless since one would be
-prevented from communicating across it since the focus rule requires
-maintenance of the channel type. Further, it would be undesirable to permit
-modification of a session type without guaranteeing exclusive access to the
-channel since this could lead to invalid communication e.g. trying to
-communicate on a closed channel. Additionally, a linear object is consumed
-upon being adopted and will be deallocated when its adopter's lifetime ends,
-although an alternative semantics could be defined which returned the linear
-resource to the context. Moreover, \citeauthor{Fahndrich:2002} do not provide
-operational semantics or soundness results for the ``adoption and focus'' type
-system.
-
-The Alms language~\cite{Aldrich:2009} provides a functional language with
-affine types based on linear logic. A notion of type \textit{kind} is used to
-distinguish between affine and unrestricted values. The typing judgements use
-two contexts; one for variables with unrestricted kind and another for
-variables of affine type and type variables. Functions are also labelled with
-a kind which may depend on their arguments e.g. a multiple argument function
-must have affine kinding if it accepts an affine typed resource as an argument
-other than the final argument. The reason for this is simply to prevent the
-function from using an affine resource more than once. Additionally, functions
-act as closures over their environments so their qualifier may be determined
-by a free variable in the environment, and a relation is provided to calculate
-the appropriate kind qualifier. Type variables are provided to support kind
-polymorphism of function qualifiers. The system is expressive enough to
-implement session types as a module using a similar approach to
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} although treating channels as
-affine rather than linearly typed. \citeauthor{Aldrich:2009} note that Alms'
-type system can support the ``adoption and focus'' system but do not specify
-details or whether any changes are necessary to their typing judgements.
-
-The work on a linear functional language with session types by
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the
-current project from the body of work on session-based $\lambda$-calculus type
-systems. They provide paper proofs of their system's type safety. Given the
-non-trivial features of the type system such as recursive session types and
-buffered channels for supporting asynchronous communication it will be
-challenging to extend these proofs manually. A mechanised version would
-provide stronger guarantees of correctness and allow others to more easily
-alter proofs as a result of changes made to the language.
-
-\subsubsection{Fusion}
-
-A number of systems incorporate both a functional language surface syntax with
-a translation to a more primitive process calculus. I shall focus on two such
-systems presented recently which have a connection to linear logic.
-
-Lolliproc~\cite{Mazurak:2010:LCC} provides a functional language with
-concurrency primitives as a surface syntax on top of an underlying process
-calculus. The process calculus provides a classical linear logic
-interpretation for the functional language, utilising double negation
-elimination to provide session type duality. The surface syntax is an
-effective abstraction layer for the programmer since it does not expose the
-underlying process calculus.
-
-Concurrency is provided by control operators which, respectively, spawn a
-child process and wait for a child process to complete. The spawning process
-creates a channel between parent and child process; a channel is represented
-as a continuation and the type denotes the communication protocol
-(i.e. session type) between parent and child.
-
-\citeauthor{Mazurak:2010:LCC} present proofs for type safety, strong
-normalisation and confluence. Confluence provides the guarantee that no race
-conditions can occur, strong normalisation prohibits non-terminating programs
-and type safety assures deadlock-freedom. Deadlock-freedom is implied by the
-acyclic communication graphs (progress of process reduction) which occur
-between parallel processes; follows from permitting only one channel between
-two halves of parallel composition construct in the process calculus. Most of
-the development is supported by Coq proof scripts with remaining issues due to
-reasoning about communication graphs. The development does not provide
-primitive send and receives; instead, these are encoded as linear functions.
-%% and classical extension to second-order logic by
-%% \citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}.
-
-\citeauthor{Wadler:2014}'s recent work~\cite{Wadler:2014} brings together a
-process calculus, CP, based on the earlier work by
-\citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
-\citeauthor{Gay:2010:LAST}. GV presents session types using the standard
-presentation as described in \S~\ref{sec:pre}. CP presents linear logical
-connectives with session type interpretations inspired by
-\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI}. \citeauthor{Wadler:2014}
-describes a continuation-passing style (CPS) translation from GV session types
-to CP classical linear logic propositions. The process calculus CP is slightly
-different to $\pi$DILL by \citeauthor{Caires:2010:STI}, apart from being
-classical in nature, the constructs for server replication are modified; CP
-provides a weakening and contraction rules whereas these are encoded in one
-rule for process reduction by \citeauthor{Caires:2010:STI}. Additionally,
-rather than a two-sided sequent calculus, \citeauthor{Wadler:2014} uses
-one-sided sequents for defining the process calculus, which provides a more
-intuitive presentation of session type duality and clearer correspondence to a
-$\lambda$-calculi surface syntax, simplifying the translation presentation. As
-presented, GV does not support all features of CP and thus the translation is
-one way (GV to CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP}
-extend GV to provide support for polymorphism and replication, following a
-similar CPS translation scheme. Lastly,
-\citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
-semantics for GV and their system has greater similarities with LAST than
-prior GV works, for example, by treating send and receive as linear functions
-rather than language primitives.
-
-\subsection{Mechanising Programming Language Metatheory}
-
-Interactive theorem proving and proof assistants have seen an upsurge in use
-in recent years. Especially within the domain of programming language
-research, there is a strong emphasis on providing mechanised proofs
-accompanying work in the area. Indeed, \citeauthor{Aydemir:2005:MMM} have
-published a series of challenges on mechanising programming languages aimed at
-providing a starting point for comparing different representation
-techniques~\cite{Aydemir:2005:MMM}. Another aim is to provide reusable
-libraries for common reasoning needed across programming language
-developments, e.g. handling of typing environments. This effort has resulted
-in a number of different approaches for representing programming languages
-within proof assistants. For instance,
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} present a representation
-of $\lambda$-calculi where bound variables are represented as \textit{de
-  Bruijn indices}~\footnote{A de Bruijn index is a number used to represent a
-  bound variable which indicates the position of its binder (starting from
-  zero for the innermost binder). For example, $\lambda~(\lambda~1)$
-  corresponds to $\lambda x. \lambda y. x$, i.e. the constant function.} and
-free variables are represented as named terms; the \textit{locally nameless}
-representation. Their work resulted in a library for handling commonly
-occurring aspects in programming language metatheory which has been used since
-in other developments~\cite{Park:2014:MMW}. Reusable libraries reduce the
-proof engineering effort needed for later works, in contrast to the
-pen-and-paper approach, and some proof assistants (as is the case with Coq)
-allow one to extract a program from the development which can provide a
-typechecker or compiler for the language formalised.~\footnote{Curry-Howard
-  strikes again; a proof of decidability of typechecking corresponds to a
-  program implementing a typechecker for the language.}
-
-Some recent mechanisation effort focuses on providing a basis for studying
-linearity within type systems. This work is of interest to the current project
-since most session-based type systems assume a strictly linear type system
-which requires re-binding of channel identifiers, such as in the functional
-setting of \citeauthor{Gay:2010:LAST}. \citeauthor{Mazurak:2010:LLT} present
-an extension to System F (termed \fpop) with a notion describing types as
-having linear or non-linear \textit{kind}~\cite{Mazurak:2010:LLT}. This
-extension is motivated by previous work on incorporating linearity into type
-systems. Overall, previous attempts either hamper the inclusion of desirable
-programming features (such as polymorphism), or do not adequately reflect
-non-linearity as the common case leading to awkward programming. \fpop avoids
-these deficiencies by categorising types into kinds whilst maintaining similar
-semantics to System F. Mechanised proofs for type soundness and parametricity
-are presented, and the semantics allow non-linear types to be treated as
-linear. Examples show the system can provide a wide range of permissions on
-interfaces, from full exclusive access to read-only shared references. The
-system is very close to the Alms language described by
-\citeauthor{Aldrich:2009} but has the advantage that the soundness results
-(type safety and parametricity) have been entirely mechanised in Coq. It would
-be interesting to see if one could encode the ``adopt'' and ``focus''
-constructs in this system to compare its expressiveness to Alms. The ``adopt''
-rule can be encoded similarly to the ShareRef
-example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to encode the
-``focus'' rule within the present semantics, as one needs to prevent
-non-linear access to its argument implying the presence of knowledge about
-aliasing within the typing judgements; absence of aliasing is not identified
-with linear kinded types in \fpop.
-
-While not related to programming language metatheory,
-\citeauthor{Gay:2001:FFP}~\cite{Gay:2001:FFP} provides a $\pi$-calculus
-framework mechanised in Isabelle/HOL. The aim is to build a mechanisaton of
-session-based type systems on top of the $\pi$-calculus framework. The use of
-de Bruijn indices for both bound and free variables causes issues with
-variable substitution often requiring permutation of typing environments (see
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} for a discussion on
-binder representations) and lifting of free variables during substitution; I
-avoid this issue in my representation choice (\S \ref{sec:approach}).
-
-The introduction of names for hypotheses during execution of custom tactics
-are an issue in the proof development. These names are automatically generated
-by the Isabelle/HOL system which creates a dependency between the names chosen
-and the tactic. Thus, one must be careful to use the exact names expected by
-the tactic. In the Coq proof assistant, defining custom tactics can be
-achieved using the Ltac language~\cite{Delahaye:2000:TLS}. Among other things,
-Ltac supports pattern matching on hypotheses and goal forms. Thus, if one
-wished to pattern match on an inductive type describing variables, one could
-do so without mentioning the actual variable name as follows:
-
-\begin{coq}
-Ltac my_tactic :=
-  match goal with
-  | [v: var |- _] =>
-    ... tactics here possibly mentioning `v' hypothesis ...
-  end.
-\end{coq}
-
-\subsubsection{Applications to Session-based Type Systems}\label{sec:asts}
-
-\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based
-type system providing session polymorphism. The system is more general than
-session polymorphism via subtyping~\cite{Gay:2005:SST} in that it is capable
-of typing a generic forwarding process between mutually dual session
-types. Defining such a process is not possible using subtyping since one
-wishes to restrict the types to be dual, not a subtype of some other
-type. This more general form of polymorphism is achieved by permitting input
-on all session types (including \lstinline{end}, the terminal session type)
-and then employing transition hypotheses to ensure the type permits such
-input. The authors mechanise their type system and its properties in the Coq
-proof assistant. However, the mechanised system is restricted to the
-$\pi$-calculus and the main focus is proving soundness properties for the
-deduction rules allowing polymorphism, rather than a more general basis for
-studying programming with session types.
-
 \section{Formalisation}
@@
 
+%\input{with-connect-rule}
+\input{cp}
+
 The locally nameless first-order representation for binders and free variables
@@
 so has less overhead in managing syntax than the locally nameless
-approach. However, it was less clear how to handle linearity using this
+approach. However, it is less clear how to handle linearity using this
 approach and has the disadvantage of being able to represent terms not valid
@@
 handle a language containing both linear and non-linear contexts. However, to
-stay true to CP/GV, I maintained a single context. Interestingly, it became
+stay true to CP/GV, I maintain a single context. Interestingly, it became
 clear that the mechanised version of CP required explicit weakening and
 contraction to facilitate the reduction rules (\S~\ref{sec:cp}).
-
-\input{with-connect-rule}
 
@@
 Currently the formalisation does not handle this. Future work.
-
-\section{Future Work}
-
-In addition to that already mentioned: there remains much to be done.
-
-\section{Conclusions}
-
-{\bf Acknowledgments.}
-This is optional; it is a location for you to thank people
-
-\sloppy
-\printbibliography
-
-\end{document}

### 2015-04-02 paper/futurework.tex
Updated paper. Separated out sections and added CP figure.

--- 2015-02-24 paper/paper.tex
@@
-\documentclass{mpaper}
-
-\usepackage{macros}
-\usepackage{hyperref}
-\usepackage{comment}
-\usepackage{listings}
-\usepackage{lstcoq}
-\usepackage{graphicx}
-\usepackage{xspace}
-\usepackage{cmll}
-\usepackage{prooftree}
-\usepackage{amsmath}
-\newcommand{\typestate}[2]{{#1}\gg{#2}}
-\usepackage[style=alphabetic,sorting=nymt,natbib=true,
-  maxbibnames=100]{biblatex}
-\addbibresource{paper.bib}
-
-\DeclareSortingScheme{nymt}{
-  \sort{
-    \field{author}
-    \field{year}
-    \field{month}
-    \field{title}
-  }
-}
-
-\input{propmacros}
-
-\newcommand{\fpop}{System F${}^\circ$\xspace}
-
-\begin{document}
-
-\title{Propositions as Sessions, Mechanically}
-\author{Craig McLaughlin}
-\matricnum{1002524}
-
-\maketitle
-
-\begin{abstract}
-\begin{comment}
-SPL (+CPL)
-
-WTD
-
-ROF
-
-IMP
-\end{comment}
-\end{abstract}
-
-\section{Introduction}\label{sec:intro}
-
-Distributed communication systems are becoming increasingly prevalent in
-modern society. To facilitate communication between concurrently executing
-agents requires complex network protocols to define the type and sequencing of
-messages. Typically these protocols are implemented in systems programming
-languages, such as C, which provide no correctness guarantees of the
-implementation. Furthermore, programming in these languages is often extremely
-error-prone, as one has a myriad of unsafe features at one's disposal. As more
-businesses begin to conduct transactions over the Web, it is imperative that
-they can reason about the communication protocols upon which they rely.
-
-Encoding communication protocols as \textit{session types} is one approach to
-achieve guarantees about correctness. One particular variant of session types
-are \textit{binary} session types~\footnote{There are also \textit{multiparty}
- session types, but I do not discuss them here.}. A binary session type
-specifies communication between two agents, and it specifies the order and
-type of messages to be sent or received by an agent in the system. The two
-agents have dual session types such that, for example, when one is expecting
-to receive a message the other is expecting to send a message; both session
-types agree on the type of message to be exchanged. In this way, correctness
-of communication can be statically assured by providing communicating agents
-with a \textit{channel} whose endpoints have dual session types. A channel is
-defined as a bi-directional conduit for exchanging messages, not unlike a
-socket in network programming. In contrast to sockets, however, a channel is
-associated with an explicit type which provides a description of its
-protocol. Thus, in session-based systems a channel is usually provided as a
-primitive type.
-
-In programming language research, work on static type systems is usually
-presented on paper complete with informal proofs of correctness for the system
-properties. Most types systems in the session types community have been
-developed in this way. However, it is more difficult to extend work using this
-approach, as it may be unclear how changes affect proofs and one will
-typically have to re-consider all proofs to be convinced the properties still
-hold. More generally, unassisted formalisms are error-prone to develop and do
-not provide a firm basis upon which other researchers can experiment.
-
-Proof assistants such as Coq~\cite{Coq:manual} offer a platform on which to
-develop programming languages and type systems and associated proofs of
-properties with strong guarantees of correctness. Not only is proving with a
-proof assistant much less error-prone than the unaided approach, it also eases
-extension of the system since the proof assistant will highlight parts of
-proofs that have been affected by changes. On the other hand, one may have to
-prove more properties using this approach since this setting does not permit
-appealing to human intuition. For example, alpha-equivalence and freshness of
-binders are usually assumed in informal presentations, whereas in the
-mechanised setting one has to explicitly encode and reason about these
-properties~\cite{Aydemir:2005:MMM, Aydemir:2008:EFM}.
-
-This project describes work towards the development of a framework for
-formalising lambda calculus type systems with session types using interactive
-theorem proving (in particular, the system Coq). The framework should enable a
-whole class of related languages and type systems to be formalised with little
-extra proof effort. Inspired by previous work on a mechanised framework for
-$\pi$-calculus type systems~\cite{Gay:2001:FFP}, the current work focuses on
-the lambda calculus to provide a basis on which session types for mainstream
-programming languages can be studied, whilst taking advantage of previous
-proof effort and specialising only for those parts unique to the particular
-system. The starting point for the framework is the calculi developed by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
-classical linear logic (CP) and a functional language with session types (GV).
-
-\section{Related Work}\label{sec:rw}
-
-The background for this project can be categorised roughly as previous work on
-session-based type systems (sub-divided into $\pi$ and $\lambda$-calculi
-variants, \S \ref{sec:pis} and \S \ref{sec:lam}, respectively) and previous
-work on formal verification using proof assistants (particularly focussed on
-programming language mechanisation). Finally, some recent work has merged
-these two disciplines by verifying a session-based $\pi$-calculus system (\S
-\ref{sec:asts}).
-
-\subsection{Session-based Type Systems and Linear Logic}\label{sec:sts}
-
-\subsubsection{Preamble}\label{sec:pre}
-
-The definition of session types will be important for the remainder:
-\input{../paper/session-types.tex}
-
-Here, I use a modified version of the definition given by
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} where I have omitted recursive
-session types. The type \lstinline{end} denotes the terminal session type and
-a channel typed as such may not perform any communication. A communication
-channel with session type $\inpt{T}{S}$ expects to receive a term of type $T$
-and then continues with session type $S$. Conversely, a channel with type
-$\outpt{T}{S}$ can send a term of type $T$ and then has type $S$. Thus, input
-and output are considered \emph{dual} session types, in the sense that a
-channel with one endpoint having type $\inpt{T}{S}$ and the other endpoint
-having type $\outpt{T}{\Sbar{S}}$, where $\Sbar{S}$ is the dual of $S$,
-permits communication between two agents, allowing one to send a term of type
-$T$ and the other to receive it with continuing endpoint types $\Sbar{S}$ and
-$S$, respectively. Similarly, the remaining types $\Branch{l_i}{S_i}{i\in I}$
-and $\Choice{l_i}{S_i}{i\in I}$ are duals of each other. A channel of type
-$\Branch{l_i}{S_i}{i\in I}$ can receive a label $l_i$ which will select the
-corresponding session type $S_i$ as the continuing type of the channel. While
-a channel with type $\Choice{l_i}{S_i}{i\in I}$ can send a label $l_i$ and
-continues as $S_i$.
-
-\subsubsection{In the \texorpdfstring{$\pi$}{pi}-calculus}\label{sec:pis}
-
-\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} present a connection
-between a $\pi$-calculus type system with session types ($\pi$DILL) and dual
-intuitionistic linear logic (DILL)~\cite{Barber:1996}. They equate linear
-propositions with session types of the process calculus via a bidirectional
-translation relation. They use a binary reduction relation and a labelled
-transition system to represent process reduction and process action with its
-environment as is standard~\cite{Sangiorgi:2001}.
-
-\input{../paper/sill-branch-choice-rules}
-
-They represent dual intuitionistic linear logic as a sequent calculus
-providing logical connectives which have session-based interpretations in
-$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL in
-Figure~\ref{fig:sill-branching}, which have only two alternatives
-(cf. multiple alternatives in the standard definition in \S~\ref{sec:pre}).
-
-Note that branch and choice both have some notion of branching (R rules for
-$\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
-a session-based setting, since the intuition for \& is that of
-\emph{offering} (receiving) a choice between alternatives rather than
-\emph{making} (sending) the choice. Here, however, the intuition should be
-extended to account for its location in the typing judgement. The channel is
-on the left of the sequent, and \citeauthor{Caires:2010:STI} suggest one
-consider the context as services being provided to the typed process. Thus,
-the type of channels in the context is really the type of the other endpoint
-of the channel, and the typed process behaves according to the dual
-protocol. For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to
-represent the dual of a protocol $A$ as defined by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process is behaving
-according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from
-alternatives (in this case, $\inrterm$) and then behaving according to the
-selected protocol (in this case, protocol $B^\bot$).
-
-As noted by Wadler~\cite{Wadler:2014} there is a certain amount of redundancy
-in the two-sided nature of the rules forcing one to commit to a certain
-convention when typing a communication system. Consider, an equivalent system
-to the buy/quote example given by
-\citeauthor{Caires:2010:STI}~\cite[\S~3]{Caires:2010:STI}:
-\begin{gather*}
-\sdef{ClientProto}{\BChoice{(\tensor{Book}
-                            {(\tensor{Card}{(\lfn{Receipt}{\1})})})}
-                           {(\tensor{Book}{(\lfn{Quote}{\1})})}}
-\end{gather*}
-
-Note, I am using $\tensor{}{}$ for session output and $\lfn{}{}$ for session
-input following the presentation by \citeauthor{Caires:2010:STI}. In the
-protocol defined above, a client selects between buying a book or receiving a
-price quote for a book. In the buy option case, the client sends the name of
-the book to the server and their card details, then the server sends the
-receipt for the purchase. In the quote option case, the client sends the name
-of the book and receives a price quote from the server. I have replaced the
-types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more meaningful
-type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate; all are
-synonyms for the type $\1$ which denotes the terminal session type in this
-system. The client and server process bodies are as follows:
-\begin{gather*}
-\sdef{ClientBody_c}
-     {\CPinr{c}{\SLoutput{1984}{c}{\SLinput{c}{quote}{\0}}}}\\
-\sdef{ServerBody_c}
-     {\CPcase{c}
-             {\SLinput{c}{book}{\SLinput{c}{card}{\SLoutput{receipt}{c}{\0}}}}
-             {\SLinput{c}{book}{\SLoutput{quote}{c}{\0}}}}
-\end{gather*}
-
-\begin{samepage}
-The typings for the client and server processes are given by:
-\begin{gather*}
-\tsequent{\cdot}{\cdot}{ClientBody_c}{\typedid{c}{ClientProto}}
-\tag{\mkTrule{ClientBody}} \label{eq:cbtyp}
-\\ \tsequent{\cdot}{\typedid{c}{ClientProto}}{ServerBody_c}{\typedid{-}{\1}}
-\tag{\mkTrule{ServerBody}} \label{eq:sbtyp}
-\end{gather*}
-\end{samepage}
-
-The derivation for \eqref{eq:cbtyp} is $T\1 R$, $\lfn{T}{R}$, $\tensor{T}{R}$,
-$\BChoice{T}{R_2}$ and \eqref{eq:sbtyp} is derivable from
-$\BChoice{T}{L}$. Except for the final rule, these are exact dual rules of the
-rules used in example by \citeauthor{Caires:2010:STI}. Indeed, consider the
-composite process:
-\begin{gather*}
-\sdef{QSimple}{\Spar{c}{ClientBody_c}{ServerBody_c}}\\
-\tsequent{\cdot}{\cdot}{QSimple}{\typedid{-}{\1}}
-\end{gather*}
-
-The final system is near identical to the one produced in the example given by
-\citeauthor{Caires:2010:STI} except the individual processes within the
-parallel composition have been swapped. Here the client communicates along the
-channel $c$ on the right-side of the sequent and the server utilises the
-channel $c$ on the left of the sequent. This typing, while correct, violates
-the intuition given by \citeauthor{Caires:2010:STI} of the context providing
-the services used by the process, since one would usually associate provision
-of services with the server not the client. Instead, the example above
-suggests the server utilises some channel provided by the client. On the other
-hand, one could imagine such a channel being setup between two communicating
-agents; the ``owner'' of the channel is largely irrelevant. Note that
-$ClientProto$ is in fact the dual of the $ServerProto$ definition by
-\citeauthor{Caires:2010:STI} using the definition of dual used in Wadler's
-classical process calculus~\cite{Wadler:2014}. In $\pi$DILL, the duality
-between endpoints of the same channel are obscured by the nature of the typing
-judgements.
-
-Process reduction steps are equated with proof reduction via cut
-elimination. The relationship with proof normalisation in DILL provides
-absence of deadlocks in session-based communication for the process
-calculus. The correspondence of session types to linear logic propositions is
-an important step in providing a solid logical foundation for a concurrent
-type theory; in the same way ``sequential'' type theory ($\lambda$-calculi)
-profitted from a logical perspective. However, for session types to become a
-more usable tool for programming distributed communication systems (the
-ultimate goal, of course, is to write more correct programs), one cannot be
-expected to program directly in the $\pi$-calculus; a more mainstream
-programming paradigm must be built upon these solid logical foundations. With
-that in mind, enter the $\lambda$-calculus!
-
-\subsubsection{In the \texorpdfstring{$\lambda$}{lambda}-calculus}
-\label{sec:lam}
-
-A series of works by \citeauthor{Gay:2003:STI}~\cite{Gay:2003:STI} and
-\citeauthor{Vasconcelos:2006:TCM}~\cite{Vasconcelos:2006:TCM} provide a
-$\lambda$-calculus based language with session types for channel
-primitives. The type system is nonstandard, featuring an additional
-environment for tracking the session types of channels; these types are
-modified during term reduction to reflect communication. The extra channel
-environment is provided to support possible aliasing of channels by treating
-channel variables as pointers or references into the channel environment. So,
-for a channel $c$ of session type $S$, a variable $x$ referring to this
-channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the
-channel environment. The typing judgements and term reduction relation must be
-augmented to handle these environments. Unfortunately, a side-effect of this
-is that function types must specify the pre and post-states of channel
-environments and therefore, the type of arguments cannot work up to
-alpha-equivalence of channel identifiers. That is, if two channels $c$ and $d$
-have the same session type, one cannot define a function to operate on
-variables of either channel type since the function definition will explicitly
-mention the identifier for the channel (in this case either, $Chan~c$ or
-$Chan~d$). However, the system does provide an attractive non-linear channel
-type which has been neglected in favour of a linear treatment of channels in
-more recent
-works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2014}. Certainly, it is the
-case that an aliased approach would fit better with mainstream programming
-languages and perhaps studying a similar system could offer some insights.
-
-The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of
-aliased access to linear resources. The type system requires special language
-features, e.g. to manage capabilities which can be thought as a form of object
-liveness mechanism, so GV typing judgements would need to be extended such
-that the capabilities can be included in soundness results. An extension to
-the translation to CP to incorporate the ``adopt'' and ``focus'' constructs
-would also need to be defined. It is not immediately clear how these features
-would assist in the management of session typed channels. The ``focus'' rule
-(and linearity generally) enforces an invariant on the focussed object
-(channel) such that the type is maintained during the focus operation:
-\input{../proposal/focus-rule}
-
-The typing judgements include \emph{capabilities} that track aliasing to heap
-objects, e.g. $\AFcap{\rho}{h}$ denotes a mapping from a static name $\rho$ to
-a heap type $h$, and a variable $x$ may be typed as $\typedid{x}{\AFtr{\rho}}$
-indicating that $x$ refers to an object of type $h$. The purpose of the
-capabilities is to enforce restrictions on when an object can be
-referenced. In other words, if $\AFcap{\rho}{h}$ is in the capabilities at
-some program point then $x$ may be used, otherwise such use is prohibited. In
-the [focus] rule, evaluating $e_1$ retrieves the object to be focussed. Note
-the type of $e_1$ is a ``guarded'' type. The guard $G$ is the static name of
-some object which ``adopted''~\footnote{I omit the adopt typing rule for
-  brevity.} an object of heap type $h$. If a program point has
-$\AFcap{G}{h_1}$ in the capabilities for some $h_1$ then all objects guarded
-by $G$ may be used; $\capsequent{\Delta}{C_1}{G}$ denotes containment of a
-such a capability in $C_1$. The guarded type may be thought of as policing
-manipulation of linear components within $h$. Treating $\otimes$ as the
-disjoint union for capability sets, the typing for $e_2$ temporarily removes
-the guard $G$ from the set of capabilities, and adds a new variable $x$, and
-associated capability, with a type permitting access to the object of type $h$
-i.e. a non-guarded type. Removing $G$ will prohibit access to allow objects
-guarded by $G$ for the during of $e_2$. The final capabilities after executing
-$e_2$ are required to be $\tensor{C_3}{\AFcap{\rho}{h}}$, enforcing that the
-type must remain as $h$. The typing of the $\letterm$ expression ensures that
-the guard $G$ (contained in $C_1$) is restored.
-
-Such a rule would appear to only benefit recursive session types. Indeed,
-focussing on a non-recursive channel would be pointless since one would be
-prevented from communicating across it since the focus rule requires
-maintenance of the channel type. Further, it would be undesirable to permit
-modification of a session type without guaranteeing exclusive access to the
-channel since this could lead to invalid communication e.g. trying to
-communicate on a closed channel. Additionally, a linear object is consumed
-upon being adopted and will be deallocated when its adopter's lifetime ends,
-although an alternative semantics could be defined which returned the linear
-resource to the context. Moreover, \citeauthor{Fahndrich:2002} do not provide
-operational semantics or soundness results for the ``adoption and focus'' type
-system.
-
-The Alms language~\cite{Aldrich:2009} provides a functional language with
-affine types based on linear logic. A notion of type \textit{kind} is used to
-distinguish between affine and unrestricted values. The typing judgements use
-two contexts; one for variables with unrestricted kind and another for
-variables of affine type and type variables. Functions are also labelled with
-a kind which may depend on their arguments e.g. a multiple argument function
-must have affine kinding if it accepts an affine typed resource as an argument
-other than the final argument. The reason for this is simply to prevent the
-function from using an affine resource more than once. Additionally, functions
-act as closures over their environments so their qualifier may be determined
-by a free variable in the environment, and a relation is provided to calculate
-the appropriate kind qualifier. Type variables are provided to support kind
-polymorphism of function qualifiers. The system is expressive enough to
-implement session types as a module using a similar approach to
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} although treating channels as
-affine rather than linearly typed. \citeauthor{Aldrich:2009} note that Alms'
-type system can support the ``adoption and focus'' system but do not specify
-details or whether any changes are necessary to their typing judgements.
-
-The work on a linear functional language with session types by
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the
-current project from the body of work on session-based $\lambda$-calculus type
-systems. They provide paper proofs of their system's type safety. Given the
-non-trivial features of the type system such as recursive session types and
-buffered channels for supporting asynchronous communication it will be
-challenging to extend these proofs manually. A mechanised version would
-provide stronger guarantees of correctness and allow others to more easily
-alter proofs as a result of changes made to the language.
-
-\subsubsection{Fusion}
-
-A number of systems incorporate both a functional language surface syntax with
-a translation to a more primitive process calculus. I shall focus on two such
-systems presented recently which have a connection to linear logic.
-
-Lolliproc~\cite{Mazurak:2010:LCC} provides a functional language with
-concurrency primitives as a surface syntax on top of an underlying process
-calculus. The process calculus provides a classical linear logic
-interpretation for the functional language, utilising double negation
-elimination to provide session type duality. The surface syntax is an
-effective abstraction layer for the programmer since it does not expose the
-underlying process calculus.
-
-Concurrency is provided by control operators which, respectively, spawn a
-child process and wait for a child process to complete. The spawning process
-creates a channel between parent and child process; a channel is represented
-as a continuation and the type denotes the communication protocol
-(i.e. session type) between parent and child.
-
-\citeauthor{Mazurak:2010:LCC} present proofs for type safety, strong
-normalisation and confluence. Confluence provides the guarantee that no race
-conditions can occur, strong normalisation prohibits non-terminating programs
-and type safety assures deadlock-freedom. Deadlock-freedom is implied by the
-acyclic communication graphs (progress of process reduction) which occur
-between parallel processes; follows from permitting only one channel between
-two halves of parallel composition construct in the process calculus. Most of
-the development is supported by Coq proof scripts with remaining issues due to
-reasoning about communication graphs. The development does not provide
-primitive send and receives; instead, these are encoded as linear functions.
-%% and classical extension to second-order logic by
-%% \citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}.
-
-\citeauthor{Wadler:2014}'s recent work~\cite{Wadler:2014} brings together a
-process calculus, CP, based on the earlier work by
-\citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
-\citeauthor{Gay:2010:LAST}. GV presents session types using the standard
-presentation as described in \S~\ref{sec:pre}. CP presents linear logical
-connectives with session type interpretations inspired by
-\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI}. \citeauthor{Wadler:2014}
-describes a continuation-passing style (CPS) translation from GV session types
-to CP classical linear logic propositions. The process calculus CP is slightly
-different to $\pi$DILL by \citeauthor{Caires:2010:STI}, apart from being
-classical in nature, the constructs for server replication are modified; CP
-provides a weakening and contraction rules whereas these are encoded in one
-rule for process reduction by \citeauthor{Caires:2010:STI}. Additionally,
-rather than a two-sided sequent calculus, \citeauthor{Wadler:2014} uses
-one-sided sequents for defining the process calculus, which provides a more
-intuitive presentation of session type duality and clearer correspondence to a
-$\lambda$-calculi surface syntax, simplifying the translation presentation. As
-presented, GV does not support all features of CP and thus the translation is
-one way (GV to CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP}
-extend GV to provide support for polymorphism and replication, following a
-similar CPS translation scheme. Lastly,
-\citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
-semantics for GV and their system has greater similarities with LAST than
-prior GV works, for example, by treating send and receive as linear functions
-rather than language primitives.
-
-\subsection{Mechanising Programming Language Metatheory}
-
-Interactive theorem proving and proof assistants have seen an upsurge in use
-in recent years. Especially within the domain of programming language
-research, there is a strong emphasis on providing mechanised proofs
-accompanying work in the area. Indeed, \citeauthor{Aydemir:2005:MMM} have
-published a series of challenges on mechanising programming languages aimed at
-providing a starting point for comparing different representation
-techniques~\cite{Aydemir:2005:MMM}. Another aim is to provide reusable
-libraries for common reasoning needed across programming language
-developments, e.g. handling of typing environments. This effort has resulted
-in a number of different approaches for representing programming languages
-within proof assistants. For instance,
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} present a representation
-of $\lambda$-calculi where bound variables are represented as \textit{de
-  Bruijn indices}~\footnote{A de Bruijn index is a number used to represent a
-  bound variable which indicates the position of its binder (starting from
-  zero for the innermost binder). For example, $\lambda~(\lambda~1)$
-  corresponds to $\lambda x. \lambda y. x$, i.e. the constant function.} and
-free variables are represented as named terms; the \textit{locally nameless}
-representation. Their work resulted in a library for handling commonly
-occurring aspects in programming language metatheory which has been used since
-in other developments~\cite{Park:2014:MMW}. Reusable libraries reduce the
-proof engineering effort needed for later works, in contrast to the
-pen-and-paper approach, and some proof assistants (as is the case with Coq)
-allow one to extract a program from the development which can provide a
-typechecker or compiler for the language formalised.~\footnote{Curry-Howard
-  strikes again; a proof of decidability of typechecking corresponds to a
-  program implementing a typechecker for the language.}
-
-Some recent mechanisation effort focuses on providing a basis for studying
-linearity within type systems. This work is of interest to the current project
-since most session-based type systems assume a strictly linear type system
-which requires re-binding of channel identifiers, such as in the functional
-setting of \citeauthor{Gay:2010:LAST}. \citeauthor{Mazurak:2010:LLT} present
-an extension to System F (termed \fpop) with a notion describing types as
-having linear or non-linear \textit{kind}~\cite{Mazurak:2010:LLT}. This
-extension is motivated by previous work on incorporating linearity into type
-systems. Overall, previous attempts either hamper the inclusion of desirable
-programming features (such as polymorphism), or do not adequately reflect
-non-linearity as the common case leading to awkward programming. \fpop avoids
-these deficiencies by categorising types into kinds whilst maintaining similar
-semantics to System F. Mechanised proofs for type soundness and parametricity
-are presented, and the semantics allow non-linear types to be treated as
-linear. Examples show the system can provide a wide range of permissions on
-interfaces, from full exclusive access to read-only shared references. The
-system is very close to the Alms language described by
-\citeauthor{Aldrich:2009} but has the advantage that the soundness results
-(type safety and parametricity) have been entirely mechanised in Coq. It would
-be interesting to see if one could encode the ``adopt'' and ``focus''
-constructs in this system to compare its expressiveness to Alms. The ``adopt''
-rule can be encoded similarly to the ShareRef
-example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to encode the
-``focus'' rule within the present semantics, as one needs to prevent
-non-linear access to its argument implying the presence of knowledge about
-aliasing within the typing judgements; absence of aliasing is not identified
-with linear kinded types in \fpop.
-
-While not related to programming language metatheory,
-\citeauthor{Gay:2001:FFP}~\cite{Gay:2001:FFP} provides a $\pi$-calculus
-framework mechanised in Isabelle/HOL. The aim is to build a mechanisaton of
-session-based type systems on top of the $\pi$-calculus framework. The use of
-de Bruijn indices for both bound and free variables causes issues with
-variable substitution often requiring permutation of typing environments (see
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} for a discussion on
-binder representations) and lifting of free variables during substitution; I
-avoid this issue in my representation choice (\S \ref{sec:approach}).
-
-The introduction of names for hypotheses during execution of custom tactics
-are an issue in the proof development. These names are automatically generated
-by the Isabelle/HOL system which creates a dependency between the names chosen
-and the tactic. Thus, one must be careful to use the exact names expected by
-the tactic. In the Coq proof assistant, defining custom tactics can be
-achieved using the Ltac language~\cite{Delahaye:2000:TLS}. Among other things,
-Ltac supports pattern matching on hypotheses and goal forms. Thus, if one
-wished to pattern match on an inductive type describing variables, one could
-do so without mentioning the actual variable name as follows:
-
-\begin{coq}
-Ltac my_tactic :=
-  match goal with
-  | [v: var |- _] => ... tactics here possibly mentioning `v' hypothesis ...
-  end.
-\end{coq}
-
-\subsubsection{Applications to Session-based Type Systems}\label{sec:asts}
-
-\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based
-type system providing session polymorphism. The system is more general than
-session polymorphism via subtyping~\cite{Gay:2005:SST} in that it is capable
-of typing a generic forwarding process between mutually dual session
-types. Defining such a process is not possible using subtyping since one
-wishes to restrict the types to be dual, not a subtype of some other
-type. This more general form of polymorphism is achieved by permitting input
-on all session types (including \lstinline{end}, the terminal session type)
-and then employing transition hypotheses to ensure the type permits such
-input. The authors mechanise their type system and its properties in the Coq
-proof assistant. However, the mechanised system is restricted to the
-$\pi$-calculus and the main focus is proving soundness properties for the
-deduction rules allowing polymorphism, rather than a more general basis for
-studying programming with session types.
-
-\section{Formalisation}
-
-The formalisation was performed using the Coq proof
-assisstant~\cite{Coq:manual}. In the spirit of the POPLMARK
-challenge~\cite{Aydemir:2005:MMM}, the development utilises available
-libraries and infrastructure as much as possible. In particular,
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a metatheory
-library for handling, among other things, association lists (environments) and
-free variable representation. The current development uses the version of the
-library distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which
-has been updated for Coq version 8.4pl4. One goal of the formalisation effort
-was to minimise the extent to which the development differs from the
-pen-and-paper presentation and avoid questions of adequacy of the
-representation. Some changes were made to ease the mechanisation and are
-highlighted where they occur.
-
-The locally nameless first-order representation for binders and free variables
-was adopted for both GV and CP, following
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. This first-order
-variant prevents variable-capture issues when compared with other first-order
-approaches. Alternatively, I could have chosen a higher-order abstract syntax
-(HOAS) approach such as the parametric version (PHOAS) described
-by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
-on the metalanguage (Coq, in this case) for handling binders and substitution
-so has less overhead in managing syntax than the locally nameless
-approach. However, it was less clear how to handle linearity using this
-approach and has the disadvantage of being able to represent terms not valid
-in the language being mechanised (so called ``exotic'' terms). Further, a
-higher-order approach would prevent the use of the Metatheory library which
-provides a number of useful tactics for manipulating typing contexts. Previous
-efforts to mechanise \fpop~\cite{Park:2014:MMW} illustrate how to handle a
-language containing both linear and non-linear contexts. However, to stay true
-to CP/GV, I maintained a single context. Interestingly, it became clear that
-the mechanised version of CP required explicit weakening and contraction to
-facilitate the reduction rules.
-
-For handling binder freshness, I chose to adopt the cofinite quantification
-approach described by~\cite{Aydemir:2008:EFM}, which excludes a finite set of
-variables from being considered as the binder. In contrast, the traditional
-``exists-fresh'' approach, where the binder is only required to be fresh
-within the abstraction's body, does not produce strong enough induction
-principles in some cases. This change will necessitate changes to the typing
-rules of GV and CP. Consider for instance the Connect rule in
-GV~\cite{Wadler:2014} (C) and its corresponding cofinite quantification
-version (CCO): \input{with-connect-rule}
-
-The conclusion specifies a variable in (C) but a type in (CCO) (de Bruijn
-index rather than a name). Thus, in the premise for (CCO) one can choose a
-suitably fresh $x$ in each premise. Note that, $L$ can be chosen such that
-$\Psi \cup \Phi \cup \fv(M) \cup \fv(N) \subseteq L$, where $\fv(M)$ are the
-set of free variables in term $M$, and $x$ can be the same in both
-premises. Similar changes are made to the rules for CP (see
-\S~\ref{sec:cp}). These changes do not alter the typing of processes nor the
-semantics of the translation from GV to CP. An equivalence between
-``exists-fresh'' and cofinite quantification definitions of the simply-typed
-$\lambda$-calculus is given by
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
-result could be derived straightforwardly (but tediously) for the CP/GV
-system.
-
-\begin{comment}
-\section{Issues}
-
-\begin{itemize}
-\item What is required of a Metatheory library for linear type systems?
-  Permutation reasoning? Weakening/Contraction for non-linear components?
-\item Discuss the different approaches that were taken to maintain close
-  correspondence with the paper presentation
-\item Make sure to note that the paper leaves details (a lot of the ``cruft'')
-  out of the proofs e.g. where weakening is applied in the typing derivations
-\item Mention the use of Metatheory and its limitations when mechanising
-  linear type systems like GV/CP
-\end{itemize}
-\end{comment}
-
-\subsection{The process calculus CP}\label{sec:cp}
-
-\begin{comment}
-FIGURES
-
-propositions
-
-processes
-
-typing judgements
-
-equivalences and axiomatic cut
-
-principal cut rules
-
-commuting conversions
-\end{comment}
-
-Formalising the process calculus CP provided many challenges for the
-metatheory library since to my knowledge, the library has never been applied
-to a process calculus before. So for instance, in the lambda calculus one
-typically does not work up to permutation of binders, whereas in a process
-calculus these kinds of structural equivalences are standard.
-
-Note that the rules for polymorphism, $\exists$ and $\forall$, are defined but
-cut elimination on them is not. Since GV (\S~\ref{sec:gv}) does not support
-polymorphism the reduction rules are unnecessary for the translation. Future
-work is to support reduction under these terms.
-
-\subsection{The functional language GV}\label{sec:gv}
-
-\begin{comment}
-FIGURES
-
-types
-
-terms
-
-typing judgements
-\end{comment}
-
-\subsection{CPS translation GV to CP}
-
-Currently the formalisation does not handle this. Future work.
-
-for an unbound index so this decision is sound. A consequence is that I must
-define a notion of opening language terms involving binders as described
-by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
-$t$ which is well-typed if $t$ is well-typed when its unbound index,
-representing the argument to the abstraction, is replaced by some fresh
-variable $x$, written $t^x$. Some terms in GV require multiple binders but I
-believe it will be straightforward to extend this notion. Unfortunately, I
-must define similar operations on CP terms and handle substitution and opening
-on them as well as propositions which can accept propositional variables. I
-hope to be able to generalise as much as possible with the help of the
-Metatheory library. say something about gmeta here and its automatic
-generation of these infrastructure lemmas
-
-While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
-may need to alter the approach slightly to avoid tricky formulations. For
-instance, instead of providing $n$-ary branch and choice in GV, I provide
-only binary versions of these operators to simplify the development. This
-change is not restrictive however, since CP itself provides only binary
-versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
-technical note, I amalgamate all types into one inductive definition and
-provide a predicate for restricting typing rules to consider only valid
-session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
-recursive definitions; session types are types which may contain types as
-subcomponents (arguments to send, for example). Unfortunately, handling
-mutually inductive definitions in Coq can be quite involved; requiring one to
-either rely on the Coq system to provide a stronger mutual induction principle
-or defining one manually. Such a definition is possible, but it complicates
-elimination via the induction principle.
-
-Another aspect of the encoding is how to define terms. The intrinsic encoding
-described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
-approach, indexing terms by their type so as to prevent ill-formed terms from
-being constructed. However, in the encoding for GV there are issues with using
-this encoding: $(1)$ it is not immediately clear how to handle linear contexts
-using the de Bruijn variable encoding, since in the intrinsic setting the
-environment supports weakening; $(2)$ an intrinsic encoding of terms would be
-complicated by the need to enforce that session types occur in certain
-instances (by use of a predicate); and $(3)$ well-typed terms require extra
-assumptions about binder freshness which cannot reasonably be expressed as a
-function type. For these reasons, an intrinsic approach does not offer much
-benefit since a well-typed term relation would still need to be
-defined. Therefore, I chose an extrinsic encoding for terms and define
-well-typed terms as a separate inductive type.
-
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
-their language proof infrastructure is proportional to the number of binding
-constructs. In the case of a simply typed lambda calculus this is not onerous,
-but GV has four binding constructs, CP has six and propositions have two. It
-may be true that this approach results in too many auxiliary definitions and
-operations not relevant to the actual properties one wishes to prove. However,
-I believe that once this infrastructure has been defined, the system can be
-extended without concern for these binding constructs.
-
-\citeauthor{Park:2014:MMW} describe a technique for
-removing nonlinear contexts from typing judgements. While their work extended
-to \fpop it is not clear how to handle an environment containing both
-non-linear and linear types as in GV. I wish to maintain as close a
-relationship as possible to the paper system presented by
-\citeauthor{Wadler:2014}, so separating out the non-linear and linear
-components (as in \fpop) is not an option at this stage.
-
-I describe the formalisation as it currently stands in this section. The use
-of the metatheory library from UPenn provided useful tactics and properties
-regarding freshness of free variables and environments.
-
-\subsection{Future Work}
+\section{Future Work}
 
 In addition to that already mentioned: there remains much to be done.
-
-\section{Conclusions}
-
-{\bf Acknowledgments.}
-This is optional; it is a location for you to thank people
-
-\printbibliography
-
-\end{document}

### 2015-04-02 paper/introduction.tex
Updated paper. Separated out sections and added CP figure.

--- 2015-02-24 paper/paper.tex
@@
-\documentclass{mpaper}
-
-\usepackage{macros}
-\usepackage{hyperref}
-\usepackage{comment}
-\usepackage{listings}
-\usepackage{lstcoq}
-\usepackage{graphicx}
-\usepackage{xspace}
-\usepackage{cmll}
-\usepackage{prooftree}
-\usepackage{amsmath}
-\newcommand{\typestate}[2]{{#1}\gg{#2}}
-\usepackage[style=alphabetic,sorting=nymt,natbib=true,
-  maxbibnames=100]{biblatex}
-\addbibresource{paper.bib}
-
-\DeclareSortingScheme{nymt}{
-  \sort{
-    \field{author}
-    \field{year}
-    \field{month}
-    \field{title}
-  }
-}
-
-\input{propmacros}
-
-\newcommand{\fpop}{System F${}^\circ$\xspace}
-
-\begin{document}
-
-\title{Propositions as Sessions, Mechanically}
-\author{Craig McLaughlin}
-\matricnum{1002524}
-
-\maketitle
-
-\begin{abstract}
-\begin{comment}
-SPL (+CPL)
-
-WTD
-
-ROF
-
-IMP
-\end{comment}
-\end{abstract}
-
 \section{Introduction}\label{sec:intro}
@@
 classical linear logic (CP) and a functional language with session types (GV).
-
-\section{Related Work}\label{sec:rw}
-
-The background for this project can be categorised roughly as previous work on
-session-based type systems (sub-divided into $\pi$ and $\lambda$-calculi
-variants, \S \ref{sec:pis} and \S \ref{sec:lam}, respectively) and previous
-work on formal verification using proof assistants (particularly focussed on
-programming language mechanisation). Finally, some recent work has merged
-these two disciplines by verifying a session-based $\pi$-calculus system (\S
-\ref{sec:asts}).
-
-\subsection{Session-based Type Systems and Linear Logic}\label{sec:sts}
-
-\subsubsection{Preamble}\label{sec:pre}
-
-The definition of session types will be important for the remainder:
-\input{../paper/session-types.tex}
-
-Here, I use a modified version of the definition given by
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} where I have omitted recursive
-session types. The type \lstinline{end} denotes the terminal session type and
-a channel typed as such may not perform any communication. A communication
-channel with session type $\inpt{T}{S}$ expects to receive a term of type $T$
-and then continues with session type $S$. Conversely, a channel with type
-$\outpt{T}{S}$ can send a term of type $T$ and then has type $S$. Thus, input
-and output are considered \emph{dual} session types, in the sense that a
-channel with one endpoint having type $\inpt{T}{S}$ and the other endpoint
-having type $\outpt{T}{\Sbar{S}}$, where $\Sbar{S}$ is the dual of $S$,
-permits communication between two agents, allowing one to send a term of type
-$T$ and the other to receive it with continuing endpoint types $\Sbar{S}$ and
-$S$, respectively. Similarly, the remaining types $\Branch{l_i}{S_i}{i\in I}$
-and $\Choice{l_i}{S_i}{i\in I}$ are duals of each other. A channel of type
-$\Branch{l_i}{S_i}{i\in I}$ can receive a label $l_i$ which will select the
-corresponding session type $S_i$ as the continuing type of the channel. While
-a channel with type $\Choice{l_i}{S_i}{i\in I}$ can send a label $l_i$ and
-continues as $S_i$.
-
-\subsubsection{In the \texorpdfstring{$\pi$}{pi}-calculus}\label{sec:pis}
-
-\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} present a connection
-between a $\pi$-calculus type system with session types ($\pi$DILL) and dual
-intuitionistic linear logic (DILL)~\cite{Barber:1996}. They equate linear
-propositions with session types of the process calculus via a bidirectional
-translation relation. They use a binary reduction relation and a labelled
-transition system to represent process reduction and process action with its
-environment as is standard~\cite{Sangiorgi:2001}.
-
-\input{../paper/sill-branch-choice-rules}
-
-They represent dual intuitionistic linear logic as a sequent calculus
-providing logical connectives which have session-based interpretations in
-$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL in
-Figure~\ref{fig:sill-branching}, which have only two alternatives
-(cf. multiple alternatives in the standard definition in \S~\ref{sec:pre}).
-
-Note that branch and choice both have some notion of branching (R rules for
-$\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
-a session-based setting, since the intuition for \& is that of
-\emph{offering} (receiving) a choice between alternatives rather than
-\emph{making} (sending) the choice. Here, however, the intuition should be
-extended to account for its location in the typing judgement. The channel is
-on the left of the sequent, and \citeauthor{Caires:2010:STI} suggest one
-consider the context as services being provided to the typed process. Thus,
-the type of channels in the context is really the type of the other endpoint
-of the channel, and the typed process behaves according to the dual
-protocol. For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to
-represent the dual of a protocol $A$ as defined by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process is behaving
-according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from
-alternatives (in this case, $\inrterm$) and then behaving according to the
-selected protocol (in this case, protocol $B^\bot$).
-
-As noted by Wadler~\cite{Wadler:2014} there is a certain amount of redundancy
-in the two-sided nature of the rules forcing one to commit to a certain
-convention when typing a communication system. Consider, an equivalent system
-to the buy/quote example given by
-\citeauthor{Caires:2010:STI}~\cite[\S~3]{Caires:2010:STI}:
-\begin{gather*}
-\sdef{ClientProto}{\BChoice{(\tensor{Book}
-                            {(\tensor{Card}{(\lfn{Receipt}{\1})})})}
-                           {(\tensor{Book}{(\lfn{Quote}{\1})})}}
-\end{gather*}
-
-Note, I am using $\tensor{}{}$ for session output and $\lfn{}{}$ for session
-input following the presentation by \citeauthor{Caires:2010:STI}. In the
-protocol defined above, a client selects between buying a book or receiving a
-price quote for a book. In the buy option case, the client sends the name of
-the book to the server and their card details, then the server sends the
-receipt for the purchase. In the quote option case, the client sends the name
-of the book and receives a price quote from the server. I have replaced the
-types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more meaningful
-type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate; all are
-synonyms for the type $\1$ which denotes the terminal session type in this
-system. The client and server process bodies are as follows:
-\begin{gather*}
-\sdef{ClientBody_c}
-     {\CPinr{c}{\SLoutput{1984}{c}{\SLinput{c}{quote}{\0}}}}\\
-\sdef{ServerBody_c}
-     {\CPcase{c}
-             {\SLinput{c}{book}{\SLinput{c}{card}{\SLoutput{receipt}{c}{\0}}}}
-             {\SLinput{c}{book}{\SLoutput{quote}{c}{\0}}}}
-\end{gather*}
-
-\begin{samepage}
-The typings for the client and server processes are given by:
-\begin{gather*}
-\tsequent{\cdot}{\cdot}{ClientBody_c}{\typedid{c}{ClientProto}}
-\tag{\mkTrule{ClientBody}} \label{eq:cbtyp}
-\\ \tsequent{\cdot}{\typedid{c}{ClientProto}}{ServerBody_c}{\typedid{-}{\1}}
-\tag{\mkTrule{ServerBody}} \label{eq:sbtyp}
-\end{gather*}
-\end{samepage}
-
-The derivation for \eqref{eq:cbtyp} is $T\1 R$, $\lfn{T}{R}$, $\tensor{T}{R}$,
-$\BChoice{T}{R_2}$ and \eqref{eq:sbtyp} is derivable from
-$\BChoice{T}{L}$. Except for the final rule, these are exact dual rules of the
-rules used in example by \citeauthor{Caires:2010:STI}. Indeed, consider the
-composite process:
-\begin{gather*}
-\sdef{QSimple}{\Spar{c}{ClientBody_c}{ServerBody_c}}\\
-\tsequent{\cdot}{\cdot}{QSimple}{\typedid{-}{\1}}
-\end{gather*}
-
-The final system is near identical to the one produced in the example given by
-\citeauthor{Caires:2010:STI} except the individual processes within the
-parallel composition have been swapped. Here the client communicates along the
-channel $c$ on the right-side of the sequent and the server utilises the
-channel $c$ on the left of the sequent. This typing, while correct, violates
-the intuition given by \citeauthor{Caires:2010:STI} of the context providing
-the services used by the process, since one would usually associate provision
-of services with the server not the client. Instead, the example above
-suggests the server utilises some channel provided by the client. On the other
-hand, one could imagine such a channel being setup between two communicating
-agents; the ``owner'' of the channel is largely irrelevant. Note that
-$ClientProto$ is in fact the dual of the $ServerProto$ definition by
-\citeauthor{Caires:2010:STI} using the definition of dual used in Wadler's
-classical process calculus~\cite{Wadler:2014}. In $\pi$DILL, the duality
-between endpoints of the same channel are obscured by the nature of the typing
-judgements.
-
-Process reduction steps are equated with proof reduction via cut
-elimination. The relationship with proof normalisation in DILL provides
-absence of deadlocks in session-based communication for the process
-calculus. The correspondence of session types to linear logic propositions is
-an important step in providing a solid logical foundation for a concurrent
-type theory; in the same way ``sequential'' type theory ($\lambda$-calculi)
-profitted from a logical perspective. However, for session types to become a
-more usable tool for programming distributed communication systems (the
-ultimate goal, of course, is to write more correct programs), one cannot be
-expected to program directly in the $\pi$-calculus; a more mainstream
-programming paradigm must be built upon these solid logical foundations. With
-that in mind, enter the $\lambda$-calculus!
-
-\subsubsection{In the \texorpdfstring{$\lambda$}{lambda}-calculus}
-\label{sec:lam}
-
-A series of works by \citeauthor{Gay:2003:STI}~\cite{Gay:2003:STI} and
-\citeauthor{Vasconcelos:2006:TCM}~\cite{Vasconcelos:2006:TCM} provide a
-$\lambda$-calculus based language with session types for channel
-primitives. The type system is nonstandard, featuring an additional
-environment for tracking the session types of channels; these types are
-modified during term reduction to reflect communication. The extra channel
-environment is provided to support possible aliasing of channels by treating
-channel variables as pointers or references into the channel environment. So,
-for a channel $c$ of session type $S$, a variable $x$ referring to this
-channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the
-channel environment. The typing judgements and term reduction relation must be
-augmented to handle these environments. Unfortunately, a side-effect of this
-is that function types must specify the pre and post-states of channel
-environments and therefore, the type of arguments cannot work up to
-alpha-equivalence of channel identifiers. That is, if two channels $c$ and $d$
-have the same session type, one cannot define a function to operate on
-variables of either channel type since the function definition will explicitly
-mention the identifier for the channel (in this case either, $Chan~c$ or
-$Chan~d$). However, the system does provide an attractive non-linear channel
-type which has been neglected in favour of a linear treatment of channels in
-more recent
-works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2014}. Certainly, it is the
-case that an aliased approach would fit better with mainstream programming
-languages and perhaps studying a similar system could offer some insights.
-
-The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of
-aliased access to linear resources. The type system requires special language
-features, e.g. to manage capabilities which can be thought as a form of object
-liveness mechanism, so GV typing judgements would need to be extended such
-that the capabilities can be included in soundness results. An extension to
-the translation to CP to incorporate the ``adopt'' and ``focus'' constructs
-would also need to be defined. It is not immediately clear how these features
-would assist in the management of session typed channels. The ``focus'' rule
-(and linearity generally) enforces an invariant on the focussed object
-(channel) such that the type is maintained during the focus operation:
-\input{../proposal/focus-rule}
-
-The typing judgements include \emph{capabilities} that track aliasing to heap
-objects, e.g. $\AFcap{\rho}{h}$ denotes a mapping from a static name $\rho$ to
-a heap type $h$, and a variable $x$ may be typed as $\typedid{x}{\AFtr{\rho}}$
-indicating that $x$ refers to an object of type $h$. The purpose of the
-capabilities is to enforce restrictions on when an object can be
-referenced. In other words, if $\AFcap{\rho}{h}$ is in the capabilities at
-some program point then $x$ may be used, otherwise such use is prohibited. In
-the [focus] rule, evaluating $e_1$ retrieves the object to be focussed. Note
-the type of $e_1$ is a ``guarded'' type. The guard $G$ is the static name of
-some object which ``adopted''~\footnote{I omit the adopt typing rule for
-  brevity.} an object of heap type $h$. If a program point has
-$\AFcap{G}{h_1}$ in the capabilities for some $h_1$ then all objects guarded
-by $G$ may be used; $\capsequent{\Delta}{C_1}{G}$ denotes containment of a
-such a capability in $C_1$. The guarded type may be thought of as policing
-manipulation of linear components within $h$. Treating $\otimes$ as the
-disjoint union for capability sets, the typing for $e_2$ temporarily removes
-the guard $G$ from the set of capabilities, and adds a new variable $x$, and
-associated capability, with a type permitting access to the object of type $h$
-i.e. a non-guarded type. Removing $G$ will prohibit access to allow objects
-guarded by $G$ for the during of $e_2$. The final capabilities after executing
-$e_2$ are required to be $\tensor{C_3}{\AFcap{\rho}{h}}$, enforcing that the
-type must remain as $h$. The typing of the $\letterm$ expression ensures that
-the guard $G$ (contained in $C_1$) is restored.
-
-Such a rule would appear to only benefit recursive session types. Indeed,
-focussing on a non-recursive channel would be pointless since one would be
-prevented from communicating across it since the focus rule requires
-maintenance of the channel type. Further, it would be undesirable to permit
-modification of a session type without guaranteeing exclusive access to the
-channel since this could lead to invalid communication e.g. trying to
-communicate on a closed channel. Additionally, a linear object is consumed
-upon being adopted and will be deallocated when its adopter's lifetime ends,
-although an alternative semantics could be defined which returned the linear
-resource to the context. Moreover, \citeauthor{Fahndrich:2002} do not provide
-operational semantics or soundness results for the ``adoption and focus'' type
-system.
-
-The Alms language~\cite{Aldrich:2009} provides a functional language with
-affine types based on linear logic. A notion of type \textit{kind} is used to
-distinguish between affine and unrestricted values. The typing judgements use
-two contexts; one for variables with unrestricted kind and another for
-variables of affine type and type variables. Functions are also labelled with
-a kind which may depend on their arguments e.g. a multiple argument function
-must have affine kinding if it accepts an affine typed resource as an argument
-other than the final argument. The reason for this is simply to prevent the
-function from using an affine resource more than once. Additionally, functions
-act as closures over their environments so their qualifier may be determined
-by a free variable in the environment, and a relation is provided to calculate
-the appropriate kind qualifier. Type variables are provided to support kind
-polymorphism of function qualifiers. The system is expressive enough to
-implement session types as a module using a similar approach to
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} although treating channels as
-affine rather than linearly typed. \citeauthor{Aldrich:2009} note that Alms'
-type system can support the ``adoption and focus'' system but do not specify
-details or whether any changes are necessary to their typing judgements.
-
-The work on a linear functional language with session types by
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the
-current project from the body of work on session-based $\lambda$-calculus type
-systems. They provide paper proofs of their system's type safety. Given the
-non-trivial features of the type system such as recursive session types and
-buffered channels for supporting asynchronous communication it will be
-challenging to extend these proofs manually. A mechanised version would
-provide stronger guarantees of correctness and allow others to more easily
-alter proofs as a result of changes made to the language.
-
-\subsubsection{Fusion}
-
-A number of systems incorporate both a functional language surface syntax with
-a translation to a more primitive process calculus. I shall focus on two such
-systems presented recently which have a connection to linear logic.
-
-Lolliproc~\cite{Mazurak:2010:LCC} provides a functional language with
-concurrency primitives as a surface syntax on top of an underlying process
-calculus. The process calculus provides a classical linear logic
-interpretation for the functional language, utilising double negation
-elimination to provide session type duality. The surface syntax is an
-effective abstraction layer for the programmer since it does not expose the
-underlying process calculus.
-
-Concurrency is provided by control operators which, respectively, spawn a
-child process and wait for a child process to complete. The spawning process
-creates a channel between parent and child process; a channel is represented
-as a continuation and the type denotes the communication protocol
-(i.e. session type) between parent and child.
-
-\citeauthor{Mazurak:2010:LCC} present proofs for type safety, strong
-normalisation and confluence. Confluence provides the guarantee that no race
-conditions can occur, strong normalisation prohibits non-terminating programs
-and type safety assures deadlock-freedom. Deadlock-freedom is implied by the
-acyclic communication graphs (progress of process reduction) which occur
-between parallel processes; follows from permitting only one channel between
-two halves of parallel composition construct in the process calculus. Most of
-the development is supported by Coq proof scripts with remaining issues due to
-reasoning about communication graphs. The development does not provide
-primitive send and receives; instead, these are encoded as linear functions.
-%% and classical extension to second-order logic by
-%% \citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}.
-
-\citeauthor{Wadler:2014}'s recent work~\cite{Wadler:2014} brings together a
-process calculus, CP, based on the earlier work by
-\citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
-\citeauthor{Gay:2010:LAST}. GV presents session types using the standard
-presentation as described in \S~\ref{sec:pre}. CP presents linear logical
-connectives with session type interpretations inspired by
-\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI}. \citeauthor{Wadler:2014}
-describes a continuation-passing style (CPS) translation from GV session types
-to CP classical linear logic propositions. The process calculus CP is slightly
-different to $\pi$DILL by \citeauthor{Caires:2010:STI}, apart from being
-classical in nature, the constructs for server replication are modified; CP
-provides a weakening and contraction rules whereas these are encoded in one
-rule for process reduction by \citeauthor{Caires:2010:STI}. Additionally,
-rather than a two-sided sequent calculus, \citeauthor{Wadler:2014} uses
-one-sided sequents for defining the process calculus, which provides a more
-intuitive presentation of session type duality and clearer correspondence to a
-$\lambda$-calculi surface syntax, simplifying the translation presentation. As
-presented, GV does not support all features of CP and thus the translation is
-one way (GV to CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP}
-extend GV to provide support for polymorphism and replication, following a
-similar CPS translation scheme. Lastly,
-\citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
-semantics for GV and their system has greater similarities with LAST than
-prior GV works, for example, by treating send and receive as linear functions
-rather than language primitives.
-
-\subsection{Mechanising Programming Language Metatheory}
-
-Interactive theorem proving and proof assistants have seen an upsurge in use
-in recent years. Especially within the domain of programming language
-research, there is a strong emphasis on providing mechanised proofs
-accompanying work in the area. Indeed, \citeauthor{Aydemir:2005:MMM} have
-published a series of challenges on mechanising programming languages aimed at
-providing a starting point for comparing different representation
-techniques~\cite{Aydemir:2005:MMM}. Another aim is to provide reusable
-libraries for common reasoning needed across programming language
-developments, e.g. handling of typing environments. This effort has resulted
-in a number of different approaches for representing programming languages
-within proof assistants. For instance,
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} present a representation
-of $\lambda$-calculi where bound variables are represented as \textit{de
-  Bruijn indices}~\footnote{A de Bruijn index is a number used to represent a
-  bound variable which indicates the position of its binder (starting from
-  zero for the innermost binder). For example, $\lambda~(\lambda~1)$
-  corresponds to $\lambda x. \lambda y. x$, i.e. the constant function.} and
-free variables are represented as named terms; the \textit{locally nameless}
-representation. Their work resulted in a library for handling commonly
-occurring aspects in programming language metatheory which has been used since
-in other developments~\cite{Park:2014:MMW}. Reusable libraries reduce the
-proof engineering effort needed for later works, in contrast to the
-pen-and-paper approach, and some proof assistants (as is the case with Coq)
-allow one to extract a program from the development which can provide a
-typechecker or compiler for the language formalised.~\footnote{Curry-Howard
-  strikes again; a proof of decidability of typechecking corresponds to a
-  program implementing a typechecker for the language.}
-
-Some recent mechanisation effort focuses on providing a basis for studying
-linearity within type systems. This work is of interest to the current project
-since most session-based type systems assume a strictly linear type system
-which requires re-binding of channel identifiers, such as in the functional
-setting of \citeauthor{Gay:2010:LAST}. \citeauthor{Mazurak:2010:LLT} present
-an extension to System F (termed \fpop) with a notion describing types as
-having linear or non-linear \textit{kind}~\cite{Mazurak:2010:LLT}. This
-extension is motivated by previous work on incorporating linearity into type
-systems. Overall, previous attempts either hamper the inclusion of desirable
-programming features (such as polymorphism), or do not adequately reflect
-non-linearity as the common case leading to awkward programming. \fpop avoids
-these deficiencies by categorising types into kinds whilst maintaining similar
-semantics to System F. Mechanised proofs for type soundness and parametricity
-are presented, and the semantics allow non-linear types to be treated as
-linear. Examples show the system can provide a wide range of permissions on
-interfaces, from full exclusive access to read-only shared references. The
-system is very close to the Alms language described by
-\citeauthor{Aldrich:2009} but has the advantage that the soundness results
-(type safety and parametricity) have been entirely mechanised in Coq. It would
-be interesting to see if one could encode the ``adopt'' and ``focus''
-constructs in this system to compare its expressiveness to Alms. The ``adopt''
-rule can be encoded similarly to the ShareRef
-example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to encode the
-``focus'' rule within the present semantics, as one needs to prevent
-non-linear access to its argument implying the presence of knowledge about
-aliasing within the typing judgements; absence of aliasing is not identified
-with linear kinded types in \fpop.
-
-While not related to programming language metatheory,
-\citeauthor{Gay:2001:FFP}~\cite{Gay:2001:FFP} provides a $\pi$-calculus
-framework mechanised in Isabelle/HOL. The aim is to build a mechanisaton of
-session-based type systems on top of the $\pi$-calculus framework. The use of
-de Bruijn indices for both bound and free variables causes issues with
-variable substitution often requiring permutation of typing environments (see
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} for a discussion on
-binder representations) and lifting of free variables during substitution; I
-avoid this issue in my representation choice (\S \ref{sec:approach}).
-
-The introduction of names for hypotheses during execution of custom tactics
-are an issue in the proof development. These names are automatically generated
-by the Isabelle/HOL system which creates a dependency between the names chosen
-and the tactic. Thus, one must be careful to use the exact names expected by
-the tactic. In the Coq proof assistant, defining custom tactics can be
-achieved using the Ltac language~\cite{Delahaye:2000:TLS}. Among other things,
-Ltac supports pattern matching on hypotheses and goal forms. Thus, if one
-wished to pattern match on an inductive type describing variables, one could
-do so without mentioning the actual variable name as follows:
-
-\begin{coq}
-Ltac my_tactic :=
-  match goal with
-  | [v: var |- _] => ... tactics here possibly mentioning `v' hypothesis ...
-  end.
-\end{coq}
-
-\subsubsection{Applications to Session-based Type Systems}\label{sec:asts}
-
-\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based
-type system providing session polymorphism. The system is more general than
-session polymorphism via subtyping~\cite{Gay:2005:SST} in that it is capable
-of typing a generic forwarding process between mutually dual session
-types. Defining such a process is not possible using subtyping since one
-wishes to restrict the types to be dual, not a subtype of some other
-type. This more general form of polymorphism is achieved by permitting input
-on all session types (including \lstinline{end}, the terminal session type)
-and then employing transition hypotheses to ensure the type permits such
-input. The authors mechanise their type system and its properties in the Coq
-proof assistant. However, the mechanised system is restricted to the
-$\pi$-calculus and the main focus is proving soundness properties for the
-deduction rules allowing polymorphism, rather than a more general basis for
-studying programming with session types.
-
-\section{Formalisation}
-
-The formalisation was performed using the Coq proof
-assisstant~\cite{Coq:manual}. In the spirit of the POPLMARK
-challenge~\cite{Aydemir:2005:MMM}, the development utilises available
-libraries and infrastructure as much as possible. In particular,
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a metatheory
-library for handling, among other things, association lists (environments) and
-free variable representation. The current development uses the version of the
-library distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which
-has been updated for Coq version 8.4pl4. One goal of the formalisation effort
-was to minimise the extent to which the development differs from the
-pen-and-paper presentation and avoid questions of adequacy of the
-representation. Some changes were made to ease the mechanisation and are
-highlighted where they occur.
-
-The locally nameless first-order representation for binders and free variables
-was adopted for both GV and CP, following
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. This first-order
-variant prevents variable-capture issues when compared with other first-order
-approaches. Alternatively, I could have chosen a higher-order abstract syntax
-(HOAS) approach such as the parametric version (PHOAS) described
-by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
-on the metalanguage (Coq, in this case) for handling binders and substitution
-so has less overhead in managing syntax than the locally nameless
-approach. However, it was less clear how to handle linearity using this
-approach and has the disadvantage of being able to represent terms not valid
-in the language being mechanised (so called ``exotic'' terms). Further, a
-higher-order approach would prevent the use of the Metatheory library which
-provides a number of useful tactics for manipulating typing contexts. Previous
-efforts to mechanise \fpop~\cite{Park:2014:MMW} illustrate how to handle a
-language containing both linear and non-linear contexts. However, to stay true
-to CP/GV, I maintained a single context. Interestingly, it became clear that
-the mechanised version of CP required explicit weakening and contraction to
-facilitate the reduction rules.
-
-For handling binder freshness, I chose to adopt the cofinite quantification
-approach described by~\cite{Aydemir:2008:EFM}, which excludes a finite set of
-variables from being considered as the binder. In contrast, the traditional
-``exists-fresh'' approach, where the binder is only required to be fresh
-within the abstraction's body, does not produce strong enough induction
-principles in some cases. This change will necessitate changes to the typing
-rules of GV and CP. Consider for instance the Connect rule in
-GV~\cite{Wadler:2014} (C) and its corresponding cofinite quantification
-version (CCO): \input{with-connect-rule}
-
-The conclusion specifies a variable in (C) but a type in (CCO) (de Bruijn
-index rather than a name). Thus, in the premise for (CCO) one can choose a
-suitably fresh $x$ in each premise. Note that, $L$ can be chosen such that
-$\Psi \cup \Phi \cup \fv(M) \cup \fv(N) \subseteq L$, where $\fv(M)$ are the
-set of free variables in term $M$, and $x$ can be the same in both
-premises. Similar changes are made to the rules for CP (see
-\S~\ref{sec:cp}). These changes do not alter the typing of processes nor the
-semantics of the translation from GV to CP. An equivalence between
-``exists-fresh'' and cofinite quantification definitions of the simply-typed
-$\lambda$-calculus is given by
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
-result could be derived straightforwardly (but tediously) for the CP/GV
-system.
-
-\begin{comment}
-\section{Issues}
-
-\begin{itemize}
-\item What is required of a Metatheory library for linear type systems?
-  Permutation reasoning? Weakening/Contraction for non-linear components?
-\item Discuss the different approaches that were taken to maintain close
-  correspondence with the paper presentation
-\item Make sure to note that the paper leaves details (a lot of the ``cruft'')
-  out of the proofs e.g. where weakening is applied in the typing derivations
-\item Mention the use of Metatheory and its limitations when mechanising
-  linear type systems like GV/CP
-\end{itemize}
-\end{comment}
-
-\subsection{The process calculus CP}\label{sec:cp}
-
-\begin{comment}
-FIGURES
-
-propositions
-
-processes
-
-typing judgements
-
-equivalences and axiomatic cut
-
-principal cut rules
-
-commuting conversions
-\end{comment}
-
-Formalising the process calculus CP provided many challenges for the
-metatheory library since to my knowledge, the library has never been applied
-to a process calculus before. So for instance, in the lambda calculus one
-typically does not work up to permutation of binders, whereas in a process
-calculus these kinds of structural equivalences are standard.
-
-Note that the rules for polymorphism, $\exists$ and $\forall$, are defined but
-cut elimination on them is not. Since GV (\S~\ref{sec:gv}) does not support
-polymorphism the reduction rules are unnecessary for the translation. Future
-work is to support reduction under these terms.
-
-\subsection{The functional language GV}\label{sec:gv}
-
-\begin{comment}
-FIGURES
-
-types
-
-terms
-
-typing judgements
-\end{comment}
-
-\subsection{CPS translation GV to CP}
-
-Currently the formalisation does not handle this. Future work.
-
-for an unbound index so this decision is sound. A consequence is that I must
-define a notion of opening language terms involving binders as described
-by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
-$t$ which is well-typed if $t$ is well-typed when its unbound index,
-representing the argument to the abstraction, is replaced by some fresh
-variable $x$, written $t^x$. Some terms in GV require multiple binders but I
-believe it will be straightforward to extend this notion. Unfortunately, I
-must define similar operations on CP terms and handle substitution and opening
-on them as well as propositions which can accept propositional variables. I
-hope to be able to generalise as much as possible with the help of the
-Metatheory library. say something about gmeta here and its automatic
-generation of these infrastructure lemmas
-
-While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
-may need to alter the approach slightly to avoid tricky formulations. For
-instance, instead of providing $n$-ary branch and choice in GV, I provide
-only binary versions of these operators to simplify the development. This
-change is not restrictive however, since CP itself provides only binary
-versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
-technical note, I amalgamate all types into one inductive definition and
-provide a predicate for restricting typing rules to consider only valid
-session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
-recursive definitions; session types are types which may contain types as
-subcomponents (arguments to send, for example). Unfortunately, handling
-mutually inductive definitions in Coq can be quite involved; requiring one to
-either rely on the Coq system to provide a stronger mutual induction principle
-or defining one manually. Such a definition is possible, but it complicates
-elimination via the induction principle.
-
-Another aspect of the encoding is how to define terms. The intrinsic encoding
-described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
-approach, indexing terms by their type so as to prevent ill-formed terms from
-being constructed. However, in the encoding for GV there are issues with using
-this encoding: $(1)$ it is not immediately clear how to handle linear contexts
-using the de Bruijn variable encoding, since in the intrinsic setting the
-environment supports weakening; $(2)$ an intrinsic encoding of terms would be
-complicated by the need to enforce that session types occur in certain
-instances (by use of a predicate); and $(3)$ well-typed terms require extra
-assumptions about binder freshness which cannot reasonably be expressed as a
-function type. For these reasons, an intrinsic approach does not offer much
-benefit since a well-typed term relation would still need to be
-defined. Therefore, I chose an extrinsic encoding for terms and define
-well-typed terms as a separate inductive type.
-
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
-their language proof infrastructure is proportional to the number of binding
-constructs. In the case of a simply typed lambda calculus this is not onerous,
-but GV has four binding constructs, CP has six and propositions have two. It
-may be true that this approach results in too many auxiliary definitions and
-operations not relevant to the actual properties one wishes to prove. However,
-I believe that once this infrastructure has been defined, the system can be
-extended without concern for these binding constructs.
-
-\citeauthor{Park:2014:MMW} describe a technique for
-removing nonlinear contexts from typing judgements. While their work extended
-to \fpop it is not clear how to handle an environment containing both
-non-linear and linear types as in GV. I wish to maintain as close a
-relationship as possible to the paper system presented by
-\citeauthor{Wadler:2014}, so separating out the non-linear and linear
-components (as in \fpop) is not an option at this stage.
-
-I describe the formalisation as it currently stands in this section. The use
-of the metatheory library from UPenn provided useful tactics and properties
-regarding freshness of free variables and environments.
-
-\subsection{Future Work}
-
-In addition to that already mentioned: there remains much to be done.
-
-\section{Conclusions}
-
-{\bf Acknowledgments.}
-This is optional; it is a location for you to thank people
-
-\printbibliography
-
-\end{document}

### 2015-04-02 paper/paper.tex
Updated paper. Separated out sections and added CP figure.

--- 2015-02-24 paper/paper.tex
@@
 
-\usepackage{macros}
-\usepackage{hyperref}
 \usepackage{comment}
-\usepackage{listings}
-\usepackage{lstcoq}
+\usepackage{lstform}
 \usepackage{graphicx}
@@
 
+%\input{cpgv}
 \input{propmacros}
@@
 
-\begin{abstract}
-\begin{comment}
-SPL (+CPL)
+\input{abstract}
+\input{introduction}
+\input{background}
+\input{formalisation}
+\input{futurework}
+\input{conclusion}
+\input{acknowledgements}
 
-WTD
-
-ROF
-
-IMP
-\end{comment}
-\end{abstract}
-
-\section{Introduction}\label{sec:intro}
-
-Distributed communication systems are becoming increasingly prevalent in
-modern society. To facilitate communication between concurrently executing
-agents requires complex network protocols to define the type and sequencing of
-messages. Typically these protocols are implemented in systems programming
-languages, such as C, which provide no correctness guarantees of the
-implementation. Furthermore, programming in these languages is often extremely
-error-prone, as one has a myriad of unsafe features at one's disposal. As more
-businesses begin to conduct transactions over the Web, it is imperative that
-they can reason about the communication protocols upon which they rely.
-
-Encoding communication protocols as \textit{session types} is one approach to
-achieve guarantees about correctness. One particular variant of session types
-are \textit{binary} session types~\footnote{There are also \textit{multiparty}
- session types, but I do not discuss them here.}. A binary session type
-specifies communication between two agents, and it specifies the order and
-type of messages to be sent or received by an agent in the system. The two
-agents have dual session types such that, for example, when one is expecting
-to receive a message the other is expecting to send a message; both session
-types agree on the type of message to be exchanged. In this way, correctness
-of communication can be statically assured by providing communicating agents
-with a \textit{channel} whose endpoints have dual session types. A channel is
-defined as a bi-directional conduit for exchanging messages, not unlike a
-socket in network programming. In contrast to sockets, however, a channel is
-associated with an explicit type which provides a description of its
-protocol. Thus, in session-based systems a channel is usually provided as a
-primitive type.
-
-In programming language research, work on static type systems is usually
-presented on paper complete with informal proofs of correctness for the system
-properties. Most types systems in the session types community have been
-developed in this way. However, it is more difficult to extend work using this
-approach, as it may be unclear how changes affect proofs and one will
-typically have to re-consider all proofs to be convinced the properties still
-hold. More generally, unassisted formalisms are error-prone to develop and do
-not provide a firm basis upon which other researchers can experiment.
-
-Proof assistants such as Coq~\cite{Coq:manual} offer a platform on which to
-develop programming languages and type systems and associated proofs of
-properties with strong guarantees of correctness. Not only is proving with a
-proof assistant much less error-prone than the unaided approach, it also eases
-extension of the system since the proof assistant will highlight parts of
-proofs that have been affected by changes. On the other hand, one may have to
-prove more properties using this approach since this setting does not permit
-appealing to human intuition. For example, alpha-equivalence and freshness of
-binders are usually assumed in informal presentations, whereas in the
-mechanised setting one has to explicitly encode and reason about these
-properties~\cite{Aydemir:2005:MMM, Aydemir:2008:EFM}.
-
-This project describes work towards the development of a framework for
-formalising lambda calculus type systems with session types using interactive
-theorem proving (in particular, the system Coq). The framework should enable a
-whole class of related languages and type systems to be formalised with little
-extra proof effort. Inspired by previous work on a mechanised framework for
-$\pi$-calculus type systems~\cite{Gay:2001:FFP}, the current work focuses on
-the lambda calculus to provide a basis on which session types for mainstream
-programming languages can be studied, whilst taking advantage of previous
-proof effort and specialising only for those parts unique to the particular
-system. The starting point for the framework is the calculi developed by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
-classical linear logic (CP) and a functional language with session types (GV).
-
-\section{Related Work}\label{sec:rw}
-
-The background for this project can be categorised roughly as previous work on
-session-based type systems (sub-divided into $\pi$ and $\lambda$-calculi
-variants, \S \ref{sec:pis} and \S \ref{sec:lam}, respectively) and previous
-work on formal verification using proof assistants (particularly focussed on
-programming language mechanisation). Finally, some recent work has merged
-these two disciplines by verifying a session-based $\pi$-calculus system (\S
-\ref{sec:asts}).
-
-\subsection{Session-based Type Systems and Linear Logic}\label{sec:sts}
-
-\subsubsection{Preamble}\label{sec:pre}
-
-The definition of session types will be important for the remainder:
-\input{../paper/session-types.tex}
-
-Here, I use a modified version of the definition given by
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} where I have omitted recursive
-session types. The type \lstinline{end} denotes the terminal session type and
-a channel typed as such may not perform any communication. A communication
-channel with session type $\inpt{T}{S}$ expects to receive a term of type $T$
-and then continues with session type $S$. Conversely, a channel with type
-$\outpt{T}{S}$ can send a term of type $T$ and then has type $S$. Thus, input
-and output are considered \emph{dual} session types, in the sense that a
-channel with one endpoint having type $\inpt{T}{S}$ and the other endpoint
-having type $\outpt{T}{\Sbar{S}}$, where $\Sbar{S}$ is the dual of $S$,
-permits communication between two agents, allowing one to send a term of type
-$T$ and the other to receive it with continuing endpoint types $\Sbar{S}$ and
-$S$, respectively. Similarly, the remaining types $\Branch{l_i}{S_i}{i\in I}$
-and $\Choice{l_i}{S_i}{i\in I}$ are duals of each other. A channel of type
-$\Branch{l_i}{S_i}{i\in I}$ can receive a label $l_i$ which will select the
-corresponding session type $S_i$ as the continuing type of the channel. While
-a channel with type $\Choice{l_i}{S_i}{i\in I}$ can send a label $l_i$ and
-continues as $S_i$.
-
-\subsubsection{In the \texorpdfstring{$\pi$}{pi}-calculus}\label{sec:pis}
-
-\citeauthor{Caires:2010:STI} \cite{Caires:2010:STI} present a connection
-between a $\pi$-calculus type system with session types ($\pi$DILL) and dual
-intuitionistic linear logic (DILL)~\cite{Barber:1996}. They equate linear
-propositions with session types of the process calculus via a bidirectional
-translation relation. They use a binary reduction relation and a labelled
-transition system to represent process reduction and process action with its
-environment as is standard~\cite{Sangiorgi:2001}.
-
-\input{../paper/sill-branch-choice-rules}
-
-They represent dual intuitionistic linear logic as a sequent calculus
-providing logical connectives which have session-based interpretations in
-$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL in
-Figure~\ref{fig:sill-branching}, which have only two alternatives
-(cf. multiple alternatives in the standard definition in \S~\ref{sec:pre}).
-
-Note that branch and choice both have some notion of branching (R rules for
-$\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
-a session-based setting, since the intuition for \& is that of
-\emph{offering} (receiving) a choice between alternatives rather than
-\emph{making} (sending) the choice. Here, however, the intuition should be
-extended to account for its location in the typing judgement. The channel is
-on the left of the sequent, and \citeauthor{Caires:2010:STI} suggest one
-consider the context as services being provided to the typed process. Thus,
-the type of channels in the context is really the type of the other endpoint
-of the channel, and the typed process behaves according to the dual
-protocol. For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to
-represent the dual of a protocol $A$ as defined by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process is behaving
-according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from
-alternatives (in this case, $\inrterm$) and then behaving according to the
-selected protocol (in this case, protocol $B^\bot$).
-
-As noted by Wadler~\cite{Wadler:2014} there is a certain amount of redundancy
-in the two-sided nature of the rules forcing one to commit to a certain
-convention when typing a communication system. Consider, an equivalent system
-to the buy/quote example given by
-\citeauthor{Caires:2010:STI}~\cite[\S~3]{Caires:2010:STI}:
-\begin{gather*}
-\sdef{ClientProto}{\BChoice{(\tensor{Book}
-                            {(\tensor{Card}{(\lfn{Receipt}{\1})})})}
-                           {(\tensor{Book}{(\lfn{Quote}{\1})})}}
-\end{gather*}
-
-Note, I am using $\tensor{}{}$ for session output and $\lfn{}{}$ for session
-input following the presentation by \citeauthor{Caires:2010:STI}. In the
-protocol defined above, a client selects between buying a book or receiving a
-price quote for a book. In the buy option case, the client sends the name of
-the book to the server and their card details, then the server sends the
-receipt for the purchase. In the quote option case, the client sends the name
-of the book and receives a price quote from the server. I have replaced the
-types $N$ and $I$ used by \citeauthor{Caires:2010:STI} with more meaningful
-type names $Book$, $Card$, $Quote$ and $Receipt$ where appropriate; all are
-synonyms for the type $\1$ which denotes the terminal session type in this
-system. The client and server process bodies are as follows:
-\begin{gather*}
-\sdef{ClientBody_c}
-     {\CPinr{c}{\SLoutput{1984}{c}{\SLinput{c}{quote}{\0}}}}\\
-\sdef{ServerBody_c}
-     {\CPcase{c}
-             {\SLinput{c}{book}{\SLinput{c}{card}{\SLoutput{receipt}{c}{\0}}}}
-             {\SLinput{c}{book}{\SLoutput{quote}{c}{\0}}}}
-\end{gather*}
-
-\begin{samepage}
-The typings for the client and server processes are given by:
-\begin{gather*}
-\tsequent{\cdot}{\cdot}{ClientBody_c}{\typedid{c}{ClientProto}}
-\tag{\mkTrule{ClientBody}} \label{eq:cbtyp}
-\\ \tsequent{\cdot}{\typedid{c}{ClientProto}}{ServerBody_c}{\typedid{-}{\1}}
-\tag{\mkTrule{ServerBody}} \label{eq:sbtyp}
-\end{gather*}
-\end{samepage}
-
-The derivation for \eqref{eq:cbtyp} is $T\1 R$, $\lfn{T}{R}$, $\tensor{T}{R}$,
-$\BChoice{T}{R_2}$ and \eqref{eq:sbtyp} is derivable from
-$\BChoice{T}{L}$. Except for the final rule, these are exact dual rules of the
-rules used in example by \citeauthor{Caires:2010:STI}. Indeed, consider the
-composite process:
-\begin{gather*}
-\sdef{QSimple}{\Spar{c}{ClientBody_c}{ServerBody_c}}\\
-\tsequent{\cdot}{\cdot}{QSimple}{\typedid{-}{\1}}
-\end{gather*}
-
-The final system is near identical to the one produced in the example given by
-\citeauthor{Caires:2010:STI} except the individual processes within the
-parallel composition have been swapped. Here the client communicates along the
-channel $c$ on the right-side of the sequent and the server utilises the
-channel $c$ on the left of the sequent. This typing, while correct, violates
-the intuition given by \citeauthor{Caires:2010:STI} of the context providing
-the services used by the process, since one would usually associate provision
-of services with the server not the client. Instead, the example above
-suggests the server utilises some channel provided by the client. On the other
-hand, one could imagine such a channel being setup between two communicating
-agents; the ``owner'' of the channel is largely irrelevant. Note that
-$ClientProto$ is in fact the dual of the $ServerProto$ definition by
-\citeauthor{Caires:2010:STI} using the definition of dual used in Wadler's
-classical process calculus~\cite{Wadler:2014}. In $\pi$DILL, the duality
-between endpoints of the same channel are obscured by the nature of the typing
-judgements.
-
-Process reduction steps are equated with proof reduction via cut
-elimination. The relationship with proof normalisation in DILL provides
-absence of deadlocks in session-based communication for the process
-calculus. The correspondence of session types to linear logic propositions is
-an important step in providing a solid logical foundation for a concurrent
-type theory; in the same way ``sequential'' type theory ($\lambda$-calculi)
-profitted from a logical perspective. However, for session types to become a
-more usable tool for programming distributed communication systems (the
-ultimate goal, of course, is to write more correct programs), one cannot be
-expected to program directly in the $\pi$-calculus; a more mainstream
-programming paradigm must be built upon these solid logical foundations. With
-that in mind, enter the $\lambda$-calculus!
-
-\subsubsection{In the \texorpdfstring{$\lambda$}{lambda}-calculus}
-\label{sec:lam}
-
-A series of works by \citeauthor{Gay:2003:STI}~\cite{Gay:2003:STI} and
-\citeauthor{Vasconcelos:2006:TCM}~\cite{Vasconcelos:2006:TCM} provide a
-$\lambda$-calculus based language with session types for channel
-primitives. The type system is nonstandard, featuring an additional
-environment for tracking the session types of channels; these types are
-modified during term reduction to reflect communication. The extra channel
-environment is provided to support possible aliasing of channels by treating
-channel variables as pointers or references into the channel environment. So,
-for a channel $c$ of session type $S$, a variable $x$ referring to this
-channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the
-channel environment. The typing judgements and term reduction relation must be
-augmented to handle these environments. Unfortunately, a side-effect of this
-is that function types must specify the pre and post-states of channel
-environments and therefore, the type of arguments cannot work up to
-alpha-equivalence of channel identifiers. That is, if two channels $c$ and $d$
-have the same session type, one cannot define a function to operate on
-variables of either channel type since the function definition will explicitly
-mention the identifier for the channel (in this case either, $Chan~c$ or
-$Chan~d$). However, the system does provide an attractive non-linear channel
-type which has been neglected in favour of a linear treatment of channels in
-more recent
-works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2014}. Certainly, it is the
-case that an aliased approach would fit better with mainstream programming
-languages and perhaps studying a similar system could offer some insights.
-
-The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of
-aliased access to linear resources. The type system requires special language
-features, e.g. to manage capabilities which can be thought as a form of object
-liveness mechanism, so GV typing judgements would need to be extended such
-that the capabilities can be included in soundness results. An extension to
-the translation to CP to incorporate the ``adopt'' and ``focus'' constructs
-would also need to be defined. It is not immediately clear how these features
-would assist in the management of session typed channels. The ``focus'' rule
-(and linearity generally) enforces an invariant on the focussed object
-(channel) such that the type is maintained during the focus operation:
-\input{../proposal/focus-rule}
-
-The typing judgements include \emph{capabilities} that track aliasing to heap
-objects, e.g. $\AFcap{\rho}{h}$ denotes a mapping from a static name $\rho$ to
-a heap type $h$, and a variable $x$ may be typed as $\typedid{x}{\AFtr{\rho}}$
-indicating that $x$ refers to an object of type $h$. The purpose of the
-capabilities is to enforce restrictions on when an object can be
-referenced. In other words, if $\AFcap{\rho}{h}$ is in the capabilities at
-some program point then $x$ may be used, otherwise such use is prohibited. In
-the [focus] rule, evaluating $e_1$ retrieves the object to be focussed. Note
-the type of $e_1$ is a ``guarded'' type. The guard $G$ is the static name of
-some object which ``adopted''~\footnote{I omit the adopt typing rule for
-  brevity.} an object of heap type $h$. If a program point has
-$\AFcap{G}{h_1}$ in the capabilities for some $h_1$ then all objects guarded
-by $G$ may be used; $\capsequent{\Delta}{C_1}{G}$ denotes containment of a
-such a capability in $C_1$. The guarded type may be thought of as policing
-manipulation of linear components within $h$. Treating $\otimes$ as the
-disjoint union for capability sets, the typing for $e_2$ temporarily removes
-the guard $G$ from the set of capabilities, and adds a new variable $x$, and
-associated capability, with a type permitting access to the object of type $h$
-i.e. a non-guarded type. Removing $G$ will prohibit access to allow objects
-guarded by $G$ for the during of $e_2$. The final capabilities after executing
-$e_2$ are required to be $\tensor{C_3}{\AFcap{\rho}{h}}$, enforcing that the
-type must remain as $h$. The typing of the $\letterm$ expression ensures that
-the guard $G$ (contained in $C_1$) is restored.
-
-Such a rule would appear to only benefit recursive session types. Indeed,
-focussing on a non-recursive channel would be pointless since one would be
-prevented from communicating across it since the focus rule requires
-maintenance of the channel type. Further, it would be undesirable to permit
-modification of a session type without guaranteeing exclusive access to the
-channel since this could lead to invalid communication e.g. trying to
-communicate on a closed channel. Additionally, a linear object is consumed
-upon being adopted and will be deallocated when its adopter's lifetime ends,
-although an alternative semantics could be defined which returned the linear
-resource to the context. Moreover, \citeauthor{Fahndrich:2002} do not provide
-operational semantics or soundness results for the ``adoption and focus'' type
-system.
-
-The Alms language~\cite{Aldrich:2009} provides a functional language with
-affine types based on linear logic. A notion of type \textit{kind} is used to
-distinguish between affine and unrestricted values. The typing judgements use
-two contexts; one for variables with unrestricted kind and another for
-variables of affine type and type variables. Functions are also labelled with
-a kind which may depend on their arguments e.g. a multiple argument function
-must have affine kinding if it accepts an affine typed resource as an argument
-other than the final argument. The reason for this is simply to prevent the
-function from using an affine resource more than once. Additionally, functions
-act as closures over their environments so their qualifier may be determined
-by a free variable in the environment, and a relation is provided to calculate
-the appropriate kind qualifier. Type variables are provided to support kind
-polymorphism of function qualifiers. The system is expressive enough to
-implement session types as a module using a similar approach to
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} although treating channels as
-affine rather than linearly typed. \citeauthor{Aldrich:2009} note that Alms'
-type system can support the ``adoption and focus'' system but do not specify
-details or whether any changes are necessary to their typing judgements.
-
-The work on a linear functional language with session types by
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the
-current project from the body of work on session-based $\lambda$-calculus type
-systems. They provide paper proofs of their system's type safety. Given the
-non-trivial features of the type system such as recursive session types and
-buffered channels for supporting asynchronous communication it will be
-challenging to extend these proofs manually. A mechanised version would
-provide stronger guarantees of correctness and allow others to more easily
-alter proofs as a result of changes made to the language.
-
-\subsubsection{Fusion}
-
-A number of systems incorporate both a functional language surface syntax with
-a translation to a more primitive process calculus. I shall focus on two such
-systems presented recently which have a connection to linear logic.
-
-Lolliproc~\cite{Mazurak:2010:LCC} provides a functional language with
-concurrency primitives as a surface syntax on top of an underlying process
-calculus. The process calculus provides a classical linear logic
-interpretation for the functional language, utilising double negation
-elimination to provide session type duality. The surface syntax is an
-effective abstraction layer for the programmer since it does not expose the
-underlying process calculus.
-
-Concurrency is provided by control operators which, respectively, spawn a
-child process and wait for a child process to complete. The spawning process
-creates a channel between parent and child process; a channel is represented
-as a continuation and the type denotes the communication protocol
-(i.e. session type) between parent and child.
-
-\citeauthor{Mazurak:2010:LCC} present proofs for type safety, strong
-normalisation and confluence. Confluence provides the guarantee that no race
-conditions can occur, strong normalisation prohibits non-terminating programs
-and type safety assures deadlock-freedom. Deadlock-freedom is implied by the
-acyclic communication graphs (progress of process reduction) which occur
-between parallel processes; follows from permitting only one channel between
-two halves of parallel composition construct in the process calculus. Most of
-the development is supported by Coq proof scripts with remaining issues due to
-reasoning about communication graphs. The development does not provide
-primitive send and receives; instead, these are encoded as linear functions.
-%% and classical extension to second-order logic by
-%% \citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}.
-
-\citeauthor{Wadler:2014}'s recent work~\cite{Wadler:2014} brings together a
-process calculus, CP, based on the earlier work by
-\citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
-\citeauthor{Gay:2010:LAST}. GV presents session types using the standard
-presentation as described in \S~\ref{sec:pre}. CP presents linear logical
-connectives with session type interpretations inspired by
-\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI}. \citeauthor{Wadler:2014}
-describes a continuation-passing style (CPS) translation from GV session types
-to CP classical linear logic propositions. The process calculus CP is slightly
-different to $\pi$DILL by \citeauthor{Caires:2010:STI}, apart from being
-classical in nature, the constructs for server replication are modified; CP
-provides a weakening and contraction rules whereas these are encoded in one
-rule for process reduction by \citeauthor{Caires:2010:STI}. Additionally,
-rather than a two-sided sequent calculus, \citeauthor{Wadler:2014} uses
-one-sided sequents for defining the process calculus, which provides a more
-intuitive presentation of session type duality and clearer correspondence to a
-$\lambda$-calculi surface syntax, simplifying the translation presentation. As
-presented, GV does not support all features of CP and thus the translation is
-one way (GV to CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP}
-extend GV to provide support for polymorphism and replication, following a
-similar CPS translation scheme. Lastly,
-\citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
-semantics for GV and their system has greater similarities with LAST than
-prior GV works, for example, by treating send and receive as linear functions
-rather than language primitives.
-
-\subsection{Mechanising Programming Language Metatheory}
-
-Interactive theorem proving and proof assistants have seen an upsurge in use
-in recent years. Especially within the domain of programming language
-research, there is a strong emphasis on providing mechanised proofs
-accompanying work in the area. Indeed, \citeauthor{Aydemir:2005:MMM} have
-published a series of challenges on mechanising programming languages aimed at
-providing a starting point for comparing different representation
-techniques~\cite{Aydemir:2005:MMM}. Another aim is to provide reusable
-libraries for common reasoning needed across programming language
-developments, e.g. handling of typing environments. This effort has resulted
-in a number of different approaches for representing programming languages
-within proof assistants. For instance,
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} present a representation
-of $\lambda$-calculi where bound variables are represented as \textit{de
-  Bruijn indices}~\footnote{A de Bruijn index is a number used to represent a
-  bound variable which indicates the position of its binder (starting from
-  zero for the innermost binder). For example, $\lambda~(\lambda~1)$
-  corresponds to $\lambda x. \lambda y. x$, i.e. the constant function.} and
-free variables are represented as named terms; the \textit{locally nameless}
-representation. Their work resulted in a library for handling commonly
-occurring aspects in programming language metatheory which has been used since
-in other developments~\cite{Park:2014:MMW}. Reusable libraries reduce the
-proof engineering effort needed for later works, in contrast to the
-pen-and-paper approach, and some proof assistants (as is the case with Coq)
-allow one to extract a program from the development which can provide a
-typechecker or compiler for the language formalised.~\footnote{Curry-Howard
-  strikes again; a proof of decidability of typechecking corresponds to a
-  program implementing a typechecker for the language.}
-
-Some recent mechanisation effort focuses on providing a basis for studying
-linearity within type systems. This work is of interest to the current project
-since most session-based type systems assume a strictly linear type system
-which requires re-binding of channel identifiers, such as in the functional
-setting of \citeauthor{Gay:2010:LAST}. \citeauthor{Mazurak:2010:LLT} present
-an extension to System F (termed \fpop) with a notion describing types as
-having linear or non-linear \textit{kind}~\cite{Mazurak:2010:LLT}. This
-extension is motivated by previous work on incorporating linearity into type
-systems. Overall, previous attempts either hamper the inclusion of desirable
-programming features (such as polymorphism), or do not adequately reflect
-non-linearity as the common case leading to awkward programming. \fpop avoids
-these deficiencies by categorising types into kinds whilst maintaining similar
-semantics to System F. Mechanised proofs for type soundness and parametricity
-are presented, and the semantics allow non-linear types to be treated as
-linear. Examples show the system can provide a wide range of permissions on
-interfaces, from full exclusive access to read-only shared references. The
-system is very close to the Alms language described by
-\citeauthor{Aldrich:2009} but has the advantage that the soundness results
-(type safety and parametricity) have been entirely mechanised in Coq. It would
-be interesting to see if one could encode the ``adopt'' and ``focus''
-constructs in this system to compare its expressiveness to Alms. The ``adopt''
-rule can be encoded similarly to the ShareRef
-example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to encode the
-``focus'' rule within the present semantics, as one needs to prevent
-non-linear access to its argument implying the presence of knowledge about
-aliasing within the typing judgements; absence of aliasing is not identified
-with linear kinded types in \fpop.
-
-While not related to programming language metatheory,
-\citeauthor{Gay:2001:FFP}~\cite{Gay:2001:FFP} provides a $\pi$-calculus
-framework mechanised in Isabelle/HOL. The aim is to build a mechanisaton of
-session-based type systems on top of the $\pi$-calculus framework. The use of
-de Bruijn indices for both bound and free variables causes issues with
-variable substitution often requiring permutation of typing environments (see
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} for a discussion on
-binder representations) and lifting of free variables during substitution; I
-avoid this issue in my representation choice (\S \ref{sec:approach}).
-
-The introduction of names for hypotheses during execution of custom tactics
-are an issue in the proof development. These names are automatically generated
-by the Isabelle/HOL system which creates a dependency between the names chosen
-and the tactic. Thus, one must be careful to use the exact names expected by
-the tactic. In the Coq proof assistant, defining custom tactics can be
-achieved using the Ltac language~\cite{Delahaye:2000:TLS}. Among other things,
-Ltac supports pattern matching on hypotheses and goal forms. Thus, if one
-wished to pattern match on an inductive type describing variables, one could
-do so without mentioning the actual variable name as follows:
-
-\begin{coq}
-Ltac my_tactic :=
-  match goal with
-  | [v: var |- _] => ... tactics here possibly mentioning `v' hypothesis ...
-  end.
-\end{coq}
-
-\subsubsection{Applications to Session-based Type Systems}\label{sec:asts}
-
-\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based
-type system providing session polymorphism. The system is more general than
-session polymorphism via subtyping~\cite{Gay:2005:SST} in that it is capable
-of typing a generic forwarding process between mutually dual session
-types. Defining such a process is not possible using subtyping since one
-wishes to restrict the types to be dual, not a subtype of some other
-type. This more general form of polymorphism is achieved by permitting input
-on all session types (including \lstinline{end}, the terminal session type)
-and then employing transition hypotheses to ensure the type permits such
-input. The authors mechanise their type system and its properties in the Coq
-proof assistant. However, the mechanised system is restricted to the
-$\pi$-calculus and the main focus is proving soundness properties for the
-deduction rules allowing polymorphism, rather than a more general basis for
-studying programming with session types.
-
-\section{Formalisation}
-
-The formalisation was performed using the Coq proof
-assisstant~\cite{Coq:manual}. In the spirit of the POPLMARK
-challenge~\cite{Aydemir:2005:MMM}, the development utilises available
-libraries and infrastructure as much as possible. In particular,
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a metatheory
-library for handling, among other things, association lists (environments) and
-free variable representation. The current development uses the version of the
-library distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which
-has been updated for Coq version 8.4pl4. One goal of the formalisation effort
-was to minimise the extent to which the development differs from the
-pen-and-paper presentation and avoid questions of adequacy of the
-representation. Some changes were made to ease the mechanisation and are
-highlighted where they occur.
-
-The locally nameless first-order representation for binders and free variables
-was adopted for both GV and CP, following
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. This first-order
-variant prevents variable-capture issues when compared with other first-order
-approaches. Alternatively, I could have chosen a higher-order abstract syntax
-(HOAS) approach such as the parametric version (PHOAS) described
-by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
-on the metalanguage (Coq, in this case) for handling binders and substitution
-so has less overhead in managing syntax than the locally nameless
-approach. However, it was less clear how to handle linearity using this
-approach and has the disadvantage of being able to represent terms not valid
-in the language being mechanised (so called ``exotic'' terms). Further, a
-higher-order approach would prevent the use of the Metatheory library which
-provides a number of useful tactics for manipulating typing contexts. Previous
-efforts to mechanise \fpop~\cite{Park:2014:MMW} illustrate how to handle a
-language containing both linear and non-linear contexts. However, to stay true
-to CP/GV, I maintained a single context. Interestingly, it became clear that
-the mechanised version of CP required explicit weakening and contraction to
-facilitate the reduction rules.
-
-For handling binder freshness, I chose to adopt the cofinite quantification
-approach described by~\cite{Aydemir:2008:EFM}, which excludes a finite set of
-variables from being considered as the binder. In contrast, the traditional
-``exists-fresh'' approach, where the binder is only required to be fresh
-within the abstraction's body, does not produce strong enough induction
-principles in some cases. This change will necessitate changes to the typing
-rules of GV and CP. Consider for instance the Connect rule in
-GV~\cite{Wadler:2014} (C) and its corresponding cofinite quantification
-version (CCO): \input{with-connect-rule}
-
-The conclusion specifies a variable in (C) but a type in (CCO) (de Bruijn
-index rather than a name). Thus, in the premise for (CCO) one can choose a
-suitably fresh $x$ in each premise. Note that, $L$ can be chosen such that
-$\Psi \cup \Phi \cup \fv(M) \cup \fv(N) \subseteq L$, where $\fv(M)$ are the
-set of free variables in term $M$, and $x$ can be the same in both
-premises. Similar changes are made to the rules for CP (see
-\S~\ref{sec:cp}). These changes do not alter the typing of processes nor the
-semantics of the translation from GV to CP. An equivalence between
-``exists-fresh'' and cofinite quantification definitions of the simply-typed
-$\lambda$-calculus is given by
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
-result could be derived straightforwardly (but tediously) for the CP/GV
-system.
-
-\begin{comment}
-\section{Issues}
-
-\begin{itemize}
-\item What is required of a Metatheory library for linear type systems?
-  Permutation reasoning? Weakening/Contraction for non-linear components?
-\item Discuss the different approaches that were taken to maintain close
-  correspondence with the paper presentation
-\item Make sure to note that the paper leaves details (a lot of the ``cruft'')
-  out of the proofs e.g. where weakening is applied in the typing derivations
-\item Mention the use of Metatheory and its limitations when mechanising
-  linear type systems like GV/CP
-\end{itemize}
-\end{comment}
-
-\subsection{The process calculus CP}\label{sec:cp}
-
-\begin{comment}
-FIGURES
-
-propositions
-
-processes
-
-typing judgements
-
-equivalences and axiomatic cut
-
-principal cut rules
-
-commuting conversions
-\end{comment}
-
-Formalising the process calculus CP provided many challenges for the
-metatheory library since to my knowledge, the library has never been applied
-to a process calculus before. So for instance, in the lambda calculus one
-typically does not work up to permutation of binders, whereas in a process
-calculus these kinds of structural equivalences are standard.
-
-Note that the rules for polymorphism, $\exists$ and $\forall$, are defined but
-cut elimination on them is not. Since GV (\S~\ref{sec:gv}) does not support
-polymorphism the reduction rules are unnecessary for the translation. Future
-work is to support reduction under these terms.
-
-\subsection{The functional language GV}\label{sec:gv}
-
-\begin{comment}
-FIGURES
-
-types
-
-terms
-
-typing judgements
-\end{comment}
-
-\subsection{CPS translation GV to CP}
-
-Currently the formalisation does not handle this. Future work.
-
-for an unbound index so this decision is sound. A consequence is that I must
-define a notion of opening language terms involving binders as described
-by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
-$t$ which is well-typed if $t$ is well-typed when its unbound index,
-representing the argument to the abstraction, is replaced by some fresh
-variable $x$, written $t^x$. Some terms in GV require multiple binders but I
-believe it will be straightforward to extend this notion. Unfortunately, I
-must define similar operations on CP terms and handle substitution and opening
-on them as well as propositions which can accept propositional variables. I
-hope to be able to generalise as much as possible with the help of the
-Metatheory library. say something about gmeta here and its automatic
-generation of these infrastructure lemmas
-
-While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
-may need to alter the approach slightly to avoid tricky formulations. For
-instance, instead of providing $n$-ary branch and choice in GV, I provide
-only binary versions of these operators to simplify the development. This
-change is not restrictive however, since CP itself provides only binary
-versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
-technical note, I amalgamate all types into one inductive definition and
-provide a predicate for restricting typing rules to consider only valid
-session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
-recursive definitions; session types are types which may contain types as
-subcomponents (arguments to send, for example). Unfortunately, handling
-mutually inductive definitions in Coq can be quite involved; requiring one to
-either rely on the Coq system to provide a stronger mutual induction principle
-or defining one manually. Such a definition is possible, but it complicates
-elimination via the induction principle.
-
-Another aspect of the encoding is how to define terms. The intrinsic encoding
-described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
-approach, indexing terms by their type so as to prevent ill-formed terms from
-being constructed. However, in the encoding for GV there are issues with using
-this encoding: $(1)$ it is not immediately clear how to handle linear contexts
-using the de Bruijn variable encoding, since in the intrinsic setting the
-environment supports weakening; $(2)$ an intrinsic encoding of terms would be
-complicated by the need to enforce that session types occur in certain
-instances (by use of a predicate); and $(3)$ well-typed terms require extra
-assumptions about binder freshness which cannot reasonably be expressed as a
-function type. For these reasons, an intrinsic approach does not offer much
-benefit since a well-typed term relation would still need to be
-defined. Therefore, I chose an extrinsic encoding for terms and define
-well-typed terms as a separate inductive type.
-
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
-their language proof infrastructure is proportional to the number of binding
-constructs. In the case of a simply typed lambda calculus this is not onerous,
-but GV has four binding constructs, CP has six and propositions have two. It
-may be true that this approach results in too many auxiliary definitions and
-operations not relevant to the actual properties one wishes to prove. However,
-I believe that once this infrastructure has been defined, the system can be
-extended without concern for these binding constructs.
-
-\citeauthor{Park:2014:MMW} describe a technique for
-removing nonlinear contexts from typing judgements. While their work extended
-to \fpop it is not clear how to handle an environment containing both
-non-linear and linear types as in GV. I wish to maintain as close a
-relationship as possible to the paper system presented by
-\citeauthor{Wadler:2014}, so separating out the non-linear and linear
-components (as in \fpop) is not an option at this stage.
-
-I describe the formalisation as it currently stands in this section. The use
-of the metatheory library from UPenn provided useful tactics and properties
-regarding freshness of free variables and environments.
-
-\subsection{Future Work}
-
-In addition to that already mentioned: there remains much to be done.
-
-\section{Conclusions}
-
-{\bf Acknowledgments.}
-This is optional; it is a location for you to thank people
-
+\sloppy
 \printbibliography

### 2015-04-03 paper/formalisation.tex
Updated paper: CP section and added figures.

@@
 library for handling, among other things, association lists (environments) and
-free variable representation. The current development uses the version of the
+variable representation. The current development uses the version of the
 library distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which
@@
 
-%\input{with-connect-rule}
-\input{cp}
-
-The locally nameless first-order representation for binders and free variables
-was adopted for both GV and CP, following
+The locally nameless cofinite quantification (LNCQ) representation for binders
+and free variables was adopted for both GV and CP, following
 \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. This first-order
@@
 fresh within the abstraction's body, does not produce strong enough induction
-principles in some cases. This change will necessitate changes to the typing
-rules of GV and CP. Consider for instance the Connect rule in GV (C) and its
-corresponding cofinite quantification version (CCO) in
-Figure~\ref{fig:connect}. The conclusion specifies a variable in (C) but a
-type in (CCO) (de Bruijn index rather than a name). Thus, in the premise for
-(CCO) one can choose a suitably fresh $x$ in each premise. Note that, $L$ can
-be chosen such that $\Psi \cup \Phi \cup \fv(M) \cup \fv(N) \subseteq L$,
-where $\fv(M)$ are the set of free variables in term $M$, and $x$ can be the
-same in both premises. Similar changes are made to the rules for CP. These
-changes do not alter the typing of processes nor the semantics of the
-translation from GV to CP. An equivalence between ``exists-fresh'' and
-cofinite quantification definitions of the simply-typed $\lambda$-calculus is
-given by \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect
-a similar result could be derived straightforwardly (but tediously) for the
-CP/GV system.
+principles in some cases, e.g. weakening~\cite{Aydemir:2008:EFM}. This
+decision impacts upon the premises of the CP/GV typing rules (\S~\ref{sec:cp}
+and \S~\ref{sec:gv}), however these changes do not alter the typing of
+processes nor the semantics of the translation from GV to CP. An equivalence
+between ``exists-fresh'' and cofinite quantification definitions of the
+simply-typed $\lambda$-calculus is given by
+\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
+result could be derived straightforwardly (but tediously) for the CP/GV
+system.
 
@@
 approach, indexing terms by their type so as to prevent ill-formed terms from
-being constructed. However, in the encoding for CP/GV there are issues with
-using this encoding: $(1)$ it is not immediately clear how to handle linear
-contexts using the de Bruijn variable encoding, since in the intrinsic setting
-the environment automatically supports weakening; $(2)$ an intrinsic encoding
-of GV terms would be complicated by the need to enforce that session types
-occur in certain instances (by use of a predicate); and $(3)$ well-typed terms
+being constructed. However, there are issues with using this encoding for
+CP/GV: $(1)$ it is not immediately clear how to handle linear contexts using
+the de Bruijn variable encoding, since in the intrinsic setting the
+environment automatically supports weakening; $(2)$ an intrinsic encoding of
+GV terms would be complicated by the need to enforce that session types occur
+in certain instances (by use of a predicate); and $(3)$ well-typed terms
 require extra assumptions about binder freshness which cannot reasonably be
@@
 
-\citeauthor{Park:2014:MMW} describe a technique for
-removing nonlinear contexts from typing judgements. While their work extended
-to \fpop it is not clear how to handle an environment containing both
-non-linear and linear types as in GV. I wish to maintain as close a
-relationship as possible to the paper system presented by
-\citeauthor{Wadler:2014}, so separating out the non-linear and linear
-components (as in \fpop) is not an option at this stage.
+\citeauthor{Park:2014:MMW} describe a technique for removing non-linear
+contexts from typing judgements. While their work extended to \fpop it is not
+clear how to handle an environment containing both non-linear and linear types
+as in CP/GV. I wish to maintain as close a relationship as possible to the
+pen-and-paper system presented by \citeauthor{Wadler:2014}, so separating out
+the non-linear and linear components (as in \fpop) is not an option at this
+stage.
 
@@
 \subsection{The process calculus CP}\label{sec:cp}
+
+\input{cp}
 
@@
 calculus these kinds of structural equivalences are standard. Additionally,
-substitution and opening are only defined on processes for names. In other
-words, one cannot substitute an arbitrary process term for a name, unlike in
-the lambda calculus. Thus frameworks which assume a lambda calculi style
-substitution operation are not immediately amenable to the CP
-setting~\cite{Lee:2012}.
+substitution and opening are only defined for names. In other words, one
+cannot substitute an arbitrary process term for a name, unlike in the lambda
+calculus. Thus frameworks which assume a lambda calculi style operations are
+not immediately amenable to the CP setting~\cite{Lee:2012}. Ordering in
+environments is ignored so the development required lemmas allowing the
+permutation of environments. Support for permutating environments is not
+currently supported by the Metatheory library but some general lemmas were
+added as a result of this work. Additionally, custom tactics were developed to
+automatatically solve some simple permutation goals, however these tactics are
+not a ``catch all'' and some generalising still needs to be done.
 
-Note that the rules for polymorphism, $\exists$ and $\forall$, are defined but
-cut elimination on them is not. Since GV (\S~\ref{sec:gv}) does not support
-polymorphism the reduction rules are unnecessary for the translation. Future
-work is to support reduction under these terms.
+% Description of the typing rules...
 
-Initially, an implicit notion of weakening and contraction were chosen so as
-to be in line with \citeauthor{Wadler:2014}'s definitions. However, explicit
-weakening and contraction were required to differentiate the reduction rules
-for the various interactions between servers and clients.
+Figure~\ref{fig:cll} presents the process calculus CP in LNCQ form (ignoring
+permutation premises). In other words bound variables are represented using de
+Brujin indices and free variables are represented using an abstract type
+provided by the Metatheory library. The rules with binders reflect this by
+specifying the type of a bound variable in the constructor rather than the
+name. Cofinite quantification is performed in the hypotheses to these rules by
+excluding a finite set of names from opening the subprocesses. Note that the
+rules for polymorphism, $\exists$ and $\forall$, were included in the final
+Coq development since GV (\S~\ref{sec:gv}) does not support polymorphism
+making the rules unnecessary for the translation. They appear in the Figure
+simply for the sake of completeness. Additionally, as noted by
+\citeauthor{Lee:2012}, support for polymorphism in type variables creates an
+explosion of the infrastructure requiring duplication of opening and
+substituting definitions and lemmas.
 
-Ordering in environments is ignored and the development required lemmas
-allowing the permutation of environments. Support for permutating environments
-is not currently supported in the Metatheory library but some general lemmas
-were added as a result of this work. Additionally, custom tactics were
-developed to automatatically solve some simple permutation goals. However,
-these tactics are not a ``catch all'' and some generalising still needs to be
-done.
+% ...leading on to the discussion of the explicit weakening...
 
-The paper presentation omits details: the principal cut rules do not specify
-weakening could occur: since weakening is implicit is tacitly assumed
-weakening can always be pushed further up the derivation tree e.g. AxCut rule
+Initially, an implicit notion of weakening was chosen so as to be in line with
+\citeauthor{Wadler:2014}'s definitions. However, explicit weakening is
+required to differentiate the principal cut elimination rules for the various
+interactions between servers and clients. contraction is not supported because
+of the difficulty in specifying its principal cut elimination rule. In order
+to specify the rule one requires to show the following:
 
-another interesting aspect which the formalism reveal is the rule for
-associativity is constructed in a way to permit associativity in a certain way
-pre-determined
+\todo{v A -> !0(A).P | Q{x/x'}
+  ==>
+v A -> !0(A).P | v A -> !0(A).P | contract (Gamma) Q}
+
+Unfortunately, the reduction relation does not have knowledge of the
+environment typing process P. This fact does not impact upon expressing
+weakening since for weakening only the names (not the bindings) within the
+environment are required. The paper presentation does not specify where
+weakening could occur; it is tacitly assumed weakening can always be pushed
+further up the derivation tree e.g. AxCut rule. a consequence of this choice
+is that for the empty choice rule there may only be a single binding in the
+context (cf. an arbitrary context in \citeauthor{Wadler:2014}'s
+presentation~\cite{Wadler:2014}).
+
+% ...leading on to presentation of principal cut elimination rules...
+
+\input{principalcp}
+
+The principal cut elimination rules are presented in Figure~\ref{??}. These
+rules are expressed to indicate some additional complexity not present in the
+original presentation. For example, structural equivalences and associativity
+are not unique and different process terms result depending on initial binder
+permutation. ${0 <~> 1}P$ indicates the permuting of binders $0$ and $1$ in
+process P. The paper presentation specifies only one possible associativity
+rule lending to the specific split of the linear environments. The reduction
+rules were complicated to prove due to being composed of multiple constructor
+instances and differed significantly to the standard operational semantics one
+typically deals with in lambda calculus formalisms.
+
+% ...commuting conversions
+
+\input{commconvonecp}
+
+\input{commconvtwocp}
+
+% ...leading on to a discussion of further CP-specific issues encountered.
 
@@
 
+\input{gv}
+
 While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
@@
 
+another issue is the way i tried to formulate types initially treating them as
+indexed families indexed by kinds. while this exploits the dependently typed
+approach it complicated later definitions and proofs. for instance decidablity
+of equality on types now relied on an axiom permitting heterogeneous
+equality. further when formulating the typing environment we must decide which
+kind each type had and most rules are polymorphic in that decision leading to
+the creation of an awkward inspection function. these issues highlighted that
+one should be careful and considerate how the representation will affect later
+expressiveness.
+
+explicit limiting and unlimiting of abstraction to ensure rules are syntax
+directed - permits inductive argument on terms when translating from gv to cp
+
 \subsection{CPS translation GV to CP}
 
+type inference of gv terms required cluttered definition of translation with
+impossible error cases. avoided a dependently typed solution due to
+difficulties encountered earlier in the development. james mckinna et
+al. suggest encoding the successful cases in an inductive type and then
+defining the function to act on these cases. will this really work? this means
+a graph of positive cases for types then a function definition for it too
+
+translation uses the coq-ext-lib from michela for monadic notation
+
 Currently the formalisation does not handle this. Future work.

### 2015-04-03 paper/formalisation.tex
Added discussion of GV.

@@
 
-typing judgements
-
 equivalences and axiomatic cut
-
-principal cut rules
-
-commuting conversions
 \end{comment}
@@
 typically deals with in lambda calculus formalisms.
+linear environments caused a problem with the splitting up of variables.
 
@@
 
+The commuting conversions are shown in Figures~\ref{??} and~\ref{??}. Notice
+that permutating binders is essential to formulating the majority of these
+rules.
+
 % ...leading on to a discussion of further CP-specific issues encountered.
-
-metatheory library and the library of tactics from software foundations are
-used extensively: metatheory does not handle multiple binders (check this?)
-maybe add a generic tactic to handle multiple binders... or perhaps add a
-lemmas for handling binders
 
@@
 terms
-
-typing judgements
 \end{comment}
@@
 
-While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
-may need to alter the approach slightly to avoid tricky formulations. For
-instance, instead of providing $n$-ary branch and choice in GV, I provide
-only binary versions of these operators to simplify the development. This
-change is not restrictive however, since CP itself provides only binary
-versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
-technical note, I amalgamate all types into one inductive definition and
-provide a predicate for restricting typing rules to consider only valid
-session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
-recursive definitions; session types are types which may contain types as
-subcomponents (arguments to send, for example). Unfortunately, handling
-mutually inductive definitions in Coq can be quite involved; requiring one to
-either rely on the Coq system to provide a stronger mutual induction principle
-or defining one manually. Such a definition is possible, but it complicates
-elimination via the induction principle.
+The typing judgements for the functional language GV are shown in
+Figure~\ref{??}. Note that I have altered the $n$-ary branch and choice
+constructs to the restricted binary versions in order to avoid tricky
+formulations. The change is not restrictive however, since CP itself provides
+only binary versions of plus ($\oplus$) and with ($\with$) constructs.
 
-another issue is the way i tried to formulate types initially treating them as
-indexed families indexed by kinds. while this exploits the dependently typed
-approach it complicated later definitions and proofs. for instance decidablity
-of equality on types now relied on an axiom permitting heterogeneous
-equality. further when formulating the typing environment we must decide which
-kind each type had and most rules are polymorphic in that decision leading to
-the creation of an awkward inspection function. these issues highlighted that
-one should be careful and considerate how the representation will affect later
-expressiveness.
+On a more technical note, all types are amalgamated into one inductive
+definition and provide a predicate for restricting typing rules to consider
+only valid session types. \citeauthor{Wadler:2014} defines GV types as a set
+of mutually recursive definitions; session types are types which may contain
+types as subcomponents (arguments to send, for example). Unfortunately,
+handling mutually inductive definitions in Coq can be quite involved;
+requiring one to either rely on the Coq system to provide a stronger mutual
+induction principle or defining one manually. Such a definition is possible,
+but it complicates applications of the induction principle.
+
+Another issue related to types is how to representing type kinds (linearity or
+non-linearity). Initially the type definition was defined as an indexed
+family, indexed by the kind. While this definition has the advantage of
+preventing certain ill-formed types, such as a non-linear session type, it
+complicated later definitions and proofs. For instance, the decidability of
+equality on types now relied on an axiom permitting heterogeneous
+equality~\cite{??}. Generally, it is better to avoid unnecessary postulates if
+possible. Further, when formulating the typing judgements the kind must be
+given to each type hypothesised and most rules (e.g. ...) are polymorphic in
+the kind, motivating an awkward inspection function to be applied per
+application of a constructor \todo{clean this description up}. A solution to
+this was to define a well-formed type relation and do away with the type
+dependency. these issues highlighted that one should be careful and
+considerate regarding when to use dependent types. these issues highlighted
+that one should be careful and considerate how the representation will affect
+later expressiveness.
 
@@
 directed - permits inductive argument on terms when translating from gv to cp
+
+metatheory library and the library of tactics from software foundations are
+used extensively: metatheory does not handle multiple binders (check this?)
+maybe add a generic tactic to handle multiple binders... or perhaps add a
+lemmas for handling binders
 

### 2015-04-04 paper/formalisation.tex
Freewriting to brainstorm about cut elim, GV, CPS translation.

@@
 The formalisation was performed using the Coq proof
-assisstant~\cite{Coq:manual}. In the spirit of the POPLMARK
+assistant~\cite{Coq:manual}. In the spirit of the POPLMARK
 challenge~\cite{Aydemir:2005:MMM}, the development utilises available
@@
 
+% ... subject reduction....
+
+subject reduction proof proceeds by induction on the reduction relation then
+is automatic by appeal to the proved well-typed reductions.
+
+% ... cut elimination ....
+
+more elaborate than first thought. complicated measure.  as it stands the
+current definitions of well-typed terms and reduction are not amenable to the
+proving of the cut elimination theorem. some method of measure is necessary in
+order to develop some notion of termination and also some relationship to the
+the the underlying subprocesses additionally there is also the issue of the
+representation which includes cases that are not immediately clear how to
+prove for instance the commuting conversions are one such example the
+reduction is only defined if the logical operator corresponds to a free
+variable but induction on the typing judgement leads to two cases one where
+the logical operator is applied to the top-level cut variable it is not
+immediately clear how to eliminate this from happening a relationship between
+the two subprocesses of the cut would need to be stated to prevent them from
+differing so that a principal cut reduction rule may be applied yes only then
+may the the rule be reduced to a cut-free thing if we had a relationship
+between them then there would be some kind of thing that would allow
+destruction on one then fixing the other in the case of a bvar equal to zero
+\todo{how to proceed?}
+
 \subsection{The functional language GV}\label{sec:gv}
@@
 formulations. The change is not restrictive however, since CP itself provides
-only binary versions of plus ($\oplus$) and with ($\with$) constructs.
+only binary versions of plus ($\oplus$) and with ($\with$)
+constructs. Explicit limiting and unlimiting of abstraction was chosen to
+enable the inclusion of these rules in the translation from GV to CP, which is
+performed by induction on GV terms.
 
@@
 
-explicit limiting and unlimiting of abstraction to ensure rules are syntax
-directed - permits inductive argument on terms when translating from gv to cp
-
-metatheory library and the library of tactics from software foundations are
-used extensively: metatheory does not handle multiple binders (check this?)
-maybe add a generic tactic to handle multiple binders... or perhaps add a
-lemmas for handling binders
+similar infrastructure for gv required as for cp copy and paste job
 
 \subsection{CPS translation GV to CP}
+
+require type inference on gv terms to construct corresponding cp terms but we
+are assured that this is okay since the theorem only provides a well-typed gv
+term in a context that is well-defined
 
@@
 defining the function to act on these cases. will this really work? this means
-a graph of positive cases for types then a function definition for it too
+a graph of positive cases for types then a function definition for it too this
+seems rather pathetic because i have to define th einference on terms and it
+seems wrong i have then to handle environments but also nonenvironments i
+can't remember what they are called unbound variables that is it they have to
+be handled as well so we have all this clutter and the types too that is the
+important part but the environments then have to be handled linearly
+unfortunately very bad because we could just use the well typed term relation
+for most of this headachey breaky stuff i'm not sure about that but what is
+can i say so for example lets see see no i also have to thing about the
+general guidelines i can be giving for this section so we have something to
+say about guidelines do we yes of course we have can say that out of this we
+realised that it is important for a representation to be good enough to
+succintly express that which we wish to say in particular the graph relation
+the inductive type that we define propositionally is better one would think
+than the other way the monadic possible failure way is not producing a valid
+thing that we can inductively induct on so we concern ourself with the proof
+of translation preserving the typed term and i think it is important to
+remember the problem we had with the inductive hypothesis we had something do
+with translating the inner term of a weakening term in a larger context than
+the one we had available and that was to do with the destruction of the term
+inference thing i think no no it was something else we had the inner term and
+we had it well typed in an environment that was smaller sure and then our
+thing our function was saying we had some environment that typed both the
+inner term and the outer weakening term but we know that is wrong so wrong so
+here is perhaps where the inductive type graph relation thing wins out in that
+we can have a hypothesis for the inner term which is sufficiently small for
+the environment and we pattern match not only on the term but also on the
+environment yes that is we will do because it seems to work only when we get
+to the messy two the messy two constructor caases will this become a bigg
+problem inthat we will then have all this stuff about environment being split
+and needing to be recombined in the composite process that is a term not a
+process but the point still stands it will need to be about the term and the
+splitting is really a duplication of the unfortunately the well typed term
+relation but otherwise i cannot think how to reuse or obtain that information
+for the translation to cp
 
-translation uses the coq-ext-lib from michela for monadic notation
+Translation uses the coq-ext-lib from michela for monadic notation
 
 Currently the formalisation does not handle this. Future work.
+
+\todo{how might we make progress on this?}
+
+\subsection{Issues and Guidelines}
+
+modularised development; tactics slow prover took a long time to reduce proofs
+so split the structure of the development \todo{picture of development
+  structure}.
+
+tactic performance is an issue to be addressed due to the slow proof search a
+future work (things like deciding equations on finite sets)
+
+One of the design goals of the development was to re-use existing libraries
+wherever possible. two such instances proved extremely useful. first is the
+Metatheory library developed at upenn. second, also from upenn, is the library
+of tactics from software foundations.
+
+software foundations helped in many ways: simplifying tacticals, ease of
+introducing proved assertions, etc. limited use of names introduced by prover
+leading to more robust development.
+
+used extensively: metatheory does not handle multiple binders (check this?)
+maybe add a generic tactic to handle multiple binders... or perhaps add a
+lemmas for handling binders
+
+Linearity create issues with typing some examples; the splitting of the
+environment was non-deterministic so required side-conditions to enforce a
+certain splitting of binders. For example the rule for associativity of
+cut. \todo{how to handle? is there a better way?}
+
+Base rules and rules with binders required permutation assumptions to allow
+any ordering of the constituent contexts. a lot of the proofs for well-typed
+reductions involved small amounts of permuting environments which were largely
+similar across the proofs. while some simple lemmas and tactics where created
+to handle the most tedious elements these did not make effective use of
+hypotheses from the context. Improved proof search for permutations is
+required to simplify such developments.

### 2015-04-05 paper/introduction.tex
Updated introduction. Removed "framework" contribution.
In line with what has actually been achieved. Will come back and
revise further later.

@@
 defined as a bi-directional conduit for exchanging messages, not unlike a
-socket in network programming. In contrast to sockets, however, a channel is
+socket in network programming. In contrast to sockets however, a channel is
 associated with an explicit type which provides a description of its
 protocol. Thus, in session-based systems a channel is usually provided as a
-primitive type.
+primitive of the language.
 
@@
 Proof assistants such as Coq~\cite{Coq:manual} offer a platform on which to
-develop programming languages and type systems and associated proofs of
+develop programming languages, type systems and associated proofs of
 properties with strong guarantees of correctness. Not only is proving with a
@@
 
-This project describes work towards the development of a framework for
-formalising lambda calculus type systems with session types using interactive
-theorem proving (in particular, the system Coq). The framework should enable a
+This project describes work towards the formalisation of calculi developed by
+\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
+classical linear logic (CP) and a functional language with session types (GV)
+using interactive theorem proving (in particular, the system Coq). The
+formalisation is a contribution to the practice of programming language
+metatheory and informs what is required of a framework for formalising lambda
+calculus type systems with session types. Such a framework should enable a
 whole class of related languages and type systems to be formalised with little
-extra proof effort. Inspired by previous work on a mechanised framework for
-$\pi$-calculus type systems~\cite{Gay:2001:FFP}, the current work focuses on
-the lambda calculus to provide a basis on which session types for mainstream
-programming languages can be studied, whilst taking advantage of previous
-proof effort and specialising only for those parts unique to the particular
-system. The starting point for the framework is the calculi developed by
-\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
-classical linear logic (CP) and a functional language with session types (GV).
+extra proof effort and is inspired by previous work on a mechanised framework
+for $\pi$-calculus type systems~\cite{Gay:2001:FFP}. The current work focuses
+on the lambda calculus to provide a basis on which session types for
+mainstream programming languages can be studied, taking advantage of previous
+proof effort through the re-use of existing libraries for programming language
+metatheory. The work has presented a number of interesting challenges for the
+formalisation of programming languages and as a result generated a number of
+useful guidelines and issues to consider.

### 2015-04-05 paper/background.tex
Formatted some typing examples and section headings.
Moved adoption and focus to future work section. Tweaking background.

@@
 The definition of session types will be important for the remainder:
-\input{../paper/session-types.tex}
+\input{session-types}
 
@@
 environment as is standard~\cite{Sangiorgi:2001}.
-
-%\input{../paper/sill-branch-choice-rules}
+\input{sill-branch-choice-rules}
 
@@
 \citeauthor{Caires:2010:STI}~\cite[\S~3]{Caires:2010:STI}:
-\begin{gather*}
-\sdef{ClientProto}{\BChoice{(\tensor{Book}
-                            {(\tensor{Card}{(\lfn{Receipt}{\1})})})}
-                           {(\tensor{Book}{(\lfn{Quote}{\1})})}}
-\end{gather*}
+\input{clientproto}
 
@@
 system. The client and server process bodies are as follows:
-\begin{gather*}
-\sdef{ClientBody_c}
-     {\CPinr{c}{\SLoutput{1984}{c}{\SLinput{c}{quote}{\0}}}}\\
-\sdef{ServerBody_c}
-     {\CPcase{c}
-             {\SLinput{c}{book}{\SLinput{c}{card}{\SLoutput{receipt}{c}{\0}}}}
-             {\SLinput{c}{book}{\SLoutput{quote}{c}{\0}}}}
-\end{gather*}
+\input{csbodies}
 
@@
 \tsequent{\cdot}{\cdot}{ClientBody_c}{\typedid{c}{ClientProto}}
-\tag{\mkTrule{ClientBody}} \label{eq:cbtyp}
+\tag{\mkTrule{Client}} \label{eq:cbtyp}
 \\ \tsequent{\cdot}{\typedid{c}{ClientProto}}{ServerBody_c}{\typedid{-}{\1}}
-\tag{\mkTrule{ServerBody}} \label{eq:sbtyp}
+\tag{\mkTrule{Server}} \label{eq:sbtyp}
 \end{gather*}
@@
 Process reduction steps are equated with proof reduction via cut
-elimination. The relationship with proof normalisation in DILL provides
-absence of deadlocks in session-based communication for the process
+elimination. The relationship with proof normalisation in DILL ensures
+freedom from deadlock in session-based communication for the process
 calculus. The correspondence of session types to linear logic propositions is
@@
 
-The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of
-aliased access to linear resources. The type system requires special language
-features, e.g. to manage capabilities which can be thought as a form of object
-liveness mechanism, so GV typing judgements would need to be extended such
-that the capabilities can be included in soundness results. An extension to
-the translation to CP to incorporate the ``adopt'' and ``focus'' constructs
-would also need to be defined. It is not immediately clear how these features
-would assist in the management of session typed channels. The ``focus'' rule
-(and linearity generally) enforces an invariant on the focussed object
-(channel) such that the type is maintained during the focus operation:
-\input{../proposal/focus-rule}
-
-The typing judgements include \emph{capabilities} that track aliasing to heap
-objects, e.g. $\AFcap{\rho}{h}$ denotes a mapping from a static name $\rho$ to
-a heap type $h$, and a variable $x$ may be typed as $\typedid{x}{\AFtr{\rho}}$
-indicating that $x$ refers to an object of type $h$. The purpose of the
-capabilities is to enforce restrictions on when an object can be
-referenced. In other words, if $\AFcap{\rho}{h}$ is in the capabilities at
-some program point then $x$ may be used, otherwise such use is prohibited. In
-the [focus] rule, evaluating $e_1$ retrieves the object to be focussed. Note
-the type of $e_1$ is a ``guarded'' type. The guard $G$ is the static name of
-some object which ``adopted''~\footnote{I omit the adopt typing rule for
-  brevity.} an object of heap type $h$. If a program point has
-$\AFcap{G}{h_1}$ in the capabilities for some $h_1$ then all objects guarded
-by $G$ may be used; $\capsequent{\Delta}{C_1}{G}$ denotes containment of a
-such a capability in $C_1$. The guarded type may be thought of as policing
-manipulation of linear components within $h$. Treating $\otimes$ as the
-disjoint union for capability sets, the typing for $e_2$ temporarily removes
-the guard $G$ from the set of capabilities, and adds a new variable $x$, and
-associated capability, with a type permitting access to the object of type $h$
-i.e. a non-guarded type. Removing $G$ will prohibit access to allow objects
-guarded by $G$ for the during of $e_2$. The final capabilities after executing
-$e_2$ are required to be $\tensor{C_3}{\AFcap{\rho}{h}}$, enforcing that the
-type must remain as $h$. The typing of the $\letterm$ expression ensures that
-the guard $G$ (contained in $C_1$) is restored.
-
-Such a rule would appear to only benefit recursive session types. Indeed,
-focussing on a non-recursive channel would be pointless since one would be
-prevented from communicating across it since the focus rule requires
-maintenance of the channel type. Further, it would be undesirable to permit
-modification of a session type without guaranteeing exclusive access to the
-channel since this could lead to invalid communication e.g. trying to
-communicate on a closed channel. Additionally, a linear object is consumed
-upon being adopted and will be deallocated when its adopter's lifetime ends,
-although an alternative semantics could be defined which returned the linear
-resource to the context. Moreover, \citeauthor{Fahndrich:2002} do not provide
-operational semantics or soundness results for the ``adoption and focus'' type
-system.
-
 The Alms language~\cite{Aldrich:2009} provides a functional language with
@@
 \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} although treating channels as
-affine rather than linearly typed. \citeauthor{Aldrich:2009} note that Alms'
-type system can support the ``adoption and focus'' system but do not specify
-details or whether any changes are necessary to their typing judgements.
+affine rather than linearly typed.
+
+%% \citeauthor{Aldrich:2009} note that Alms' type system can support the
+%% ``adoption and focus'' system but do not specify details or whether any
+%% changes are necessary to their typing judgements.
 
@@
 
-\subsection{Mechanising Programming Language Metatheory}
+\subsection{Mechanising PL Metatheory}
 
@@
 \citeauthor{Aldrich:2009} but has the advantage that the soundness results
-(type safety and parametricity) have been entirely mechanised in Coq. It would
-be interesting to see if one could encode the ``adopt'' and ``focus''
-constructs in this system to compare its expressiveness to Alms. The ``adopt''
-rule can be encoded similarly to the ShareRef
-example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to encode the
-``focus'' rule within the present semantics, as one needs to prevent
-non-linear access to its argument implying the presence of knowledge about
-aliasing within the typing judgements; absence of aliasing is not identified
-with linear kinded types in \fpop.
+(type safety and parametricity) have been entirely mechanised in Coq.
+
+%%  It would be interesting to see if one could encode the ``adopt'' and
+%% ``focus'' constructs in this system to compare its expressiveness to
+%% Alms. The ``adopt'' rule can be encoded similarly to the ShareRef
+%% example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to encode
+%% the ``focus'' rule within the present semantics, as one needs to prevent
+%% non-linear access to its argument implying the presence of knowledge about
+%% aliasing within the typing judgements; absence of aliasing is not
+%% identified with linear kinded types in \fpop.
 

### 2015-04-05 paper/futurework.tex
Updated paper: CP section and added figures.
Moved adoption and focus to future work section. Tweaking background.

@@
 
-In addition to that already mentioned: there remains much to be done.
+Some areas of future work to look into would be the following.
+
+\subsection{Besides Linear}
+
+The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of
+aliased access to linear resources. The type system requires special language
+features, e.g. to manage capabilities which can be thought as a form of object
+liveness mechanism, so GV typing judgements would need to be extended such
+that the capabilities can be included in soundness results. An extension to
+the translation to CP to incorporate the ``adopt'' and ``focus'' constructs
+would also need to be defined. It is not immediately clear how these features
+would assist in the management of session typed channels. The ``focus'' rule
+(and linearity generally) enforces an invariant on the focussed object
+(channel) such that the type is maintained during the focus operation:
+\input{../proposal/focus-rule}
+
+The typing judgements include \emph{capabilities} that track aliasing to heap
+objects, e.g. $\AFcap{\rho}{h}$ denotes a mapping from a static name $\rho$ to
+a heap type $h$, and a variable $x$ may be typed as $\typedid{x}{\AFtr{\rho}}$
+indicating that $x$ refers to an object of type $h$. The purpose of the
+capabilities is to enforce restrictions on when an object can be
+referenced. In other words, if $\AFcap{\rho}{h}$ is in the capabilities at
+some program point then $x$ may be used, otherwise such use is prohibited. In
+the [focus] rule, evaluating $e_1$ retrieves the object to be focussed. Note
+the type of $e_1$ is a ``guarded'' type. The guard $G$ is the static name of
+some object which ``adopted''~\footnote{I omit the adopt typing rule for
+  brevity.} an object of heap type $h$. If a program point has
+$\AFcap{G}{h_1}$ in the capabilities for some $h_1$ then all objects guarded
+by $G$ may be used; $\capsequent{\Delta}{C_1}{G}$ denotes containment of a
+such a capability in $C_1$. The guarded type may be thought of as policing
+manipulation of linear components within $h$. Treating $\otimes$ as the
+disjoint union for capability sets, the typing for $e_2$ temporarily removes
+the guard $G$ from the set of capabilities, and adds a new variable $x$, and
+associated capability, with a type permitting access to the object of type $h$
+i.e. a non-guarded type. Removing $G$ will prohibit access to allow objects
+guarded by $G$ for the during of $e_2$. The final capabilities after executing
+$e_2$ are required to be $\tensor{C_3}{\AFcap{\rho}{h}}$, enforcing that the
+type must remain as $h$. The typing of the $\letterm$ expression ensures that
+the guard $G$ (contained in $C_1$) is restored.
+
+Such a rule would appear to only benefit recursive session types. Indeed,
+focussing on a non-recursive channel would be pointless since one would be
+prevented from communicating across it since the focus rule requires
+maintenance of the channel type. Further, it would be undesirable to permit
+modification of a session type without guaranteeing exclusive access to the
+channel since this could lead to invalid communication e.g. trying to
+communicate on a closed channel. Additionally, a linear object is consumed
+upon being adopted and will be deallocated when its adopter's lifetime ends,
+although an alternative semantics could be defined which returned the linear
+resource to the context. Moreover, \citeauthor{Fahndrich:2002} do not provide
+operational semantics or soundness results for the ``adoption and focus'' type
+system.
+
+\citeauthor{??} extend \citeauthor{Wadler:2014}'s work to support polymorphism
+in GV. (polymorphism in CP) Future work is to support reduction under these
+terms.

### 2015-04-06 paper/background.tex
Tidying up background. Adding figures for CP/GV types and terms.
I think this is important background for understanding the
languages I am aiming to formalise.

@@
 
-The work on a linear functional language with session types by
-\citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the
-current project from the body of work on session-based $\lambda$-calculus type
-systems. They provide paper proofs of their system's type safety. Given the
-non-trivial features of the type system such as recursive session types and
-buffered channels for supporting asynchronous communication it will be
-challenging to extend these proofs manually. A mechanised version would
-provide stronger guarantees of correctness and allow others to more easily
-alter proofs as a result of changes made to the language.
+The work on an linear functional language with asynchronous session types
+(LAST) by \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main
+influence for the current project from the body of work on session-based
+$\lambda$-calculus type systems. They provide paper proofs of their system's
+type safety. Given the non-trivial features of the type system such as
+recursive session types and buffered channels for supporting asynchronous
+communication it will be challenging to extend these proofs manually. A
+mechanised version would provide stronger guarantees of correctness and allow
+others to more easily alter proofs as a result of changes made to the
+language.
 
@@
 as a continuation and the type denotes the communication protocol
-(i.e. session type) between parent and child.
+(i.e. session type) between parent and child \todo{what part of session type
+  is being discussed here? spawning process or spawning operation?}.
 
@@
 acyclic communication graphs (progress of process reduction) which occur
-between parallel processes; follows from permitting only one channel between
-two halves of parallel composition construct in the process calculus. Most of
-the development is supported by Coq proof scripts with remaining issues due to
-reasoning about communication graphs. The development does not provide
-primitive send and receives; instead, these are encoded as linear functions.
-%% and classical extension to second-order logic by
-%% \citeauthor{Mazurak:2013:LPP}'s PhD thesis~\cite{Mazurak:2013:LPP}.
+between parallel processes where progress follows from permitting only one
+shared channel between the two halves of a parallel composition in the process
+calculus. Most of the development is supported by Coq proof scripts with the
+exception being reasoning about communication graphs. The development does not
+provide primitive send and receives; instead, these are encoded as linear
+functions.
+
+\input{gv-types-terms}
 
@@
 \citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
-\citeauthor{Gay:2010:LAST}. GV presents session types using the standard
-presentation as described in \S~\ref{sec:pre}. CP presents linear logical
-connectives with session type interpretations inspired by
-\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI}. \citeauthor{Wadler:2014}
-describes a continuation-passing style (CPS) translation from GV session types
-to CP classical linear logic propositions. The process calculus CP is slightly
-different to $\pi$DILL by \citeauthor{Caires:2010:STI}, apart from being
-classical in nature, the constructs for server replication are modified; CP
-provides a weakening and contraction rules whereas these are encoded in one
-rule for process reduction by \citeauthor{Caires:2010:STI}. Additionally,
-rather than a two-sided sequent calculus, \citeauthor{Wadler:2014} uses
-one-sided sequents for defining the process calculus, which provides a more
-intuitive presentation of session type duality and clearer correspondence to a
-$\lambda$-calculi surface syntax, simplifying the translation presentation. As
-presented, GV does not support all features of CP and thus the translation is
-one way (GV to CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP}
-extend GV to provide support for polymorphism and replication, following a
-similar CPS translation scheme. Lastly,
+\citeauthor{Gay:2010:LAST}. GV presents session types using a modified version
+of the standard definition presented in \S~\ref{sec:pre}. Figure~\ref{??}
+presents the types of GV with session types ranged over by $S$. Note that
+\lstinline{end} has been replaced with $\outend$ and $\inend$ denoting the
+terminating session type for $\cdots$ and $\cdots$ respectively \todo{what are
+  the interpretations for $\outend$ and $\inend$? Need a figure for GV
+  sessions}. CP presents linear logical connectives with session type
+interpretations inspired by
+\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI} shown in Figure~\ref{??}
+\todo{figure for CP types with session type
+  interpretation}. \citeauthor{Wadler:2014} describes a continuation-passing
+style (CPS) translation from GV session types to CP classical linear logic
+propositions.  \todo{terms and interpretations for CP/GV?} The process
+calculus CP is slightly different to $\pi$DILL by
+\citeauthor{Caires:2010:STI}, apart from being classical in nature, the
+constructs for server replication are modified; CP provides separate weakening
+and contraction rules whereas these are encoded in one rule for process
+reduction in $\pi$DILL. Additionally, rather than a two-sided sequent
+calculus, \citeauthor{Wadler:2014} uses one-sided sequents for defining CP,
+which provides a more intuitive presentation of session type duality and a
+clearer correspondence to a $\lambda$-calculi surface syntax, simplifying the
+translation presentation. As presented, GV does not support all features of CP
+and thus the translation is one way (GV to
+CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to
+provide support for polymorphism and replication, following a similar CPS
+translation scheme. Lastly,
 \citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational
@@
 
-\subsection{Mechanising PL Metatheory}
+\subsection{Mechanising Metatheory}
 
-Interactive theorem proving and proof assistants have seen an upsurge in use
-in recent years. Especially within the domain of programming language
+Interactive theorem proving and proof assistants have experienced an upsurge
+in use in recent years. Especially within the domain of programming language
 research, there is a strong emphasis on providing mechanised proofs
@@
   zero for the innermost binder). For example, $\lambda~(\lambda~1)$
-  corresponds to $\lambda x. \lambda y. x$, i.e. the constant function.} and
-free variables are represented as named terms; the \textit{locally nameless}
-representation. Their work resulted in a library for handling commonly
-occurring aspects in programming language metatheory which has been used since
-in other developments~\cite{Park:2014:MMW}. Reusable libraries reduce the
-proof engineering effort needed for later works, in contrast to the
+  corresponds to $\lambda x. \lambda y. x$, i.e. the \textbf{K} combinator.}
+and free variables are represented as names; the \textit{locally nameless}
+representation~\cite{Char:2012}. Their work resulted in a library for handling
+commonly occurring aspects in programming language metatheory which has been
+used since in other developments~\cite{Park:2014:MMW}. Reusable libraries
+reduce the proof engineering effort needed for later works, in contrast to the
 pen-and-paper approach, and some proof assistants (as is the case with Coq)
 allow one to extract a program from the development which can provide a
-typechecker or compiler for the language formalised.~\footnote{Curry-Howard
-  strikes again; a proof of decidability of typechecking corresponds to a
-  program implementing a typechecker for the language.}
+typechecker or compiler for the language formalised.
 
@@
 an extension to System F (termed \fpop) with a notion describing types as
-having linear or non-linear \textit{kind}~\cite{Mazurak:2010:LLT}. This
-extension is motivated by previous work on incorporating linearity into type
-systems. Overall, previous attempts either hamper the inclusion of desirable
-programming features (such as polymorphism), or do not adequately reflect
-non-linearity as the common case leading to awkward programming. \fpop avoids
-these deficiencies by categorising types into kinds whilst maintaining similar
-semantics to System F. Mechanised proofs for type soundness and parametricity
-are presented, and the semantics allow non-linear types to be treated as
-linear. Examples show the system can provide a wide range of permissions on
-interfaces, from full exclusive access to read-only shared references. The
-system is very close to the Alms language described by
-\citeauthor{Aldrich:2009} but has the advantage that the soundness results
-(type safety and parametricity) have been entirely mechanised in Coq.
+having linear or non-linear kind~\cite{Mazurak:2010:LLT}. This extension is
+motivated by previous work on incorporating linearity into type systems which
+either hampered the inclusion of desirable programming features (such as
+polymorphism), or did not adequately reflect non-linearity as the common case
+leading to awkward programming. \fpop avoids these deficiencies by
+categorising types into kinds whilst maintaining similar semantics to System
+F. Mechanised proofs for type soundness and parametricity are presented, and
+the semantics allow non-linear types to be treated as linear. Examples show
+the system can provide a wide range of permissions on interfaces, from full
+exclusive access to read-only shared references. The system is very close to
+the Alms language but has the advantage that the soundness results (type
+safety and parametricity) have been entirely mechanised in Coq.
 
-%%  It would be interesting to see if one could encode the ``adopt'' and
-%% ``focus'' constructs in this system to compare its expressiveness to
-%% Alms. The ``adopt'' rule can be encoded similarly to the ShareRef
-%% example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to encode
-%% the ``focus'' rule within the present semantics, as one needs to prevent
-%% non-linear access to its argument implying the presence of knowledge about
-%% aliasing within the typing judgements; absence of aliasing is not
-%% identified with linear kinded types in \fpop.
-
-While not related to programming language metatheory,
 \citeauthor{Gay:2001:FFP}~\cite{Gay:2001:FFP} provides a $\pi$-calculus
@@
 de Bruijn indices for both bound and free variables causes issues with
-variable substitution often requiring permutation of typing environments (see
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} for a discussion on
+variable substitution often requiring permutation of typing environments
+(\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} provide a discussion on
 binder representations) and lifting of free variables during substitution; I

### 2015-04-07 paper/background.tex
Added beginnings of CP propositions and processes.
Need to ask Simon what he thinks about all these figures.
Some further discussion of permutation issues.
Perhaps an example of permutation?

@@
 
+\input{cp-types-terms}
+
 \citeauthor{Wadler:2014}'s recent work~\cite{Wadler:2014} brings together a
 process calculus, CP, based on the earlier work by
-\citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by
-\citeauthor{Gay:2010:LAST}. GV presents session types using a modified version
-of the standard definition presented in \S~\ref{sec:pre}. Figure~\ref{??}
-presents the types of GV with session types ranged over by $S$. Note that
-\lstinline{end} has been replaced with $\outend$ and $\inend$ denoting the
-terminating session type for $\cdots$ and $\cdots$ respectively \todo{what are
-  the interpretations for $\outend$ and $\inend$? Need a figure for GV
-  sessions}. CP presents linear logical connectives with session type
+\citeauthor{Caires:2010:STI}, and a functional language, GV, based on LAST by
+\citeauthor{Gay:2010:LAST}. Figure~\ref{fig:gvtnts} presents the types and
+terms GV with types ranged over by $T$, session types ranged over by $S$ and
+terms ranged over by $L$, $M$ and $N$. GV session types are a modified version
+of the definition presented in \S~\ref{sec:pre} where \lstinline{end} has been
+replaced with $\outend$ and $\inend$. The intention is that $\outend$
+($\inend$) terminates a session type whose last operation is an output
+(input), and they correspond to the units of the respective linear logical
+connectives \todo{Do I need to elaborate on the translation? Yet another
+  figure?}. CP provides linear logical connectives with session type
 interpretations inspired by
-\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI} shown in Figure~\ref{??}
-\todo{figure for CP types with session type
-  interpretation}. \citeauthor{Wadler:2014} describes a continuation-passing
-style (CPS) translation from GV session types to CP classical linear logic
-propositions.  \todo{terms and interpretations for CP/GV?} The process
-calculus CP is slightly different to $\pi$DILL by
-\citeauthor{Caires:2010:STI}, apart from being classical in nature, the
-constructs for server replication are modified; CP provides separate weakening
-and contraction rules whereas these are encoded in one rule for process
-reduction in $\pi$DILL. Additionally, rather than a two-sided sequent
-calculus, \citeauthor{Wadler:2014} uses one-sided sequents for defining CP,
-which provides a more intuitive presentation of session type duality and a
-clearer correspondence to a $\lambda$-calculi surface syntax, simplifying the
-translation presentation. As presented, GV does not support all features of CP
-and thus the translation is one way (GV to
+\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI} shown in
+Figure~\ref{fig:cptnts} along with the terms of CP. The process calculus CP is
+slightly different to $\pi$DILL by \citeauthor{Caires:2010:STI}, apart from
+being classical in nature, the constructs for server replication are modified:
+CP provides separate weakening and contraction rules whereas these are encoded
+in one rule for process reduction in $\pi$DILL. Additionally, rather than a
+two-sided sequent calculus, \citeauthor{Wadler:2014} uses one-sided sequents
+for defining CP, which provides a more intuitive presentation of session type
+duality and a clearer correspondence to a $\lambda$-calculi surface syntax
+like GV. \citeauthor{Wadler:2014} describes a continuation-passing style (CPS)
+translation from GV types and terms to CP classical linear logic propositions
+and processes, respectively. As presented, GV does not support all features of
+CP and thus the translation is one way (GV to
 CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to
-provide support for polymorphism and replication, following a similar CPS
+provide support for polymorphism and replication following a similar CPS
 translation scheme. Lastly,

### 2015-04-07 paper/formalisation.tex
Moved adoption and focus to future work section. Tweaking background.
Some further discussion of permutation issues.
Perhaps an example of permutation?

@@
 their language proof infrastructure is proportional to the number of binding
-constructs. In the case of a simply typed lambda calculus this is not onerous,
-but GV has four binding constructs, CP has six and propositions have
-two. Indeed, a lot of effort was expended setting up this
+constructs. In the case of a simply typed $\lambda$-calculus ($\stlc$) this is
+not onerous, but GV has four binding constructs, CP has six and propositions
+have two. Indeed, a lot of effort was expended setting up this
 infrastructure. However, I believe one of the contributions of this work is to
@@
 
-propositions
-
-processes
-
 equivalences and axiomatic cut
@@
 
-Formalising the process calculus CP provided many challenges for the
-metatheory library since to my knowledge, the library has never been applied
-to a process calculus before. So for instance, in the lambda calculus one
-typically does not work up to permutation of binders, whereas in a process
-calculus these kinds of structural equivalences are standard. Additionally,
-substitution and opening are only defined for names. In other words, one
-cannot substitute an arbitrary process term for a name, unlike in the lambda
-calculus. Thus frameworks which assume a lambda calculi style operations are
-not immediately amenable to the CP setting~\cite{Lee:2012}. Ordering in
-environments is ignored so the development required lemmas allowing the
-permutation of environments. Support for permutating environments is not
-currently supported by the Metatheory library but some general lemmas were
-added as a result of this work. Additionally, custom tactics were developed to
-automatatically solve some simple permutation goals, however these tactics are
-not a ``catch all'' and some generalising still needs to be done.
-
 % Description of the typing rules...
 
-Figure~\ref{fig:cll} presents the process calculus CP in LNCQ form (ignoring
-permutation premises). In other words bound variables are represented using de
-Brujin indices and free variables are represented using an abstract type
-provided by the Metatheory library. The rules with binders reflect this by
-specifying the type of a bound variable in the constructor rather than the
-name. Cofinite quantification is performed in the hypotheses to these rules by
-excluding a finite set of names from opening the subprocesses. Note that the
-rules for polymorphism, $\exists$ and $\forall$, were included in the final
-Coq development since GV (\S~\ref{sec:gv}) does not support polymorphism
-making the rules unnecessary for the translation. They appear in the Figure
-simply for the sake of completeness. Additionally, as noted by
-\citeauthor{Lee:2012}, support for polymorphism in type variables creates an
-explosion of the infrastructure requiring duplication of opening and
-substituting definitions and lemmas.
+Figure~\ref{fig:cll} presents the process calculus CP in LNCQ form with some
+tedious details relating to environment uniqueness and ordering (see below)
+elided. In other words, bound variables are represented using de Brujin
+indices and free variables are represented using an abstract type provided by
+the Metatheory library. The rules with binders reflect this by specifying the
+type of a bound variable in the constructor rather than the name. Cofinite
+quantification is performed in the hypotheses to these rules by excluding a
+finite set of names from opening the subprocesses where opening a process $P$
+with name $x$ is denoted $P^x$. Note that the rules for polymorphism,
+$\exists$ and $\forall$, were included in the final Coq development since GV
+(\S~\ref{sec:gv}) does not support polymorphism making the rules unnecessary
+for the translation. Further, as noted by \citeauthor{Lee:2012}, support for
+polymorphism in type variables creates an explosion of the infrastructure
+requiring additional of opening and substituting definitions and lemmas. They
+appear in the Figure simply for the sake of completeness.
+
+Formalising the process calculus CP provided many challenges for the
+Metatheory library since to my knowledge, the library has never been applied
+to a process calculus before, and the treatment of names, binding and
+substitution differ to that present in $\lambda$-calculi. For instance, in the
+$\stlc$ one typically does not work up to permutation of binders, whereas in a
+process calculus these kinds of structural equivalences are
+standard. Additionally, substitution and opening are only defined for
+names. In other words, one cannot substitute an arbitrary process term for a
+name, unlike in the $\lambda$-calculus. Thus frameworks which assume
+$\lambda$-calculi style operations are not immediately amenable to the CP
+setting~\cite{Lee:2012}. Ordering in environments is ignored so the
+development required lemmas allowing the permutation of
+environments. Permutating environments is not currently supported by the
+Metatheory library but some general lemmas were added as a result of this
+work, and custom tactics were developed to automatically solve certain
+permutation goals via simple proof search. However, there are still many cases
+where manual sequences of primitive lemmas, such as appending, commutativity
+and transitivity, need to be used. A lot of these cases are similar but there
+would not appear to be an advantage in developing a tactic to do the
+work. Doing so would create a dependency between a hard-to-debug and obscure
+Ltac definition and the proofs which utilise it, becoming a burden to
+change. This development philosophy treats Ltac definitions akin to macros, a
+shorthand for expressing an oft-repeated sequence of tactics, and it is
+doubtful much can be gained from it unless the use of Ltac definitions can be
+made more transparent.
 
@@
 of the difficulty in specifying its principal cut elimination rule. In order
-to specify the rule one requires to show the following:
+to specify the rule one requires to show the following: \todo{fix binders of
+  Q}
+\begin{gather*}
+\tm{\nu A.(!0(A).P \mid ?z[z',z''].Q)}
+\quad \becomes \quad \\
+\tm{\nu A.(!0(A).P \mid \nu A.(!0(A).P \mid \key{contract}_{\Gamma} Q))}
+\end{gather*}
 
-\todo{v A -> !0(A).P | Q{x/x'}
-  ==>
-v A -> !0(A).P | v A -> !0(A).P | contract (Gamma) Q}
+where $\tm{P^y} \tpvdash \tp{{?\Gamma} \comma \tmof{y}A}$ for some
+sufficiently fresh name $y$ and $\key{contract}_{\Gamma} \tm{Q}$ contracts
+$\tm{Q}$ with the names in $\Gamma$.
 
@@
 context (cf. an arbitrary context in \citeauthor{Wadler:2014}'s
-presentation~\cite{Wadler:2014}).
+presentation~\cite{Wadler:2014}). one can compare alms with this version of cp
+since it weakening rule corresponds to affine logic
 
@@
 
-The principal cut elimination rules are presented in Figure~\ref{??}. These
-rules are expressed to indicate some additional complexity not present in the
-original presentation. For example, structural equivalences and associativity
-are not unique and different process terms result depending on initial binder
-permutation. ${0 <~> 1}P$ indicates the permuting of binders $0$ and $1$ in
-process P. The paper presentation specifies only one possible associativity
-rule lending to the specific split of the linear environments. The reduction
-rules were complicated to prove due to being composed of multiple constructor
-instances and differed significantly to the standard operational semantics one
-typically deals with in lambda calculus formalisms.
-linear environments caused a problem with the splitting up of variables.
+The structural equvialences and principal cut elimination rules are presented
+in Figures~\ref{??} and \ref{??} respectively. These rules are expressed to
+indicate some additional complexity not present in the original
+presentation. For example, structural equivalences are not unique and
+different process terms result depending on initial binder permutation. ${0
+  <~> 1}P$ indicates the permuting of binders $0$ and $1$ in process P. The
+paper presentation specifies only one possible associativity rule lending to
+the specific split of the linear environments. The reduction rules were
+complicated to prove due to being composed of multiple constructor instances
+and differed significantly to the standard operational semantics one typically
+deals with in lambda calculus formalisms.  linear environments caused a
+problem with the splitting up of variables.
 
@@
 
-% ... subject reduction....
+% ... subject reduction ...
 
@@
 \subsection{The functional language GV}\label{sec:gv}
-
-\begin{comment}
-FIGURES
-
-types
-
-terms
-\end{comment}
 

### 2015-04-07 paper/introduction.tex
Some further discussion of permutation issues.
Perhaps an example of permutation?

@@
 formalisation is a contribution to the practice of programming language
-metatheory and informs what is required of a framework for formalising lambda
-calculus type systems with session types. Such a framework should enable a
-whole class of related languages and type systems to be formalised with little
-extra proof effort and is inspired by previous work on a mechanised framework
-for $\pi$-calculus type systems~\cite{Gay:2001:FFP}. The current work focuses
-on the lambda calculus to provide a basis on which session types for
-mainstream programming languages can be studied, taking advantage of previous
-proof effort through the re-use of existing libraries for programming language
-metatheory. The work has presented a number of interesting challenges for the
-formalisation of programming languages and as a result generated a number of
-useful guidelines and issues to consider.
+metatheory and informs what is required of a framework for formalising
+$\lambda$-calculus type systems with session types. Such a framework should
+enable a whole class of related languages and type systems to be formalised
+with little extra proof effort and is inspired by previous work on a
+mechanised framework for $\pi$-calculus type systems~\cite{Gay:2001:FFP}. The
+current work focuses on the $\lambda$-calculus to provide a basis on which
+session types for mainstream programming languages can be studied, taking
+advantage of previous proof effort through the re-use of existing libraries
+for programming language metatheory. The work has presented a number of
+interesting challenges for the formalisation of programming languages and as a
+result generated a number of useful guidelines and issues to consider.

### 2015-04-07 paper/formalisation.tex
Further additions/structural changes to formalisation section.

@@
 processes nor the semantics of the translation from GV to CP. An equivalence
-between ``exists-fresh'' and cofinite quantification definitions of the
-simply-typed $\lambda$-calculus is given by
+between ``exists-fresh'' and cofinite quantification definitions for the
+simply-typed $\lambda$-calculus ($\stlc$) is given by
 \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
@@
 being constructed. However, there are issues with using this encoding for
-CP/GV: $(1)$ it is not immediately clear how to handle linear contexts using
-the de Bruijn variable encoding, since in the intrinsic setting the
-environment automatically supports weakening; $(2)$ an intrinsic encoding of
-GV terms would be complicated by the need to enforce that session types occur
-in certain instances (by use of a predicate); and $(3)$ well-typed terms
-require extra assumptions about binder freshness which cannot reasonably be
-expressed as a function type. For these reasons, an intrinsic approach does
-not offer much benefit since a well-typed term relation would still need to be
-defined. Therefore, I chose an extrinsic encoding for terms and define
-well-typed terms as a separate inductive type.
+CP/GV:
+\begin{enumerate}
+\item It is not immediately clear how to handle linear contexts using the de
+  Bruijn variable encoding, since in the intrinsic setting the environment
+  automatically supports weakening
+\item An intrinsic encoding of GV terms would be complicated by the need to
+  enforce that session types occur in certain instances (by use of a
+  predicate)
+\item Well-typed terms require extra assumptions about binder freshness which
+  cannot reasonably be expressed as a function type.
+\end{enumerate}
+
+For these reasons, an intrinsic approach does not offer much benefit since a
+well-typed term relation would still need to be defined. Therefore, I chose an
+extrinsic encoding for terms and define well-typed terms as a separate
+inductive type.
 
@@
 their language proof infrastructure is proportional to the number of binding
-constructs. In the case of a simply typed $\lambda$-calculus ($\stlc$) this is
-not onerous, but GV has four binding constructs, CP has six and propositions
-have two. Indeed, a lot of effort was expended setting up this
-infrastructure. However, I believe one of the contributions of this work is to
-articulate what one needs from state of the art metatheory libraries in order
-to handle session-based languages, and more generally, linear type systems.
+constructs. In the case of a $\stlc$ this is not onerous, but GV has four
+binding constructs, CP has six and propositions have two. Indeed, a lot of
+effort was expended setting up this infrastructure. However, I believe one of
+the contributions of this work is to articulate what one needs from state of
+the art metatheory libraries in order to handle session-based languages, and
+more generally, linear type systems.
 
@@
 the Metatheory library. The rules with binders reflect this by specifying the
-type of a bound variable in the constructor rather than the name. Cofinite
+type of a bound variable in the constructor rather than a name. Cofinite
 quantification is performed in the hypotheses to these rules by excluding a
@@
 with name $x$ is denoted $P^x$. Note that the rules for polymorphism,
-$\exists$ and $\forall$, were included in the final Coq development since GV
-(\S~\ref{sec:gv}) does not support polymorphism making the rules unnecessary
-for the translation. Further, as noted by \citeauthor{Lee:2012}, support for
-polymorphism in type variables creates an explosion of the infrastructure
-requiring additional of opening and substituting definitions and lemmas. They
-appear in the Figure simply for the sake of completeness.
+$\exists$ and $\forall$, were not included in the final Coq development since
+GV (\S~\ref{sec:gv}) does not support polymorphism making the rules
+unnecessary for the translation. Further, as noted by \citeauthor{Lee:2012},
+support for polymorphism in type variables creates an explosion of the
+infrastructure requiring additional of opening and substituting definitions
+and lemmas. They appear in the Figure simply for the sake of completeness.
 
@@
 $\lambda$-calculi style operations are not immediately amenable to the CP
-setting~\cite{Lee:2012}. Ordering in environments is ignored so the
-development required lemmas allowing the permutation of
-environments. Permutating environments is not currently supported by the
-Metatheory library but some general lemmas were added as a result of this
-work, and custom tactics were developed to automatically solve certain
-permutation goals via simple proof search. However, there are still many cases
-where manual sequences of primitive lemmas, such as appending, commutativity
-and transitivity, need to be used. A lot of these cases are similar but there
-would not appear to be an advantage in developing a tactic to do the
-work. Doing so would create a dependency between a hard-to-debug and obscure
-Ltac definition and the proofs which utilise it, becoming a burden to
-change. This development philosophy treats Ltac definitions akin to macros, a
-shorthand for expressing an oft-repeated sequence of tactics, and it is
-doubtful much can be gained from it unless the use of Ltac definitions can be
-made more transparent.
+setting~\cite{Lee:2012}.
 
-% ...leading on to the discussion of the explicit weakening...
+Ordering in environments is ignored so the development required operations for
+handling permutations of environments. The Metatheory library does not
+currently support permutating environments, however some basic lemmas were
+added as a result of this work, and custom tactics were developed to
+automatically solve certain permutation goals via simple proof search. Even
+so, there are still many cases where sequences of primitive lemmas, such as
+those handling append, commutativity and transitivity, need to be applied
+manually. A lot of these cases are similar but there would not appear to be an
+advantage in developing a tactic to do the work. Doing so would create a
+dependency between a hard-to-debug and obscure Ltac definition and the proofs
+which utilise it, becoming a burden to change. Further, such a development
+philosophy treats Ltac tactics akin to macros, a shorthand for expressing an
+oft-repeated sequence of tactics, and it is doubtful much can be gained from
+it unless the execution of Ltac tactics can be made more transparent.
+
+\input{principalcp}
+
+The principal cut elimination rules are presented in Figure~\ref{??}. The
+reduction rules were complicated to prove due to being composed of multiple
+constructor instances and differ significantly to the standard operational
+semantics one typically deals with in $\lambda$-calculus formalisms. As a
+result, these well-typed reductions were not proved ``in-place'', instead each
+rule is proved as a separate lemma. Most proofs followed a pattern but it was
+difficult to automate satisfactorily for the reasons stated in the previous
+paragraph. However, some tasks could be delegated to a custom tactic, for
+example instantiating cofinitely quantified hypotheses with a newly introduced
+fresh variable.
 
@@
 required to differentiate the principal cut elimination rules for the various
-interactions between servers and clients. contraction is not supported because
-of the difficulty in specifying its principal cut elimination rule. In order
-to specify the rule one requires to show the following: \todo{fix binders of
-  Q}
+interactions between servers and clients ($\beta_{!?}$ and $\beta_{!W}$). The
+paper presentation does not specify where weakening could occur; it is tacitly
+assumed weakening can always be pushed further up the derivation tree to a
+base rule e.g. $\Ax\Cut$ rule. Contraction is not supported in the
+formalisation due to the difficulty in specifying its principal cut
+elimination rule. In order to specify the rule one requires to show the
+following:
 \begin{gather*}
-\tm{\nu A.(!0(A).P \mid ?z[z',z''].Q)}
+\tm{\nu A.(!0(A).P \mid ?z[[A]].Q)}
 \quad \becomes \quad \\
@@
 
-where $\tm{P^y} \tpvdash \tp{{?\Gamma} \comma \tmof{y}A}$ for some
+where $\tm{?z[[A]].Q}$ denotes contraction by $z$ of two channels of type $A$
+in $Q$, $\tm{P^y} \tpvdash \tp{{?\Gamma} \comma \tmof{y}A}$ for some
 sufficiently fresh name $y$ and $\key{contract}_{\Gamma} \tm{Q}$ contracts
@@
 Unfortunately, the reduction relation does not have knowledge of the
-environment typing process P. This fact does not impact upon expressing
-weakening since for weakening only the names (not the bindings) within the
-environment are required. The paper presentation does not specify where
-weakening could occur; it is tacitly assumed weakening can always be pushed
-further up the derivation tree e.g. AxCut rule. a consequence of this choice
-is that for the empty choice rule there may only be a single binding in the
-context (cf. an arbitrary context in \citeauthor{Wadler:2014}'s
-presentation~\cite{Wadler:2014}). one can compare alms with this version of cp
+environment, $\Gamma$, typing process $\tm{P}$. This fact does not impact upon
+expressing weakening because only the names (not the types in the bindings)
+within the environment are required for weakening. one can compare alms with this version of cp
 since it weakening rule corresponds to affine logic
 
-% ...leading on to presentation of principal cut elimination rules...
-
-\input{principalcp}
-
-The structural equvialences and principal cut elimination rules are presented
-in Figures~\ref{??} and \ref{??} respectively. These rules are expressed to
-indicate some additional complexity not present in the original
-presentation. For example, structural equivalences are not unique and
-different process terms result depending on initial binder permutation. ${0
-  <~> 1}P$ indicates the permuting of binders $0$ and $1$ in process P. The
-paper presentation specifies only one possible associativity rule lending to
-the specific split of the linear environments. The reduction rules were
-complicated to prove due to being composed of multiple constructor instances
-and differed significantly to the standard operational semantics one typically
-deals with in lambda calculus formalisms.  linear environments caused a
-problem with the splitting up of variables.
+In the empty choice rule there may only be a single binding in the context
+(cf. an arbitrary context in \citeauthor{Wadler:2014}'s
+presentation~\cite{Wadler:2014}). While this diverges with the presentation of
+linear logic it became necessary to do this because otherwise there would be
+no relationship between the free variables within a process and its
+environment. no there is a clear iff relationship which permits reduction
+rules to apply such as i think weakening but it should be any rule which takes
+some fvs of proc and infers something about environments from that...otherwise
+what it is not at all clear how to encode the reductions in a relation, the
+subprocesses environments are created from destructing the environment
 
 % ...commuting conversions
+
+\input{cpstructeq}
 
@@
 
-The commuting conversions are shown in Figures~\ref{??} and~\ref{??}. Notice
-that permutating binders is essential to formulating the majority of these
-rules.
-
-% ...leading on to a discussion of further CP-specific issues encountered.
+The structural equivalences and commuting conversions are shown in
+Figure~\ref{??} and Figures~\ref{??} and~\ref{??}, respectively. These rules
+are expressed to indicate some additional complexity not present in the
+original presentation. For example, structural equivalences are not unique and
+different process terms result depending on initial binder
+permutation. $\permb{0}{1}$ indicates the permuting of binders $0$ and $1$ in
+process $\tm{P}$. The paper presentation specifies only one possible
+associativity rule lending to the specific split of the linear environments.
+Notice that permutating binders is essential to formulating the majority of
+these rules. Analysing permutations containing singleton bindings was
+necessary for the commuting conversion rules. Singleton bindings can occur
+frequently during a proof and it is sometimes necessary to analyse a
+permutation of the form P++x~A ~~ Q++R to determine whether x occurs in Q or
+R. the rule for associativity requires this and so do quite a few reduction
+rules. since this involves some element of searching the context followed by
+destructing hypotheses it could be effectively automated in permutations in
+subprocesses which use a binder of the parent. the rule for associativity w
+linear environments caused a problem with the splitting up of variables.
 

### 2015-04-08 paper/formalisation.tex
Added Coq code to be described.
Also removed dependency of paper folder on the proposal items.

@@
 Figure~\ref{fig:cll} presents the process calculus CP in LNCQ form with some
-tedious details relating to environment uniqueness and ordering (see below)
-elided. In other words, bound variables are represented using de Brujin
-indices and free variables are represented using an abstract type provided by
-the Metatheory library. The rules with binders reflect this by specifying the
-type of a bound variable in the constructor rather than a name. Cofinite
-quantification is performed in the hypotheses to these rules by excluding a
-finite set of names from opening the subprocesses where opening a process $P$
-with name $x$ is denoted $P^x$. Note that the rules for polymorphism,
-$\exists$ and $\forall$, were not included in the final Coq development since
-GV (\S~\ref{sec:gv}) does not support polymorphism making the rules
-unnecessary for the translation. Further, as noted by \citeauthor{Lee:2012},
-support for polymorphism in type variables creates an explosion of the
-infrastructure requiring additional of opening and substituting definitions
-and lemmas. They appear in the Figure simply for the sake of completeness.
+tedious details relating to environment uniqueness elided. In other words,
+bound variables are represented using de Bruijn indices and free variables are
+represented using an abstract type provided by the Metatheory library. The
+rules with binders reflect this by specifying the type of a bound variable in
+the constructor rather than a name. Cofinite quantification is performed in
+the hypotheses to these rules by excluding a finite set of names from opening
+the subprocesses where opening a process $P$ with name $x$ is denoted
+$P^x$. Note that the rules for polymorphism, $\exists$ and $\forall$, were not
+included in the final Coq development since GV (\S~\ref{sec:gv}) does not
+support polymorphism making the rules unnecessary for the
+translation. Further, as noted by \citeauthor{Lee:2012}, support for
+polymorphism in type variables creates an explosion of the infrastructure
+requiring additional of opening and substituting definitions and lemmas. They
+appear in the Figure simply for the sake of completeness.
 
@@
 
+\subsubsection{Subject Reduction Theorem}
+
+The subject reduction theorem can be expressed in Coq as follows:
+\begin{coq}
+Theorem proc_sub_red:
+  forall $\Gamma$ P Q
+         (WT: P $\tpvdash_{cp}$ $\Gamma$)
+         (RED: P $\becomes$ Q),
+    Q $\tpvdash_{cp}$ $\Gamma$.
+Proof. ii; gen $\Gamma$; induction RED; subst; eauto. Qed.
+\end{coq}
+
+\coqe$forall$ quantifies over hypotheses in the statement; everything before
+the comma. \key{WT} and \key{RED} introduce names for the corresponding
+hypotheses. The statement to prove is that if \coqe$P$ is a well-typed process
+in a context $\Gamma$ and \coqe$P$ reduces to \coqe$Q$ then \coqe$Q$ is
+well-typed in $\Gamma$. The proof environment is started using \coqe$Proof.$
+and then followed by a sequence of tactics until all goals have a been
+proven. After execution of \coqe$Qed.$ the Coq system inspects the validity of
+the proof object constructed by the tactics and, if valid, will added the
+theorem to the environment.
+
+Some tactics used above are shorthands for commonly combined tactics, defined
+using Ltac, a language for defining custom tactics and decision
+procedures. For example, \coqe$ii$ repeatedly performs hypothesis introduction
+then simplification until no more can be peformed, and \coqe$gen$ $\Gamma$
+generalises the following induction over the environment $\Gamma$.
+
 \input{principalcp}
 
-The principal cut elimination rules are presented in Figure~\ref{??}. The
+The last step in the proof \coqe$eauto$ attempts to solve the goals generated
+by the induction using lemmas known to the system. The goals to be proved
+corresponding exactly to lemmas satisfying the reduction relation. The
 reduction rules were complicated to prove due to being composed of multiple
 constructor instances and differ significantly to the standard operational
-semantics one typically deals with in $\lambda$-calculus formalisms. As a
-result, these well-typed reductions were not proved ``in-place'', instead each
-rule is proved as a separate lemma. Most proofs followed a pattern but it was
-difficult to automate satisfactorily for the reasons stated in the previous
-paragraph. However, some tasks could be delegated to a custom tactic, for
-example instantiating cofinitely quantified hypotheses with a newly introduced
-fresh variable.
+semantics one typically deals with in $\lambda$-calculus formalisms. Most
+proofs followed a pattern but it was difficult to automate satisfactorily for
+the similar reasons stated for permutation automation. However, some tasks
+could be delegated to a custom tactic, for example instantiating cofinitely
+quantified hypotheses with a newly introduced fresh variable. An example of a
+principal cut reduction statement in Coq is shown in Figure~\ref{principal}.
+The hypothesis contains explicit de Bruijn
 
@@
 expressing weakening because only the names (not the types in the bindings)
-within the environment are required for weakening. one can compare alms with this version of cp
-since it weakening rule corresponds to affine logic
+within the environment are required for weakening. Without contraction, this
+version of CP corresponds to affine logic. alms calculus vs gv calculus w/o
+contraction?
 
@@
 
-\input{cpstructeq}
+%\input{cpstructeq}
 
-\input{commconvonecp}
+%\input{commconvonecp}
 
-\input{commconvtwocp}
+%\input{commconvtwocp}
 
@@
 
-\input{gv}
+%\input{gv}
 

### 2015-04-08 paper/formalisation.tex
Added a section discussing the failed attempt at cut elimination.

--- 2015-04-07 paper/formalisation.tex
@@
 
-% Description of the typing rules...
-
-Figure~\ref{fig:cll} presents the process calculus CP in LNCQ form with some
-tedious details relating to environment uniqueness and ordering (see below)
-elided. In other words, bound variables are represented using de Brujin
-indices and free variables are represented using an abstract type provided by
-the Metatheory library. The rules with binders reflect this by specifying the
-type of a bound variable in the constructor rather than a name. Cofinite
-quantification is performed in the hypotheses to these rules by excluding a
-finite set of names from opening the subprocesses where opening a process $P$
-with name $x$ is denoted $P^x$. Note that the rules for polymorphism,
+Figure~\ref{fig:cll} presents the well-typed process relation for CP as an
+inductive type in Coq. Each typing rule in CP is represented as a constructor
+of the inductive type \coqe$cp_rule$. The inductive type is indexed by a
+process (\coqe$proc$) and an environment (\coqe$penv$) indicated by them
+appearing after the colon on the first line. \coqe$forall$ quantifies over
+hypotheses in a constructor; everything before the comma. Compound hypotheses
+are given labels, such as \key{PER}, which is the default name used to
+introduce them into the proof context during theorem proving. Some details
+such as environment uniqueness have been elided, but notice the use of LNCQ
+form within the hypotheses of the rules with binders. These rules exclude a
+finite set of names \coqe$L$ from opening the subprocesses where opening a
+process \coqe$P$ with name \coqe$x$ is denoted by \coqe$open_proc P x$. Bound
+variables are represented using de Bruijn indices and free variables are
+represented using an abstract type, \coqe$atom$, provided by the Metatheory
+library. The rules with binders specify the type of a bound variable in the
+constructor rather than a name. Note that the rules for polymorphism,
 $\exists$ and $\forall$, were not included in the final Coq development since
@@
 infrastructure requiring additional of opening and substituting definitions
-and lemmas. They appear in the Figure simply for the sake of completeness.
+and lemmas.
 
@@
 names. In other words, one cannot substitute an arbitrary process term for a
-name, unlike in the $\lambda$-calculus. Thus frameworks which assume
+name, unlike in the $\lambda$-calculus. Thus, frameworks which assume
 $\lambda$-calculi style operations are not immediately amenable to the CP
@@
 Ordering in environments is ignored so the development required operations for
-handling permutations of environments. The Metatheory library does not
-currently support permutating environments, however some basic lemmas were
-added as a result of this work, and custom tactics were developed to
-automatically solve certain permutation goals via simple proof search. Even
-so, there are still many cases where sequences of primitive lemmas, such as
-those handling append, commutativity and transitivity, need to be applied
-manually. A lot of these cases are similar but there would not appear to be an
-advantage in developing a tactic to do the work. Doing so would create a
-dependency between a hard-to-debug and obscure Ltac definition and the proofs
-which utilise it, becoming a burden to change. Further, such a development
-philosophy treats Ltac tactics akin to macros, a shorthand for expressing an
-oft-repeated sequence of tactics, and it is doubtful much can be gained from
-it unless the execution of Ltac tactics can be made more transparent.
+handling permutations of environments as shown in the \coqe$cp_rule$
+definition. The Metatheory library does not currently support permutating
+environments, however some basic lemmas were added as a result of this work,
+and custom tactics were developed to automatically solve certain permutation
+goals via simple proof search. Even so, there are still many cases where
+sequences of primitive lemmas, such as those handling append, commutativity
+and transitivity, need to be applied manually. A lot of these cases are
+similar but there would not appear to be an advantage in developing a tactic
+to do the work. Doing so would create a dependency between a hard-to-debug and
+obscure Ltac definition and the proofs which utilise it, becoming a burden to
+change. Further, such a development philosophy treats Ltac tactics akin to
+macros, a shorthand for expressing an oft-repeated sequence of tactics, and it
+is doubtful much can be gained from it unless the execution of Ltac tactics
+can be made more transparent.
+
+\subsubsection{Subject Reduction Theorem}
+
+The subject reduction theorem can be expressed in Coq as follows:
+\begin{coq}
+Theorem proc_sub_red:
+  forall $\Gamma$ P Q
+         (WT: P $\tpvdash_{cp}$ $\Gamma$)
+         (RED: P $\becomes$ Q),
+    Q $\tpvdash_{cp}$ $\Gamma$.
+Proof. ii; gen $\Gamma$; induction RED; subst; eauto. Qed.
+\end{coq}
+
+\coqe$Theorem$ or \coqe$Lemma$ indicate the start of a proposition to be
+proved (Coq makes no distinction between the two keywords). The statement to
+prove is that if \coqe$P$ is a well-typed process in a context $\Gamma$ and
+\coqe$P$ reduces to \coqe$Q$ then \coqe$Q$ is well-typed in $\Gamma$. The
+proof environment is started using \coqe$Proof.$ and then followed by a
+sequence of tactics until all goals have been proven. After execution of
+\coqe$Qed.$ the Coq system inspects the validity of the proof object
+constructed by the tactics and, if valid, will add the theorem to the
+environment.
+
+Some tactics used above are shorthands for commonly combined tactics, for
+example \coqe$ii$ repeatedly performs hypothesis introduction then
+simplification until no more can be peformed, and \coqe$gen$ $\Gamma$
+generalises the goal over the typing environment $\Gamma$.
 
@@
 
-The principal cut elimination rules are presented in Figure~\ref{??}. The
-reduction rules were complicated to prove due to being composed of multiple
-constructor instances and differ significantly to the standard operational
-semantics one typically deals with in $\lambda$-calculus formalisms. As a
-result, these well-typed reductions were not proved ``in-place'', instead each
-rule is proved as a separate lemma. Most proofs followed a pattern but it was
-difficult to automate satisfactorily for the reasons stated in the previous
-paragraph. However, some tasks could be delegated to a custom tactic, for
+The \key{RED} hypothesis refers to an inductive type defining the principal
+cut reductions and commuting conversions on processes. The inductive cases in
+\coqe$proc_sub_red$ were complicated to prove due to the process terms being
+composed of multiple constructor instances, differing significantly from the
+standard operational semantics typically dealt with in $\lambda$-calculus
+formalisms. To mitigate the complexity, the proof for each case was written as
+a separate lemma (termed ``reduction lemmas''). The last step in the proof of
+\coqe$proc_sub_red$, \coqe$eauto$, solves the goals generated by the induction
+using the reduction lemmas.
+
+An example of a reduction lemma is shown in Figure~\ref{fig:principal}. Cut of
+server accept against client request involves three constructor instances, the
+outer cut and the two dual constructors as the subprocesses. As shown, to
+specify this compound process requires the explicit use of de Bruijn
+indices. These indices will be replaced in the proof by a sufficiently fresh
+name after destructing \key{WT} to obtain the derivations for \coqe$P$ and
+\coqe$Q$. Most proofs followed a pattern but it was difficult to automate
+satisfactorily for similar reasons to those stated for automating permutation
+manipulation. However, some tasks were delegated to custom tactics, for
 example instantiating cofinitely quantified hypotheses with a newly introduced
@@
 \citeauthor{Wadler:2014}'s definitions. However, explicit weakening is
-required to differentiate the principal cut elimination rules for the various
-interactions between servers and clients ($\beta_{!?}$ and $\beta_{!W}$). The
-paper presentation does not specify where weakening could occur; it is tacitly
+required to differentiate the principal cut elimination rules for the
+interactions between servers and clients (spawning and weakening). The paper
+presentation does not specify where weakening could occur; it is tacitly
 assumed weakening can always be pushed further up the derivation tree to a
-base rule e.g. $\Ax\Cut$ rule. Contraction is not supported in the
+base rule e.g. \coqe$cp_fwd$ rule. Contraction is not supported in the
 formalisation due to the difficulty in specifying its principal cut
@@
 Unfortunately, the reduction relation does not have knowledge of the
-environment, $\Gamma$, typing process $\tm{P}$. This fact does not impact upon
-expressing weakening because only the names (not the types in the bindings)
-within the environment are required for weakening. one can compare alms with this version of cp
-since it weakening rule corresponds to affine logic
+environment, $\Gamma$, typing process $\tm{P}$ because this information is
+only available after the compound cut is destructed, yielding a derivation for
+$\tm{P}$. This fact does not impact upon expressing weakening because only the
+names (not the types in the bindings) within the environment are required for
+weakening.
 
@@
 (cf. an arbitrary context in \citeauthor{Wadler:2014}'s
-presentation~\cite{Wadler:2014}). While this diverges with the presentation of
-linear logic it became necessary to do this because otherwise there would be
-no relationship between the free variables within a process and its
-environment. no there is a clear iff relationship which permits reduction
-rules to apply such as i think weakening but it should be any rule which takes
-some fvs of proc and infers something about environments from that...otherwise
-what it is not at all clear how to encode the reductions in a relation, the
-subprocesses environments are created from destructing the environment
+presentation~\cite{Wadler:2014}). \todo{look into a fix for the empty choice
+  rule} While this diverges with the presentation of linear logic it became
+necessary to do this because otherwise there would be no relationship between
+the free variables within a process and its environment. no there is a clear
+iff relationship which permits reduction rules to apply such as i think
+weakening but it should be any rule which takes some fvs of proc and infers
+something about environments from that...otherwise what it is not at all clear
+how to encode the reductions in a relation, the subprocesses environments are
+created from destructing the environment
 
@@
 
-more elaborate than first thought. complicated measure.  as it stands the
-current definitions of well-typed terms and reduction are not amenable to the
-proving of the cut elimination theorem. some method of measure is necessary in
-order to develop some notion of termination and also some relationship to the
-the the underlying subprocesses additionally there is also the issue of the
-representation which includes cases that are not immediately clear how to
-prove for instance the commuting conversions are one such example the
-reduction is only defined if the logical operator corresponds to a free
-variable but induction on the typing judgement leads to two cases one where
-the logical operator is applied to the top-level cut variable it is not
-immediately clear how to eliminate this from happening a relationship between
-the two subprocesses of the cut would need to be stated to prevent them from
-differing so that a principal cut reduction rule may be applied yes only then
-may the the rule be reduced to a cut-free thing if we had a relationship
-between them then there would be some kind of thing that would allow
-destruction on one then fixing the other in the case of a bvar equal to zero
-\todo{how to proceed?}
+\subsubsection{Top-level Cut Elimination}
+
+The theorem for top-level cut elimination can be expressed as:
+\begin{coq}
+Theorem proc_cut_elim:
+  forall P $\Gamma$
+         (WT: P $\tpcp$ $\Gamma$),
+    exists Q, P $\becomes^\star$ Q /\ $\sim$ is_cut Q.
+\end{coq}
+
+where $\becomes^\star$ denotes the reflexive and transitive closure of
+$\becomes$ and $\sim$ \coqe$is_cut Q$ expresses that Q is not a cut.
+
+Unfortunately, the theorem has eluded attempts at mechanised proof so far. The
+hypothesis of a well-typed process \coqe$P$ is not sufficient to prove the
+theorem. For instance, the \coqe$cp_cut$ constructor does not ensure that
+there is a duality relationship between the subprocesses. Following informal
+proofs, it is likely to require a termination measure involving the length of
+the derivation is needed. Additionally, the locally nameless representation
+creates subgoals for which it is not clear how to prove. For instance, a
+commuting conversion is only defined if the logical operator corresponds to a
+free variable but induction on \key{WT} yields two cases one of which applies
+the logical operator to the top-level cut variable. Since there is no
+relationship between the subprocesses, in general it will not be possible to
+apply a principal cut rule.
 

### 2015-04-09 paper/formalisation.tex
Working on the issues related to reduction relation definition.

@@
 
-Initially, an implicit notion of weakening was chosen so as to be in line with
-\citeauthor{Wadler:2014}'s definitions. However, explicit weakening is
-required to differentiate the principal cut elimination rules for the
-interactions between servers and clients (spawning and weakening). The paper
-presentation does not specify where weakening could occur; it is tacitly
-assumed weakening can always be pushed further up the derivation tree to a
-base rule e.g. \coqe$cp_fwd$ rule. Contraction is not supported in the
-formalisation due to the difficulty in specifying its principal cut
-elimination rule. In order to specify the rule one requires to show the
-following:
+\subsubsection{Issues for Reduction}
+
+A consequence of defining the reduction relation with respect to compound
+process terms is that the subprocesses have implicit derivations which are
+unavailable in the conclusion. For example, \coqe$reduce_spawn$ provides no
+information about the environment which types \coqe$P$ in the statement of the
+lemma. The derivation for \coqe$P$ (similarly, \coqe$Q$) is obtained by
+inverting the hypothesis \key{WT} and specialising the resulting assumptions
+to a newly introduced fresh variable (to replace the de Bruijn index). In the
+case of \coqe$reduce_spawn$, two inversions are required to get derivations
+for \coqe$P$ and \coqe$Q$. This property of the reduction relation affects the
+definitions of the typing rules for CP, though it was not obvious at the time.
+Notable examples are the structural rules, weakening and contraction, and
+empty choice.
+
+\textbf{Weakening and Contraction} Initially, an implicit notion of weakening
+was chosen so as to be in line with \citeauthor{Wadler:2014}'s
+definitions. However, explicit weakening is required to differentiate the
+principal cut elimination rules for the interactions between servers and
+clients; spawning and garbage collection (weakening). The paper presentation
+does not specify where weakening could occur; it is tacitly assumed weakening
+can always be pushed further up the derivation tree to a base rule
+e.g. \coqe$cp_fwd$ rule.
+
+\begin{samepage}
+Contraction is not supported in the formalisation due to the difficulty in
+specifying its principal cut elimination rule.  In order to specify the rule
+one requires knowledge of the environment typing process \coqe$P$ with the
+rule of the form:
 \begin{gather*}
@@
 
-where $\tm{?z[[A]].Q}$ denotes contraction by $z$ of two channels of type $A$
-in $Q$, $\tm{P^y} \tpvdash \tp{{?\Gamma} \comma \tmof{y}A}$ for some
-sufficiently fresh name $y$ and $\key{contract}_{\Gamma} \tm{Q}$ contracts
-$\tm{Q}$ with the names in $\Gamma$.
+where $\Gamma$ is the assumed environment typing \coqe$P$, $\tm{?z[[A]].Q}$
+denotes contraction by $z$ of two channels of type $A$ in $Q$ and
+$\key{contract}_{\Gamma} \tm{Q}$ contracts $\tm{Q}$ with the bindings in
+$\Gamma$. Unfortunately, as explained above, there is no way for the reduction
+relation to obtain the environment $\Gamma$.
+\end{samepage}
 
-Unfortunately, the reduction relation does not have knowledge of the
-environment, $\Gamma$, typing process $\tm{P}$ because this information is
-only available after the compound cut is destructed, yielding a derivation for
-$\tm{P}$. This fact does not impact upon expressing weakening because only the
-names (not the types in the bindings) within the environment are required for
-weakening.
+The reader may wonder how weakening is supported if not
+contraction. Fortunately, weakening only requires the names (not the types in
+the bindings) within the environment, and these may be obtained by relying on
+the following lemma relating the free variables of a process and the domain of
+its environment:
+\begin{coq}
+Lemma in_env_fv:
+  forall P $\Gamma$ x (WT: P $\tpcp$ $\Gamma$),
+    x $\in$ dom $\Gamma$ $\iff$ x $\in$ fv_proc P.
+\end{coq}
 
-In the empty choice rule there may only be a single binding in the context
-(cf. an arbitrary context in \citeauthor{Wadler:2014}'s
-presentation~\cite{Wadler:2014}). \todo{look into a fix for the empty choice
-  rule} While this diverges with the presentation of linear logic it became
-necessary to do this because otherwise there would be no relationship between
-the free variables within a process and its environment. no there is a clear
-iff relationship which permits reduction rules to apply such as i think
-weakening but it should be any rule which takes some fvs of proc and infers
-something about environments from that...otherwise what it is not at all clear
-how to encode the reductions in a relation, the subprocesses environments are
-created from destructing the environment
+\textbf{Empty Choice} The empty choice rule may only have a single binding in
+the context (cf. an arbitrary context in \citeauthor{Wadler:2014}'s
+presentation~\cite{Wadler:2014}). Adding an arbitrary environment to the empty
+choice constructor would prohibit \coqe$in_env_fv$ since its process no longer
+contains all the names of its environment. Not only would weakening be
+inadmissible if \coqe$in_env_fv$ did not hold, but also axiom cut reduction:
+\begin{coq}
+Lemma reduce_axcut:
+  forall P A (w : atom) $\Gamma$
+         (NFV: w $\notin$ fv_proc P)
+         (WT: $\nu$ A.(w $\link$ 0 $\mid$ P) $\tpcp$ $\Gamma$),
+    (open_proc P w) $\tpcp$ $\Gamma$.
+\end{coq}
+
+The proof of \coqe$reduce_axcut$ needs to rename the variable chosen to open
+the process \coqe$P$ in the conclusion. Initially, the variable is a fresh
+\coqe$x$ introduced during the proof, but using the following
+\coqe$typing_rename$ lemma, \coqe$w$ can replace \coqe$x$ by appeal to
+\coqe$in_env_fv$:
+\begin{coq}
+Lemma typing_rename:
+  forall $\Gamma$ P k (x y : atom) A
+         (NINX: x $\notin$ dom $\Gamma$ $\cup$ fv_proc P)
+         (NINY: y $\notin$ dom $\Gamma$ $\cup$ fv_proc P)
+         (WTX: $\open{k}{x}{P}$ $\tpcp$ (x $\sim$ A) ++ $\Gamma$),
+    $\open{k}{y}{P}$ $\tpcp$ (y $\sim$ A) ++ $\Gamma$.
+\end{coq}
+
+where $\open{k}{x}{P}$ denotes replacing de Bruijn index $k$ with free
+variable $x$ in process $P$.
+
+While this diverges with the linear logic rule for top, I believe it to be
+sound within the CP/GV eco-system such that this variant of CP still leads to
+a valid translation to GV. In a private communication, Philip Wadler echoed
+these remarks but pointed out that this is likely more to do with GV
+\todo{what's another way to say ``too small'' vs. ``too large''}.
 
@@
 linear environments caused a problem with the splitting up of variables.
-
-% ... subject reduction ...
-
-subject reduction proof proceeds by induction on the reduction relation then
-is automatic by appeal to the proved well-typed reductions.
-
-% ... cut elimination ....
 
@@
 
-modularised development; tactics slow prover took a long time to reduce proofs
-so split the structure of the development \todo{picture of development
-  structure}.
+\textbf{Modularised development} Not only is modular development an important
+programming practice in general, but in the context of theorem proving it has
+the advantage of reducing interactive load time. Executing proof scripts can
+be time-consuming especially if those scripts make frequent use of decision
+procedures performing proof search, then the execution time is proportional to
+the size of the proof context and complexity of the goal. There were cases in
+which the system crashed unexpectedly during interactive mode and it is
+conceivable that having many large proofs in-memory could be a contributing
+factor. For these reasons, the development is partitioned into GV and CP
+sections each with sub-divided into definition, infrastructure and typing
+modules (CP also has an additional module for reduction lemmas). The CPS
+translation is also defined separately. Load time is considerably improved and
+the consequences of a crash are mitigated.
+
+\textbf{Representation} It is important not to go overboard in the
+when developing not to use dependent types all the time without working out if
+it will be to the advantage of the prover use dependent types and other power
+with care for example we could don't want to use mutual inductive definitions
+for the types construct since this is overly complicated for what we wish to
+achieve and we don't want to index types with kinds since this complicates
+further definitions it is about trying to use dependent types responsibly
+without using them in an uncritical or thinking manner with great power comes
+great responsibility
+
+\textbf{Library reuse} One of the design goals of the development was to reuse
+existing libraries wherever possible. Two such instances proved extremely
+useful, first is the Metatheory library as already mentioned. Second, is the
+library of tactics provided by the Software Foundations online book~\cite{SF}.
+The tactics helped in many ways including simplifying introduction of
+assertions, providing convenient shorthands and ... Traditionally, assertions
+in Coq have to be explicitly written within a proof leading to the use of
+empheremal names introduced by prover. If the proof of the assertion follows
+from an already proved lemma and some assumptions in the context this is
+unnecessary verbage and this is what the library allows to omit \todo{example
+  of using the forwards tactic} an example of the so called forwards tactic is
+given below the $\sim$ is a a shorthand for immediately applying \coqe$auto$
+to all subgoals generated by the tactic.
+
+\textbf{Linear environments} Linearity create issues with typing some
+examples; the splitting of the environment was non-deterministic so required
+side-conditions to enforce a certain splitting of binders. For example the
+rule for associativity of cut. \todo{how to handle? is there a better way?}
 
@@
 future work (things like deciding equations on finite sets)
-
-One of the design goals of the development was to re-use existing libraries
-wherever possible. two such instances proved extremely useful. first is the
-Metatheory library developed at upenn. second, also from upenn, is the library
-of tactics from software foundations.
-
-software foundations helped in many ways: simplifying tacticals, ease of
-introducing proved assertions, etc. limited use of names introduced by prover
-leading to more robust development.
-
-used extensively: metatheory does not handle multiple binders (check this?)
-maybe add a generic tactic to handle multiple binders... or perhaps add a
-lemmas for handling binders
-
-Linearity create issues with typing some examples; the splitting of the
-environment was non-deterministic so required side-conditions to enforce a
-certain splitting of binders. For example the rule for associativity of
-cut. \todo{how to handle? is there a better way?}
 

### 2015-04-09 paper/formalisation.tex
Mostly finished CP section (good enough for second draft).

@@
 finite set of names \coqe$L$ from opening the subprocesses where opening a
-process \coqe$P$ with name \coqe$x$ is denoted by \coqe$open_proc P x$. Bound
-variables are represented using de Bruijn indices and free variables are
-represented using an abstract type, \coqe$atom$, provided by the Metatheory
-library. The rules with binders specify the type of a bound variable in the
-constructor rather than a name. Note that the rules for polymorphism,
-$\exists$ and $\forall$, were not included in the final Coq development since
-GV (\S~\ref{sec:gv}) does not support polymorphism making the rules
-unnecessary for the translation. Further, as noted by \citeauthor{Lee:2012},
-support for polymorphism in type variables creates an explosion of the
-infrastructure requiring additional of opening and substituting definitions
-and lemmas.
+process \coqe$P$ with name \coqe$x$ is denoted by \coqe$open_proc P x$ and
+corresponds to replacing de Bruijn index 0 with \coqe!x!. Bound variables are
+represented using de Bruijn indices and free variables are represented using
+an abstract type, \coqe$atom$, provided by the Metatheory library. The rules
+with binders specify the type of a bound variable in the constructor rather
+than a name. Note that the rules for polymorphism, $\exists$ and $\forall$,
+were not included in the final Coq development since GV (\S~\ref{sec:gv}) does
+not support polymorphism making the rules unnecessary for the
+translation. Further, as noted by \citeauthor{Lee:2012}, support for
+polymorphism in type variables creates an explosion of the infrastructure
+requiring additional of opening and substituting definitions and lemmas.
 
@@
 where $\open{k}{x}{P}$ denotes replacing de Bruijn index $k$ with free
-variable $x$ in process $P$.
+variable $x$ in process $P$~\footnote{For example, \coqe!$\open{0}{x}{P}
+  \defeq$ open_proc P x!}.
 
@@
 
-% ...commuting conversions
+The structural equivalences and commuting conversions have some additional
+complexity not present in the original presentation. For instance, structural
+equivalences are not unique and different process terms result depending on
+the initial binder permutation. Consider the associativity rule for which the
+paper presentation specifies only one rule:
+\begin{coq}
+Lemma assoc_1:
+  forall P Q R A B $\Gamma$
+         (LCP: forall x, lc_proc (open_proc P x))
+         (WT: $\nu$ B.($\nu$ A.(P $\mid$ Q) $\mid$ R) $\tpcp$ $\Gamma$),
+    $\nu$ A.(P $\mid$ $\nu$ B.($\permb{0}{1}{Q}$ $\mid$ R)) $\tpcp$ $\Gamma$.
+\end{coq}
 
-%\input{cpstructeq}
+where \coqe$lc_proc P$ specifies that \coqe$P$ is a \textit{local closure}, in
+other words contains no unbound de Bruijn indices and $\permb{0}{1}Q$
+indicates the permuting of binders 0 and 1 in process $\tm{Q}$. The typing
+rule for \coqe$cp_cut$ does not specify any particular split of the linear
+environment in the conclusion; splitting is non-deterministic. So, in order to
+prove the above statement, it must be true in \key{WT} that \coqe$Q$ takes the
+binder of type \coqe$B$ and not \coqe$P$, hence the local closure assumption
+on \coqe$open_proc P x$. Similarly, to prove the converse statement
+\coqe$forall x, open_proc R x$ must be assumed. For commuting conversions,
+permutating binders is essential in formulating the majority of the rules.
 
-%\input{commconvonecp}
+Regarding commuting conversions, many require analysing permutations
+containing singleton bindings. Consider the following commuting conversion:
+\begin{coq}
+  Lemma reduce_cc_input:
+    forall P Q (x:atom) A B $\Gamma$
+           (WT: $\nu$ A.(x(B).P $\mid$ Q) $\tpcp$ $\Gamma$),
+      x(B).($\nu$ A.($\open{0}{1}{P}$ $\mid$ Q)) $\tpcp$ $\Gamma$.
+\end{coq}
 
-%\input{commconvtwocp}
+Afer inverting \key{WT} and specialising the resulting hypotheses, regarding
+the subprocesses, with fresh variables the proof state is as follows:
 
-The structural equivalences and commuting conversions are shown in
-Figure~\ref{??} and Figures~\ref{??} and~\ref{??}, respectively. These rules
-are expressed to indicate some additional complexity not present in the
-original presentation. For example, structural equivalences are not unique and
-different process terms result depending on initial binder
-permutation. $\permb{0}{1}$ indicates the permuting of binders $0$ and $1$ in
-process $\tm{P}$. The paper presentation specifies only one possible
-associativity rule lending to the specific split of the linear environments.
-Notice that permutating binders is essential to formulating the majority of
-these rules. Analysing permutations containing singleton bindings was
-necessary for the commuting conversion rules. Singleton bindings can occur
-frequently during a proof and it is sometimes necessary to analyse a
-permutation of the form P++x~A ~~ Q++R to determine whether x occurs in Q or
-R. the rule for associativity requires this and so do quite a few reduction
-rules. since this involves some element of searching the context followed by
-destructing hypotheses it could be effectively automated in permutations in
-subprocesses which use a binder of the parent. the rule for associativity w
-linear environments caused a problem with the splitting up of variables.
+\begin{tabular}{l}
+\begin{coq}
+1 subgoal
+(* ... other elements of the proof context ... *)
+CPQ : open_proc Q y $\tpcp$ y$\sim$A++$\Delta$Q
+CPP0 : $\open{0}{z}{\open{1}{y}{P}}$ $\tpcp$ z$\sim$B++x$\sim$C++$\Delta$P0
+PER : Permutation $\Gamma$ ($\Delta$P++$\Delta$Q)
+PER0 : Permutation ($\Delta$P++y$\sim$A)]) (x$\sim$B$\parr$C++$\Delta$P0)
+\end{coq}\\ \hline
+\begin{coq}
+x(B).($\nu$ A.($\open{0}{1}{P}$ $\mid$ Q)) $\tpcp$ $\Gamma$.
+\end{coq}
+\end{tabular}
+\\~\\
+
+In order to prove the goal, $\Gamma$ must be replaced with the sublists typing
+\coqe!P! and \coqe!Q!. Therefore, \coqe!PER0! needs to modified such that the
+left-hand side is $\Delta$P and the right-hand side has \coqe!y$\sim$A!
+removed. In other words, the binding of \coqe!y! within \coqe!$\Delta$P0! must
+be extracted. Since many of the commuting conversion rules required this sort
+of reasoning separating the work out into a tactic comes to mind. Indeed, Ltac
+is well-suited for this type of work, searching the context and manipulating
+hypotheses of a particular form. The Ltac script in Figure~\ref{fig:extract}
+does the job\todo{keep (and explain) this?}.
+
+\begin{figure}
+\begin{coq}
+Ltac extract_bnd x A :=
+  match goal with
+    | [H: Permutation (_++x~A) ?F |- _] =>
+      let BNDS := fresh "BNDS" in
+      forwards~ BNDS: Perm_binds x A H
+      ; analyze_binds_uniq BNDS
+      ; try (by applys~ uniq_perm H)
+      ; try (match goal with
+               | [H: binds x A ?E |- _] =>
+                 let E1 := fresh "E1" in
+                 let E2 := fresh "E2" in
+                 let EQ := fresh "EQ" in
+                 apply binds_env_split in H
+                 ; inversion_clear H as (E1 & E2 & EQ)
+                 ; substs~; des_reqs
+             end)
+  end.
+\end{coq}
+\caption{Tactic finds binding in permutation RHS}
+\label{fig:extract}
+\end{figure}
+
+However, the task of removing the binding from the permutation is not
+automated to keep information loss within the context transparent to an
+individual proof. The tactic appears more general in this form and allows more
+than one ordering of tactic sequences. feels like two jobs to remove it as
+well as find it. although perhaps extract is not the right term if it doesn't
+remove it
 
@@
 
-The theorem for top-level cut elimination can be expressed as:
+\begin{figure}
 \begin{coq}
@@
 \end{coq}
+\caption{Cut elimination theorem}
+\label{fig:cutelim}
+\end{figure}
 
+The theorem for top-level cut elimination appears in Figure~\ref{fig:cutelim}
 where $\becomes^\star$ denotes the reflexive and transitive closure of

### 2015-04-10 paper/formalisation.tex
Cleaned up discussion of GV. More to add.

@@
 types as subcomponents (arguments to send, for example). Unfortunately,
-handling mutually inductive definitions in Coq can be quite involved;
+handling mutually inductive definitions in Coq can be quite involved,
 requiring one to either rely on the Coq system to provide a stronger mutual
@@
 
-Another issue related to types is how to representing type kinds (linearity or
-non-linearity). Initially the type definition was defined as an indexed
-family, indexed by the kind. While this definition has the advantage of
-preventing certain ill-formed types, such as a non-linear session type, it
-complicated later definitions and proofs. For instance, the decidability of
-equality on types now relied on an axiom permitting heterogeneous
-equality~\cite{??}. Generally, it is better to avoid unnecessary postulates if
-possible. Further, when formulating the typing judgements the kind must be
-given to each type hypothesised and most rules (e.g. ...) are polymorphic in
-the kind, motivating an awkward inspection function to be applied per
-application of a constructor \todo{clean this description up}. A solution to
-this was to define a well-formed type relation and do away with the type
-dependency. these issues highlighted that one should be careful and
-considerate regarding when to use dependent types. these issues highlighted
-that one should be careful and considerate how the representation will affect
-later expressiveness.
-
-similar infrastructure for gv required as for cp copy and paste job
+Another issue related to types is how to represent type kinds, linear and
+unlimited kind. Initially the type definition was indexed by the kind and
+while this definition has the advantage of preventing certain ill-formed
+types, such as a non-linear session type, it complicated later definitions and
+proofs. For instance, the decidability of equality on types now relied on an
+axiom permitting heterogeneous equality~\cite{McBride:1999}. Generally, it is
+better to avoid unnecessary postulates if possible. Further, when formulating
+the typing rules, the kind must be given to each type hypothesised and most
+rules (e.g. \coqe!wt_tm_labs!) are polymorphic in the kind, requring an
+awkward kind inspection function to be applied to each type within a
+binding. The alternative approach taken was to define a well-formed type
+relation and remove the kind index from the type definition. These issues
+highlighted that one should be careful and considerate regarding when to use
+dependent types. These issues highlighted that one should take care over the
+representation and its effect on later expressiveness.
 

### 2015-04-10 paper/formalisation.tex
Added select rules regarding GV and some initial discussion.

@@
 
-%\input{gv}
+\input{gv}
 
 The typing judgements for the functional language GV are shown in
-Figure~\ref{??}. Note that I have altered the $n$-ary branch and choice
+Figure~\ref{fig:gv}. Note that I have altered the $n$-ary branch and choice
 constructs to the restricted binary versions in order to avoid tricky
@@
 enable the inclusion of these rules in the translation from GV to CP, which is
-performed by induction on GV terms.
+performed by induction on GV terms. \todo{could take about opening the
+  multiple binder let. multi but not as in literature}
 
@@
 
-\subsection{CPS translation GV to CP}
+\subsection{CPS translation from GV to CP}
 
-require type inference on gv terms to construct corresponding cp terms but we
-are assured that this is okay since the theorem only provides a well-typed gv
-term in a context that is well-defined
+Following the original work by \citeauthor{Wadler:2014}, the translation from
+GV to CP requires defining translations on types and terms of GV, which for
+terms is in continuation-passing style.
 
-type inference of gv terms required cluttered definition of translation with
-impossible error cases. avoided a dependently typed solution due to
-difficulties encountered earlier in the development. james mckinna et
-al. suggest encoding the successful cases in an inductive type and then
-defining the function to act on these cases. will this really work? this means
-a graph of positive cases for types then a function definition for it too this
-seems rather pathetic because i have to define th einference on terms and it
-seems wrong i have then to handle environments but also nonenvironments i
-can't remember what they are called unbound variables that is it they have to
-be handled as well so we have all this clutter and the types too that is the
-important part but the environments then have to be handled linearly
-unfortunately very bad because we could just use the well typed term relation
-for most of this headachey breaky stuff i'm not sure about that but what is
-can i say so for example lets see see no i also have to thing about the
-general guidelines i can be giving for this section so we have something to
-say about guidelines do we yes of course we have can say that out of this we
-realised that it is important for a representation to be good enough to
-succintly express that which we wish to say in particular the graph relation
-the inductive type that we define propositionally is better one would think
-than the other way the monadic possible failure way is not producing a valid
-thing that we can inductively induct on so we concern ourself with the proof
-of translation preserving the typed term and i think it is important to
-remember the problem we had with the inductive hypothesis we had something do
-with translating the inner term of a weakening term in a larger context than
-the one we had available and that was to do with the destruction of the term
-inference thing i think no no it was something else we had the inner term and
-we had it well typed in an environment that was smaller sure and then our
-thing our function was saying we had some environment that typed both the
-inner term and the outer weakening term but we know that is wrong so wrong so
-here is perhaps where the inductive type graph relation thing wins out in that
-we can have a hypothesis for the inner term which is sufficiently small for
-the environment and we pattern match not only on the term but also on the
-environment yes that is we will do because it seems to work only when we get
-to the messy two the messy two constructor caases will this become a bigg
-problem inthat we will then have all this stuff about environment being split
-and needing to be recombined in the composite process that is a term not a
-process but the point still stands it will need to be about the term and the
-splitting is really a duplication of the unfortunately the well typed term
-relation but otherwise i cannot think how to reuse or obtain that information
-for the translation to cp
+\subsubsection{Types}
 
-Translation uses the coq-ext-lib from michela for monadic notation
+The translation of from GV types to CP propositons is shown in Figure~\ref{??}
+\todo{Coq version or regular version?}. Translation of session types is
+noteworthy for taking a GV session type to its notional dual in CP, reflecting
+the semantic difference between channels in GV and those in CP.
 
-Currently the formalisation does not handle this. Future work.
+\subsubsection{Terms}
 
-\todo{how might we make progress on this?}
+Due to the locally nameless representation, the translation of GV terms
+required type inference of terms to construct corresponding CP terms in order
+to specify the propositions. So, the translation takes the environment, in
+which the term was typed, as an argument to query the type of free
+variables. Unfortunately, this query returns a type of \coqe!option typ!
+(\coqe!option! is equivalent to \tm{Maybe} in Haskell) since it may not be
+found, but in fact since only well-typed GV terms are translated it is
+guaranteed to return something! Therefore, the definition is unnecessarily
+cluttered with impossible error cases.
+
+The term translation was initially designed using a monadic approach to
+account for the possible error cases in retreiving names from the
+environment. The coq-ext-lib
+library~\footnote{https://github.com/coq-ext-lib/coq-ext-lib} provides
+definitions for monad, functor and applicative type classes and associated
+instances for basic Coq types. It is similar in style to Haskell's support for
+these objects. Unfortunately, this definition did not yield a suitable
+function to prove the translation well-typed because the function needs to be
+able to operate structurally on not only terms, but also the environment and
+the processes being constructed.
+
+An alternative formulation, which avoids the need to handle error cases, is to
+first define only the positive cases using an inductive type, the so-called
+\textit{graph relation}, and then define a dependent function which only
+considers the positive cases within the relation~\cite{McKinna:2009}. However,
+several issues were encountered when attempting to specify the
+translation. The graph relation requires much of the clutter present in both
+the \coqe!cp_rule! and \coqe!wt_tm! definition such as linear enviornments and
+cofinite quanitifcation.
+
+i'm not sure about that but what is can i say so for example lets see see no i
+also have to thing about the general guidelines i can be giving for this
+section so we have something to say about guidelines do we yes of course we
+have can say that out of this we realised that it is important for a
+representation to be good enough to succintly express that which we wish to
+say in particular the graph relation the inductive type that we define
+propositionally is better one would think than the other way the monadic
+possible failure way is not producing a valid thing that we can inductively
+induct on so we concern ourself with the proof of translation preserving the
+typed term and i think it is important to remember the problem we had with the
+inductive hypothesis we had something do with translating the inner term of a
+weakening term in a larger context than the one we had available and that was
+to do with the destruction of the term inference thing i think no no it was
+something else we had the inner term and we had it well typed in an
+environment that was smaller sure and then our thing our function was saying
+we had some environment that typed both the inner term and the outer weakening
+term but we know that is wrong so wrong so here is perhaps where the inductive
+type graph relation thing wins out in that we can have a hypothesis for the
+inner term which is sufficiently small for the environment and we pattern
+match not only on the term but also on the environment yes that is we will do
+because it seems to work only when we get to the messy two the messy two
+constructor caases will this become a bigg problem inthat we will then have
+all this stuff about environment being split and needing to be recombined in
+the composite process that is a term not a process but the point still stands
+it will need to be about the term and the splitting is really a duplication of
+the unfortunately the well typed term relation but otherwise i cannot think
+how to reuse or obtain that information for the translation to cp
+
+Roughly, the theorm aiming to prove is:
 
@@
 
-\textbf{Modularised development} Not only is modular development an important
+\textbf{Modularised development.} Not only is modular development an important
 programming practice in general, but in the context of theorem proving it has
@@
 
-\textbf{Representation} It is important not to go overboard in the
+\textbf{Representation.} It is important not to go overboard in the
 when developing not to use dependent types all the time without working out if
@@
 
-\textbf{Library reuse} One of the design goals of the development was to reuse
-existing libraries wherever possible. Two such instances proved extremely
-useful, first is the Metatheory library as already mentioned. Second, is the
-library of tactics provided by the Software Foundations online book~\cite{SF}.
-The tactics helped in many ways including simplifying introduction of
-assertions, providing convenient shorthands and ... Traditionally, assertions
-in Coq have to be explicitly written within a proof leading to the use of
-empheremal names introduced by prover. If the proof of the assertion follows
-from an already proved lemma and some assumptions in the context this is
-unnecessary verbage and this is what the library allows to omit \todo{example
-  of using the forwards tactic} an example of the so called forwards tactic is
-given below the $\sim$ is a a shorthand for immediately applying \coqe$auto$
-to all subgoals generated by the tactic.
+\textbf{Library reuse.} One of the design goals of the development was to
+reuse existing libraries wherever possible. Two such instances proved
+extremely useful, first is the Metatheory library as already
+mentioned. Second, is the library of tactics provided by the Software
+Foundations online book~\cite{SF}.  The tactics helped in many ways including
+simplifying introduction of assertions, providing convenient shorthands and
+... Traditionally, assertions in Coq have to be explicitly written within a
+proof leading to the use of empheremal names introduced by prover. If the
+proof of the assertion follows from an already proved lemma and some
+assumptions in the context this is unnecessary verbage and this is what the
+library allows to omit \todo{example of using the forwards tactic} an example
+of the so called forwards tactic is given below the $\sim$ is a a shorthand
+for immediately applying \coqe$auto$ to all subgoals generated by the tactic.
 
-\textbf{Linear environments} Linearity create issues with typing some
+\textbf{Linear environments.} Linearity create issues with typing some
 examples; the splitting of the environment was non-deterministic so required

### 2015-04-10 paper/background.tex
Minor fixes. Resolved a todo regarding Lolliproc.

@@
 Concurrency is provided by control operators which, respectively, spawn a
-child process and wait for a child process to complete. The spawning process
-creates a channel between parent and child process; a channel is represented
-as a continuation and the type denotes the communication protocol
-(i.e. session type) between parent and child \todo{what part of session type
-  is being discussed here? spawning process or spawning operation?}.
+child process and wait for a child process to complete. The spawning operator
+creates a channel between the parent and child processes; a channel is
+represented by a pair of endpoints with dual types denoting the communication
+protocol (i.e. session type) between parent and child.
 

### 2015-04-10 paper/formalisation.tex
Minor fixes. Resolved a todo regarding Lolliproc.

@@
 
-\textbf{Weakening and Contraction} Initially, an implicit notion of weakening
+\textbf{Weakening and Contraction.} Initially, an implicit notion of weakening
 was chosen so as to be in line with \citeauthor{Wadler:2014}'s
@@
 
-\textbf{Empty Choice} The empty choice rule may only have a single binding in
+\textbf{Empty Choice.} The empty choice rule may only have a single binding in
 the context (cf. an arbitrary context in \citeauthor{Wadler:2014}'s
@@
 cofinite quanitifcation.
-
+\begin{comment}
 i'm not sure about that but what is can i say so for example lets see see no i
@@
 how to reuse or obtain that information for the translation to cp
-
+\end{comment}
 Roughly, the theorm aiming to prove is:

### 2015-04-10 paper/introduction.tex
Minor fixes. Resolved a todo regarding Lolliproc.

@@
 session types for mainstream programming languages can be studied, taking
-advantage of previous proof effort through the re-use of existing libraries
+advantage of previous proof effort through the reuse of existing libraries
 for programming language metatheory. The work has presented a number of

### 2015-04-10 paper/background.tex
Commuting conversion example (Wadler's style). Fixed small issues.

@@
 (input), and they correspond to the units of the respective linear logical
-connectives \todo{Do I need to elaborate on the translation? Yet another
-  figure?}. CP provides linear logical connectives with session type
-interpretations inspired by
-\citeauthor{Caires:2010:STI}~\cite{Caires:2010:STI} shown in
-Figure~\ref{fig:cptnts} along with the terms of CP. The process calculus CP is
-slightly different to $\pi$DILL by \citeauthor{Caires:2010:STI}, apart from
-being classical in nature, the constructs for server replication are modified:
-CP provides separate weakening and contraction rules whereas these are encoded
-in one rule for process reduction in $\pi$DILL. Additionally, rather than a
+connectives (see \S~\ref{sec:trans}). CP provides linear logical connectives
+with session type interpretations shown in Figure~\ref{fig:cptnts} along with
+the terms of CP.
+
+The process calculus CP is slightly different to $\pi$DILL by
+\citeauthor{Caires:2010:STI}, besides from being classical (rather than
+intuitionistic), the constructs for server replication are modified: CP
+provides separate weakening and contraction rules whereas these are encoded in
+one rule for process reduction in $\pi$DILL. Additionally, rather than a
 two-sided sequent calculus, \citeauthor{Wadler:2014} uses one-sided sequents
@@
 duality and a clearer correspondence to a $\lambda$-calculi surface syntax
-like GV. \citeauthor{Wadler:2014} describes a continuation-passing style (CPS)
+like GV.
+
+\input{ccoutput}
+
+Subject reduction in CP proceeds by cut reductions of the logical operators
+against their duals and \textit{commuting conversions}. Commuting conversions
+push a cut inside a logical operator and the intuition is that they permit
+further communication to occur which was guarded by a cut. An example of a
+commuting conversion on output is shown in Figure~\ref{fig:ccoutput}. On the
+left-hand side $R$ can interact with its environment, but commuting the output
+and the cut prevents $R$ from doing this. As explained by
+\citeauthor{Wadler:2014}, the semantics of the rule are either $x$ is bound by
+an outer cut and will be eliminated eventually by cut reduction, or the
+process is blocked waiting for external communication on channel $x$. This
+differs from the lambda calculus where the evaluation strategy is
+deterministic (e.g. left-to-right call-by-value semantics).
+
+\citeauthor{Wadler:2014} describes a continuation-passing style (CPS)
 translation from GV types and terms to CP classical linear logic propositions
-and processes, respectively. As presented, GV does not support all features of
-CP and thus the translation is one way (GV to
+and processes, respectively. Freedom from deadlock within GV is assured
+through the translation of CP and the top-level cut elimination property of CP
+(see \S~\ref{sec:cutelim}).
+
+\citeauthor{Wadler:2014}'s GV does not support all features of CP and thus the
+translation is one way (GV to
 CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to

### 2015-04-10 paper/formalisation.tex
Commuting conversion example (Wadler's style). Fixed small issues.

@@
 
-While this diverges with the linear logic rule for top, I believe it to be
-sound within the CP/GV eco-system such that this variant of CP still leads to
-a valid translation to GV. In a private communication, Philip Wadler echoed
-these remarks but pointed out that this is likely more to do with GV
-\todo{what's another way to say ``too small'' vs. ``too large''}.
+While this diverges with the linear logic rule for top, I believe this variant
+of CP still admits a valid translation from GV. In a private communication,
+Philip Wadler echoed these remarks but pointed out that this is likely because
+the GV-to-CP translation does not require the full expressiveness of CP.
 
@@
 
-\subsubsection{Top-level Cut Elimination}
+\subsubsection{Top-level Cut Elimination}\label{sec:cutelim}
 
@@
 
-\subsection{CPS translation from GV to CP}
+\subsection{CPS translation from GV to CP}\label{sec:trans}
 

### 2015-04-11 paper/formalisation.tex
Described Coq version of ccoutput (added its forgotten figure).

@@
 clients; spawning and garbage collection (weakening). The paper presentation
-does not specify where weakening could occur; it is tacitly assumed weakening
-can always be pushed further up the derivation tree to a base rule
-e.g. \coqe$cp_fwd$ rule.
+does not specify where weakening could occur due to its implicit nature; it is
+tacitly assumed weakening can always be pushed further up the derivation tree
+to a base rule e.g. \coqe$cp_fwd$ rule. The same is true of contraction where
+it is defined as a substitution of names on process terms, not a
+syntax-directed rule.
 
@@
 
-\textbf{Empty Choice.} The empty choice rule may only have a single binding in
-the context (cf. an arbitrary context in \citeauthor{Wadler:2014}'s
-presentation~\cite{Wadler:2014}). Adding an arbitrary environment to the empty
-choice constructor would prohibit \coqe$in_env_fv$ since its process no longer
-contains all the names of its environment. Not only would weakening be
-inadmissible if \coqe$in_env_fv$ did not hold, but also axiom cut reduction:
+\textbf{Empty Choice.} The development provides a restricted form of the empty
+choice rule. In contrast to the arbitrary context in
+\citeauthor{Wadler:2014}'s presentation~\cite{Wadler:2014}, \coqe!cp_empcho!
+may only have a single binding in the context. Adding an arbitrary environment
+to the empty choice constructor would prohibit \coqe$in_env_fv$ since its
+process no longer contains all the names of its environment. Not only would
+weakening be inadmissible if \coqe$in_env_fv$ did not hold, but also axiom cut
+reduction:
 \begin{coq}
@@
 
+\subsubsection{Equivalences and Conversions}
+
 The structural equivalences and commuting conversions have some additional
@@
 Regarding commuting conversions, many require analysing permutations
-containing singleton bindings. Consider the following commuting conversion:
+containing singleton bindings. Consider the following the Coq encoding of the
+commuting conversion for output (cf. Figure~\ref{fig:ccoutput}):
 \begin{coq}
-  Lemma reduce_cc_input:
-    forall P Q (x:atom) A B $\Gamma$
-           (WT: $\nu$ A.(x(B).P $\mid$ Q) $\tpcp$ $\Gamma$),
-      x(B).($\nu$ A.($\open{0}{1}{P}$ $\mid$ Q)) $\tpcp$ $\Gamma$.
+Lemma reduce_cc_multi_one:
+  forall P Q R (x:atom) A B $\Gamma$ (LCQ: lc_proc Q)
+         (WT: $\nu$ A.([B] x.(P $\mid$ Q) $\mid$ R) $\tpcp$ $\Gamma$),
+    [B] x.($\nu$ A.($\permb{0}{1}{P}$ $\mid$ R) $\mid$ Q) $\tpcp$ $\Gamma$.
 \end{coq}
 
-Afer inverting \key{WT} and specialising the resulting hypotheses, regarding
+Notice again, as in the case for \coqe!assoc_1!, it necessary to stipulate
+that \coqe!Q! is locally closed to ensure \coqe!P! receives the channel
+created by the outer cut as part of its environment.
+
+After inverting \key{WT} and specialising the resulting hypotheses, regarding
 the subprocesses, with fresh variables the proof state is as follows:
@@
 (* ... other elements of the proof context ... *)
-CPQ : open_proc Q y $\tpcp$ y$\sim$A++$\Delta$Q
-CPP0 : $\open{0}{z}{\open{1}{y}{P}}$ $\tpcp$ z$\sim$B++x$\sim$C++$\Delta$P0
-PER : Permutation $\Gamma$ ($\Delta$P++$\Delta$Q)
-PER0 : Permutation ($\Delta$P++y$\sim$A)]) (x$\sim$B$\parr$C++$\Delta$P0)
+CPR : $\open{0}{y}{R}~\tpcp$ y$\sim$A++$\GD$Q
+CPQ : Q $\tpcp$ x$\sim$C++$\GD$Q0
+CPP : $\open{0}{z}{\open{1}{y}{P}}$ $\tpcp$ z$\sim$B++$\GD$P0
+NIN : y $\notin$ {x} $\cup$ dom $\GD$Q0
+PER : Permutation $\Gamma$ ($\GD$P++$\GD$Q)
+PER0 : Permutation ($\GD$P++y$\sim$A) (x$\sim$B$\otimes$C++$\GD$P0++$\GD$Q0)
 \end{coq}\\ \hline
 \begin{coq}
-x(B).($\nu$ A.($\open{0}{1}{P}$ $\mid$ Q)) $\tpcp$ $\Gamma$.
+x(B).($\nu$ A.($\open{0}{1}{P}$ $\mid$ Q)) $\tpcp$ $\Gamma$
 \end{coq}
@@
 In order to prove the goal, $\Gamma$ must be replaced with the sublists typing
-\coqe!P! and \coqe!Q!. Therefore, \coqe!PER0! needs to modified such that the
-left-hand side is $\Delta$P and the right-hand side has \coqe!y$\sim$A!
-removed. In other words, the binding of \coqe!y! within \coqe!$\Delta$P0! must
-be extracted. Since many of the commuting conversion rules required this sort
-of reasoning separating the work out into a tactic comes to mind. Indeed, Ltac
-is well-suited for this type of work, searching the context and manipulating
-hypotheses of a particular form. The Ltac script in Figure~\ref{fig:extract}
-does the job\todo{keep (and explain) this?}.
+\coqe!P!, \coqe!Q! and \coqe!R!. Therefore, \coqe!PER0! needs to be modified
+such that the left-hand side is $\GD$P and the right-hand side has
+\coqe!y$\sim$A!  removed. In other words, the binding of \coqe!y! within
+\coqe!$\GD$P0! must be extracted (by \coqe!NIN!, \coqe!y! must be in
+\coqe!$\GD$P0!). Since many of the commuting conversion rules required this
+sort of reasoning, separating the work out into a tactic appears
+beneficial. Indeed, Ltac is well-suited for this type of work, searching the
+context and manipulating hypotheses of a particular form. The Ltac script in
+Figure~\ref{fig:extract} does the job and after executing
+\coqe!extract_bnd y A! \key{PER0} becomes:
+\begin{coq}
+Permutation ($\GD$P++y$\sim$A) (x$\sim$B$\otimes$C++E1++y$\sim$A++E2++$\GD$Q0)
+\end{coq}
+where E1 and E2 are sublists of $\GD$P0.
 
@@
 automated to keep information loss within the context transparent to an
-individual proof. The tactic appears more general in this form and allows more
-than one ordering of tactic sequences. feels like two jobs to remove it as
-well as find it. although perhaps extract is not the right term if it doesn't
-remove it
+individual proof. The tactic is more modular and general in this form,
+permitting composition and permutation of tactic sequences.
 

### 2015-04-11 paper/formalisation.tex
Added tilde to tactics in extract_bnd Ltac script.

@@
 
+\input{extract}
+
 Regarding commuting conversions, many require analysing permutations
@@
 where E1 and E2 are sublists of $\GD$P0.
-
-\begin{figure}
-\begin{coq}
-Ltac extract_bnd x A :=
-  match goal with
-    | [H: Permutation (_++x~A) ?F |- _] =>
-      let BNDS := fresh "BNDS" in
-      forwards~ BNDS: Perm_binds x A H
-      ; analyze_binds_uniq BNDS
-      ; try (by applys~ uniq_perm H)
-      ; try (match goal with
-               | [H: binds x A ?E |- _] =>
-                 let E1 := fresh "E1" in
-                 let E2 := fresh "E2" in
-                 let EQ := fresh "EQ" in
-                 apply binds_env_split in H
-                 ; inversion_clear H as (E1 & E2 & EQ)
-                 ; substs~; des_reqs
-             end)
-  end.
-\end{coq}
-\caption{Tactic finds binding in permutation RHS}
-\label{fig:extract}
-\end{figure}
 

### 2015-04-11 paper/formalisation.tex
Added informal cut rule for comparison with formalised variant.
Changed cut reduction figure to contrast with the informal rule.
Described Wadler-style CP rule and cut reduction.

@@
 \input{cp}
-
-\begin{comment}
-FIGURES
-
-equivalences and axiomatic cut
-\end{comment}
+\input{papercut}
 
@@
 with binders specify the type of a bound variable in the constructor rather
-than a name. Note that the rules for polymorphism, $\exists$ and $\forall$,
-were not included in the final Coq development since GV (\S~\ref{sec:gv}) does
-not support polymorphism making the rules unnecessary for the
+than a name.
+
+Note that the rules for polymorphism, $\exists$ and $\forall$, were not
+included in the final Coq development since GV (\S~\ref{sec:gv}) does not
+support polymorphism making the rules unnecessary for the
 translation. Further, as noted by \citeauthor{Lee:2012}, support for
@@
 can be made more transparent.
+
+As an illustrative example of the differences between the Coq encoding of the
+rules and the paper presentation consider the pen-and-paper version of the cut
+rule shown in Figure~\ref{fig:papercut}. There is no overhead when introducing
+binders to the subprocesses or for environment ordering, but apart from these
+representation issues the rules look similar which could assist the uptake of
+such formal tools by programming language researchers.
 
@@
 
-An example of a reduction lemma is shown in Figure~\ref{fig:principal}. Cut of
-server accept against client request involves three constructor instances, the
-outer cut and the two dual constructors as the subprocesses. As shown, to
-specify this compound process requires the explicit use of de Bruijn
-indices. These indices will be replaced in the proof by a sufficiently fresh
-name after destructing \key{WT} to obtain the derivations for \coqe$P$ and
-\coqe$Q$. Most proofs followed a pattern but it was difficult to automate
-satisfactorily for similar reasons to those stated for automating permutation
-manipulation. However, some tasks were delegated to custom tactics, for
-example instantiating cofinitely quantified hypotheses with a newly introduced
-fresh variable.
+An example of a reduction lemma, and its counterpart in
+\citeauthor{Wadler:2014}'s presentation, is shown in
+Figure~\ref{fig:principal}. While \citeauthor{Wadler:2014} describes the cut
+reductions using derivation trees, this structure is left implicit in the Coq
+encoding. Cut of server accept against client request involves three
+constructor instances, the outer cut and the two dual constructors as the
+subprocesses. As shown, to specify this compound process requires the explicit
+use of de Bruijn indices. These indices will be replaced in the proof by a
+sufficiently fresh name after destructing \key{WT} to obtain the derivations
+for \coqe!P! and \coqe!Q!. Most proofs followed a pattern but it was difficult
+to automate satisfactorily for similar reasons to those stated for automating
+permutation manipulation. However, some tasks were delegated to custom
+tactics, for example instantiating cofinitely quantified hypotheses with a
+newly introduced fresh variable.
 
@@
 Figure~\ref{fig:extract} does the job and after executing
-\coqe!extract_bnd y A! \key{PER0} becomes:
+\coqe!extract_bnd y A!, \key{PER0} becomes:
 \begin{coq}

### 2015-04-11 paper/formalisation.tex
Added Wadler let rule version.

@@
 \input{gv}
+\input{letscompare}
 
@@
 enable the inclusion of these rules in the translation from GV to CP, which is
-performed by induction on GV terms. \todo{could take about opening the
-  multiple binder let. multi but not as in literature}
+performed by induction on GV terms and is therefore
+syntax-directed. Figure~\ref{fig:letscompare} presents the informal version of
+the well-typed let rule. While \citeauthor{Char:2012} suggests a two-indices
+approach for handling for multiple binders~\cite{Char:2012}, such a strategy
+seems excessive for the GV let rule and the only rule to benefit from multiple
+binders. For that reason, the simple solution of performing two openings was
+chosen.
 

### 2015-04-11 paper/formalisation.tex
Added figure to discussion on type kind representation.

@@
 
-On a more technical note, all types are amalgamated into one inductive
-definition and provide a predicate for restricting typing rules to consider
-only valid session types. \citeauthor{Wadler:2014} defines GV types as a set
-of mutually recursive definitions; session types are types which may contain
-types as subcomponents (arguments to send, for example). Unfortunately,
-handling mutually inductive definitions in Coq can be quite involved,
-requiring one to either rely on the Coq system to provide a stronger mutual
-induction principle or defining one manually. Such a definition is possible,
-but it complicates applications of the induction principle.
+All types are amalgamated into one inductive definition and provide a
+predicate for restricting typing rules to consider only valid session
+types. As shown in \S~\ref{sec:...}, \citeauthor{Wadler:2014} defines GV types
+as a set of mutually recursive definitions. Unfortunately, handling mutually
+inductive definitions in Coq can be quite involved, requiring one to either
+rely on the Coq system to provide a stronger mutual induction principle or
+define one manually. Such a definition is possible, but it complicates
+applications of the induction principle.
+
+\input{wft}
 
@@
 the typing rules, the kind must be given to each type hypothesised and most
-rules (e.g. \coqe!wt_tm_labs!) are polymorphic in the kind, requring an
-awkward kind inspection function to be applied to each type within a
-binding. The alternative approach taken was to define a well-formed type
-relation and remove the kind index from the type definition. These issues
+rules are polymorphic in the kind, requring an awkward kind inspection
+function to be applied to each type within a binding. The alternative approach
+taken was to define a well-formed type relation and remove the kind index from
+the type definition. Figure~\ref{fig:wft} shows the well-formedness for the
+linear function type. Notice, that even though the conclusion must have linear
+kind, as in rule \coqe!wt_tm_labs!, the argument and result types can have any
+kind, highlighting the aforementioned kind polymorphism. These issues
 highlighted that one should be careful and considerate regarding when to use

### 2015-04-12 paper/formalisation.tex
Sketched some discussion of cut elimination progress.

@@
 there is a duality relationship between the subprocesses. Following informal
-proofs, it is likely to require a termination measure involving the length of
-the derivation is needed. Additionally, the locally nameless representation
-creates subgoals for which it is not clear how to prove. For instance, a
-commuting conversion is only defined if the logical operator corresponds to a
-free variable but induction on \key{WT} yields two cases one of which applies
-the logical operator to the top-level cut variable. Since there is no
-relationship between the subprocesses, in general it will not be possible to
-apply a principal cut rule.
+proofs, it initially seemed likely that the proof required a termination
+measure involving the length of the derivation. However due to the
+difficulties in trying to prove the above rule some doubt was
+seeded. additionally previous attempts in the literature formalised cut
+elimination using a deep embedding perhaps an alternative formulation of the
+reduction relation as a deep embedding expressed using derivations would
+permit progress on the proof but this particular avenue is left for future
+work
+
+Formalised efforts for proving cut elimination for other logics have adopted a
+deep embedding~\cite{??,??,??}, where sequents and derivations are represented
+explicitly rather than just the conclusion as in the shallow embedding adopted
+here. It is unclear whether the cut elimination theorem can be rescued in the
+current setting or not, but previous work suggests it is better to use other
+approaches. It is worth noting that \citeauthor{Tews:2013}'s Coq development
+significant; about 36,000 lines.
+
+the locally nameless
+representation creates subgoals for which it is not clear how to prove. For
+instance, a commuting conversion is only defined if the logical operator
+corresponds to a free variable, but induction on \key{WT} yields two cases one
+of which applies the logical operator to the top-level cut variable and since
+there is no relationship between the subprocesses, in general it will not be
+possible to apply a principal cut rule.
 

### 2015-04-12 paper/formalisation.tex
Added the references.

@@
 Formalised efforts for proving cut elimination for other logics have adopted a
-deep embedding~\cite{??,??,??}, where sequents and derivations are represented
-explicitly rather than just the conclusion as in the shallow embedding adopted
-here. It is unclear whether the cut elimination theorem can be rescued in the
-current setting or not, but previous work suggests it is better to use other
-approaches. It is worth noting that \citeauthor{Tews:2013}'s Coq development
-significant; about 36,000 lines.
+deep embedding~\cite{Tews:2013,Dawson:2010,Dawson:2002}, where sequents and
+derivations are represented explicitly rather than just the conclusion as in
+the shallow embedding adopted here. It is unclear whether the cut elimination
+theorem can be rescued in the current setting or not, but previous work
+suggests it is better to use other approaches. It is worth noting that
+\citeauthor{Tews:2013}'s Coq development significant; about 36,000 lines.
 

### 2015-04-12 paper/formalisation.tex
Made my peace with top-level cut elimination.

--- 2015-04-11 paper/formalisation.tex
@@
 polymorphism in type variables creates an explosion of the infrastructure
-requiring additional of opening and substituting definitions and lemmas.
+requiring additional of opening and substituting definitions and lemmata.
 
@@
 definition. The Metatheory library does not currently support permutating
-environments, however some basic lemmas were added as a result of this work,
+environments, however some basic lemmata were added as a result of this work,
 and custom tactics were developed to automatically solve certain permutation
 goals via simple proof search. Even so, there are still many cases where
-sequences of primitive lemmas, such as those handling append, commutativity
+sequences of primitive lemmata, such as those handling append, commutativity
 and transitivity, need to be applied manually. A lot of these cases are
@@
 formalisms. To mitigate the complexity, the proof for each case was written as
-a separate lemma (termed ``reduction lemmas''). The last step in the proof of
+a separate lemma (termed ``reduction lemmata''). The last step in the proof of
 \coqe$proc_sub_red$, \coqe$eauto$, solves the goals generated by the induction
-using the reduction lemmas.
+using the reduction lemmata.
 
@@
 $\becomes$ and $\sim$ \coqe$is_cut Q$ expresses that Q is not a cut.
-
 Unfortunately, the theorem has eluded attempts at mechanised proof so far. The
 hypothesis of a well-typed process \coqe$P$ is not sufficient to prove the
-theorem. For instance, the \coqe$cp_cut$ constructor does not ensure that
-there is a duality relationship between the subprocesses. Following informal
-proofs, it is likely to require a termination measure involving the length of
-the derivation is needed. Additionally, the locally nameless representation
-creates subgoals for which it is not clear how to prove. For instance, a
-commuting conversion is only defined if the logical operator corresponds to a
-free variable but induction on \key{WT} yields two cases one of which applies
-the logical operator to the top-level cut variable. Since there is no
-relationship between the subprocesses, in general it will not be possible to
-apply a principal cut rule.
+theorem. For instance, the \coqe$cp_cut$ constructor requires a
+``relatedness'' condition on the subprocesses to ensure that they eventually
+operate dually on the bound variable. The condition is required after
+induction on \key{WT} yields a goal similar to the following:
+\begin{coq}
+exists R, ($\nu$A.([B]0.($P_1$ $\mid$ $P_2$) $\mid$ Q)) $\becomes^\star$ R
+               /\   
+          (is_cut R -> False)
+\end{coq}
+
+The left subprocess has been destructed to a logical operator which operates
+on the bound variable of the cut, but since there is no relationship between
+the two subprocesses it is impossible to prove the goal.
+
+Following informal proofs, it seems likely that the proof requires a
+termination measure involving the length of the derivation. In hindsight,
+however, I do not believe the current encoding is the best for this sort of
+reasoning. Formalised efforts for proving cut elimination for other logics
+have adopted an approach which permits reasoning on the structure of
+derivations represented explicitly as
+trees~\cite{Tews:2013,Dawson:2010,Dawson:2002}. Whereas, in the approach taken
+here, that information is limited and ephemeral in that it is only available
+during the proving of a derivation e.g. a reduction lemma. It is future work
+to investigate these and other strategies which will assist in proving
+cut-elimination for CP.
 
@@
 sections each with sub-divided into definition, infrastructure and typing
-modules (CP also has an additional module for reduction lemmas). The CPS
+modules (CP also has an additional module for reduction lemmata). The CPS
 translation is also defined separately. Load time is considerably improved and
@@
 reductions involved small amounts of permuting environments which were largely
-similar across the proofs. while some simple lemmas and tactics where created
+similar across the proofs. while some simple lemmata and tactics where created
 to handle the most tedious elements these did not make effective use of

### 2015-04-12 paper/formalisation.tex
Minor tweaks.
Added the rest of GV's terms to figure.

@@
 exists R, ($\nu$A.([B]0.($P_1$ $\mid$ $P_2$) $\mid$ Q)) $\becomes^\star$ R
-               /\   
-          (is_cut R -> False)
+          /\ (is_cut R -> False)
 \end{coq}
@@
 during the proving of a derivation e.g. a reduction lemma. It is future work
-to investigate these and other strategies which will assist in proving
+to investigate these and other strategies which will assist in formalising
 cut-elimination for CP.
@@
 predicate for restricting typing rules to consider only valid session
-types. As shown in \S~\ref{sec:...}, \citeauthor{Wadler:2014} defines GV types
-as a set of mutually recursive definitions. Unfortunately, handling mutually
-inductive definitions in Coq can be quite involved, requiring one to either
-rely on the Coq system to provide a stronger mutual induction principle or
-define one manually. Such a definition is possible, but it complicates
+types. As shown in \S~\ref{sec:fusion}, \citeauthor{Wadler:2014} defines GV
+types as a set of mutually recursive definitions. Unfortunately, handling
+mutually inductive definitions in Coq can be quite involved, requiring one to
+either rely on the Coq system to provide a stronger mutual induction principle
+or define one manually. Such a definition is possible, but it complicates
 applications of the induction principle.

### 2015-04-13 paper/formalisation.tex
Tweakening the GV section.

@@
 syntax-directed. Figure~\ref{fig:letscompare} presents the informal version of
-the well-typed let rule. While \citeauthor{Char:2012} suggests a two-indices
-approach for handling for multiple binders~\cite{Char:2012}, such a strategy
+the well-typed let rule. While \citeauthor{Char:2012} suggests using
+two-indices for handling multiple binders~\cite{Char:2012}, such a strategy
 seems excessive for the GV let rule and the only rule to benefit from multiple
@@
 
-All types are amalgamated into one inductive definition and provide a
-predicate for restricting typing rules to consider only valid session
-types. As shown in \S~\ref{sec:fusion}, \citeauthor{Wadler:2014} defines GV
-types as a set of mutually recursive definitions. Unfortunately, handling
-mutually inductive definitions in Coq can be quite involved, requiring one to
-either rely on the Coq system to provide a stronger mutual induction principle
-or define one manually. Such a definition is possible, but it complicates
-applications of the induction principle.
+As shown in \S~\ref{sec:fusion}, \citeauthor{Wadler:2014} defines GV types as
+a set of mutually recursive definitions. Unfortunately, handling mutually
+inductive definitions in Coq can be quite involved, requiring one to either
+rely on the Coq system to provide a stronger mutual induction principle or
+define one manually. Such a definition is possible, but it complicates
+applications of the induction principle. Therefore, all types are amalgamated
+into one inductive definition.
 
@@
 kind, as in rule \coqe!wt_tm_labs!, the argument and result types can have any
-kind, highlighting the aforementioned kind polymorphism. These issues
-highlighted that one should be careful and considerate regarding when to use
-dependent types. These issues highlighted that one should take care over the
-representation and its effect on later expressiveness.
+kind, highlighting the aforementioned kind polymorphism. An additional
+predicate was defined to ensure typing rules consider only valid session
+types. The predicate appears in the well-formedness constructors for session
+types, ensuring the continuation is a session type and not, for example, a
+linear function type. These issues highlight that one should be mindful of the
+chosen representation and its effect on later expressiveness.
 

### 2015-04-13 paper/formalisation.tex
Added figures for the sections on GV and the CPS translation.

@@
 Following the original work by \citeauthor{Wadler:2014}, the translation from
-GV to CP requires defining translations on types and terms of GV, which for
-terms is in continuation-passing style.
+GV to CP requires defining translations on types and terms of GV.
 
@@
 
-The translation of from GV types to CP propositons is shown in Figure~\ref{??}
-\todo{Coq version or regular version?}. Translation of session types is
-noteworthy for taking a GV session type to its notional dual in CP, reflecting
-the semantic difference between channels in GV and those in CP.
+\input{gv-trans-types}
+
+The translation of from GV types to CP propositons is shown in
+Figure~\ref{fig:typtrans}. Translation of session types is noteworthy for
+taking a GV session type to its notional dual in CP, reflecting the semantic
+difference between channels in GV and those in CP. For GV, the type refers to
+the action on the channel endpoint. For CP, the proposition refers to the
+\textit{interface} exposed to a user of the channel. For example, output in GV
+corresponds to input in CP; the action on the channel endpoint is to output a
+value, and to do this the channel must input the value.
 
@@
 
-Due to the locally nameless representation, the translation of GV terms
-required type inference of terms to construct corresponding CP terms in order
-to specify the propositions. So, the translation takes the environment, in
-which the term was typed, as an argument to query the type of free
-variables. Unfortunately, this query returns a type of \coqe!option typ!
-(\coqe!option! is equivalent to \tm{Maybe} in Haskell) since it may not be
-found, but in fact since only well-typed GV terms are translated it is
-guaranteed to return something! Therefore, the definition is unnecessarily
-cluttered with impossible error cases.
+\input{gv-cps-cp}
+
+The translation of GV terms to CP processes is given, following
+\citeauthor{Wadler:2014}, using continuation-passing style. In this style a
+well-typed term in GV is parameterised by a name representing the channel
+which communicates the result of the translation (a CP process). Due to the
+locally nameless representation, the translation of GV terms required a
+procedure to infer the type of terms in order to specify the binding type for
+the corresponding CP process. An example, translating choice, is shown in
+Figure~\ref{fig:trans}, note that I am using \citeauthor{Wadler:2014}'s style
+but using an LNCQ notation for clarity. In order to construct the cut and
+output processes, the type of \tm{M} and \tm{N} must be computed. So, the
+translation takes the environment, in which the term is typed, as an argument
+to query the type of free variables. Unfortunately, this query returns a type
+of \coqe!option typ!  (\coqe!option! is equivalent to \tm{Maybe} in Haskell)
+since it may not be found, but in fact since only well-typed GV terms are
+translated it is guaranteed to return something!  Therefore, the definition is
+unnecessarily cluttered with impossible error cases.
 
@@
 able to operate structurally on not only terms, but also the environment and
-the processes being constructed.
+the processes being constructed. Consider again the translation of case. here
+we require to consider nl and nr in a separate environment than m and type the
+resulting case term in the combined environment we also need to open these
+terms using cofinite quantification to yield a suitable induction
+principle. all told the function is not readily definable. further problems do
+arise as we shall see.
 
@@
 first define only the positive cases using an inductive type, the so-called
-\textit{graph relation}, and then define a dependent function which only
-considers the positive cases within the relation~\cite{McKinna:2009}. However,
-several issues were encountered when attempting to specify the
-translation. The graph relation requires much of the clutter present in both
-the \coqe!cp_rule! and \coqe!wt_tm! definition such as linear enviornments and
-cofinite quanitifcation.
-\begin{comment}
-i'm not sure about that but what is can i say so for example lets see see no i
-also have to thing about the general guidelines i can be giving for this
-section so we have something to say about guidelines do we yes of course we
-have can say that out of this we realised that it is important for a
-representation to be good enough to succintly express that which we wish to
-say in particular the graph relation the inductive type that we define
-propositionally is better one would think than the other way the monadic
-possible failure way is not producing a valid thing that we can inductively
-induct on so we concern ourself with the proof of translation preserving the
-typed term and i think it is important to remember the problem we had with the
-inductive hypothesis we had something do with translating the inner term of a
-weakening term in a larger context than the one we had available and that was
-to do with the destruction of the term inference thing i think no no it was
-something else we had the inner term and we had it well typed in an
-environment that was smaller sure and then our thing our function was saying
-we had some environment that typed both the inner term and the outer weakening
-term but we know that is wrong so wrong so here is perhaps where the inductive
-type graph relation thing wins out in that we can have a hypothesis for the
-inner term which is sufficiently small for the environment and we pattern
-match not only on the term but also on the environment yes that is we will do
-because it seems to work only when we get to the messy two the messy two
-constructor caases will this become a bigg problem inthat we will then have
-all this stuff about environment being split and needing to be recombined in
-the composite process that is a term not a process but the point still stands
-it will need to be about the term and the splitting is really a duplication of
-the unfortunately the well typed term relation but otherwise i cannot think
-how to reuse or obtain that information for the translation to cp
-\end{comment}
+\textit{graph relation}, and then define a dependent function that only
+considers the positive cases for which the relation is
+defined~\cite{McKinna:2009}. However, several issues were encountered when
+attempting to specify the translation. The graph relation requires much of the
+clutter present in both the \coqe!cp_rule! and \coqe!wt_tm! definition such as
+linear enviornments and cofinite quantification. the proof of cps translation
+using the graph relation is a work in progress. the main difficulty is that
+the representation requires opening both GV terms and CP processes in
+hypotheses which would appear to duplicate a lot of the overhead of their
+respective rule definitions.  again, representation is important even if the
+proof has still eluded us.  while it is remains to be seen whether or not the
+graph relation assists us in these matters it is nonetheless worth trying to
+express a property in multiple ways
+
 Roughly, the theorm aiming to prove is:
+\begin{coq}
+Theorem cps_trans_wt:
+  forall $\Phi$ M T z P $\Gamma$ $\GD$
+         (WT: $\Phi$ $\tpgv$ M$\of{T}$)
+         (NIN: z $\notin$ GVFV M $\cup$ dom $\Phi$)
+         (ENV: trans_env $\Phi$ = $\GD$)
+         (PER: Permutation $\Gamma$ ($\GD$ ++ z $\sim$ $\sem{T}$t))
+         (CP: transTm M $\Phi$ z P),
+    P $\tpcp$ $\Gamma$.
+\end{coq}
 
@@
 factor. For these reasons, the development is partitioned into GV and CP
-sections each with sub-divided into definition, infrastructure and typing
+sections each further sub-divided into definition, infrastructure and typing
 modules (CP also has an additional module for reduction lemmata). The CPS
@@
 
-\textbf{Representation.} It is important not to go overboard in the
-when developing not to use dependent types all the time without working out if
-it will be to the advantage of the prover use dependent types and other power
-with care for example we could don't want to use mutual inductive definitions
-for the types construct since this is overly complicated for what we wish to
-achieve and we don't want to index types with kinds since this complicates
-further definitions it is about trying to use dependent types responsibly
-without using them in an uncritical or thinking manner with great power comes
-great responsibility
+\textbf{Representation.} As discussed in the previous sections representation
+choice is a central issue in formalising programming language
+metatheory. Decisions on how to define, for instance, types, calculi and
+translations impact on future definitions and properties dependent on
+them. Considerations range from exploiting the full expressivity of the
+theorem prover by using dependent types in the case of Coq. to how to encode
+the properties of interest and being able to assess accurately the most
+fruitful encoding upfront. avoiding features that would provide a more faitful
+relationship between formal and informal systems is desirable if those
+features would unnecessarily complicated the reasoning. it is a tradeoff.
 
@@
 
-tactic performance is an issue to be addressed due to the slow proof search a
-future work (things like deciding equations on finite sets)
-
-Base rules and rules with binders required permutation assumptions to allow
-any ordering of the constituent contexts. a lot of the proofs for well-typed
-reductions involved small amounts of permuting environments which were largely
-similar across the proofs. while some simple lemmata and tactics where created
-to handle the most tedious elements these did not make effective use of
+\textbf{Effective Automation.} It is worth considering situations where
+automation would simplify proofs. Automation is already an important
+consideration in the interactive theorem proving community. Base rules and
+rules with binders required permutation assumptions to allow any ordering of
+the constituent contexts. a lot of the proofs for well-typed reductions
+involved small amounts of permuting environments which were largely similar
+across the proofs. while some simple lemmata and tactics where created to
+handle the most tedious elements these did not make effective use of
 hypotheses from the context. Improved proof search for permutations is

### 2015-04-13 paper/formalisation.tex
Added forwards tactic example to Library reuse section.

@@
 Foundations online book~\cite{SF}. The tactics helped in many ways including
-simplifying introduction of assertions, providing convenient shorthands and
-... Traditionally, assertions in Coq have to be explicitly written within a
-proof leading to the use of empheremal names introduced by prover. If the
-proof of the assertion follows from an already proved lemma and some
-assumptions in the context this is unnecessary verbage and this is what the
-library allows to omit \todo{example of using the forwards tactic} an example
-of the so called forwards tactic is given below the $\sim$ is a a shorthand
-for immediately applying \coqe$auto$ to all subgoals generated by the tactic.
+simplifying the introduction of assertions and providing convenient shorthands
+for commonly executed sequences of tactics. An example for simplifying
+assertions would be the \coqe!forwards! tactic permitting lemma
+instantiation. Consider the following example (during the proof of $assoc_1$):
+\begin{tabular}{l}
+\begin{coq}
+1 subgoal
+(* ... other elements of the proof context ... *)
+PER : Permutation (x $\sim$ B ++ $\GD$P) ($\GD$P0++$\GD$Q0)
+UN : uniq (x $\sim$ B ++ $\GD$P)
+\end{coq}\\ \hline
+\begin{coq}
+$\nu$ A.(P $\mid$ $\nu$ B.($\open{0}{1}{Q}$ $\mid$ R)) $\tpcp$ $\Gamma$
+\end{coq}
+\end{tabular}
+\\~\\
+
+In order to proceed with proving the goal, it is necessary to show that
+$\GD$P0++$\GD$Q0 is unique. The proof follows from \key{PER} and \key{UN} upon
+application of an appropriate lemma, and there are two possible ways to prove
+this using plain Coq. One way is to apply the lemma to one of \key{PER} and
+\key{UN} thereby replacing the chosen hypothesis with the conclusion of the
+lemma (termed \textit{forward reasoning}). The opposite of forward reasoning,
+namely backward reasoning, does not apply since the goal has nothing to do
+uniqueness of $\GD$P0++$\GD$Q0. However, using the \coqe!assert! tactic one
+can utilise backward reasoning. An assertion temporarily focuses a different
+goal (the asserted goal) which may then be tackled using tactics in the
+standard way. Unfortunately, this approach requires the explicit writing of
+the goal, namely \coqe!uniq ($\GD$P0++$\GD$Q0)!, into the proof script of the
+lemma cluttering the proof script sensitive to change because \coqe!$\GD$P0!
+and \coqe!$\GD$Q0! are emphemeral names introduced by prover.
+
+The \coqe!forwards! tactic avoids the problem of explicit names by allowing
+the uniqueness property to be proven succiently:
+\begin{coq}
+forwards UN1: uniq_perm PER UN.
+\end{coq}
+
+where \coqe!uniq_perm! is the appropriate lemma and \coqe!UN1! is the name
+given to the desired uniqueness property. Note that \key{PER} and \key{UN} are
+mentioned in the proof with \coqe!assert! so the use of \coqe!forwards! has
+reduced the dependencies on introduced names. The improvement is more
+pronounced in cases where the goal is particularly long involving many
+variables from the context.
 

### 2015-04-13 paper/formalisation.tex
Semi-finalised issues and guidelines.

@@
 polymorphism in type variables creates an explosion of the infrastructure
-requiring additional of opening and substituting definitions and lemmata.
+requiring additional open and substitution definitions and related lemmata.
 
@@
 the corresponding CP process. An example, translating choice, is shown in
-Figure~\ref{fig:trans}, note that I am using \citeauthor{Wadler:2014}'s style
-but using an LNCQ notation for clarity. In order to construct the cut and
-output processes, the type of \tm{M} and \tm{N} must be computed. So, the
-translation takes the environment, in which the term is typed, as an argument
-to query the type of free variables. Unfortunately, this query returns a type
-of \coqe!option typ!  (\coqe!option! is equivalent to \tm{Maybe} in Haskell)
-since it may not be found, but in fact since only well-typed GV terms are
-translated it is guaranteed to return something!  Therefore, the definition is
-unnecessarily cluttered with impossible error cases.
+Figure~\ref{fig:trans}, unfortunately this definition is wrong because the
+process P is not being opened and therefore during the proof we are unable to
+reason about the typability of P opened with a sufficiently fresh variable the
+reason is that P is a subprocess and therefore is not locally closed note that
+I am using \citeauthor{Wadler:2014}'s style but using an LNCQ notation for
+clarity. In order to construct the cut and output processes, the type of
+\tm{M} and \tm{N} must be computed. So, the translation takes the environment,
+in which the term is typed, as an argument to query the type of free
+variables. Unfortunately, this query returns a type of \coqe!option typ!
+(\coqe!option! is equivalent to \tm{Maybe} in Haskell) since it may not be
+found, but in fact since only well-typed GV terms are translated it is
+guaranteed to return something!  Therefore, the definition is unnecessarily
+cluttered with impossible error cases.
 
@@
 \textbf{Representation.} As discussed in the previous sections representation
-choice is a central issue in formalising programming language
+choice is a central issue when formalising programming language
 metatheory. Decisions on how to define, for instance, types, calculi and
 translations impact on future definitions and properties dependent on
-them. Considerations range from exploiting the full expressivity of the
-theorem prover by using dependent types in the case of Coq. to how to encode
-the properties of interest and being able to assess accurately the most
-fruitful encoding upfront. avoiding features that would provide a more faitful
-relationship between formal and informal systems is desirable if those
-features would unnecessarily complicated the reasoning. it is a tradeoff.
+them. Considerations range from how much expressive power provided by the
+theorem prover to exploit to the choice of encoding to use for the properties
+of interest. Some of these decisions may impact the correspondence between the
+formal and informal systems e.g. the empty choice rule and omission of
+contraction in CP, and the omission of n-ary branch and choice from GV, and
+such things are necessarily a tradeoff between simplicity and
+adequacy.
+
+Interplay between different decisions is an additional factor to consider. For
+example, the representation choice for linearity makes environment splitting
+non-deterministic. Applying inversion on well-typed process terms requires, in
+some cases such as \coqe!assoc_1!, to enforce a certain split between
+subprocesses. The choice of LNCQ representation forces a local closure
+assumption in these cases. Alternative representations for the CP calculus,
+such as separating the rules (link, cut, etc.) from the derivations
+(associativity, reductions, etc.)~\cite{Tews:2013,Dawson:2010,Dawson:2002},
+may ameliorate the handling of linearity by allowing one to explicate the
+particular splitting for a derivation.
 
@@
 application of an appropriate lemma, and there are two possible ways to prove
-this using plain Coq. One way is to apply the lemma to one of \key{PER} and
+this using plain Coq. One way is to apply the lemma to one of \key{PER} or
 \key{UN} thereby replacing the chosen hypothesis with the conclusion of the
@@
 lemma cluttering the proof script sensitive to change because \coqe!$\GD$P0!
-and \coqe!$\GD$Q0! are emphemeral names introduced by prover.
+and \coqe!$\GD$Q0! are ephemeral names introduced by prover.
 
@@
 
-\textbf{Linear environments.} Linearity create issues with typing some
-examples; the splitting of the environment was non-deterministic so required
-side-conditions to enforce a certain splitting of binders. For example the
-rule for associativity of cut. \todo{how to handle? is there a better way?}
-
 \textbf{Effective Automation.} It is worth considering situations where
 automation would simplify proofs. Automation is already an important
-consideration in the interactive theorem proving community. Base rules and
-rules with binders required permutation assumptions to allow any ordering of
-the constituent contexts. a lot of the proofs for well-typed reductions
-involved small amounts of permuting environments which were largely similar
-across the proofs. while some simple lemmata and tactics where created to
-handle the most tedious elements these did not make effective use of
-hypotheses from the context. Improved proof search for permutations is
-required to simplify such developments.
+consideration in the interactive theorem proving community. Examples of
+automation from the current development have been highlighted, but particular
+emphasis should be placed on tactics involving non-trivial proof search of the
+context because these tend to be more widely applicable.

### 2015-04-13 paper/formalisation.tex
Tweaking the description of CPS term translation.

@@
 procedure to infer the type of terms in order to specify the binding type for
-the corresponding CP process. An example, translating choice, is shown in
-Figure~\ref{fig:trans}, unfortunately this definition is wrong because the
-process P is not being opened and therefore during the proof we are unable to
-reason about the typability of P opened with a sufficiently fresh variable the
-reason is that P is a subprocess and therefore is not locally closed note that
-I am using \citeauthor{Wadler:2014}'s style but using an LNCQ notation for
-clarity. In order to construct the cut and output processes, the type of
-\tm{M} and \tm{N} must be computed. So, the translation takes the environment,
-in which the term is typed, as an argument to query the type of free
+the corresponding CP process. So, the translation takes the environment, in
+which the term is typed, as an argument to query the type of free
 variables. Unfortunately, this query returns a type of \coqe!option typ!
-(\coqe!option! is equivalent to \tm{Maybe} in Haskell) since it may not be
-found, but in fact since only well-typed GV terms are translated it is
-guaranteed to return something!  Therefore, the definition is unnecessarily
-cluttered with impossible error cases.
+(\coqe!option! is equivalent to \tm{Maybe} in Haskell) to account for it
+possibly not appearing in the environment, but in fact since only well-typed
+GV terms are translated it is guaranteed to return something! Therefore, the
+definition is unnecessarily cluttered with impossible error cases.
 
 The term translation was initially designed using a monadic approach to
-account for the possible error cases in retreiving names from the
+streamline handling of the possible error cases in retreiving names from the
 environment. The coq-ext-lib
@@
 able to operate structurally on not only terms, but also the environment and
-the processes being constructed. Consider again the translation of case. here
-we require to consider nl and nr in a separate environment than m and type the
-resulting case term in the combined environment we also need to open these
-terms using cofinite quantification to yield a suitable induction
-principle. all told the function is not readily definable. further problems do
-arise as we shall see.
+the processes being constructed. An example, translating choice, is shown in
+Figure~\ref{fig:trans}. The translation of case considers \coqe!NL! and
+\coqe!NR! in a separate environment to \coqe!M! and types the resulting case
+term in the combined environment. Additionally, the subterms need to be
+considered as opened terms using cofinite quantification in order to yield a
+suitable induction principle.
 
@@
 attempting to specify the translation. The graph relation requires much of the
-clutter present in both the \coqe!cp_rule! and \coqe!wt_tm! definition such as
-linear enviornments and cofinite quantification. the proof of cps translation
-using the graph relation is a work in progress. the main difficulty is that
-the representation requires opening both GV terms and CP processes in
-hypotheses which would appear to duplicate a lot of the overhead of their
-respective rule definitions.  again, representation is important even if the
-proof has still eluded us.  while it is remains to be seen whether or not the
-graph relation assists us in these matters it is nonetheless worth trying to
-express a property in multiple ways
+clutter present in both the \coqe!cp_rule! and \coqe!wt_tm! definitions such
+as linear enviornments and cofinite quantification.  Consider again
+Figure~\ref{fig:trans}. unfortunately this definition is wrong because the
+process P is not being opened and therefore during the proof we are unable to
+reason about the typability of P opened with a sufficiently fresh variable the
+reason is that P is a subprocess and therefore is not locally closed note that
+I am using \citeauthor{Wadler:2014}'s style but using an LNCQ notation for
+clarity. In order to construct the cut and output processes, the type of
+\tm{M} and \tm{N} must be computed. the proof of cps translation using the
+graph relation is a work in progress. the main difficulty is that the
+representation requires opening both GV terms and CP processes in hypotheses
+which would appear to duplicate a lot of the overhead of their respective rule
+definitions.  again, representation is important even if the proof has still
+eluded us.  while it is remains to be seen whether or not the graph relation
+assists us in these matters it is nonetheless worth trying to express a
+property in multiple ways
 

### 2015-04-14 paper/formalisation.tex
Polished CPS term translation section.

@@
 well-typed term in GV is parameterised by a name representing the channel
-which communicates the result of the translation (a CP process). Due to the
-locally nameless representation, the translation of GV terms required a
-procedure to infer the type of terms in order to specify the binding type for
-the corresponding CP process. So, the translation takes the environment, in
-which the term is typed, as an argument to query the type of free
-variables. Unfortunately, this query returns a type of \coqe!option typ!
-(\coqe!option! is equivalent to \tm{Maybe} in Haskell) to account for it
-possibly not appearing in the environment, but in fact since only well-typed
-GV terms are translated it is guaranteed to return something! Therefore, the
-definition is unnecessarily cluttered with impossible error cases.
-
-The term translation was initially designed using a monadic approach to
-streamline handling of the possible error cases in retreiving names from the
-environment. The coq-ext-lib
-library~\footnote{https://github.com/coq-ext-lib/coq-ext-lib} provides
-definitions for monad, functor and applicative type classes and associated
-instances for basic Coq types. It is similar in style to Haskell's support for
-these objects. Unfortunately, this definition did not yield a suitable
-function to prove the translation well-typed because the function needs to be
-able to operate structurally on not only terms, but also the environment and
-the processes being constructed. An example, translating choice, is shown in
-Figure~\ref{fig:trans}. The translation of case considers \coqe!NL! and
-\coqe!NR! in a separate environment to \coqe!M! and types the resulting case
-term in the combined environment. Additionally, the subterms need to be
-considered as opened terms using cofinite quantification in order to yield a
-suitable induction principle.
-
-An alternative formulation, which avoids the need to handle error cases, is to
-first define only the positive cases using an inductive type, the so-called
-\textit{graph relation}, and then define a dependent function that only
-considers the positive cases for which the relation is
-defined~\cite{McKinna:2009}. However, several issues were encountered when
-attempting to specify the translation. The graph relation requires much of the
-clutter present in both the \coqe!cp_rule! and \coqe!wt_tm! definitions such
-as linear enviornments and cofinite quantification.  Consider again
-Figure~\ref{fig:trans}. unfortunately this definition is wrong because the
-process P is not being opened and therefore during the proof we are unable to
-reason about the typability of P opened with a sufficiently fresh variable the
-reason is that P is a subprocess and therefore is not locally closed note that
-I am using \citeauthor{Wadler:2014}'s style but using an LNCQ notation for
-clarity. In order to construct the cut and output processes, the type of
-\tm{M} and \tm{N} must be computed. the proof of cps translation using the
-graph relation is a work in progress. the main difficulty is that the
-representation requires opening both GV terms and CP processes in hypotheses
-which would appear to duplicate a lot of the overhead of their respective rule
-definitions.  again, representation is important even if the proof has still
-eluded us.  while it is remains to be seen whether or not the graph relation
-assists us in these matters it is nonetheless worth trying to express a
-property in multiple ways
-
-Roughly, the theorm aiming to prove is:
+which communicates the result of its translation (a CP process). The aim is to
+prove the following theorem about the translation:
 \begin{coq}
@@
 \end{coq}
+where \coqe!z! is the channel parameterising the translation and
+\coqe!transTm! is described below. The theorem states that if a GV term
+\coqe!M! is well-typed then its translation will be a well-typed process in CP
+with some accompanying translations on environments.
+
+Due to the locally nameless representation, the translation of GV terms
+requires a procedure to infer the type of terms in order to specify binder(s)
+in the corresponding CP process (see the right-hand side of
+Figure~\ref{fig:trans}). So the translation takes the environment, in which
+the term is typed, as an argument to query the type of free
+variables. Unfortunately, this query returns a type of \coqe!option typ!
+(\coqe!option! is equivalent to \tm{Maybe} in Haskell) to account for the name
+possibly not appearing in the environment, but in fact since only well-typed
+GV terms are translated it is guaranteed to return something! Therefore, the
+definition is unnecessarily cluttered with impossible error cases.
+
+The term translation was initially designed using a monadic approach to
+streamline handling of the possible error cases. The coq-ext-lib
+library~\footnote{https://github.com/coq-ext-lib/coq-ext-lib} provides
+definitions for monad, functor and applicative type classes and associated
+instances for basic Coq types. It is similar in style to Haskell's support for
+these objects. Unfortunately, this definition did not yield a suitable
+function to prove the translation well-typed because the function needs to be
+able to operate structurally on not only terms, but also the environment and
+the processes being constructed.
+
+An alternative formulation, which avoids the need to handle error cases, is to
+first define only the positive cases using an inductive type, the so-called
+\textit{graph relation}, and then define a dependent function that only
+considers the positive cases for which the relation is
+defined~\cite{McKinna:2009}. An example, translating choice, is shown in
+Figure~\ref{fig:trans}. The translation of choice considers \coqe!NL! and
+\coqe!NR! in a separate environment to \coqe!M! and types the resulting case
+term in the combined environment. The \coqe!NL! and \coqe!NR! subterms need to
+be considered as opened terms using cofinite quantification in order to yield
+suitable induction hypotheses. The graph relation is able to specify these
+constraints in contrast to the monadic function definition. However, several
+issues were encountered when attempting to specify the translation. The graph
+relation requires much of the clutter present in both the \coqe!cp_rule! and
+\coqe!wt_tm! definitions such as linear enviornments and cofinite
+quantification, and unfortunately the specification of choice in
+Figure~\ref{fig:trans} is not quite correct. The process \coqe!P! is not being
+opened in the \key{TR} assumption yet in the conclusion it is not locally
+closed. It would appear that, in addition to opening GV subterms, the graph
+relation must also open CP subprocesses even in cases where there was no
+binder in the original GV term e.g. function
+application~\cite{Wadler:2014}. Attempts to resolve this issue are left for
+future work as is the proof of \coqe!cps_trans_wt!, but it appears the graph
+relation approach is the most promising. These difficulties again illustrate
+the importance of the representation choice.
 

### 2015-04-14 paper/conclusion.tex
Added conclusion.

\section{Conclusion}

The current paper presented work towards a formalisation of
\citeauthor{Wadler:2014}'s two-tier GV/CP system. The Coq development defines
the process calculus CP and the functional language GV using state-of-the-art
techniques from the programming language and interactivee theorem proving
communities. Proof of subject reduction for CP has been formalised along with
work towards proving sound both cut elimination and the GV to CP CPS
translation. While ultimately there remains work to be done to complete the
development, several issues and guidelines have been identified to be
considered by future endeavours in the area. Finally, some ideas for future
work, on GV/CP specifically but also generally in the area of session-based
type system formalisations, were proffered.

### 2015-04-14 paper/abstract.tex
Tidying up background. Adding figures for CP/GV types and terms.
I think this is important background for understanding the
languages I am aiming to formalise.
Added abstract.

@@
 \begin{abstract}
-\begin{comment}
-SPL (+CPL)
-
-WTD
-
-ROF
-
-IMP
-\end{comment}
+Session types ensure that communicating agents adhere to an agreed upon
+protocol. Recent work has presented a number of related systems with
+increasingly elaborate session type features including asynchronous buffered
+communication, type polymorphism and recursion. That these features are sound
+is crucial to maintain the assurances provided by session types. However, the
+metatheory of these systems has not been formally studied and proofs are
+typically developed by-hand and checked manually. The current paper describes
+work towards a formalisation of such a type system based on recent work on a
+two-tier system connecting classical linear logic and session types. The
+formalisation, developed in Coq, proves subject reduction for one of the tiers
+and makes strides towards the other system properties. This work illustrates
+the limitations of the state-of-the-art techniques for formalising programming
+language metatheory and identifies a number of issues and guidelines of which
+practitioners in the field may wish to take account. The paper concludes by
+highlighting some areas of future work.
 \end{abstract}

### 2015-04-14 paper/futurework.tex
Tidying up background. Adding figures for CP/GV types and terms.
I think this is important background for understanding the
languages I am aiming to formalise.
Semi-finalised issues and guidelines.
Added future work and hacked away at the biblography.

@@
 
-Some areas of future work to look into would be the following.
+\textbf{CP-GV system.} The issues regarding cut elimination suggest there
+should be greater study into calculi based on the sequent calculus. It is
+unclear to what extent process calculi such as CP and lambda calculi can be
+unified to use the same framework or library. The different considerations of
+sequent and process calculi perhaps mandate a different library or
+framework. It would be interesting to consider CP in the setting described by
+\citeauthor{Tews:2013,Dawson:2010,Dawson:2002}, where explicit representation
+is given to derivations. The aim here would be to make progress in proving the
+cut elimination theorem for CP, but it is unclear how linearity and
+permutation of environments would be handled in this setting. Otherwise,
+defining an appropriate termination measure for the current version of CP may
+be possible, though it remains unclear how to obtain the desired relationship
+between subprocesses of cut (as explained in \S~\ref{sec:cutelim}).
 
-\subsection{Besides Linear}
+Further progress with the graph relation version of the translation could be
+made in the current setting aiming to limit the duplication discussed
+previously The translation would be substantially affected by the
+aforementioned change to CP because the derivations in the conlusions of the
+rules would have to be constructed. In this case, it may be worth
+investigating representing derivations in GV explicitly.
 
-The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of
-aliased access to linear resources. The type system requires special language
-features, e.g. to manage capabilities which can be thought as a form of object
-liveness mechanism, so GV typing judgements would need to be extended such
-that the capabilities can be included in soundness results. An extension to
-the translation to CP to incorporate the ``adopt'' and ``focus'' constructs
-would also need to be defined. It is not immediately clear how these features
-would assist in the management of session typed channels. The ``focus'' rule
-(and linearity generally) enforces an invariant on the focussed object
-(channel) such that the type is maintained during the focus operation:
+\citeauthor{Lindley:2014:SAP} extend \citeauthor{Wadler:2014}'s work to
+support polymorphism in GV and n-ary branch and choice in CP. Future work is
+to support these constructs.
+
+\textbf{Beyond Linear.} Most session-based type systems treat channels as
+linear resources which may not be aliased. The ``adoption and focus''
+system~\cite{Fahndrich:2002} provides a notion of aliased access to linear
+resources. The type system requires special language features, e.g. to manage
+capabilities. \citeauthor{Fahndrich:2002} do not provide operational semantics
+or soundness results for the type system. It would be interesting to
+investigate these properties formally in a session type setting. The ``focus''
+rule (and linearity generally) enforces an invariant on the focussed object
+(channel) such that the type is maintained during the focus operation. Such a
+rule would appear to only benefit recursive session types. Indeed, focussing
+on a non-recursive channel would appear pointless because one is prevented
+from communicating across it for the entire scope of the focus. Moreover, it
+would be undesirable to permit modification of a session type without
+guaranteeing exclusive access to the channel since this could lead to invalid
+communication.
+
+\begin{comment}
  \input{../proposal/focus-rule}
@@
 the guard $G$ (contained in $C_1$) is restored.
-
-Such a rule would appear to only benefit recursive session types. Indeed,
-focussing on a non-recursive channel would be pointless since one would be
-prevented from communicating across it since the focus rule requires
-maintenance of the channel type. Further, it would be undesirable to permit
-modification of a session type without guaranteeing exclusive access to the
-channel since this could lead to invalid communication e.g. trying to
-communicate on a closed channel. Additionally, a linear object is consumed
-upon being adopted and will be deallocated when its adopter's lifetime ends,
-although an alternative semantics could be defined which returned the linear
-resource to the context. Moreover, \citeauthor{Fahndrich:2002} do not provide
-operational semantics or soundness results for the ``adoption and focus'' type
-system.
-
-\citeauthor{??} extend \citeauthor{Wadler:2014}'s work to support polymorphism
-in GV. (polymorphism in CP) Future work is to support reduction under these
-terms.
+\end{comment}

### 2015-04-14 paper/abstract.tex
Polished abstract.

--- 2015-04-02 paper/abstract.tex
@@
 \begin{abstract}
-\begin{comment}
-SPL (+CPL)
-
-WTD
-
-ROF
-
-IMP
-\end{comment}
+Session types ensure that communicating agents adhere to an agreed upon
+protocol. The soundness proofs of session-based type systems are typically
+developed by hand and checked manually. The correctness of these proofs is
+crucial to maintain the assurances provided by session types. However,
+informal proofs are error-prone to develop and extend. The current paper
+describes work towards a formalisation of an existing two-tier session-based
+type system connecting classical linear logic and session types. The
+formalisation, developed in Coq, proves subject reduction for one of the tiers
+and makes strides towards the other system properties. This work illustrates
+the limitations of state-of-the-art techniques for formalising programming
+language metatheory and identifies a number of issues and guidelines of which
+practitioners in the field may wish to take account. The paper concludes by
+highlighting some areas of future work.
 \end{abstract}

### 2015-04-14 paper/introduction.tex
Introduction tweaked.

@@
 with a \textit{channel} whose endpoints have dual session types. A channel is
-defined as a bi-directional conduit for exchanging messages, not unlike a
+defined as a bidirectional conduit for exchanging messages, not unlike a
 socket in network programming. In contrast to sockets however, a channel is
@@
 approach, as it may be unclear how changes affect proofs and one will
-typically have to re-consider all proofs to be convinced the properties still
+typically have to reconsider all proofs to be convinced the properties still
 hold. More generally, unassisted formalisms are error-prone to develop and do
@@
 session types for mainstream programming languages can be studied, taking
-advantage of previous proof effort through the reuse of existing libraries
-for programming language metatheory. The work has presented a number of
-interesting challenges for the formalisation of programming languages and as a
-result generated a number of useful guidelines and issues to consider.
+advantage of previous proof effort through the reuse of existing libraries for
+programming language metatheory. This work has presented a number of
+interesting challenges for the formalisation of programming languages coupled
+with process calculi, and as a result generated a number of useful guidelines
+and issues to consider.

### 2015-04-14 paper/background.tex
Added the rest of GV's terms to figure.
Tweaked background (removing superfluous commas, etc.)

@@
 $\Branch{l_i}{S_i}{i\in I}$ can receive a label $l_i$ which will select the
-corresponding session type $S_i$ as the continuing type of the channel. While
-a channel with type $\Choice{l_i}{S_i}{i\in I}$ can send a label $l_i$ and
+corresponding session type $S_i$ as the continuing type of the channel. A
+channel with type $\Choice{l_i}{S_i}{i\in I}$ can send a label $l_i$ and
 continues as $S_i$.
@@
 $\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
-a session-based setting, since the intuition for \& is that of
-\emph{offering} (receiving) a choice between alternatives rather than
-\emph{making} (sending) the choice. Here, however, the intuition should be
-extended to account for its location in the typing judgement. The channel is
-on the left of the sequent, and \citeauthor{Caires:2010:STI} suggest one
-consider the context as services being provided to the typed process. Thus,
-the type of channels in the context is really the type of the other endpoint
-of the channel, and the typed process behaves according to the dual
-protocol. For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to
-represent the dual of a protocol $A$ as defined by
+a session-based setting, since the intuition for \& is that of \emph{offering}
+(receiving) a choice between alternatives rather than \emph{making} (sending)
+the choice. Here, however, the intuition should be extended to account for its
+location in the typing judgement. The channel is in the context (on the left
+of the sequent) and \citeauthor{Caires:2010:STI} suggest one consider the
+context as services being provided to the typed process. Thus, the type of
+channels in the context is really the type of the other endpoint of the
+channel, and the typed process behaves according to the dual protocol. For
+example, consider $\SILLbranchLTwo$, taking $A^\bot$ to represent the
+dual of a protocol $A$ as defined by
 \citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process is behaving
-according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from
+according to the protocol $\BChoice{A^\bot}{B^\bot}$ selecting from
 alternatives (in this case, $\inrterm$) and then behaving according to the
@@
 in the two-sided nature of the rules forcing one to commit to a certain
-convention when typing a communication system. Consider, an equivalent system
+convention when typing a communication system. Consider an equivalent system
 to the buy/quote example given by
@@
 $\BChoice{T}{R_2}$ and \eqref{eq:sbtyp} is derivable from
-$\BChoice{T}{L}$. Except for the final rule, these are exact dual rules of the
-rules used in example by \citeauthor{Caires:2010:STI}. Indeed, consider the
-composite process:
+$\BChoice{T}{L}$. Except for rule $T\1 R$, these are exact dual rules of the
+rules used in the example by \citeauthor{Caires:2010:STI}. Indeed, consider
+the composite process:
 \begin{gather*}
@@
 
-The final system is near identical to the one produced in the example given by
-\citeauthor{Caires:2010:STI} except the individual processes within the
-parallel composition have been swapped. Here the client communicates along the
-channel $c$ on the right-side of the sequent and the server utilises the
-channel $c$ on the left of the sequent. This typing, while correct, violates
-the intuition given by \citeauthor{Caires:2010:STI} of the context providing
-the services used by the process, since one would usually associate provision
-of services with the server not the client. Instead, the example above
-suggests the server utilises some channel provided by the client. On the other
-hand, one could imagine such a channel being setup between two communicating
-agents; the ``owner'' of the channel is largely irrelevant. Note that
-$ClientProto$ is in fact the dual of the $ServerProto$ definition by
-\citeauthor{Caires:2010:STI} using the definition of dual used in Wadler's
-classical process calculus~\cite{Wadler:2014}. In $\pi$DILL, the duality
-between endpoints of the same channel are obscured by the nature of the typing
-judgements.
+The final system is near identical to the original example except the
+individual processes within the parallel composition have been swapped. Here
+the client communicates along the channel $c$ on the right-side of the sequent
+and the server utilises the channel $c$ on the left of the sequent. This
+typing, while correct, violates the intuition given by
+\citeauthor{Caires:2010:STI} of the context providing the services used by the
+process, since one would usually associate provision of services with the
+server not the client. Instead, the example above suggests the server utilises
+some channel provided by the client. On the other hand, one could imagine such
+a channel being setup between two communicating agents; the ``owner'' of the
+channel is largely irrelevant. Note that $ClientProto$ is in fact the dual of
+the $ServerProto$ definition by \citeauthor{Caires:2010:STI} using the
+definition of dual from Wadler's classical process
+calculus~\cite{Wadler:2014}. In $\pi$DILL, the duality between endpoints of
+the same channel are obscured by the nature of the typing judgements.
 
@@
 
-\subsubsection{Fusion}
+\subsubsection{Fusion}\label{sec:fusion}
 

### 2015-04-14 paper/formalisation.tex
Tweaking some sections.

@@
 them. Considerations range from how much expressive power provided by the
-theorem prover to exploit to the choice of encoding to use for the properties
-of interest. Some of these decisions may impact the correspondence between the
-formal and informal systems e.g. the empty choice rule and omission of
-contraction in CP, and the omission of n-ary branch and choice from GV, and
-such things are necessarily a tradeoff between simplicity and
+theorem prover to exploit to the choice of which encoding to use for the
+properties of interest. Some of these decisions may impact the correspondence
+between the formal and informal systems e.g. the empty choice rule and
+omission of contraction in CP, and the omission of $n$-ary branch and choice
+from GV, and such things are necessarily a tradeoff between simplicity and
 adequacy.
@@
 assertions would be the \coqe!forwards! tactic permitting lemma
-instantiation. Consider the following example (during the proof of $assoc_1$):
+instantiation. Consider the following example (during the proof of
+\coqe!assoc_1!):
 \begin{tabular}{l}
@@
 $\GD$P0++$\GD$Q0 is unique. The proof follows from \key{PER} and \key{UN} upon
-application of an appropriate lemma, and there are two possible ways to prove
+application of an appropriate lemma and there are two possible ways to prove
 this using plain Coq. One way is to apply the lemma to one of \key{PER} or
@@
 namely backward reasoning, does not apply since the goal has nothing to do
-uniqueness of $\GD$P0++$\GD$Q0. However, using the \coqe!assert! tactic one
-can utilise backward reasoning. An assertion temporarily focuses a different
-goal (the asserted goal) which may then be tackled using tactics in the
-standard way. Unfortunately, this approach requires the explicit writing of
-the goal, namely \coqe!uniq ($\GD$P0++$\GD$Q0)!, into the proof script of the
-lemma cluttering the proof script sensitive to change because \coqe!$\GD$P0!
-and \coqe!$\GD$Q0! are ephemeral names introduced by prover.
+with the uniqueness of $\GD$P0++$\GD$Q0. However, using the \coqe!assert!
+tactic one can utilise backward reasoning. An assertion temporarily focuses a
+different goal (the asserted goal) which may then be tackled using tactics in
+the standard way. Unfortunately, this approach requires the explicit writing
+of the goal, namely \coqe!uniq ($\GD$P0++$\GD$Q0)!, into the proof script
+cluttering it with unnecessary details and making it sensitive to change
+because \coqe!$\GD$P0! and \coqe!$\GD$Q0! are ephemeral names introduced by
+theorem prover.
 
 The \coqe!forwards! tactic avoids the problem of explicit names by allowing
-the uniqueness property to be proven succiently:
+the uniqueness property to be proven succinctly:
 \begin{coq}

### 2015-04-14 paper/futurework.tex
Tweaking some sections.

@@
 \textbf{CP-GV system.} The issues regarding cut elimination suggest there
-should be greater study into calculi based on the sequent calculus. It is
-unclear to what extent process calculi such as CP and lambda calculi can be
+should be greater study of calculi based on the sequent calculus. It is
+unclear to what extent process calculi such as CP and $\lambda$ calculi can be
 unified to use the same framework or library. The different considerations of
-sequent and process calculi perhaps mandate a different library or
-framework. It would be interesting to consider CP in the setting described by
-\citeauthor{Tews:2013,Dawson:2010,Dawson:2002}, where explicit representation
-is given to derivations. The aim here would be to make progress in proving the
-cut elimination theorem for CP, but it is unclear how linearity and
-permutation of environments would be handled in this setting. Otherwise,
+sequent and process calculi perhaps mandate a different approach for each. It
+would be interesting to consider CP in the setting described by
+\citeauthor{Tews:2013} and \citeauthor{Dawson:2010}, where explicit
+representation is given to derivations. The aim here would be to make progress
+on proving the cut elimination theorem for CP, but it is unclear how linearity
+and permutation of environments would be handled in this setting. Otherwise,
 defining an appropriate termination measure for the current version of CP may
@@
 made in the current setting aiming to limit the duplication discussed
-previously The translation would be substantially affected by the
+previously. The translation would be substantially affected by the
 aforementioned change to CP because the derivations in the conlusions of the
@@
 \citeauthor{Lindley:2014:SAP} extend \citeauthor{Wadler:2014}'s work to
-support polymorphism in GV and n-ary branch and choice in CP. Future work is
+support polymorphism in GV and $n$-ary branch and choice in CP. Future work is
 to support these constructs.
@@
 system~\cite{Fahndrich:2002} provides a notion of aliased access to linear
-resources. The type system requires special language features, e.g. to manage
-capabilities. \citeauthor{Fahndrich:2002} do not provide operational semantics
-or soundness results for the type system. It would be interesting to
-investigate these properties formally in a session type setting. The ``focus''
-rule (and linearity generally) enforces an invariant on the focussed object
-(channel) such that the type is maintained during the focus operation. Such a
-rule would appear to only benefit recursive session types. Indeed, focussing
-on a non-recursive channel would appear pointless because one is prevented
-from communicating across it for the entire scope of the focus. Moreover, it
-would be undesirable to permit modification of a session type without
-guaranteeing exclusive access to the channel since this could lead to invalid
+resources.\todo{say what adopt and focus do} The type system requires special
+language features, e.g. to manage capabilities. \citeauthor{Fahndrich:2002} do
+not provide operational semantics or soundness results for the type system. It
+would be interesting to investigate these properties formally in a session
+type setting. The ``focus'' rule enforces an invariant on the focussed object
+(channel) such that the type is maintained during the focus operation. The
+object may be aliased while it is under a focus operation. Such a rule would
+appear to only benefit recursive session types. Indeed, focussing on a
+non-recursive channel would appear pointless because one is prevented from
+communicating across it for the entire scope of the focus. Moreover, it would
+be undesirable to permit modification of a session type without guaranteeing
+exclusive access to the channel since this could lead to invalid
 communication.

### 2015-04-15 paper/formalisation.tex
Finalised the first complete draft.

@@
 approach and has the disadvantage of being able to represent terms not valid
-in the language being mechanised (so called ``exotic'' terms). Further, a
+in the language being mechanised (so called ``exotic'' terms). Furthermore, a
 higher-order approach would prevent the use of the Metatheory library which
@@
 handle a language containing both linear and non-linear contexts. However, to
-stay true to CP/GV, I maintain a single context. Interestingly, it became
-clear that the mechanised version of CP required explicit weakening and
-contraction to facilitate the reduction rules (\S~\ref{sec:cp}).
+stay true to CP/GV, I maintain a single context. \citeauthor{Park:2014:MMW}
+describe a technique for removing non-linear contexts from typing
+judgements. While their work extended to \fpop it is not clear how to modify
+their technique to apply to an environment containing both non-linear and
+linear types.
 
@@
 fresh within the abstraction's body, does not produce strong enough induction
-principles in some cases, e.g. weakening~\cite{Aydemir:2008:EFM}. This
-decision impacts upon the premises of the CP/GV typing rules (\S~\ref{sec:cp}
-and \S~\ref{sec:gv}), however these changes do not alter the typing of
-processes nor the semantics of the translation from GV to CP. An equivalence
-between ``exists-fresh'' and cofinite quantification definitions for the
-simply-typed $\lambda$-calculus ($\stlc$) is given by
-\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
-result could be derived straightforwardly (but tediously) for the CP/GV
-system.
+principles in some cases, e.g. environment
+weakening~\cite{Aydemir:2008:EFM}. This decision impacts upon the premises of
+the CP/GV typing rules (\S~\ref{sec:cp} and \S~\ref{sec:gv}), however these
+changes do not alter the typing of terms or processes nor the semantics of the
+translation from GV to CP. An equivalence between ``exists-fresh'' and
+cofinite quantification definitions for the simply-typed $\lambda$-calculus
+($\stlc$) is given by \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM},
+and I suspect a similar result could be derived straightforwardly (but
+tediously) for the CP/GV system.
 
@@
 
-\citeauthor{Park:2014:MMW} describe a technique for removing non-linear
-contexts from typing judgements. While their work extended to \fpop it is not
-clear how to handle an environment containing both non-linear and linear types
-as in CP/GV. I wish to maintain as close a relationship as possible to the
-pen-and-paper system presented by \citeauthor{Wadler:2014}, so separating out
-the non-linear and linear components (as in \fpop) is not an option at this
-stage.
-
-\begin{comment}
-\section{Issues}
-
-\begin{itemize}
-\item What is required of a Metatheory library for linear type systems?
-  Permutation reasoning? Weakening/Contraction for non-linear components?
-\item Discuss the different approaches that were taken to maintain close
-  correspondence with the paper presentation
-\item Make sure to note that the paper leaves details (a lot of the ``cruft'')
-  out of the proofs e.g. where weakening is applied in the typing derivations
-\item Mention the use of Metatheory and its limitations when mechanising
-  linear type systems like GV/CP
-\end{itemize}
-\end{comment}
-
 \subsection{The process calculus CP}\label{sec:cp}
@@
 inductive type in Coq. Each typing rule in CP is represented as a constructor
-of the inductive type \coqe$cp_rule$. The inductive type is indexed by a
-process (\coqe$proc$) and an environment (\coqe$penv$) indicated by them
-appearing after the colon on the first line. \coqe$forall$ quantifies over
+of the inductive type \coqe!cp_rule!. The inductive type is indexed by a
+process (\coqe!proc!) and an environment (\coqe!penv!) indicated by them
+appearing after the colon on the first line. \coqe!forall! quantifies over
 hypotheses in a constructor; everything before the comma. Compound hypotheses
@@
 introduce them into the proof context during theorem proving. Some details
-such as environment uniqueness have been elided, but notice the use of LNCQ
-form within the hypotheses of the rules with binders. These rules exclude a
-finite set of names \coqe$L$ from opening the subprocesses where opening a
-process \coqe$P$ with name \coqe$x$ is denoted by \coqe$open_proc P x$ and
+such as environment uniqueness have been elided, but LNCQ form within
+hypotheses for the rules with binders has been kept. These rules exclude a
+finite set of names \coqe!L! from opening the subprocesses where opening a
+process \coqe!P! with name \coqe!x! is denoted by \coqe!open_proc P x! and
 corresponds to replacing de Bruijn index 0 with \coqe!x!. Bound variables are
 represented using de Bruijn indices and free variables are represented using
-an abstract type, \coqe$atom$, provided by the Metatheory library. The rules
+an abstract type, \coqe!atom!, provided by the Metatheory library. The rules
 with binders specify the type of a bound variable in the constructor rather
@@
 and transitivity, need to be applied manually. A lot of these cases are
-similar but there would not appear to be an advantage in developing a tactic
-to do the work. Doing so would create a dependency between a hard-to-debug and
-obscure Ltac definition and the proofs which utilise it, becoming a burden to
+similar but it would not be advantageous to develop a tactic to do the
+work. Doing so would create a dependency between a hard-to-debug and obscure
+Ltac definition and the proofs which utilise it, becoming a burden to
 change. Further, such a development philosophy treats Ltac tactics akin to
@@
 formalisms. To mitigate the complexity, the proof for each case was written as
-a separate lemma (termed ``reduction lemmata''). The last step in the proof of
-\coqe$proc_sub_red$, \coqe$eauto$, solves the goals generated by the induction
-using the reduction lemmata.
+a separate lemma (collectively termed ``reduction lemmata''). The last step in
+the proof of \coqe$proc_sub_red$, \coqe$eauto$, solves the goals generated by
+the induction using the reduction lemmata.
 
@@
 principal cut elimination rules for the interactions between servers and
-clients; spawning and garbage collection (weakening). The paper presentation
-does not specify where weakening could occur due to its implicit nature; it is
-tacitly assumed weakening can always be pushed further up the derivation tree
-to a base rule e.g. \coqe$cp_fwd$ rule. The same is true of contraction where
-it is defined as a substitution of names on process terms, not a
-syntax-directed rule.
+clients; spawning and garbage collection of a server with no clients
+(weakening). The paper presentation does not specify where weakening could
+occur due to its implicit nature and it is tacitly assumed weakening can
+always be pushed further up the derivation tree to a base rule
+e.g. \coqe$cp_fwd$ rule. The same is true of contraction where it is defined
+as a substitution of names on process terms, not a syntax-directed rule.
 
@@
 \begin{gather*}
-\tm{\nu A.(!0(A).P \mid ?z[[A]].Q)}
+\colored
+\tm{\nu \tp{A}.(!0(\tp{A}).P \mid ?z[[\tp{A}]].Q)}
 \quad \becomes \quad \\
-\tm{\nu A.(!0(A).P \mid \nu A.(!0(A).P \mid \key{contract}_{\Gamma} Q))}
+\colored
+\tm{\nu \tp{A}.(!0(\tp{A}).P \mid \nu
+                \tp{A}.(!0(\tp{A}).P \mid \key{contract}_{\Gamma} Q))}
 \end{gather*}
 
-where $\Gamma$ is the assumed environment typing \coqe$P$, $\tm{?z[[A]].Q}$
-denotes contraction by $z$ of two channels of type $A$ in $Q$ and
-$\key{contract}_{\Gamma} \tm{Q}$ contracts $\tm{Q}$ with the bindings in
-$\Gamma$. Unfortunately, as explained above, there is no way for the reduction
-relation to obtain the environment $\Gamma$.
+where $\Gamma$ is the assumed environment typing $P$,
+$\colored\tm{?z[[\tp{A}]].Q}$ denotes contraction by $z$ of two channels of
+type $A$ in $Q$ and $\key{contract}_{\Gamma} \tm{Q}$ contracts $\tm{Q}$ with
+the bindings in $\Gamma$. Unfortunately, as explained above, there is no way
+for the reduction relation to obtain the environment $\Gamma$.
 \end{samepage}
@@
 on \coqe$open_proc P x$. Similarly, to prove the converse statement
-\coqe$forall x, open_proc R x$ must be assumed. For commuting conversions,
-permutating binders is essential in formulating the majority of the rules.
+\coqe$forall x, open_proc R x$ must be assumed. A different associativity rule
+would occur if \coqe!Q! was locally closed in the above. For commuting
+conversions, permutating binders is essential in formulating the majority of
+the rules.
 
@@
 Regarding commuting conversions, many require analysing permutations
-containing singleton bindings. Consider the following the Coq encoding of the
+containing singleton bindings. Consider the following Coq encoding of the
 commuting conversion for output (cf. Figure~\ref{fig:ccoutput}):
@@
   forall P Q R (x:atom) A B $\Gamma$ (LCQ: lc_proc Q)
-         (WT: $\nu$ A.([B] x.(P $\mid$ Q) $\mid$ R) $\tpcp$ $\Gamma$),
-    [B] x.($\nu$ A.($\permb{0}{1}{P}$ $\mid$ R) $\mid$ Q) $\tpcp$ $\Gamma$.
+         (WT: $\nu$ A.(x[B].(P $\mid$ Q) $\mid$ R) $\tpcp$ $\Gamma$),
+    x[B].($\nu$ A.($\permb{0}{1}{P}$ $\mid$ R) $\mid$ Q) $\tpcp$ $\Gamma$.
 \end{coq}
@@
 
-After inverting \key{WT} and specialising the resulting hypotheses, regarding
-the subprocesses, with fresh variables the proof state is as follows:
+After inverting \key{WT} and specialising the resulting hypotheses regarding
+the subprocesses with fresh variables the proof state is as follows:
 
@@
 \begin{coq}
-x(B).($\nu$ A.($\open{0}{1}{P}$ $\mid$ Q)) $\tpcp$ $\Gamma$
+x[B].($\nu$ A.($\open{0}{1}{P}$ $\mid$ Q)) $\tpcp$ $\Gamma$
 \end{coq}
@@
 \coqe!$\GD$P0!). Since many of the commuting conversion rules required this
-sort of reasoning, separating the work out into a tactic appears
+sort of reasoning, separating the work out into a tactic proved
 beneficial. Indeed, Ltac is well-suited for this type of work, searching the
@@
 individual proof. The tactic is more modular and general in this form,
-permitting composition and permutation of tactic sequences.
+permitting composition and permutation with other tactics.
 
@@
 \begin{coq}
-exists R, ($\nu$A.([B]0.($P_1$ $\mid$ $P_2$) $\mid$ Q)) $\becomes^\star$ R
+exists R, ($\nu$A.(0[B].($P_1$ $\mid$ $P_2$) $\mid$ Q)) $\becomes^\star$ R
           /\ (is_cut R -> False)
@@
 Following informal proofs, it seems likely that the proof requires a
-termination measure involving the length of the derivation. In hindsight,
+termination measure involving the length of the derivation. In hindsight
 however, I do not believe the current encoding is the best for this sort of
@@
 Following the original work by \citeauthor{Wadler:2014}, the translation from
-GV to CP requires defining translations on types and terms of GV.
+GV to CP requires defining translations on types, environments and terms of
+GV. The translation on environments simply maps the translation on types over
+the bindings. The other two translations require more detailed explanation.
 
@@
 
-The translation of from GV types to CP propositons is shown in
+The translation of from GV types to CP propositions is shown in
 Figure~\ref{fig:typtrans}. Translation of session types is noteworthy for
@@
 corresponds to input in CP; the action on the channel endpoint is to output a
-value, and to do this the channel must input the value.
+value, and to do this the channel must input the value provided by the
+programmer.
 
@@
 The term translation was initially designed using a monadic approach to
-streamline handling of the possible error cases. The coq-ext-lib
+streamline handling of the error cases. The coq-ext-lib
 library~\footnote{https://github.com/coq-ext-lib/coq-ext-lib} provides
@@
 cluttering it with unnecessary details and making it sensitive to change
-because \coqe!$\GD$P0! and \coqe!$\GD$Q0! are ephemeral names introduced by
+because \coqe!$\GD$P0! and \coqe!$\GD$Q0! are mercurial names introduced by
 theorem prover.

### 2015-04-15 paper/futurework.tex
Finalised the first complete draft.

@@
 system~\cite{Fahndrich:2002} provides a notion of aliased access to linear
-resources.\todo{say what adopt and focus do} The type system requires special
-language features, e.g. to manage capabilities. \citeauthor{Fahndrich:2002} do
-not provide operational semantics or soundness results for the type system. It
-would be interesting to investigate these properties formally in a session
-type setting. The ``focus'' rule enforces an invariant on the focussed object
-(channel) such that the type is maintained during the focus operation. The
-object may be aliased while it is under a focus operation. Such a rule would
-appear to only benefit recursive session types. Indeed, focussing on a
-non-recursive channel would appear pointless because one is prevented from
-communicating across it for the entire scope of the focus. Moreover, it would
-be undesirable to permit modification of a session type without guaranteeing
-exclusive access to the channel since this could lead to invalid
-communication.
+resources. The ``adopt'' permits aliases on linear resources and ``focus'' is
+used to access a linear resource through one of the aliases. The type system
+requires special language features to manage
+capabilities. \citeauthor{Fahndrich:2002} do not provide operational semantics
+or soundness results for the type system. It would be interesting to
+investigate these properties formally in a session type setting. The ``focus''
+rule enforces an invariant on the focussed object (channel) such that the type
+is maintained during the focus operation. The object may be aliased while it
+is under a focus operation. Such a rule would appear to only benefit recursive
+session types. Indeed, focussing on a non-recursive channel would appear
+pointless because one is prevented from communicating across it for the entire
+scope of the focus. Moreover, it would be undesirable to permit modification
+of a session type without guaranteeing exclusive access to the channel since
+this could lead to invalid communication.
 

### 2015-04-18 paper/conclusion.tex
Spell checking and minor tweaks.

@@
 the process calculus CP and the functional language GV using state-of-the-art
-techniques from the programming language and interactivee theorem proving
+techniques from the programming language and interactive theorem proving
 communities. Proof of subject reduction for CP has been formalised along with
-work towards proving sound both cut elimination and the GV to CP CPS
+work towards formalising both cut elimination and the GV to CP CPS
 translation. While ultimately there remains work to be done to complete the
@@
 considered by future endeavours in the area. Finally, some ideas for future
-work, on GV/CP specifically but also generally in the area of session-based
-type system formalisations, were proffered.
+work on GV/CP specifically but also generally in the area of session-based
+type system formalisations were proffered.

### 2015-04-18 paper/formalisation.tex
Spell checking and minor tweaks.

@@
 effort was expended setting up this infrastructure. However, I believe one of
-the contributions of this work is to articulate what one needs from state of
-the art metatheory libraries in order to handle session-based languages, and
-more generally, linear type systems.
+the contributions of this work is to articulate what one needs from
+state-of-the-art metatheory libraries in order to handle session-based
+languages, and more generally, linear type systems.
 
@@
 handling permutations of environments as shown in the \coqe$cp_rule$
-definition. The Metatheory library does not currently support permutating
+definition. The Metatheory library does not currently support permuting
 environments, however some basic lemmata were added as a result of this work,
@@
 \coqe$P$ reduces to \coqe$Q$ then \coqe$Q$ is well-typed in $\Gamma$. The
-proof environment is started using \coqe$Proof.$ and then followed by a
-sequence of tactics until all goals have been proven. After execution of
-\coqe$Qed.$ the Coq system inspects the validity of the proof object
-constructed by the tactics and, if valid, will add the theorem to the
-environment.
+proof environment is started using \coqe$Proof.$ then followed by a sequence
+of tactics until all goals have been proven. After execution of \coqe$Qed.$
+the Coq system inspects the validity of the proof object constructed by the
+tactics and, if valid, will add the theorem to the environment.
 
@@
 example \coqe$ii$ repeatedly performs hypothesis introduction then
-simplification until no more can be peformed, and \coqe$gen$ $\Gamma$
+simplification until no more can be performed, and \coqe$gen$ $\Gamma$
 generalises the goal over the typing environment $\Gamma$.
@@
 inverting the hypothesis \key{WT} and specialising the resulting assumptions
-to a newly introduced fresh variable (to replace the de Bruijn index). In the
+to a newly introduced fresh variable (replacing the de Bruijn index). In the
 case of \coqe$reduce_spawn$, two inversions are required to get derivations
@@
 principal cut elimination rules for the interactions between servers and
-clients; spawning and garbage collection of a server with no clients
+clients: spawning and garbage collection of a server with no clients
 (weakening). The paper presentation does not specify where weakening could
@@
 the following lemma relating the free variables of a process and the domain of
-its environment:
+its environment:\pagebreak
 \begin{coq}
@@
 would occur if \coqe!Q! was locally closed in the above. For commuting
-conversions, permutating binders is essential in formulating the majority of
+conversions, permuting binders is essential in formulating the majority of
 the rules.
@@
 After inverting \key{WT} and specialising the resulting hypotheses regarding
-the subprocesses with fresh variables the proof state is as follows:
+the subprocesses to fresh variables the proof state is as follows:
 
@@
 however, I do not believe the current encoding is the best for this sort of
-reasoning. Formalised efforts for proving cut elimination for other logics
-have adopted an approach which permits reasoning on the structure of
-derivations represented explicitly as
+reasoning. Formalised efforts for proving cut elimination for logics, rather
+than process calculi, have adopted an approach which permits reasoning on the
+structure of derivations represented explicitly as
 trees~\cite{Tews:2013,Dawson:2010,Dawson:2002}. Whereas, in the approach taken
@@
 during the proving of a derivation e.g. a reduction lemma. It is future work
-to investigate these and other strategies which will assist in formalising
-cut-elimination for CP.
+to investigate these and other strategies for formalising cut-elimination for
+CP.
 
@@
 
-The typing judgements for the functional language GV are shown in
+Some typing judgements for the functional language GV are shown in
 Figure~\ref{fig:gv}. Note that I have altered the $n$-ary branch and choice
@@
 proofs. For instance, the decidability of equality on types now relied on an
-axiom permitting heterogeneous equality~\cite{McBride:1999}. Generally, it is
-better to avoid unnecessary postulates if possible. Further, when formulating
-the typing rules, the kind must be given to each type hypothesised and most
-rules are polymorphic in the kind, requring an awkward kind inspection
+axiom for heterogeneous equality~\cite{McBride:1999}. Generally, it is better
+to avoid unnecessary postulates if possible. Further, when formulating the
+typing rules, the kind had to be given to each type hypothesised and most
+rules are polymorphic in the kind, requiring an awkward kind inspection
 function to be applied to each type within a binding. The alternative approach
@@
 
-The translation of from GV types to CP propositions is shown in
+The translation from GV types to CP propositions is shown in
 Figure~\ref{fig:typtrans}. Translation of session types is noteworthy for
@@
 first define only the positive cases using an inductive type, the so-called
-\textit{graph relation}, and then define a dependent function that only
-considers the positive cases for which the relation is
-defined~\cite{McKinna:2009}. An example, translating choice, is shown in
-Figure~\ref{fig:trans}. The translation of choice considers \coqe!NL! and
-\coqe!NR! in a separate environment to \coqe!M! and types the resulting case
-term in the combined environment. The \coqe!NL! and \coqe!NR! subterms need to
-be considered as opened terms using cofinite quantification in order to yield
-suitable induction hypotheses. The graph relation is able to specify these
-constraints in contrast to the monadic function definition. However, several
-issues were encountered when attempting to specify the translation. The graph
-relation requires much of the clutter present in both the \coqe!cp_rule! and
-\coqe!wt_tm! definitions such as linear enviornments and cofinite
+\textit{graph relation}, then define a dependent function that only considers
+the positive cases for which the relation is defined~\cite{McKinna:2009}. An
+example, translating choice, is shown in Figure~\ref{fig:trans}. The
+translation of choice considers \coqe!NL! and \coqe!NR! in a separate
+environment to \coqe!M! and types the resulting case term in the combined
+environment. The \coqe!NL! and \coqe!NR! subterms need to be considered as
+opened terms using cofinite quantification in order to yield suitable
+induction hypotheses. The graph relation is able to specify these constraints
+in contrast to the monadic function definition. However, several issues were
+encountered when attempting to specify the translation. The graph relation
+requires much of the clutter present in both the \coqe!cp_rule! and
+\coqe!wt_tm! definitions such as linear environments and cofinite
 quantification, and unfortunately the specification of choice in
-Figure~\ref{fig:trans} is not quite correct. The process \coqe!P! is not being
-opened in the \key{TR} assumption yet in the conclusion it is not locally
-closed. It would appear that, in addition to opening GV subterms, the graph
-relation must also open CP subprocesses even in cases where there was no
+Figure~\ref{fig:trans} is not quite correct. The process \coqe!P! in the
+conclusion is not locally closed yet it is not being opened in the \key{TR}
+assumption. It would appear that, in addition to opening GV subterms, the
+graph relation must also open CP subprocesses even in cases where there was no
 binder in the original GV term e.g. function

### 2015-04-18 paper/futurework.tex
Spell checking and minor tweaks.

@@
 previously. The translation would be substantially affected by the
-aforementioned change to CP because the derivations in the conlusions of the
+aforementioned change to CP because the derivations in the conclusions of the
 rules would have to be constructed. In this case, it may be worth
@@
 rule enforces an invariant on the focussed object (channel) such that the type
-is maintained during the focus operation. The object may be aliased while it
-is under a focus operation. Such a rule would appear to only benefit recursive
-session types. Indeed, focussing on a non-recursive channel would appear
-pointless because one is prevented from communicating across it for the entire
-scope of the focus. Moreover, it would be undesirable to permit modification
-of a session type without guaranteeing exclusive access to the channel since
-this could lead to invalid communication.
+is maintained during the focus operation. Such a rule would appear to only
+benefit recursive session types. Indeed, focussing on a non-recursive channel
+would appear pointless because one is prevented from communicating across it
+for the entire scope of the focus. Moreover, it would be undesirable to permit
+modification of a session type without guaranteeing exclusive access to the
+channel since this could lead to invalid communication.
 

### 2015-04-18 paper/background.tex
Spell checking and minor tweaks.
Profitted -> Profited.

@@
 type theory; in the same way ``sequential'' type theory ($\lambda$-calculi)
-profitted from a logical perspective. However, for session types to become a
+profited from a logical perspective. However, for session types to become a
 more usable tool for programming distributed communication systems (the
@@
 \citeauthor{Gay:2001:FFP}~\cite{Gay:2001:FFP} provides a $\pi$-calculus
-framework mechanised in Isabelle/HOL. The aim is to build a mechanisaton of
+framework mechanised in Isabelle/HOL. The aim is to build a mechanisation of
 session-based type systems on top of the $\pi$-calculus framework. The use of

### 2015-04-18 paper/presentation.tex
Added presentation file with my initial musings on script.

\documentclass{beamer}

\usepackage[latin1]{inputenc}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{comment}
\usepackage{tikz}
\usetikzlibrary{arrows,positioning}
\usepackage[style=alphabetic,natbib=true]{biblatex}

\usetheme{Warsaw}

\title[Project Presentation]{Propositions as Sessions, Mechanically}
\author{Craig McLaughlin}
\institute{University Of Glasgow}
\date{\today}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

\begin{frame}
Hand-written (informal) proofs $=$ Error-prone $+$ Difficult to update
\end{frame}

today i am going to talk about a formalisation effort based on an existing
type system whose properties have previously only been studied
informally. first in order to motivate the previous and current work i shall
make two general observations here. one: distributed communication systems are
becoming increasingly prevalent in modern society. two: the correct
functioning of these systems hinges on the protocol between any two
communicating agents to be well-defined and correctly implemented.

one approach to achieve such correctness guarantees is by using binary session
types. [INTRODUCE BINARY SESSION TYPES HERE]. specifying a communication
protocol then involves providing mutual dual types to the communicating
parties. [a simple example with two processes is displayed].

there has been a lot of activity in this area producing a variety of different
session-based type systems. many systems build on top of previous work adding
more complex features such as asynchronous, polymorphic and recursive session
types. extending informal proofs is error-prone and difficult because it is
not always clear which parts of a system's metatheory is affected by a
change. a formalised system would be interesting to consider. the community
would benefit from a formalised system to aid extension and reduce likelihood
of errors in the proofs.

recent work has provided a logical foundation for understanding session
types. session types correspond to classical linear logic connectives. where
negation in the logic corresponds to the dual property of a session
type. [present example slide of the connectives in classical linear logic and
  their session type interpretation]. the formalisation i will describe today
regards this connection, the system GV/CP originally described by Phil Wadler
was the aim of this project to mechanise it. it provides a basis for further
additions and also provides some properties that may be desirable such as
deadlock freedom. the formalisation provides a means to gain more confidence
in the proofs but also assists in extending them to account for the effects of
additions.

\begin{comment}
whats the problem
what have others done
why are we doing it
what did we do
what have we learned
\end{comment}
\begin{frame}
\frametitle{Preamble}
\item Propositions as Sessions
\item system features
\item Coq
\item encoding
\item subject reduction theorem
\item difficulties in formalisation
\item implications
\item future work
\end{frame}

\printbibliography

\end{document}

### 2015-04-18 paper/presentation.tex
Working out how to introduce CP/GV.

@@
 not always clear which parts of a system's metatheory is affected by a
-change. a formalised system would be interesting to consider. the community
-would benefit from a formalised system to aid extension and reduce likelihood
-of errors in the proofs.
+change. a formalised system would aid extension and reduce likelihood
+of errors in the proofs since the theorem prover would highlight the effects
 
 recent work has provided a logical foundation for understanding session
-types. session types correspond to classical linear logic connectives. where
-negation in the logic corresponds to the dual property of a session
-type. [present example slide of the connectives in classical linear logic and
-  their session type interpretation]. the formalisation i will describe today
-regards this connection, the system GV/CP originally described by Phil Wadler
-was the aim of this project to mechanise it. it provides a basis for further
-additions and also provides some properties that may be desirable such as
-deadlock freedom. the formalisation provides a means to gain more confidence
-in the proofs but also assists in extending them to account for the effects of
-additions.
+types. Wadler's work on the CP/GV system describes a two-tier type system with
+session types. [present it on screen.what?] the process calculus CP interprets
+linear logical connectives as session types and the surface functional
+language GV can be translated into CP. The system enjoys freedom from
+deadlock. surface language.
+
+session types correspond to classical linear logic connectives. where negation
+in the logic corresponds to the dual property of a session type. [present
+  example slide of the connectives in classical linear logic and their session
+  type interpretation]. the formalisation i will describe today regards this
+connection, I will present work towards a formalisation of the system GV/CP
+originally described by Phil Wadler was the aim of this project to mechanise
+it. it provides a basis for further additions and also provides some
+properties that may be desirable such as deadlock freedom. the formalisation
+provides a means to gain more confidence in the proofs but also assists in
+extending them to account for the effects of additions.
 

### 2015-04-20 paper/presentation.tex
Added some ideas for the presentation script.

@@
 
-\begin{frame}
-Hand-written (informal) proofs $=$ Error-prone $+$ Difficult to update
-\end{frame}
-
 today i am going to talk about a formalisation effort based on an existing
@@
 
+\begin{frame}
+Session types by example
+\end{frame}
+
 one approach to achieve such correctness guarantees is by using binary session
@@
 parties. [a simple example with two processes is displayed].
+
+\begin{frame}
+Hand-written (informal) proofs $=$ Error-prone $+$ Difficult to update
+\end{frame}
 
@@
 not always clear which parts of a system's metatheory is affected by a
-change. a formalised system would aid extension and reduce likelihood
-of errors in the proofs since the theorem prover would highlight the effects
+change. a formalised system would reduce likelihood of errors in the proofs
+and aid extension since the theorem prover would highlight changes to the
+metatheory
 
-recent work has provided a logical foundation for understanding session
-types. Wadler's work on the CP/GV system describes a two-tier type system with
-session types. [present it on screen.what?] the process calculus CP interprets
-linear logical connectives as session types and the surface functional
-language GV can be translated into CP. The system enjoys freedom from
-deadlock. surface language.
+\begin{frame}
+CP - process calculus with operators from classical linear logic
+GV - functional language with session type
+CP -> GV translation (interprets classical linear logic operators as session
+types)
+\end{frame}
 
-session types correspond to classical linear logic connectives. where negation
-in the logic corresponds to the dual property of a session type. [present
-  example slide of the connectives in classical linear logic and their session
-  type interpretation]. the formalisation i will describe today regards this
+the formalisation i present here is based on recent work providing a logical
+foundation for understanding session types. in particular i present work
+towards a formalisation of Wadler's original CP/GV system which describes a
+two-tier type system with session types. A high-level overview of the system
+is as follows: we have a process calculus CP with operators from classical
+linear logic, a high-level functional language GV with session types, and a
+translation from GV to CP which interprets operators in classical linear logic
+as session types.
+
+\begin{frame}
+an example of CP classical linear logic operators and their session type
+interpretation
+\end{frame}
+
+by classical linear logic i mean a logic where negation is an involution which
+restricts the use of certain rules on resources, namely that all linear
+resources must be used (corresponding to prohibiting weakening) and that one
+cannot duplicate a linear resource (corresponding to contraction). some
+operators in the CP calculus are presented. the interpretation of tensor
+product is output of A and continue as B. The dual of this is input of A and
+continue as the negation of B. Thus negation in classical linear logic
+corresponds to session duality.
+
+\begin{frame}
+an example of GV and a translation???
+\end{frame}
+
+GV has built-in support for session types. consider the rule for sending a
+value of type T along a channel N of type !T.S. the translation to CP is
+presented. blah blah something about the session types being a little odd or
+overlook this probably
+
+\begin{frame}
+- subject reduction of CP
+- top-level cut elimination
+- translation preserving well-typedness
+\end{frame}
+
+the system has three main properties. subject reduction of CP, top-level
+cut-elimination and translation preserving well-typedness. One out of the
+three has been formalised and some issues and guidelines have been developed
+accordingly. some comments on progress towards the other two is offered
+
+The system
+enjoys freedom from deadlock cut elimination in CP corresponds to
+communication in GV.
+
+the formalisation i will describe today regards this
 connection, I will present work towards a formalisation of the system GV/CP

### 2015-04-20 paper/presentation.tex
Added some more details for presentation script.

@@
 
-the formalisation i present here is based on recent work providing a logical
-foundation for understanding session types. in particular i present work
-towards a formalisation of Wadler's original CP/GV system which describes a
-two-tier type system with session types. A high-level overview of the system
-is as follows: we have a process calculus CP with operators from classical
-linear logic, a high-level functional language GV with session types, and a
+the formalisation i present here is an encoding of a system in the Coq proof
+assistant. it is based on a two-tier system providing a logical foundation for
+understanding session types. A high-level overview of the system is as
+follows: we have a process calculus CP with operators from classical linear
+logic, a high-level functional language GV with session types, and a
 translation from GV to CP which interprets operators in classical linear logic
@@
 cannot duplicate a linear resource (corresponding to contraction). some
-operators in the CP calculus are presented. the interpretation of tensor
-product is output of A and continue as B. The dual of this is input of A and
-continue as the negation of B. Thus negation in classical linear logic
-corresponds to session duality.
+operators in the CP calculus are presented as they appear in the Coq encoding
+(note that they are very similar to their informal counterparts). the
+interpretation of tensor product is output of A and continue as B. The dual of
+this is input of A and continue as the negation of B. Thus negation in
+classical linear logic corresponds to session duality.
 
@@
 
-The system
-enjoys freedom from deadlock cut elimination in CP corresponds to
+The system enjoys freedom from deadlock cut elimination in CP corresponds to
 communication in GV.
 
-the formalisation i will describe today regards this
-connection, I will present work towards a formalisation of the system GV/CP
-originally described by Phil Wadler was the aim of this project to mechanise
-it. it provides a basis for further additions and also provides some
-properties that may be desirable such as deadlock freedom. the formalisation
-provides a means to gain more confidence in the proofs but also assists in
-extending them to account for the effects of additions.
+subject reduction is formalised using a relation on processes
 
-\begin{comment}
-whats the problem
-what have others done
-why are we doing it
-what did we do
-what have we learned
-\end{comment}
+cut elimination is not currently proven since we require a termination measure
+on the length of a derivation and a relationship between the subprocesses of a
+cut. other work based on formalising logics where the derivations are explicit
+may be inspirational
+
+the translation seems to require a duplication of the specifications for cp
+and gv typing rules not done
+
+what can we show them for the failures?
+
+here are some issues and things learnt! -- our guidelines are all about
+modularisation, representation, library reuse, automation!
+
 \begin{frame}

### 2015-04-21 paper/presentation.tex
Some boilerplate for slides.
Working on the slide content for presentation.

@@
 \usepackage[latin1]{inputenc}
+\usepackage{lstmacros}
 \usepackage{graphicx}
+\usepackage{array}
 \usepackage{subfigure}
 \usepackage{amsmath}
-\usepackage{listings}
 \usepackage{comment}
+\usepackage{etoolbox}
 \usepackage{tikz}
@@
 \usetheme{Warsaw}
+
+\newtoggle{tdlp}
+\toggletrue{tdlp}
+\input{propmacros}
+\newcommand{\altt}{~|~}
 
@@
 
+\begin{comment}
 today i am going to talk about a formalisation effort based on an existing
@@
 communicating agents to be well-defined and correctly implemented.
+\end{comment}
 
 \begin{frame}
-Session types by example
+\frametitle{Session Types}
+\begin{math}
+  \begin{array}{rclc}
+    S & \bnf  & \inpt{T}{S} & \mbox{input value of type T, continue as S}\\
+      & \altt & \outpt{T}{S} & \mbox{output value of type T, continue as S}\\
+      & \altt & \Branch{l_i}{S_i}{i\in I} &
+                           \mbox{offer between $i$ alternatives}\\
+      & \altt & \Choice{l_i}{S_i}{i\in I} &
+                           \mbox{choose between $i$ alternatives}\\
+      & \altt & \mbox{\lstinline{end}} & \mbox{terminate session}
+ \end{array}
+\end{math}
 \end{frame}
 
+\begin{comment}
 one approach to achieve such correctness guarantees is by using binary session
@@
 parties. [a simple example with two processes is displayed].
+\end{comment}
 
@@
 
+\begin{comment}
 there has been a lot of activity in this area producing a variety of different
@@
 metatheory
+\end{comment}
 
@@
 
+\begin{comment}
 the formalisation i present here is an encoding of a system in the Coq proof
@@
 as session types.
+\end{comment}
 
 \begin{frame}
+negation as involution
 an example of CP classical linear logic operators and their session type
-interpretation
+interpretation (times/with)
 \end{frame}
 
+\begin{comment}
 by classical linear logic i mean a logic where negation is an involution which
@@
 resources must be used (corresponding to prohibiting weakening) and that one
-cannot duplicate a linear resource (corresponding to contraction). some
-operators in the CP calculus are presented as they appear in the Coq encoding
+cannot duplicate a linear resource (corresponding to contraction). some typing
+rules for the CP calculus are presented as they appear in the Coq encoding
 (note that they are very similar to their informal counterparts). the
@@
 classical linear logic corresponds to session duality.
+\end{comment}
 
 \begin{frame}
-an example of GV and a translation???
+an example of GV
 \end{frame}
 
+\begin{comment}
 GV has built-in support for session types. consider the rule for sending a
-value of type T along a channel N of type !T.S. the translation to CP is
+value of type T along a channel N of type !T.S.
+\end{comment}
+
+\begin{frame}
+something about a translation
+\end{frame}
+
+\begin{comment}
+the translation to CP is
 presented. blah blah something about the session types being a little odd or
 overlook this probably
+\end{comment}
 
@@
 
+\begin{comment}
 the system has three main properties. subject reduction of CP, top-level
-cut-elimination and translation preserving well-typedness. One out of the
-three has been formalised and some issues and guidelines have been developed
-accordingly. some comments on progress towards the other two is offered
+cut-elimination and translation preserving well-typedness. of which subject
+reduction is the only one fully formalised. the other two presented some
+challenges but some comments on progress towards the other two is
+offered. i also highlight some issues and guidelines
+\end{comment}
 
-The system enjoys freedom from deadlock cut elimination in CP corresponds to
-communication in GV.
+\begin{frame}
+subject reduction
+\end{frame}
 
+\begin{comment}
 subject reduction is formalised using a relation on processes
+\end{comment}
+
+\begin{frame}
+problems for cut elimination
+\end{frame}
+
+\begin{comment}
+sessions correspond to propositions
+processes correspond to proofs
+cut elimination corresponds to communication
 
@@
 may be inspirational
+\end{comment}
 
+\begin{frame}
+problems of the translation
+\end{frame}
+
+\begin{comment}
 the translation seems to require a duplication of the specifications for cp
@@
 modularisation, representation, library reuse, automation!
+\end{comment}
 
@@
 \frametitle{Preamble}
+\begin{itemize}
 \item Propositions as Sessions
@@
 \item future work
+\end{itemize}
 \end{frame}

### 2015-04-22 paper/conclusion.tex
GV/CP -> CP/GV.

@@
 The current paper presented work towards a formalisation of
-\citeauthor{Wadler:2014}'s two-tier GV/CP system. The Coq development defines
+\citeauthor{Wadler:2014}'s two-tier CP/GV system. The Coq development defines
 the process calculus CP and the functional language GV using state-of-the-art
@@
 considered by future endeavours in the area. Finally, some ideas for future
-work on GV/CP specifically but also generally in the area of session-based
+work on CP/GV specifically but also generally in the area of session-based
 type system formalisations were proffered.

### 2015-04-22 paper/presentation.tex
I have some more ideas on how to proceed (script and slide-wise).

@@
 \usepackage[latin1]{inputenc}
+\usepackage{cmll}
 \usepackage{lstmacros}
@@
 
+\let\oldframetitle\frametitle
+\renewcommand{\frametitle}[1]{
+  \oldframetitle{#1}\vspace{-3.5cm}
+}
+
 \title[Project Presentation]{Propositions as Sessions, Mechanically}
@@
 \end{frame}
+
+\section{Preamble}
 
@@
 \begin{frame}
-\frametitle{Session Types}
+\oldframetitle{Session Types}
 \begin{math}
-  \begin{array}{rclc}
-    S & \bnf  & \inpt{T}{S} & \mbox{input value of type T, continue as S}\\
-      & \altt & \outpt{T}{S} & \mbox{output value of type T, continue as S}\\
+  \begin{array}{cccc}
+    S & \bnf  & \inpt{T}{S} & \mbox{receive value of type T, continue as S}\\
+      & \altt & \outpt{T}{S} & \mbox{send value of type T, continue as S}\\
       & \altt & \Branch{l_i}{S_i}{i\in I} &
-                           \mbox{offer between $i$ alternatives}\\
+                           \mbox{offer between $|I|$ alternatives}\\
       & \altt & \Choice{l_i}{S_i}{i\in I} &
-                           \mbox{choose between $i$ alternatives}\\
-      & \altt & \mbox{\lstinline{end}} & \mbox{terminate session}
+                           \mbox{choose between $|I|$ alternatives}\\
+      & \altt & \mbox{\lstinline{end}} & \mbox{terminate session}\\
+ \dual{S} & \bnf & \mbox{dual of S}\\
+ \dual{(\inpt{T}{S})} & \bnf & \outpt{T}{\dual{S}}\\
+ \dual{\Branch{l_i}{S_i}{i\in I}} & \bnf & \Choice{l_i}{\dual{S_i}}{i\in I}\\
+            \vdots                & \cdots & \vdots
  \end{array}
@@
 \begin{frame}
+\frametitle{Motivation}
 Hand-written (informal) proofs $=$ Error-prone $+$ Difficult to update
@@
 \begin{frame}
-CP - process calculus with operators from classical linear logic
-GV - functional language with session type
-CP -> GV translation (interprets classical linear logic operators as session
-types)
+\frametitle{System Overview}
+CP $=$ process calculus with operators from classical linear logic
+
+GV $=$ functional language with session types
+
+GV $\to$ CP translation
 \end{frame}
@@
 \begin{frame}
-negation as involution
-an example of CP classical linear logic operators and their session type
-interpretation (times/with)
+\frametitle{CP Propositions}
+\begin{math}
+\begin{array}{rclc}
+    \tp{A,B,C} & \bnf
+          & \tp{A \otimes B} & \key{times: output A then behave as B}\\
+          & \altt & \tp{A \parr B} & \key{par: input A then behave as B}\\
+          & \altt & \cdots
+\end{array}
+\end{math}
+
+For all propositions $A$, $\dualbt{(\dualbt{A})} = A$ (negation is an
+involution)
 \end{frame}
@@
 classical linear logic corresponds to session duality.
+\end{comment}
+
+\section{Formalisation}
+
+\begin{frame}
+\frametitle{Goals}
+\begin{itemize}
+\item Reuse where possible
+\item Minimise difference between informal/formal variants
+\item Avoid known pitfalls
+\end{itemize}
+\end{frame}
+
+\begin{comment}
+The encoding follows practices in formalising programming language metatheory
+with a principal design goal of reusing existing libraries from previous
+developments. Additionally, I want to keep as close a correspondence as
+possible between the development and the pen-and-paper presentation in order
+to mitigate questions of adequacy, you know, whether the formal system is what
+you intended to formalise. And lastly, I aimed to exploit the features of the
+Coq system without over complicating the development such as taking careful
+consideration over representation choices.
+\end{comment}
+
+\begin{frame}
+\oldframetitle{Preliminaries}
+\begin{itemize}
+\item Locally Nameless
+  \begin{itemize}
+  \item Avoids variable capture
+  \item de Bruijn indices for bound variables
+    ($\lambda~(\lambda~1) \equiv \lambda x. \lambda y. x$)
+  \end{itemize}
+\item Cofinite Quantification
+  \begin{itemize}
+  \item stronger induction principle
+  \item
+    \begin{math}
+    \begin{array}{lcl}
+      \mbox{Exists-Fresh} & & \mbox{Cofinite}\\
+      \inference{
+        x \notin FV(N) &
+        \Tp{\Gamma \comma \Tmof{x}T} \Tpvdash \Tmof{N^x}\Tp{U}
+      }{}{
+        \Tp{\Gamma} \Tpvdash \Tmof{abs~N}\Tp{T \to U}
+      }
+    & \mbox{vs}. &
+      \inference{
+        \forall x \notin L.
+        (\Tp{\Gamma \comma \Tmof{x}T} \Tpvdash \Tmof{N^x}\Tp{U})
+      }{}{
+        \Tp{\Gamma} \Tpvdash \Tmof{abs~N}\Tp{T \to U}
+      }
+    \end{array}
+    \end{math}
+  \end{itemize}
+\item Metatheory library for representation
+\end{itemize}
+\end{frame}
+
+\begin{comment}
+one of the key issues when formalising programming language metatheory is how
+to represent variables both bound and free. I chose the locally nameless
+representation which separates bound and free variables into distinct
+entities. Thus it avoids the variable capture problem in contrast to other
+approaches.
+
+In this approach, a de Bruijn index is a number used to represent a bound
+variable which indicates the position of its binder (starting from zero for
+the innermost binder).
+
+Constructs with binders have to open the term with a free variable and this
+requires some technique for handling freshness of the opening variable. I
+chose to adopt the cofinite quantification approach which allows to exclude a
+finite set of names from consideration. in contrast to the sometimes weaker
+notion of a single sufficiently fresh name
+\end{comment}
+
+\begin{frame}
+another frame for cp this time its well typing time
+\end{frame}
+
+\begin{comment}
+lets take a look at cp's typing rules. here we used a well typed relation on
+the process terms of cp. we require additional assumptions for environment
+ordering and
 \end{comment}
@@
 \begin{frame}
-\frametitle{Preamble}
 \begin{itemize}

### 2015-04-22 paper/presentation.tex
Working on comparing CP typing rules (informal vs. formal).

@@
 
+\defverbatim\cprule{
+\begin{coq}
+Inductive cp_rule : proc -> penv -> Prop :=
+  | $\cdots$
+  | cp_cut :
+      forall (L:atoms) P Q A $\Gamma$ $\Delta$P $\Delta$Q
+             (PER: Permutation $\Gamma$ ($\Delta$P ++ $\Delta$Q))
+             (CPP: forall (x:atom) (NL: x $\notin$ L),
+                     (open_proc P x) $\tpcp$ (x $\sim$ A) ++ $\Delta$P)
+             (CPQ: forall (x:atom) (NL: x $\notin$ L),
+                (open_proc Q x) $\tpcp$ (x $\sim$ $\dualbt{A}$) ++ $\Delta$Q),
+        $\nu$ A.(P $\mid$ Q) $\tpcp$ $\Gamma$
+  | $\cdots$
+\end{coq}
+}
+
 \begin{frame}
-another frame for cp this time its well typing time
+\cprule
 \end{frame}
@@
 the process terms of cp. we require additional assumptions for environment
-ordering and
+ordering and more attention is paid to the opening of subprocesses. this
+presents the informal and formal definitions of cut.
 \end{comment}

### 2015-04-23 paper/presentation.tex
Progress being made on presentation structure.

--- 2015-04-22 paper/presentation.tex
@@
 
-\begin{frame}
-another frame for cp this time its well typing time
+\subsection{CP}
+
+\begin{frame}[fragile]
+\oldframetitle{CP Rules: Formal vs. Informal Comparison}
+\begin{tabular}{lc}
+\begin{coqp}
+Inductive cp_rule : proc -> penv -> Prop :=
+  | $\cdots$
+  | cp_cut :
+      forall (L:atoms) P Q A $\Gamma$ $\Delta$P $\Delta$Q
+             (PER: Permutation $\Gamma$ ($\Delta$P ++ $\Delta$Q))
+             (CPP: forall (x:atom) (NL: x $\notin$ L),
+                     (open_proc P x) $\tpcp$ (x $\sim$ A) ++ $\Delta$P)
+             (CPQ: forall (x:atom) (NL: x $\notin$ L),
+                (open_proc Q x) $\tpcp$ (x $\sim$ $\dualbt{A}$) ++ $\Delta$Q),
+        $\nu$ A.(P $\mid$ Q) $\tpcp$ $\Gamma$
+  | $\cdots$
+\end{coqp} &
+\begin{math}
+\colored
+\inference{
+  \tm{P} \tpvdash \tp{\Gamma \comma \tmof{x}A}
+  &
+  \tm{Q} \tpvdash \tp{\Delta \comma \tmof{x}\dualbt{A}}
+}{}{
+  \tm{\nu x \of{A}.(P \mid Q)} \tpvdash \tp{\Gamma \comma \Delta}
+}
+\end{math}
+\end{tabular}
+
+%% \begin{coqp}
+%% Inductive cp_rule : proc -> penv -> Prop :=
+%%   | $\cdots$
+%%   | cp_cut :
+%%       forall (L:atoms) P Q A $\Gamma$ $\Delta$P $\Delta$Q
+%%              (PER: Permutation $\Gamma$ ($\Delta$P ++ $\Delta$Q))
+%%              (CPP: forall (x:atom) (NL: x $\notin$ L),
+%%                      (open_proc P x) $\tpcp$ (x $\sim$ A) ++ $\Delta$P)
+%%              (CPQ: forall (x:atom) (NL: x $\notin$ L),
+%%                 (open_proc Q x) $\tpcp$ (x $\sim$ $\dualbt{A}$) ++ $\Delta$Q),
+%%         $\nu$ A.(P $\mid$ Q) $\tpcp$ $\Gamma$
+%%   | $\cdots$
+%% \end{coqp}
 \end{frame}
@@
 the process terms of cp. we require additional assumptions for environment
-ordering and
+ordering and more attention is paid to the opening of subprocesses. this
+presents the informal and formal definitions of cut. There is no overhead when
+introducing binders to the subprocesses or for environment ordering, but apart
+from these representation issues the rules look similar which could assist the
+uptake of such formal tools by programming language researchers.
 \end{comment}
 
+\defverbatim[colored]\subred{
+\begin{coq}
+Theorem proc_sub_red:
+  forall $\Gamma$ P Q
+         (WT: P $\tpvdash_{cp}$ $\Gamma$)
+         (RED: P $\becomes$ Q),
+    Q $\tpvdash_{cp}$ $\Gamma$.
+\end{coq}
+}
+
 \begin{frame}
-an example of GV
+\oldframetitle{Subject Reduction in CP}
+\subred
+\begin{itemize}
+\item {\fontsize{6}{7.2}\selectfont blah one}
+\item blah two
+\item blah three
+\item blah four
+\item blah five
+\item blah six
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Subject Reduction in CP}
+\[\colored
+\begin{array}{c}
+\mbox{Subject reduction} = \left\{ \begin{array}{ll}
+                                   \mbox{Principal cut reductions}~(\beta)\\
+                                   \mbox{Commuting Conversions}~(\kappa)
+                                   \end{array} \right.
+\\~\\
+(\kappa_{\otimes1}) \quad
+\tm{\nu z\Of{C}.(x[y].(P \mid Q) \mid R)}
+\quad \becomes \hfill
+\hfill \tm{x[y].(\nu z\Of{C}.(P \mid R) \mid Q)}
+\end{array}
+\]
+\end{frame}
+
+\begin{comment}
+Subject reduction in CP proceeds by cut reductions of the logical operators
+against their duals and \textit{commuting conversions} as well as some other
+rules. Commuting conversions push a cut inside a logical operator and the
+intuition is that they permit further communication to occur which was guarded
+by a cut.
+\end{comment}
+
+\subsection{GV}
+
+\begin{frame}
+GV
 \end{frame}
@@
 \end{comment}
+
+\subsection{Translation}
 
@@
 \end{frame}
+
+\subsection{Issues}
 
@@
 
+\section{Future Work}
+
 \begin{frame}

### 2015-04-23 paper/presentation.tex
Added more slides and figures (emulating my paper presentation).

@@
 \end{tabular}
-
-%% \begin{coqp}
-%% Inductive cp_rule : proc -> penv -> Prop :=
-%%   | $\cdots$
-%%   | cp_cut :
-%%       forall (L:atoms) P Q A $\Gamma$ $\Delta$P $\Delta$Q
-%%              (PER: Permutation $\Gamma$ ($\Delta$P ++ $\Delta$Q))
-%%              (CPP: forall (x:atom) (NL: x $\notin$ L),
-%%                      (open_proc P x) $\tpcp$ (x $\sim$ A) ++ $\Delta$P)
-%%              (CPQ: forall (x:atom) (NL: x $\notin$ L),
-%%                 (open_proc Q x) $\tpcp$ (x $\sim$ $\dualbt{A}$) ++ $\Delta$Q),
-%%         $\nu$ A.(P $\mid$ Q) $\tpcp$ $\Gamma$
-%%   | $\cdots$
-%% \end{coqp}
 \end{frame}
@@
 \begin{comment}
-lets take a look at cp's typing rules. here we used a well typed relation on
-the process terms of cp. we require additional assumptions for environment
-ordering and more attention is paid to the opening of subprocesses. this
+CP rules were encoded as an inductive in Coq and an example for cut is shown
+here. CP provided challenges for metatheory library since it differ in many
+respects to lambda calculi. for example, in the lambda calculus one does not
+work up to permutation of binders whereas in a process calculus these kinds of
+structural equivalences are standard. we require additional assumptions for
+environment ordering and more attention is paid to the opening of
+subprocesses. metatheory did not support permutations and as a result of this
+work some basic operations and custom tactics were developed. the custom
+tactics can solve simple permutation goals via simple proof search. this slide
 presents the informal and formal definitions of cut. There is no overhead when
@@
 
-\begin{frame}
+\begin{frame}[fragile]
 \oldframetitle{Subject Reduction in CP}
-\subred
+\begin{tabular}{ll}
+\begin{coq}
+Theorem proc_sub_red:
+  forall $\Gamma$ P Q
+         (WT: P $\tpvdash_{cp}$ $\Gamma$)
+         (RED: P $\becomes$ Q),
+    Q $\tpvdash_{cp}$ $\Gamma$.
+\end{coq} &
+\begin{math}
+\becomes \defeq \left\{ \begin{array}{ll}
+                   \mbox{Principal cut reductions}~(\beta)\\
+                   \mbox{Commuting Conversions}~(\kappa)
+                   \end{array} \right.
+\end{math}
+\end{tabular}
 \begin{itemize}
@@
 
-\begin{frame}
-\frametitle{Subject Reduction in CP}
-\[\colored
-\begin{array}{c}
-\mbox{Subject reduction} = \left\{ \begin{array}{ll}
-                                   \mbox{Principal cut reductions}~(\beta)\\
-                                   \mbox{Commuting Conversions}~(\kappa)
-                                   \end{array} \right.
-\\~\\
-(\kappa_{\otimes1}) \quad
-\tm{\nu z\Of{C}.(x[y].(P \mid Q) \mid R)}
-\quad \becomes \hfill
-\hfill \tm{x[y].(\nu z\Of{C}.(P \mid R) \mid Q)}
-\end{array}
-\]
-\end{frame}
-
 \begin{comment}
@@
 \end{comment}
+
+\begin{frame}[fragile]
+\oldframetitle{Principal Cut Reduction ($\beta_{!?}$)}
+\begin{tabular}{c}
+\begin{coq}
+Lemma reduce_spawn:
+  forall P Q A dA $\Gamma$
+         (DUA: dual_props A dA)
+         (WT: $\nu$ !A.(!0(A).P $\mid$ ?0[dA].Q) $\tpcp$ $\Gamma$),
+    $\nu$ A.(P $\mid$ Q) $\tpcp$ $\Gamma$.
+\end{coq}
+\\~\\
+\begin{math}
+\begin{array}{c}
+\small
+\colored
+\inference{
+  \inference{
+    \tm{P} \tpvdash \tp{{?\Gamma} \comma \tmof{y}A}
+  }{!}{
+    \tm{!x(y).P} \tpvdash \tp{{?\Gamma} \comma \tmof{x}{!A}}
+  }
+  &
+  \inference{
+    \tm{Q} \tpvdash \tp{\Delta \comma \tmof{y}\dualbt{A}}
+  }{?}{
+    \tm{?x[y].Q} \tpvdash \tp{\Delta \comma \tmof{x}{?\dualbt{A}}}
+  }
+}{}{
+  \tm{\nu x \Of{!A}.(!x(y).P \mid {?x[y].Q})} \tpvdash
+    \tp{{?\Gamma} \comma \Delta}
+}
+\becomes
+\inference{
+  \tm{P} \tpvdash \tp{{?\Gamma} \comma \tmof{y}A}
+  \\
+  \tm{Q} \tpvdash \tp{\Delta \comma \tmof{y}\dualbt{A}}
+}{}{
+  \tm{\nu y\Of{A}.(P \mid Q)} \tpvdash \tp{{?\Gamma} \comma \Delta}
+}
+\normalsize
+\end{array}
+\end{math}
+\end{tabular}
+\end{frame}
+
+\begin{frame}
+\[\colored
+\begin{array}{c}
+(\beta_{!?}) \quad
+\tm{\nu x \Of{!A}.(!x(y).P \mid {?x[y].Q})}
+\quad \becomes \quad
+\tm{\nu y\Of{A}.(P \mid Q)}
+\\~\\
+(\kappa_{\otimes1}) \quad
+\tm{\nu z\Of{C}.(x[y].(P \mid Q) \mid R)}
+\quad \becomes \quad
+\tm{x[y].(\nu z\Of{C}.(P \mid R) \mid Q)}
+\end{array}
+\]
+\end{frame}
 

### 2015-04-23 paper/presentation.tex
Adding more script and slides (CP-centric). Using \note command.

@@
 
+\setbeameroption{show notes}
+\setbeamertemplate{note page}[plain]
+
 \title[Project Presentation]{Propositions as Sessions, Mechanically}
@@
 \titlepage
-\end{frame}
-
-\section{Preamble}
-
-\begin{comment}
+\note{
 today i am going to talk about a formalisation effort based on an existing
@@
 communicating agents to be well-defined and correctly implemented.
-\end{comment}
+}
+
+\end{frame}
+
+\section{Preamble}
 
@@
 \end{math}
-\end{frame}
-
-\begin{comment}
+\note{
 one approach to achieve such correctness guarantees is by using binary session
@@
 protocol then involves providing mutual dual types to the communicating
-parties. [a simple example with two processes is displayed].
-\end{comment}
+parties.
+}
+\end{frame}
 
@@
 Hand-written (informal) proofs $=$ Error-prone $+$ Difficult to update
-\end{frame}
-
-\begin{comment}
+\note{
 there has been a lot of activity in this area producing a variety of different
@@
 metatheory
-\end{comment}
+}
+\end{frame}
 
@@
 GV $\to$ CP translation
-\end{frame}
-
-\begin{comment}
+\note{
 the formalisation i present here is an encoding of a system in the Coq proof
@@
 as session types.
-\end{comment}
+}
+\end{frame}
 
@@
 involution)
+
+\note{
+Here are some exmaples of linear logic connectives with session type
+interpretations. The operator times corresponds to output and the operator par
+corresponds to input. Since it is based on the classical variant of linear
+logic we have that negation is an involution.
+}
 \end{frame}
-
-\begin{comment}
-by classical linear logic i mean a logic where negation is an involution which
-restricts the use of certain rules on resources, namely that all linear
-resources must be used (corresponding to prohibiting weakening) and that one
-cannot duplicate a linear resource (corresponding to contraction). some typing
-rules for the CP calculus are presented as they appear in the Coq encoding
-(note that they are very similar to their informal counterparts). the
-interpretation of tensor product is output of A and continue as B. The dual of
-this is input of A and continue as the negation of B. Thus negation in
-classical linear logic corresponds to session duality.
-\end{comment}
 
@@
 \end{itemize}
-\end{frame}
-
-\begin{comment}
+\note{
 The encoding follows practices in formalising programming language metatheory
@@
 consideration over representation choices.
-\end{comment}
+}
+\end{frame}
 
@@
 \end{itemize}
-\end{frame}
-
-\begin{comment}
+\note{
 one of the key issues when formalising programming language metatheory is how
@@
 chose to adopt the cofinite quantification approach which allows to exclude a
-finite set of names from consideration. in contrast to the sometimes weaker
+finite set of names from consideration in contrast to the sometimes weaker
 notion of a single sufficiently fresh name
-\end{comment}
+
+I utilise the metatheory library developed by a team of researchers at the
+university of pennslevennia which handles these representation choices.
+}
+\end{frame}
 
@@
 \end{tabular}
+\note{
+
+CP rules were encoded as an inductive in Coq and an example for cut is shown
+here. CP provided challenges for the metatheory library since it differs in
+many respects to lambda calculi. for example, in the lambda calculus one does
+not work up to permutation of binders whereas in a process calculus these
+kinds of structural equivalences are standard as we shall see. we require
+additional assumptions for environment ordering and more attention is paid to
+the opening of subprocesses. the metatheory library did not support
+permutations and as a result of this work some basic operations and custom
+tactics were developed. the custom tactics can solve simple permutation goals
+via simple proof search. this slide presents the informal and formal
+definitions of cut. notice there is no overhead when introducing binders to
+the subprocesses or for environment ordering, but apart from these
+representation issues the rules look similar which could assist the uptake of
+such formal tools by programming language researchers.
+
+}
 \end{frame}
-
-\begin{comment}
-CP rules were encoded as an inductive in Coq and an example for cut is shown
-here. CP provided challenges for metatheory library since it differ in many
-respects to lambda calculi. for example, in the lambda calculus one does not
-work up to permutation of binders whereas in a process calculus these kinds of
-structural equivalences are standard. we require additional assumptions for
-environment ordering and more attention is paid to the opening of
-subprocesses. metatheory did not support permutations and as a result of this
-work some basic operations and custom tactics were developed. the custom
-tactics can solve simple permutation goals via simple proof search. this slide
-presents the informal and formal definitions of cut. There is no overhead when
-introducing binders to the subprocesses or for environment ordering, but apart
-from these representation issues the rules look similar which could assist the
-uptake of such formal tools by programming language researchers.
-\end{comment}
-
-\defverbatim[colored]\subred{
-\begin{coq}
-Theorem proc_sub_red:
-  forall $\Gamma$ P Q
-         (WT: P $\tpvdash_{cp}$ $\Gamma$)
-         (RED: P $\becomes$ Q),
-    Q $\tpvdash_{cp}$ $\Gamma$.
-\end{coq}
-}
 
@@
 \end{tabular}
-\begin{itemize}
-\item {\fontsize{6}{7.2}\selectfont blah one}
-\item blah two
-\item blah three
-\item blah four
-\item blah five
-\item blah six
-\end{itemize}
-\end{frame}
-
-\begin{comment}
+\[\colored
+(\kappa_{\otimes1}) \quad
+\tm{\nu z\Of{C}.(x[y].(P \mid Q) \mid R)}
+\quad \becomes \quad
+\tm{x[y].(\nu z\Of{C}.(P \mid R) \mid Q)}
+\]
+\note{
 Subject reduction in CP proceeds by cut reductions of the logical operators
@@
 by a cut.
-\end{comment}
+}
+\end{frame}
 
@@
 \end{tabular}
+\note{
+An example of a principal cut reduction: cut of server accept against client
+request. Comparing the formal and informal versions, the informal version
+describes the cut reductions using derivation trees while this structure is
+left implicit in the Coq encoding.
+}
 \end{frame}
@@
 
+\begin{frame}[fragile]
+\begin{coq}
+Theorem proc_cut_elim:
+  forall P $\Gamma$
+         (WT: P $\tpcp$ $\Gamma$),
+    exists Q, P $\becomes^\star$ Q /\ $\sim$ is_cut Q.
+\end{coq}
+\note{
+mention star being the reflexive and transitive closure of reduction
+
+Well-typed relation alone is insufficient to prove the theorem. The cut typing
+rule does not specify the subprocesses are related i.e. are eventually dual
+logical operators which would be necessary to proceed using the current
+definition.
+
+It seems necessary to define a termination measure on the length of the
+derivation but this has not yet been done. Additionally, work on formalising
+logics with cut elimination take an entirely different approach favouring an
+explicit representation of derivations as trees. Adopting this approach for CP
+may prove fruitful not just in this instance but also in other areas of the
+development (what? contraction?)
+
+}
+\end{frame}
+
 \subsection{GV}
@@
 GV
-\end{frame}
-
-\begin{comment}
+\note{
 GV has built-in support for session types. consider the rule for sending a
 value of type T along a channel N of type !T.S.
-\end{comment}
+}
+\end{frame}
 
@@
 something about a translation
-\end{frame}
-
-\begin{comment}
+\note{
 the translation to CP is
@@
 overlook this probably
-\end{comment}
-
-\begin{frame}
-- subject reduction of CP
-- top-level cut elimination
-- translation preserving well-typedness
+}
 \end{frame}
@@
 
-\begin{comment}
-the system has three main properties. subject reduction of CP, top-level
-cut-elimination and translation preserving well-typedness. of which subject
-reduction is the only one fully formalised. the other two presented some
-challenges but some comments on progress towards the other two is
-offered. i also highlight some issues and guidelines
-\end{comment}
-
-\begin{frame}
-subject reduction
-\end{frame}
-
-\begin{comment}
-subject reduction is formalised using a relation on processes
-\end{comment}
-
 \begin{frame}
 problems for cut elimination
-\end{frame}
-
-\begin{comment}
+\note{
 sessions correspond to propositions
@@
 may be inspirational
-\end{comment}
+}
+\end{frame}
 
@@
 problems of the translation
-\end{frame}
-
-\begin{comment}
+\note{
 the translation seems to require a duplication of the specifications for cp
@@
 modularisation, representation, library reuse, automation!
-\end{comment}
+}
+\end{frame}
 
@@
 \begin{itemize}
-\item Propositions as Sessions
-\item system features
-\item Coq
-\item encoding
-\item subject reduction theorem
 \item difficulties in formalisation

### 2015-04-25 paper/presentation.tex
Commuting conversions more in-depth (too much?).
Added GV example plus some draft bits that I forgot to commit.

@@
 \setbeamertemplate{note page}[plain]
+
+% GV/CP example listing details
+\lstdefinelanguage{GV}{
+    mathescape=true,
+    morekeywords=[1]{case, connect, in, let, of, receive, select, send,
+      terminate, to, with},
+    sensitive=false,
+    morecomment=[l]{//},
+    morecomment=[s]{/*}{*/},
+    morestring=[b]",
+    basicstyle=\HappyMed,
+    keywordstyle=[1]{\color{magenta}},
+}
+
+\lstnewenvironment{gv}{\lstset{language=GV}}{}
 
@@
 }
+\end{frame}
+
+\begin{frame}[fragile]
+\oldframetitle{GV Example}
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} cc}
+Process A ($A_x$) & Process B ($B_x$)\\
+$\colored\tp{\inpt{Int}{\outpt{String}{\key{end}_!}}}$ ($\colored\tp{S}$) &
+$\colored\tp{\outpt{Int}{\inpt{String}{\key{end}_?}}}$
+  ($\colored\tp{\overline{S}}$)\\
+\begin{gv}
+let (m,$x_1$) = receive $x$ in
+send thanks $x_1$
+\end{gv} &
+\begin{gv}
+let $x_1$ = send gift $x$ in
+let (resp,$x_2$) = receive $x_1$ in
+terminate $x_2$
+\end{gv}
+\end{tabular*}
+\[\colored
+\inference{
+  \Tp{thanks:\tp{String} \comma x:\tp{S}} \Tpvdash A_x:\tp{\key{end}_!} &
+  \Tp{gift:\tp{Int} \comma x:\tp{\overline{S}}} \Tpvdash B_x:\tp{\key{Unit}}
+}{Connect}{
+  \Tp{thanks:\tp{String} \comma gift:\tp{Int}} \Tpvdash
+    \Tm{with}~x~\Tm{connect}~A_x~\Tm{to}~B_x : \tp{Unit}
+}
+\]
 \end{frame}
@@
 \begin{frame}[fragile]
-\oldframetitle{Principal Cut Reduction ($\beta_{!?}$)}
+\oldframetitle{Principal Cut Reduction ($\beta_{!?}$): Formal vs. Informal}
 \begin{tabular}{c}
@@
 
-\begin{frame}
-\[\colored
+\begin{frame}[fragile]
+\oldframetitle{Commuting Conversion ($\kappa_{\otimes1}$): Formal vs.
+  Informal}
+\begin{tabular}{c}
+\begin{coq}
+Lemma reduce_cc_multi_one:
+  forall P Q R (x:atom) A B $\Gamma$ (LCQ: lc_proc Q)
+         (WT: $\nu$ A.(x[B].(P $\mid$ Q) $\mid$ R) $\tpcp$ $\Gamma$),
+    x[B].($\nu$ A.($\permb{0}{1}{P}$ $\mid$ R) $\mid$ Q) $\tpcp$ $\Gamma$.
+\end{coq}
+\\~\\
+\begin{math}
+\colored
 \begin{array}{c}
-(\beta_{!?}) \quad
-\tm{\nu x \Of{!A}.(!x(y).P \mid {?x[y].Q})}
-\quad \becomes \quad
-\tm{\nu y\Of{A}.(P \mid Q)}
-\\~\\
-(\kappa_{\otimes1}) \quad
-\tm{\nu z\Of{C}.(x[y].(P \mid Q) \mid R)}
-\quad \becomes \quad
-\tm{x[y].(\nu z\Of{C}.(P \mid R) \mid Q)}
+
+% \spacer
+\vcenter{
+  \inference{
+    \inference{
+      \tm{P} \tpvdash \tp{\Gamma \comma \tmof{y}A \comma \tmof{z}C}
+      &
+      \tm{Q} \tpvdash \tp{\Delta \comma \tmof{x}B}
+    }{\otimes}{
+      \tm{x[y].(P \mid Q)} \tpvdash
+        \tp{\Gamma \comma \Delta \comma \tmof{x}A \otimes B \comma \tmof{z}C}
+    }
+    &
+    \tm{R} \tpvdash \tp{\Theta \comma \tmof{z}C^\bot}
+  }{Cut}{
+    \tm{\nu z\Of{C}.(x[y].(P \mid Q) \mid R)} \tpvdash
+      \tp{\Gamma \comma \Delta \comma \Theta \comma \tmof{x}A \otimes B}
+  }
+}
+\quad \becomes \hfill \\~\\
+\hfill
+\vcenter{
+  \inference{
+    \inference{
+      \tm{P} \tpvdash \tp{\Gamma \comma \tmof{y}A \comma \tmof{z}C}
+      &
+      \tm{R} \tpvdash \tp{\Theta \comma \tmof{z}C^\bot}
+    }{Cut}{
+      \tm{\nu z\Of{C}.(P \mid R)} \tpvdash
+        \tp{\Gamma \comma \Theta \comma \tmof{y}A}
+    }
+    &
+    \tm{Q} \tpvdash \tp{\Delta \comma \tmof{x}B}
+  }{\otimes}{
+    \tm{x[y].(\nu z\Of{C}.(P \mid R) \mid Q)} \tpvdash
+      \tp{\Gamma \comma \Delta \comma \Theta \comma \tmof{x}A \otimes B}
+  }
+}
+
 \end{array}
-\]
+%% \colored
+%% (\kappa_{\otimes1}) \quad
+%% \tm{\nu z\Of{C}.(x[y].(P \mid Q) \mid R)}
+%% \quad \becomes \quad
+%% \tm{x[y].(\nu z\Of{C}.(P \mid R) \mid Q)}
+\end{math}
+\end{tabular}
+\note{
+
+(well, that looks ridiculous.)\\~\\
+
+Comparing formal and informal versions of a commuting conversion for output
+reveals some issues. it is necessary to stipulate that Q is locally closed to
+ensure that the P subprocess receives the channel from the outer cut as part
+of its environment -- since the environment splitting is in general
+non-deterministic. Additionally, we need to permute the binders of P in the
+conclusion and this is necessary for the majority of the commuting conversion
+rules.
+
+}
 \end{frame}
@@
 \begin{frame}[fragile]
+\oldframetitle{Top-Level Cut Elimination}
 \begin{coq}
@@
 
-Well-typed relation alone is insufficient to prove the theorem. The cut typing
-rule does not specify the subprocesses are related i.e. are eventually dual
+Well-typed relation alone is insufficient to prove the theorem. The rule for
+cut does not specify the subprocesses are related i.e. are eventually dual
 logical operators which would be necessary to proceed using the current
@@
 may prove fruitful not just in this instance but also in other areas of the
-development (what? contraction?)
+development such as specifying environment splitting as in the rules for
+commuting conversions.
 
@@
 \note{
+
+Typing rules again? Very similar to CP.\\~\\
+Representation issues?\\~\\
 GV has built-in support for session types. consider the rule for sending a
 value of type T along a channel N of type !T.S.
+
 }
@@
 something about a translation
+
 \note{
+
 the translation to CP is
@@
 overlook this probably
+
+the translation seems to require a duplication of the specifications for cp
+and gv typing rules not done
+
 }
@@
 \begin{frame}
-problems for cut elimination
+\frametitle{Issues and Guidelines}
+\begin{itemize}
+\item representation
+\item modularisation
+\item automation
+\item reuse
+\end{itemize}
 \note{
-sessions correspond to propositions
-processes correspond to proofs
-cut elimination corresponds to communication
-
-cut elimination is not currently proven since we require a termination measure
-on the length of a derivation and a relationship between the subprocesses of a
-cut. other work based on formalising logics where the derivations are explicit
-may be inspirational
-}
-\end{frame}
-
-\begin{frame}
-problems of the translation
-\note{
-the translation seems to require a duplication of the specifications for cp
-and gv typing rules not done
-
-what can we show them for the failures?
-
-here are some issues and things learnt! -- our guidelines are all about
-modularisation, representation, library reuse, automation!
+  representation, modularisation(?), automation, reuse
 }
@@
 \begin{frame}
+\oldframetitle{Future Work}
 \begin{itemize}
@@
 \end{itemize}
+\note{
+
+}
 \end{frame}

### 2015-04-25 paper/presentation.tex
Fixed some issues with the frame title macro (more generic command).

@@
 }
+\newcommand{\ftitle}[2]{\oldframetitle{#1}\vspace{-#2cm}}
 
@@
 \begin{frame}
-\oldframetitle{Session Types}
+\ftitle{Session Types}{1.5}
 \begin{math}
@@
 
+\begin{frame}
+\ftitle{System Overview}{2}
+CP $=$ process calculus with operators from classical linear logic
+
+GV $=$ functional language with session types
+
+GV $\to$ CP translation
+\\~\\
+Relies on Curry-Howard variant:
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}ccc}
+Propositions & correspond to & Session Types\\
+Proofs & correspond to & Processes\\
+Cut Elimination & correponds to & Communication
+\end{tabular*}
+\note{
+the formalisation i present here is an encoding of a system in the Coq proof
+assistant. it is based on a two-tier system providing a logical foundation for
+understanding session types. A high-level overview of the system is as
+follows: we have a process calculus CP with operators from classical linear
+logic, a high-level functional language GV with session types, and a
+translation from GV to CP which interprets operators in classical linear logic
+as session types.
+}
+\end{frame}
+
 \begin{frame}[fragile]
-\oldframetitle{GV Example}
+\ftitle{GV Example}{2}
 \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} cc}
@@
 \frametitle{Motivation}
-Hand-written (informal) proofs $=$ Error-prone $+$ Difficult to update
+Informal proofs $=$ Error-prone $+$ Difficult to update
 \note{
@@
 metatheory
-}
-\end{frame}
-
-\begin{frame}
-\frametitle{System Overview}
-CP $=$ process calculus with operators from classical linear logic
-
-GV $=$ functional language with session types
-
-GV $\to$ CP translation
-\note{
-the formalisation i present here is an encoding of a system in the Coq proof
-assistant. it is based on a two-tier system providing a logical foundation for
-understanding session types. A high-level overview of the system is as
-follows: we have a process calculus CP with operators from classical linear
-logic, a high-level functional language GV with session types, and a
-translation from GV to CP which interprets operators in classical linear logic
-as session types.
 }
@@
 \begin{frame}
-GV
+GV is similar
 \note{

### 2015-04-27 paper/presentation.tex
Added a CP Example (not a translation from GV however).

@@
 \usepackage{etoolbox}
+\usepackage[cal=cm,calscaled=.80]{mathalfa}
 \usepackage{tikz}
@@
 
-% GV/CP example listing details
+\newcommand\CPFSize{\fontsize{7}{8.4}\selectfont}
+% CP/GV example listing details
+\lstdefinelanguage{CP}{
+    mathescape=true,
+    morekeywords=[1]{case, inl, inr},
+    sensitive=false,
+    morecomment=[l]{//},
+    morecomment=[s]{/*}{*/},
+    morestring=[b]",
+    basicstyle=\CPFSize,
+    keywordstyle=[1]{\color{red}},
+}
 \lstdefinelanguage{GV}{
@@
 
+\lstnewenvironment{cp}{\lstset{language=CP}}{}
 \lstnewenvironment{gv}{\lstset{language=GV}}{}
@@
       & \altt & \outpt{T}{S} & \mbox{send value of type T, continue as S}\\
-      & \altt & \Branch{l_i}{S_i}{i\in I} &
+      & \altt & \Branch{\mathcal{L}_i}{S_i}{i\in I} &
                            \mbox{offer between $|I|$ alternatives}\\
-      & \altt & \Choice{l_i}{S_i}{i\in I} &
+      & \altt & \Choice{\mathcal{L}_i}{S_i}{i\in I} &
                            \mbox{choose between $|I|$ alternatives}\\
@@
  \dual{(\inpt{T}{S})} & \bnf & \outpt{T}{\dual{S}}\\
- \dual{\Branch{l_i}{S_i}{i\in I}} & \bnf & \Choice{l_i}{\dual{S_i}}{i\in I}\\
+ \dual{\Branch{\mathcal{L}_i}{S_i}{i\in I}} & \bnf &
+       \Choice{\mathcal{L}_i}{\dual{S_i}}{i\in I}\\
             \vdots                & \cdots & \vdots
@@
 \begin{frame}[fragile]
+\ftitle{CP Example}{0.5}
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} cc}
+Process A ($A_x$) & Process B ($B_x$)\\
+$\colored\tp{Int \parr (String \otimes \1)}$
+  ($\colored\tp{\dualbt{S}}$) &
+$\colored\tp{Int \otimes (String \parr \bot)}$ ($\colored\tp{S}$)\\
+\begin{cp}
+$\colored\tm{x(y).x[w].(\mbox{put-thanks}_w \mid x[\,].0)}$
+\end{cp} &
+\begin{cp}
+$\colored\tm{x[y].(\mbox{put-gift}_y \mid x(w).x().\mbox{get-response}_w)}$
+\end{cp}
+\end{tabular*}
+\[\colored
+\inference{
+  \tm{B_x} \tpvdash \tp{\Gamma \comma \tmof{x}S}
+  &
+  \tm{A_x} \tpvdash \tp{\Delta \comma \tmof{x}\dualbt{S}}
+}{Cut}{
+  \tm{\nu x \of{S}.(B_x \mid A_x)} \tpvdash \tp{\Gamma \comma \Delta}
+}
+\]
+\CPFSize
+\colored
+\begin{align*}
+  \tm{\nu x.(B_x \mid A_x)} & =
+    \tm{\nu x.(x[y].(\mbox{put-gift}_y \mid x(w).x().\mbox{get-response}_w)
+          \mid x(y).x[w].(\mbox{put-thanks}_w \mid x[\,].0))}
+    \\
+                            & \becomes
+    \tm{\nu y.(\mbox{put-gift}_y \mid
+               \nu x.(x(w).x().\mbox{get-response}_w) \mid
+                      x[w].(\mbox{put-thanks}_w \mid x[\,].0))}
+    \\
+                            & \becomes
+    \tm{\nu y.(\mbox{put-gift}_y \mid
+               \nu w.(\mbox{put-thanks}_w \mid
+                      \nu x.(x().\mbox{get-response}_w \mid x[\,].0)))}
+    \\
+                            & \becomes
+    \tm{\nu y.(\mbox{put-gift}_y \mid
+               \nu w.(\mbox{put-thanks}_w \mid \mbox{get-response}_w))}
+\end{align*}
+\note{
+Here are some exmaples of linear logic connectives with session type
+interpretations. The operator times corresponds to output and the operator par
+corresponds to input. Since it is based on the classical variant of linear
+logic we have that negation is an involution.
+}
+\end{frame}
+
+\begin{frame}[fragile]
 \ftitle{GV Example}{2}
@@
 Informal proofs $=$ Error-prone $+$ Difficult to update
+\begin{itemize}
+\item Subject Reduction: If $m : T$ and $m \rightarrow m'$ then $m' : T$
+\end{itemize}
 \note{
@@
 metatheory
-}
-\end{frame}
-
-\begin{frame}
-\frametitle{CP Propositions}
-\begin{math}
-\begin{array}{rclc}
-    \tp{A,B,C} & \bnf
-          & \tp{A \otimes B} & \key{times: output A then behave as B}\\
-          & \altt & \tp{A \parr B} & \key{par: input A then behave as B}\\
-          & \altt & \cdots
-\end{array}
-\end{math}
-
-For all propositions $A$, $\dualbt{(\dualbt{A})} = A$ (negation is an
-involution)
-
-\note{
-Here are some exmaples of linear logic connectives with session type
-interpretations. The operator times corresponds to output and the operator par
-corresponds to input. Since it is based on the classical variant of linear
-logic we have that negation is an involution.
 }

### 2015-04-27 paper/presentation.tex
Working on GV/Translation parts of talk now.
Almost there.

@@
 \begin{frame}
-\oldframetitle{Preliminaries}
+\ftitle{Preliminaries}{2}
 \begin{itemize}
@@
 
-\begin{frame}
-GV is similar
+\begin{frame}[fragile]
+\ftitle{GV}{2.5}
+\begin{itemize}
+\item GV similar to CP (types, terms, well-typed relation, $\ldots$)
+\item Representation choices (Kindedness, Sessions)
+\end{itemize}
+\begin{coqp}
+Inductive wf_typ : typ -> kind -> Prop :=
+  | $\ldots$
+  | wf_labs : forall kt ku T U
+                     (WFT: wf_typ T kt)
+                     (WFU: wf_typ U ku),
+                wf_typ (T $\lolli$ U) lin
+  | $\ldots$
+\end{coqp}
 \note{
@@
 
-\begin{frame}
-something about a translation
-
+\begin{frame}[fragile]
+\ftitle{Translation from GV to CP : Work in Progress}{1}
+\begin{theorem}[Translation]
+If a GV term is well-typed then its translation is well-typed in CP.
+\end{theorem}
+Issues
+\begin{itemize}
+\item type inference, positive cases and duplication
+\end{itemize}
+\begin{coqp}
+trans_case:
+  forall L $S_1$ $S_2$ M NL NR P QL QR G H z
+         (INFM: inferTy M G ($S_1$ $\with$ $S_2$))
+         (TR: transTm M G 0 P)
+         (TRNL: forall x (NIN: x $\notin$ L),
+                  transTm (open NL x) (x $\sim$ $S_1$ ++ H) z QL)
+         (TRNR: forall x (NIN: x $\notin$ L),
+                  transTm (open NR x) (x $\sim$ $S_2$ ++ H) z QR),
+  transTm (tm_case M NL NR) (G++H) z
+          ($\nu \sem{S_1 \with S_2}$t.(P $\mid$ 0 CASE QL OR QR))
+\end{coqp}
 \note{
 
-the translation to CP is
-presented. blah blah something about the session types being a little odd or
-overlook this probably
+continuation-passing style.
 
-the translation seems to require a duplication of the specifications for cp
-and gv typing rules not done
+A well-typed term in GV is parameterised by a name representing the channel
+which communicates the result of its translation
 

### 2015-04-28 paper/presentation.tex
Finished slides (minus some revision).

@@
 \begin{theorem}[Translation]
-If a GV term is well-typed then its translation is well-typed in CP.
+If a GV term is well-typed then its translation is well-typed in CP:
+If $\Phi \tpgv~M\of{T}$ then
+$\sem{M}x \tpcp~\dualbt{\sem{\Phi}},x\of{\sem{T}}$
 \end{theorem}
@@
 
-continuation-passing style.
+A well-typed term in GV is parameterised by a name representing the channel
+which communicates the result of its translation\\~\\
 
-A well-typed term in GV is parameterised by a name representing the channel
-which communicates the result of its translation
+due to locally nameless translation requires type inference procedure to
+obtain binders for corresponding CP process. environment may not have variable
+(impossible, only well-typed terms are translated) \\~\\
+
+solution: instead of defining functionally, use inductive type to specify only
+positive cases
+
+problem: requires much of the details from the CP/GV definitions such as
+linear environment splitting and cofinite quantification (for both CP and GV!)
+even if GV term with no binders e.g. function application.
 
@@
 \begin{itemize}
-\item representation
-\item modularisation
-\item automation
-\item reuse
+\item Representation
+\item Modularisation
+\item Automation
+\item Reuse
 \end{itemize}
 \note{
-  representation, modularisation(?), automation, reuse
+
+  representation is a central issue when formalising languages and choices
+  impact future definitions and properties dependent on them. Considerations
+  range from how much expressive power provided by the theorem prover to
+  exploit to the choice of which encoding to use for the properties of
+  interest. Some decisions may impact upon the correspondence between the
+  formal and informal systems.
+
+  Modularising the development was essential to reduce the interactive load
+  time and mitigate crashes.
+
+  Automation advise use of non-trivial tactics favour proof search over
+  in-place macros for oft-repeated sequences. It is worth considering
+  situations where automation would simplify proofs. Automation is already an
+  important consideration in the interactive theorem proving
+  community. Examples of automation from the current development have been
+  highlighted, but particular emphasis should be placed on tactics involving
+  non-trivial proof search of the context because these tend to be more widely
+  applicable.
+
+  Reuse libraries wherever possible: programming language assistance but also
+  helpful tactics not present in standard Coq (added some to solve permutation
+  goals)
+
 }
@@
 \begin{frame}
-\oldframetitle{Future Work}
+\ftitle{Future Work}{3}
 \begin{itemize}
-\item difficulties in formalisation
-\item implications
-\item future work
+\item Further investigate top-level cut elimination
+\item Make progress on translation
+\item Support polymorphism and $n$-ary branch and choice
+\item Alternate forms of alias control (adoption and focus)
 \end{itemize}
 \note{
+
+  resolve issues with proving top-level cut elimination either by constructing
+  explicit derivations or by defining a suitable induction on the current
+  definition. however, for the latter approach it is unclear how to obtain the
+  desired relationship between subprocesses of cut
+
+  Further progress on the current inductive definition of the translation can
+  be made and then limiting the extent of duplication would be another aim
+
+  the development currently does not support polymorphism and has only binary
+  branch and choice. extending the system to support these features would
+  bring it in line with extensions of the original cp/gv system.
+
+  lastly, investigating alternate forms of alias control such as the adoption
+  and focus system which allows controlled aliasing of linear objects may be
+  worth formal study in the context of session types.
 
